# 4 DoubleML for Difference-in-Differences {.unnumbered}

This tutorial shows how **DoubleML** can be used with the **did** package in R to estimate group-time average treatment effects in **difference-in-differences (DiD)** models with multiple time periods. We follow the [DoubleML R DiD example](https://docs.doubleml.org/stable/examples/R_double_ml_did.html). Using **simulated data** from the {did} package and **real-world data** (county-level teen employment, `mpdta`), we compare default regression-based DiD estimation with **ML-based doubly robust DiD** via DoubleML.

## Overview

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Difference-in-Differences (DiD) | Compares before-after changes between treated and untreated groups. | "Compare employment change in treated states vs untreated states." |
| ATT (Average Treatment effect on the Treated) | Average effect for units that actually received treatment. | "Policy effect for states that raised minimum wage." |
| Parallel trends assumption | Without treatment, treated and control groups would follow similar trends. | "Before policy, both groups move similarly over time." |
| Propensity score | Probability of receiving treatment given covariates. | "Estimated chance a county gets treated based on baseline features." |
| Doubly robust estimator | Stays consistent if either outcome model or treatment model is correct. | "Even if one model is wrong, effect can still be valid." |
| Cross-fitting | Uses held-out folds for predictions to reduce overfitting bias. | "Train on fold A, predict on fold B, then swap." |

**Double Machine Learning for Difference-in-Differences (DML-DiD)** combines the credibility of DiD identification with the flexibility of modern machine learning to estimate causal effects in high-dimensional, observational panel data settings [[8]].

Traditional DiD assumes:

$$Y_{it} = \alpha_i + \gamma_t + \theta (D_{it} \times Post_t) + \varepsilon_{it}$$

**DML-DiD** extends this by allowing flexible, nonparametric control for high-dimensional confounders $X_{it}$:
$$\Delta Y_{it} = \theta D_{it} + g(X_{it}) + \zeta_{it}$$

where $\Delta Y_{it} = Y_{it} - Y_{i,t-1}$ is the first-differenced outcome, and $g(\cdot)$ is estimated via ML [[1]].

### How It Works (4 Key Steps)

| Step | Component | Purpose |
|------|-----------|---------|
| **1** | **Nuisance estimation** | Use ML (RF, XGBoost, Lasso) to estimate: <br>• Propensity score: $\hat{g}(X) = P(D=1\|X)$ <br>• Outcome regression: $\hat{\ell}(X) = E[\Delta Y\|D=0,X]$ |
| **2** | **Cross-fitting** | Split data into $K$ folds; train nuisance models on $K-1$ folds, predict on held-out fold to avoid overfitting [[10]] |
| **3** | **Neyman-orthogonal score** | Construct debiased score: <br>$$\psi_i = \frac{D_i - \hat{g}(X_i)}{\hat{p}(1-\hat{g}(X_i))} \cdot (\Delta Y_i - \hat{\ell}(X_i))$$ <br>This ensures small errors in $\hat{g}, \hat{\ell}$ have negligible impact on $\hat{\theta}$ [[9]] |
| **4** | **Final estimation** | $\hat{\theta}_{DML} = \frac{1}{n}\sum_i \psi_i$ with valid standard errors via influence function asymptotics |

###  Why Combine DML with DiD?

| Challenge | Traditional DiD | DML-DiD Solution |
|-----------|---------------|----------------|
| **High-dimensional controls** | OLS fails when $p \gg n$ | ML handles 100s–1000s of covariates |
| **Nonlinear confounding** | Linear $g(X)$ may be misspecified | Flexible ML captures complex $g(X)$ |
| **Regularization bias** | Lasso shrinks coefficients → biased $\theta$ | Neyman orthogonality "debias" the final estimate |
| **Valid inference** | No built-in SEs for ML-adjusted DiD | Asymptotic normality → confidence intervals, p-values |

### Supported Data Structures

DML-DiD accommodates:

- **Two-period DiD**: Classic treatment vs. control, pre/post design
- **Staggered adoption**: Units treated at different times (group-time ATT) [[19]]
- **Continuous treatments**: Dose-response DiD using kernel weighting + GPS [[6]]
- **Repeated cross-sections or balanced panels**

The `DoubleML` package implements these via `DoubleMLPLR`, `DoubleMLIRM`, and custom wrappers for the `did` package [[10]].

### Limitations & Caveats

-  **Parallel trends still required**: DML adjusts for *observed* $X$; unobserved time-varying confounders remain a threat  
-️  **Sample size**: Cross-fitting requires sufficient observations per fold; small panels may yield unstable nuisance estimates  
-️  **Interpretability**: While $\theta$ is causal, the ML nuisance functions $g(X), \ell(X)$ remain black boxes  
-️  **Implementation complexity**: Requires careful tuning of ML hyperparameters and cross-fitting folds

## Implementation in R

DoubleML can be integrated into the {did} package to flexibly estimate nuisance functions (outcome and propensity) while still leveraging did's infrastructure for group-time effects, aggregation, and inference. The key steps are: 

| Topic | What we cover |
|-------|----------------|
| **Multi-period DiD** | Group-time ATT(g,t) with staggered treatment (Callaway & Sant'Anna 2021) |
| **Doubly robust DiD** | Sant'Anna & Zhao (2020): ML for outcome and propensity in DiD |
| **DoubleML + did** | `RCausalML::doubleml_did_linear` / `_rf` / `_xgboost` (DoubleML IRM, score `"ATTE"`) |
| **Learners** | Defaults: LM + logit; **ranger**; **xgboost** (see `?doubleml_did_linear`) |
| **Evaluation** | `doubleml_did_eval_*` + `doubleml_did_eval_preds`; compare `aggte` and RMSE across wrappers |
| **did tools** | `ggdid`, `aggte`, pre-test of parallel trends |

**Key idea**: The {did} package's default estimator is doubly robust and uses (unpenalized) linear and logistic regression. By passing a **custom estimation function** that uses DoubleML internally, we can use flexible ML for the nuisance functions while still using did's infrastructure for group-time effects, aggregation, and inference.

### Setup and Libraries

### Load and Check Required Libraries

```{r packages-list}
packages <- c(
  "DoubleML",
  "RCausalML",
  "did",
  "mlr3",
  "mlr3learners",
  "mlr3measures",
  "xgboost",
  "tidyverse",
  "lgr"
)
```

### Install Missing Packages

```{r install-packages}
#| eval: false
#| warning: false
#| error: false
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)
```

### Load Libraries and Options

```{r setup-load-and-helpers}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

# Suppress mlr3 messages during fitting
lgr::get_logger("mlr3")$set_threshold("warn")

#' Adapt RCausalML DoubleML DiD wrappers for did::att_gt
#'
#' `att_gt()` passes `i.weights`, `inffunc`, etc. to a custom `est_method`.
#' The RCausalML helpers only need `y1`, `y0`, `D`, and `covariates`; extra
#' arguments are ignored here (weights are not used in this notebook).
as_did_est_method <- function(fn) {
  function(y1, y0, D, covariates, ...) {
    fn(y1, y0, D, covariates)
  }
}

#' Row-bind `aggte(..., type = "simple")` summaries for several `att_gt` fits
summarize_dml_wrappers <- function(named_attgt, alp = 0.05) {
  z <- stats::qnorm(1 - alp / 2)
  purrr::imap_dfr(named_attgt, function(obj, wrapper) {
    ag <- did::aggte(obj, type = "simple", na.rm = TRUE)
    tibble::tibble(
      wrapper = wrapper,
      overall_att = ag$overall.att,
      overall_se = ag$overall.se,
      z = ag$overall.att / ag$overall.se,
      ci_low = ag$overall.att - z * ag$overall.se,
      ci_high = ag$overall.att + z * ag$overall.se
    )
  })
}
```


## Part A: Double Did With Simulated Data

We use the introductory example from the **did** package: a simulated dataset with multiple time periods and staggered treatment. This mirrors the [DoubleML DiD documentation](https://docs.doubleml.org/stable/examples/R_double_ml_did.html).

### Data

The simulation uses `reset.sim()` and `build_sim_dataset()` from the did package. We set 4 time periods and dynamic treatment effects; units first treated in period 1 are dropped so we can recover ATT(g,t) for later cohorts.

```{r sim-build-dataset}
# Original code from did vignette: https://github.com/bcallaway11/did/blob/master/vignettes/did-basics.Rmd
time.periods <- 4
sp <- reset.sim()
sp$te <- 0

set.seed(1814)

# Add dynamic effects (treatment effect varies by period)
sp$te.e <- 1:time.periods

# Generate dataset; drops "always-treated" (period 1) units
dta <- build_sim_dataset(sp)

# Number of observations
nrow(dta)
```


```{r sim-head-dta}
head(dta)
```

The data is in **long format** (one row per unit-period). Key variables:

- **G**: first period when the unit is treated (cohort)
- **period**: time period
- **Y**: outcome
- **treat**: 1 if unit is treated in this period
- **X**: covariate for adjustment
- **id**, **cluster**: unit and cluster identifiers

### Comparison to Default did Package

By default, **did** estimates group-time ATTs using (unpenalized) linear and logistic regression. We run the default `att_gt` first as a benchmark.

```{r attgt-default-sim}
# Estimate group-time average treatment effects with default (regression-based) method
example_attgt <- att_gt(yname = "Y",
                        tname = "period",
                        idname = "id",
                        gname = "G",
                        xformla = ~X,
                        data = dta
                        )

summary(example_attgt)
```

The output shows ATT(g,t) for each group-time pair, standard errors, simultaneous confidence bands, and a **pre-test p-value for parallel trends**.

### Using ML for DiD: RCausalML + DoubleML

Sant'Anna and Zhao (2020) give a **doubly robust DiD** estimator that works with flexible nuisances. The **did** package accepts a custom `est_method` that receives `y1`, `y0`, `D`, and `covariates` for each internal 2×2 DiD cell. **RCausalML** provides ready-made wrappers that:

1. Form the first difference $\Delta Y = Y_1 - Y_0$
2. Fit **DoubleMLIRM** with score `"ATTE"` (binary treatment) using default **mlr3** learners for each wrapper
3. Return `ATT` and `att.inf.func` in the format **did** expects

| Wrapper | Outcome nuisance `ml_g` | Propensity `ml_m` |
|---------|-------------------------|-------------------|
| `doubleml_did_linear` | `regr.lm` | `classif.log_reg` |
| `doubleml_did_rf` | `regr.ranger` | `classif.ranger` |
| `doubleml_did_xgboost` | `regr.xgboost` | `classif.xgboost` |

See `?RCausalML::doubleml_did_linear` for arguments (custom learners, `n_folds`, `score`, etc.).

### Run DiD with the three RCausalML wrappers

```{r attgt-dml-three-wrappers-sim}
set.seed(1234)
example_attgt_dml_linear <- att_gt(
  yname = "Y",
  tname = "period",
  idname = "id",
  gname = "G",
  xformla = ~X,
  data = dta,
  est_method = as_did_est_method(doubleml_did_linear)
)

example_attgt_dml_rf <- att_gt(
  yname = "Y",
  tname = "period",
  idname = "id",
  gname = "G",
  xformla = ~X,
  data = dta,
  est_method = as_did_est_method(doubleml_did_rf)
)

example_attgt_dml_xgb <- att_gt(
  yname = "Y",
  tname = "period",
  idname = "id",
  gname = "G",
  xformla = ~X,
  data = dta,
  est_method = as_did_est_method(doubleml_did_xgboost)
)

summary(example_attgt_dml_linear)
```

```{r summary-dml-rf-sim}
summary(example_attgt_dml_rf)
```

```{r summary-dml-xgb-sim}
summary(example_attgt_dml_xgb)
```

For this simulation the DGP is close to linear, so **linear nuisances** typically track the default **did** doubly robust estimator most closely; **random forest** and **XGBoost** add flexibility and may show somewhat larger standard errors in some cells because of cross-fitting and nonparametric bias–variance trade-offs.

### Compare wrapper performance (Part A)

We summarize **aggregated** group-time effects with `aggte(..., type = "simple")`, and compare **cell-by-cell** agreement between wrappers (root mean squared difference in $\widehat{\mathrm{ATT}}(g,t)$).

```{r compare-wrappers-part-a}
cmp_agg_a <- summarize_dml_wrappers(
  list(linear = example_attgt_dml_linear, rf = example_attgt_dml_rf, xgboost = example_attgt_dml_xgb)
)
knitr::kable(cmp_agg_a, digits = 4, caption = "Part A: simple aggregate ATT by nuisance learner wrapper")

gt_mat_a <- tibble::tibble(
  linear = example_attgt_dml_linear$att,
  rf = example_attgt_dml_rf$att,
  xgboost = example_attgt_dml_xgb$att
)

tibble::tibble(
  comparison = c("linear vs rf", "linear vs xgboost", "rf vs xgboost"),
  rmse_att_gt = c(
    sqrt(mean((gt_mat_a$linear - gt_mat_a$rf)^2, na.rm = TRUE)),
    sqrt(mean((gt_mat_a$linear - gt_mat_a$xgboost)^2, na.rm = TRUE)),
    sqrt(mean((gt_mat_a$rf - gt_mat_a$xgboost)^2, na.rm = TRUE))
  )
) |>
  knitr::kable(digits = 5, caption = "Part A: RMSE between group-time ATTs across wrappers")
```

### Exploiting the Functionalities of did

Once we have group-time estimates (from default or DoubleML), we can use all **did** tools: plotting and aggregation.

### Plot Group-Time Average Treatment Effects

```{r plot-ggdid-dml-linear}
#| fig-width: 7
#| fig-height: 4
ggdid(example_attgt_dml_linear)
```

### Aggregate Effects

We can aggregate the group-time ATTs into a single **simple average** (or other aggregation types).

```{r aggte-simple-sim}
agg.simple <- aggte(example_attgt_dml_linear, type = "simple")
summary(agg.simple)
```

### Nuisance fit: `doubleml_did_eval_*` and `doubleml_did_eval_preds`

RCausalML exposes **`doubleml_did_eval_linear`**, **`doubleml_did_eval_rf`**, and **`doubleml_did_eval_xgboost`**, which fit with **`store_predictions = TRUE`** and attach **`eval_predictions`** from **`doubleml_did_eval_preds`** (cross-fitted RMSE for `ml_g0` on control units, accuracy for `ml_m`, etc.). The low-level scorer is **`?doubleml_did_eval_preds`**.

Each full `att_gt` fit calls the chosen wrapper on many internal 2×2 slices (`y1`, `y0`, treatment indicator, `covariates`). To compare **nuisance** quality in one place—without re-running the full grid—we simulate a single cross-section with the same first-difference structure (outcome change, binary treatment, one covariate) and report **`doubleml_did_eval_*`** metrics (`ml_g0` RMSE on controls, `ml_m` accuracy).

```{r eval-nuisance-part-a}
set.seed(99)
n_ev <- 600
X_ev <- matrix(stats::rnorm(n_ev), ncol = 1L)
colnames(X_ev) <- "X1"
Dev <- stats::rbinom(n_ev, 1L, stats::plogis(0.35 * X_ev[, 1L]))
delta_ev <- 0.5 * Dev + 0.25 * X_ev[, 1L] + stats::rnorm(n_ev)
y0_ev <- stats::rnorm(n_ev)
y1_ev <- y0_ev + delta_ev

ev_lin <- doubleml_did_eval_linear(y1_ev, y0_ev, Dev, X_ev, print_eval = FALSE)
ev_rf <- doubleml_did_eval_rf(y1_ev, y0_ev, Dev, X_ev, print_eval = FALSE)
ev_xgb <- doubleml_did_eval_xgboost(y1_ev, y0_ev, Dev, X_ev, print_eval = FALSE)

dplyr::bind_rows(
  tibble::as_tibble_row(unlist(ev_lin$eval_predictions)) |> dplyr::mutate(wrapper = "linear"),
  tibble::as_tibble_row(unlist(ev_rf$eval_predictions)) |> dplyr::mutate(wrapper = "rf"),
  tibble::as_tibble_row(unlist(ev_xgb$eval_predictions)) |> dplyr::mutate(wrapper = "xgboost")
) |>
  dplyr::relocate(wrapper) |>
  knitr::kable(digits = 4, caption = "Part A: illustrative nuisance scores (one simulated DiD slice)")

```

On a **linear** DGP, **`ml_g0`** RMSE is often lowest for the **linear** wrapper; flexible methods can still yield similar **ATT** if the Neyman-orthogonal score limits first-stage bias.


##  Part B: Use Case with `mpdta`

Callaway and Sant’Anna (2021) examined the causal impact of **higher state minimum wages** (above the federal level) on **teen employment** across U.S. counties, using data from roughly 2001–2007.

This demonstration uses a cleaned, balanced panel dataset from their study (called `mpdta` in the **did** R package; see the package documentation for details). The data is employed here purely as an example to illustrate key differences between the **DoubleML** and **did** packages. A similar notebook using the same data appears in the **did** package documentation.

Following the original example, we report results under two identification strategies:
- **Unconditional** parallel trends assumption
- **Conditional** parallel trends assumption (adjusting for covariates)

For **Double Machine Learning (DML)** DiD we fit **three RCausalML wrappers**—`doubleml_did_linear`, `doubleml_did_rf`, and `doubleml_did_xgboost`—and compare aggregated ATTs, dispersion of group-time estimates, and illustrative nuisance metrics (via `doubleml_did_eval_*`). Small differences versus default **did** are expected because of **cross-fitting** and flexible nuisances.

The **did** package includes the **mpdta** dataset: county-level teen employment (log employment) from 2003–2007, with variation in the timing of minimum-wage increases across states. We use it to illustrate DoubleML for DiD on real data.

### Load and Prepare mpdta

```{r mpdta-load-head}
data(mpdta, package = "did")

# Variables: countyreal (id), year (time), lemp (outcome), first.treat (cohort), lpop (covariate)
# first.treat = 0 means never treated; otherwise first period when treated
head(mpdta)
```

### Default did on mpdta

```{r attgt-default-mpdta}
attgt_mpdta <- att_gt(yname = "lemp",
                     tname = "year",
                     idname = "countyreal",
                     gname = "first.treat",
                     xformla = ~lpop,
                     data = mpdta)

summary(attgt_mpdta)
```

### DoubleML wrappers on mpdta (linear, RF, XGBoost)

```{r attgt-dml-three-wrappers-mpdta}
set.seed(1234)
attgt_mpdta_dml_lin <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~lpop,
  data = mpdta,
  est_method = as_did_est_method(doubleml_did_linear)
)

attgt_mpdta_dml_rf <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~lpop,
  data = mpdta,
  est_method = as_did_est_method(doubleml_did_rf)
)

attgt_mpdta_dml_xgb <- att_gt(
  yname = "lemp",
  tname = "year",
  idname = "countyreal",
  gname = "first.treat",
  xformla = ~lpop,
  data = mpdta,
  est_method = as_did_est_method(doubleml_did_xgboost)
)

summary(attgt_mpdta_dml_lin)
```

```{r summary-mpdta-dml-rf}
summary(attgt_mpdta_dml_rf)
```

```{r summary-mpdta-dml-xgb}
summary(attgt_mpdta_dml_xgb)
```

### Compare wrapper performance (Part B)

```{r compare-wrappers-part-b}
cmp_agg_b <- summarize_dml_wrappers(
  list(
    linear = attgt_mpdta_dml_lin,
    rf = attgt_mpdta_dml_rf,
    xgboost = attgt_mpdta_dml_xgb
  )
)
knitr::kable(cmp_agg_b, digits = 4, caption = "Part B: simple aggregate ATT by wrapper")

gt_mat_b <- tibble::tibble(
  linear = attgt_mpdta_dml_lin$att,
  rf = attgt_mpdta_dml_rf$att,
  xgboost = attgt_mpdta_dml_xgb$att
)

tibble::tibble(
  comparison = c("linear vs rf", "linear vs xgboost", "rf vs xgboost"),
  rmse_att_gt = c(
    sqrt(mean((gt_mat_b$linear - gt_mat_b$rf)^2, na.rm = TRUE)),
    sqrt(mean((gt_mat_b$linear - gt_mat_b$xgboost)^2, na.rm = TRUE)),
    sqrt(mean((gt_mat_b$rf - gt_mat_b$xgboost)^2, na.rm = TRUE))
  )
) |>
  knitr::kable(digits = 5, caption = "Part B: RMSE between group-time ATTs across wrappers")
```

### Nuisance fit (illustrative, two covariates like `~ lpop`)

```{r eval-nuisance-part-b}
set.seed(100)
n_ev2 <- 600
X_ev2 <- cbind(stats::rnorm(n_ev2), stats::rnorm(n_ev2))
colnames(X_ev2) <- c("x1", "x2")
D2 <- stats::rbinom(n_ev2, 1L, stats::plogis(0.2 * X_ev2[, 1L] - 0.15 * X_ev2[, 2L]))
delta2 <- 0.35 * D2 + 0.2 * X_ev2[, 1L] + stats::rnorm(n_ev2)
y0_2 <- stats::rnorm(n_ev2)
y1_2 <- y0_2 + delta2

ev2_lin <- doubleml_did_eval_linear(y1_2, y0_2, D2, X_ev2, print_eval = FALSE)
ev2_rf <- doubleml_did_eval_rf(y1_2, y0_2, D2, X_ev2, print_eval = FALSE)
ev2_xgb <- doubleml_did_eval_xgboost(y1_2, y0_2, D2, X_ev2, print_eval = FALSE)

dplyr::bind_rows(
  tibble::as_tibble_row(unlist(ev2_lin$eval_predictions)) |> dplyr::mutate(wrapper = "linear"),
  tibble::as_tibble_row(unlist(ev2_rf$eval_predictions)) |> dplyr::mutate(wrapper = "rf"),
  tibble::as_tibble_row(unlist(ev2_xgb$eval_predictions)) |> dplyr::mutate(wrapper = "xgboost")
) |>
  dplyr::relocate(wrapper) |>
  knitr::kable(digits = 4, caption = "Part B: illustrative nuisance scores (two covariates)")
```

### Aggregate effect (reference: linear wrapper)

```{r aggte-simple-mpdta}
agg_mpdta <- aggte(attgt_mpdta_dml_lin, type = "simple", na.rm = TRUE)
summary(agg_mpdta)
```

Interpretation: the aggregate ATT summarizes the average effect of minimum-wage treatment on log teen employment across group-time cells, with covariate adjustment (`lpop`) and cross-fitted nuisances via **RCausalML** + DoubleML. Compare `cmp_agg_b` for the same aggregation under RF and XGBoost nuisances.

### SHAP analysis

**`did::att_gt()`** with **`doubleml_did_*`** runs many internal 2×2 slices and returns **group-time ATTs** plus influence functions—it does **not** expose one global object with **`predict(object, newdata)`**. So **`RCausalML::explain_cate()`** (see **`R/shapviz_integration.R`**) does not apply *directly* to the `att_gt` fit. That helper wraps **`kernelshap::kernelshap()`** / **`kernelshap::permshap()`** with a default **`pred_fun`** based on **`rcausalml_predict_numeric()`**; you pass the result to **`shapviz::shapviz()`** for importance, dependence, waterfall, and force plots.

To **illustrate** the same workflow on data with a **first-difference outcome**, **binary treatment**, and **covariates** as in a single DiD slice, we fit **`LinearDML`** on the **Part B** toy sample (`x1`, `x2`, `D2`, and `y1_2 - y0_2` from the nuisance illustration above). SHAP values describe contributions to **predicted** $\\hat{\\tau}(X)$ from that DML learner—a **descriptive** heterogeneity summary, **not** the group-time ATT reported by **`did`**.

This section mirrors the **dynamic DML** notebook: we use **`explain_cate()`** with **{kernelshap}** (including **`permshap`**) and **{shapviz}** for visualization. Because SHAP is computationally expensive, we subsample rows for explanation and for the background (integration) distribution; use **non-overlapping** index sets when possible to avoid leakage. Parallel SHAP uses **{future}** when that package is available (`parallel = TRUE` in **`explain_cate()`**); otherwise estimation runs sequentially.

#### 1. SHAP estimation

SHAP values explain how each covariate contributes to the individual CATE—e.g. which features make the model believe the first-difference “effect” is larger or smaller for that row. We compute them with **`explain_cate()`**:

- **`X`** — rows you want to explain  
- **`bg_X`** — background (integration) distribution  
- Prefer **non-overlapping** explanation vs. background rows when the pool is large enough  

```{r shap-did-check}
#| message: false
has_ks <- requireNamespace("kernelshap", quietly = TRUE)
has_sv <- requireNamespace("shapviz", quietly = TRUE)
run_shap <- has_ks && has_sv
if (!run_shap) {
  message("Skipping SHAP: install.packages(c(\"kernelshap\", \"shapviz\")) to run this section.")
}
```

```{r shap-did-estimation-setup}
#| eval: !expr run_shap
has_future <- requireNamespace("future", quietly = TRUE)
if (has_future) {
  library(future)
  plan(multisession, workers = max(1L, parallel::detectCores() - 1L))
}

dy_shap <- as.numeric(y1_2 - y0_2)
X_df <- as.data.frame(X_ev2)
X_shap_pool <- X_df[!duplicated(X_df), , drop = FALSE]

set.seed(102)
ldml_shap <- LinearDML(model_y = "lm", model_t = "lm", n_fold = 5L, seed = 102)
ldml_shap <- fit(ldml_shap, X_df, D2, dy_shap)

nexplain <- min(80L, nrow(X_shap_pool))
nbg <- min(50L, nrow(X_shap_pool))

if (nrow(X_shap_pool) > nexplain + nbg) {
  idxexplain <- sample(nrow(X_shap_pool), nexplain)
  idxbg <- sample(setdiff(seq_len(nrow(X_shap_pool)), idxexplain), nbg)
} else {
  idxexplain <- sample(nrow(X_shap_pool), nexplain)
  idxbg <- sample(nrow(X_shap_pool), nbg)
}

Xexplain <- X_shap_pool[idxexplain, , drop = FALSE]
bg_X_df <- X_shap_pool[idxbg, , drop = FALSE]
```

```{r shap-did-estimation}
#| eval: !expr run_shap

fg_prev <- getOption("future.globals.maxSize", default = 500 * 1024^2)
if (has_future) {
  options(future.globals.maxSize = max(2 * 1024^3, fg_prev))
}
ks_args <- list(
  object = ldml_shap,
  X = Xexplain,
  bg_X = bg_X_df,
  use_permshap = TRUE,
  parallel = has_future,
  verbose = FALSE
)
if (has_future) {
  ks_args$parallel_args <- list(packages = c("RCausalML"))
}
ks <- tryCatch(
  do.call(explain_cate, ks_args),
  finally = {
    if (has_future) {
      future::plan(future::sequential)
      options(future.globals.maxSize = fg_prev)
    }
  }
)
```

#### 2. Visualize SHAP values

**2.1 Beeswarm importance plot** — distribution of SHAP values per feature; large absolute values mean a stronger influence on the predicted CATE.

```{r shap-did-importance-beeswarm}
#| fig-width: 6
#| fig-height: 5
#| eval: !expr run_shap

shp <- shapviz::shapviz(ks, X = Xexplain)
shapviz::sv_importance(shp, kind = "beeswarm") +
  ggplot2::labs(title = "SHAP (beeswarm): predicted CATE, LinearDML on DiD-style slice")
```

**2.2 Bar importance plot** — average |SHAP| per feature, useful for ranking.

```{r shap-did-importance-bar}
#| fig-width: 6
#| fig-height: 5
#| eval: !expr run_shap

shapviz::sv_importance(shp, show_numbers = TRUE) +
  ggplot2::labs(title = "SHAP importance (bar): predicted CATE, LinearDML on DiD-style slice")
```

#### 3. SHAP dependence plots

How each feature drives the predicted treatment effect (colour often encodes another variable automatically).

```{r shap-did-dependence-plots}
#| fig-width: 8
#| fig-height: 4
#| eval: !expr run_shap

xvars <- colnames(shp$X)
shapviz::sv_dependence(shp, v = xvars, share_y = TRUE)
```

For this toy DGP, larger **x1** shifts both treatment propensity and $\\Delta Y$; you may see SHAP for **x1** and **x2** align with those roles.

#### 4. SHAP for a single prediction

Waterfall and force plots show why the model assigns a high or low predicted CATE for one row (baseline, feature contributions, final value).

```{r shap-did-single-prediction}
#| fig-width: 6
#| fig-height: 5
#| eval: !expr run_shap

shapviz::sv_waterfall(shp, row_id = 1)
shapviz::sv_force(shp, row_id = 1)
```

#### 5. Subgroup-specific SHAP (optional)

Example: waterfall for the observation with the **largest x1** in the explanation sample (illustrative heterogeneity).

```{r shap-did-subgroup-waterfall}
#| fig-width: 6
#| fig-height: 5
#| eval: !expr run_shap

row_show <- if (nrow(shp$X) >= 1L) as.integer(which.max(shp$X$x1)) else 1L
shapviz::sv_waterfall(shp, row_id = row_show)
```

## Summary and Conclusion

This tutorial showed **DoubleML**-based DiD in R via **RCausalML** wrappers passed to **did**’s `att_gt()`. We compared default doubly robust **did** with three nuisance engines—**linear**, **random forest**, and **XGBoost**—and summarized performance with **`aggte(..., type = "simple")`**, **pairwise RMSE** of group-time ATTs, and **illustrative** `doubleml_did_eval_*` nuisance scores.

| Step | Description |
|------|-------------|
| **Data** | Long-format panel with cohort (`G` / `first.treat`), time, outcome, and covariates |
| **Default did** | `att_gt(..., xformla = ~ ...)` with built-in DR / regression / IPW |
| **DoubleML DiD** | `est_method = as_did_est_method(doubleml_did_*)` so **did** receives `ATT` and `att.inf.func` |
| **Wrappers** | `doubleml_did_linear`, `doubleml_did_rf`, `doubleml_did_xgboost`; eval variants `doubleml_did_eval_*` |
| **did tools** | `ggdid()`, `aggte()`, pre-test of parallel trends unchanged |


## Resources

- Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics*, 225(2), 200–230. <https://doi.org/10.1016/j.jeconom.2020.12.001>

- Sant'Anna, P. H. C., & Zhao, J. (2020). Doubly robust difference-in-differences estimators. *Journal of Econometrics*, 219(1), 101–122.

- [DoubleML R: DoubleML for Difference-in-Differences](https://docs.doubleml.org/stable/examples/R_double_ml_did.html)

- [did package (Callaway & Sant'Anna)](https://bcallaway11.github.io/did/)

- [DoubleML User Guide](https://docs.doubleml.org/stable/guide/index.html)
