# 8. Cluster Robust Double Machine Learning {.unnumbered}

This tutorial shows how to perform **cluster robust double machine learning (DML)** in R with the **DoubleML** package, while using **RCausalML**’s DoubleML-facing helpers from **`R/DMLearner.R`** wherever possible (notably **`doubleml_data_from_data_frame()`** to build `DoubleMLClusterData` and `DoubleMLData` from a `data.frame`). When errors are correlated within clusters (e.g., within regions, firms, or markets), the usual i.i.d. assumption fails and inference must account for clustering. We follow the [DoubleML R Cluster Robust example](https://docs.doubleml.org/stable/examples/R_double_ml_multiway_cluster.html) and implement **one-way** and **two-way** cluster robust DML using **simulated data** from Chiang et al. (2021) and **real-world data** (U.S. automobile demand, BLP 1995). 


## Overview

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Clustered data | Observations grouped in units where errors may be correlated. | "Students nested within schools." |
| Cluster-robust standard errors | Standard errors adjusted for within-cluster correlation. | "Do not treat students in same school as independent." |
| Intra-cluster correlation (ICC) | Similarity of observations within the same cluster. | "People in same village have similar outcomes." |
| Effective sample size | Independent information amount after accounting for clustering. | "1,000 rows may behave like only 400 independent units." |
| Design effect | Variance inflation caused by clustered sampling. | "Ignoring clustering gives confidence intervals that are too narrow." |
| Multiway clustering | Robust errors allowing correlation across multiple cluster dimensions. | "Cluster by firm and by year at the same time." |

When data are clustered, observations within the same cluster may be correlated, violating the i.i.d. assumption. For example, in a panel dataset, errors may be correlated within individuals over time (one-way clustering), or in a multi-dimensional setting, errors may be correlated within two dimensions (e.g., firms and markets) leading to two-way clustering.

Clustering affects DML in two ways:

1.  **Variance–covariance**: Standard errors, p-values, and confidence intervals must use a cluster robust variance estimator (as in Cameron et al. 2011).
2.  **Sample splitting**: Standard cross-fitting would put observations from the same cluster in both train and test folds, so errors would be correlated across folds. Chiang et al. (2021) propose splitting by **cluster**: for two-way clustering with $K$ folds per dimension, the data are split into $K^2$ folds so that nuisance functions are estimated on clusters disjoint from those used to evaluate the score.

**Cluster Robust Double Machine Learning (Cluster Robust DML)** is an extension of the standard Double Machine Learning framework designed for data where observations are **not independent**.

In standard DML, it is assumed that every data point (row) is independent and identically distributed (i.i.d.). However, in many real-world datasets, observations are grouped into **clusters** (e.g., students within schools, patients within hospitals, firms within industries, or repeated time periods for the same individual). Within these clusters, errors are correlated.

If you apply standard DML to clustered data, your **standard errors will be underestimated**, leading to invalid confidence intervals and false positives (Type I errors). Cluster Robust DML fixes this by adjusting both the **cross-fitting procedure** and the **variance estimation**.

### Why Is It Needed? (The Problem)

| Scenario | Assumption | Consequence if Ignored |
| :--- | :--- | :--- |
| **Standard DML** | Observations are independent (i.i.d.). | Valid inference. |
| **Clustered Data** | Observations within a group are correlated (e.g., $\text{Cov}(\epsilon_{i}, \epsilon_{j}) \neq 0$ if $i, j$ are in same cluster). | **Standard errors are too small.** <br> You might conclude a treatment effect is significant ($p < 0.05$) when it is actually noise. |


###  How It Works

Cluster Robust DML modifies two specific steps of the standard DML algorithm:

####  Clustered Cross-Fitting

In standard DML, you split data into $K$ folds randomly by row. In Cluster Robust DML, **you must split by Cluster ID.**

*   **Standard:** Fold 1 might contain Student A from Classroom 1, and Fold 2 might contain Student B from Classroom 1. This causes **data leakage** because the ML model learns information about Classroom 1 in Fold 1 and predicts on Fold 2.
*   **Cluster Robust:** All students from Classroom 1 must be in the *same* fold. This ensures that the nuisance models ($\hat{g}, \hat{m}$) are trained on completely independent clusters from the ones they are predicting on.

#### Cluster-Robust Variance Estimation

In the final step, when calculating the standard error for $\theta$, you cannot sum the scores across individual observations. You must **aggregate the scores at the cluster level first**.

1.  Calculate the orthogonal score $\psi_i$ for every observation.
2.  Sum these scores within each cluster: $\Psi_c = \sum_{i \in \text{cluster } c} \psi_i$.
3.  Calculate the variance based on the variation between $\Psi_c$ across clusters, not $\psi_i$ across observations.

$$ \hat{\sigma}^2_{cluster} = \frac{1}{C(C-1)} \sum_{c=1}^{C} (\Psi_c - \bar{\Psi})^2 $$

*(Where $C$ is the number of clusters, not observations)*



### When to Use Cluster Robust DML

You should use this variant whenever your data has a hierarchical or panel structure:

1.  **Panel Data:** Multiple time periods for the same unit (Cluster = Individual ID).
2.  **Multi-Site Experiments:** Participants nested within locations (Cluster = Site ID).
3.  **Network Data:** Individuals connected within a network component.
4.  **Survey Data:** Complex survey designs with stratification or clustering.
5.  **Difference-in-Differences:** Especially with staggered adoption where units are observed over time (Cluster = Unit ID).

### Advantages

*   **Valid Inference:** Prevents false positives caused by intra-cluster correlation.
*   **Flexible Nuisance Control:** Still allows you to use ML (XGBoost, RF) to control for high-dimensional confounders within the clusters.
*   **Easy Implementation:** In the `DoubleML` package, it requires only one additional argument (`cluster_cols`).

### Limitations

*   **Number of Clusters:** You need a sufficient number of **clusters**, not just observations.
    *   *Rule of Thumb:* If you have 10,000 observations but only 10 clusters, Cluster Robust DML will **not** work well. The asymptotic theory relies on $C \to \infty$ (number of clusters), not $N \to \infty$.
    *   Generally, you want at least 30–50 clusters for reliable inference.
*   **Imbalanced Clusters:** If cluster sizes vary wildly (e.g., one cluster has 5 observations, another has 5,000), variance estimation can become unstable.
*   **Computational Cost:** Clustered cross-fitting can be slightly slower if clusters are large, as the training/test splits are less granular.

## Implementation in R

The **DoubleML** package provides `DoubleMLClusterData`, `DoubleMLPLIV`, and the simulator **`make_pliv_multiway_cluster_CKMS2021()`** for the Chiang et al. (2021) DGP. **RCausalML** implements thin, documented wrappers around the DoubleML data backend in **`R/DMLearner.R`**: **`doubleml_data_from_data_frame()`** (and **`doubleml_data_from_matrix()`**) delegate to `DoubleML::double_ml_data_from_data_frame()` / `double_ml_data_from_matrix()`. This notebook **constructs BLP analysis data with `doubleml_data_from_data_frame()`** so the workflow is explicitly tied to **DMLearner.R**; the simulated CKMS2021 block still calls **DoubleML**’s generator directly (that DGP is not duplicated in RCausalML).

This tutorial covers:


| Topic                    | What we cover                                                                       |
|----------------------|-------------------------------------------------|
| **Clustered errors**     | Why i.i.d. inference fails; one-way vs. two-way clustering                          |
| **Cluster robust DML**   | Adjusted variance estimator and cluster-aware sample splitting (Chiang et al. 2021) |
| **RCausalML `R/DMLearner.R`** | **`doubleml_data_from_data_frame()`** for `DoubleMLClusterData` / `DoubleMLData` from a `data.frame` |
| **DoubleML**             | `make_pliv_multiway_cluster_CKMS2021()`, `DoubleMLPLIV`, cluster cross-fitting        |
| **Real data**            | BLP automobile data (hdm): clustering by product and market                         |

### Setup and Libraries


```{r}
#| label: setup-packages

packages <- c(
  "RCausalML",
  "DoubleML",
  "mlr3",
  "mlr3learners",
  "hdm",
  "ggplot2",
  "reshape2",
  "gridExtra",
  "lgr"
)
```

### Install Missing Packages

```{r}
#| label: setup-install-packages
#| warning: false
#| error: false

new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)
```

### Load Libraries and Options

```{r}
#| label: setup-load-libraries-source-dmlearner

invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

lgr::get_logger("mlr3")$set_threshold("warn")

# --- RCausalML R/DMLearner.R (doubleml_data_from_*, doubleml_plr, doubleml_pliv, …) ---
# When rendering from the package source tree, reload DMLearner.R so edits apply without reinstalling.
.qmd_dir <- tryCatch({
  .inp <- knitr::current_input()
  if (is.null(.inp) || !nzchar(.inp)) stop("no knitr input")
  dirname(normalizePath(.inp, winslash = "/", mustWork = FALSE))
}, error = function(e) NA_character_)
if (!is.na(.qmd_dir) && nzchar(.qmd_dir)) {
  .pkg_root <- normalizePath(file.path(.qmd_dir, "..", "..", ".."), mustWork = FALSE)
  .dml_r <- file.path(.pkg_root, "R", "DMLearner.R")
  if (file.exists(.dml_r)) {
    sys.source(.dml_r, envir = .GlobalEnv)
    message("Sourced RCausalML R/DMLearner.R from: ", .dml_r)
  }
} else {
  message("Using RCausalML from library(); DMLearner.R helpers already attached.")
}
stopifnot(exists("doubleml_data_from_data_frame", mode = "function"))
```


## Part 1: Two-Way Cluster Robust DML (Simulated Data)

We use the **partially linear IV (PLIV)** model and data generating process from Chiang et al. (2021). Data are double-indexed: $\{W_{ij}: i=1,\ldots,N,\, j=1,\ldots,M\}$.

**Model:**

$$Y_{ij} = D_{ij} \theta_0 + g_0(X_{ij}) + \varepsilon_{ij}, \quad \mathbb{E}(\varepsilon_{ij} \mid X_{ij}, Z_{ij}) = 0$$

$$Z_{ij} = m_0(X_{ij}) + V_{ij}, \quad \mathbb{E}(V_{ij} \mid X_{ij}) = 0$$

Errors and covariates can be correlated within dimension $i$, within dimension $j$, or both (two-way clustering).

### Simulate Two-Way Cluster Data

DoubleML provides `make_pliv_multiway_cluster_CKMS2021()` to generate data from this DGP. We use the default parameters (e.g. $\theta=1$, $N=M=25$, $p_x=100$).

```{r}
#| label: sim-twoway-ckms-generate

# Simulation parameters (Chiang et al. 2021, Section 5)
N <- 25   # clusters, first dimension
M <- 25   # clusters, second dimension
dim_X <- 100
set.seed(3141)

obj_dml_data <- make_pliv_multiway_cluster_CKMS2021(N, M, dim_X)
```

### Data Backend for Cluster Data

Cluster robust DML uses the **DoubleMLClusterData** backend. Cluster variables are specified at construction (or set later). The object then carries cluster information for both sample splitting and variance estimation.

```{r}
#| label: sim-twoway-print-backend

# Simulated data is already DoubleMLClusterData with two cluster variables
print(obj_dml_data)
```

```{r}
#| label: sim-twoway-head-cluster-cols

# Cluster variables are in the data
head(obj_dml_data$data[, c("Y", "D", "Z", "cluster_var_i", "cluster_var_j")])
```

### Initialize `DoubleMLPLIV` with Cluster Robust Cross-Fitting

We set ML learners for the three nuisances (l, m, r) and use $K=3$ folds per cluster dimension, giving $K^2=9$ folds in total.

```{r}
#| label: sim-twoway-pliv-init

# ML methods for l, m, r (cross-validated lasso)
ml_l <- lrn("regr.cv_glmnet", nfolds = 10, s = "lambda.min")
ml_m <- lrn("regr.cv_glmnet", nfolds = 10, s = "lambda.min")
ml_r <- lrn("regr.cv_glmnet", nfolds = 10, s = "lambda.min")

dml_pliv_obj <- DoubleMLPLIV$new(obj_dml_data,
                                 ml_l, ml_m, ml_r,
                                 n_folds = 3)

print(dml_pliv_obj)
```

The output shows **No. folds: 9** (i.e. $3^2$) and **Cluster variable(s): cluster_var_i, cluster_var_j**.

### Fit and Cluster Robust Inference

`fit()` uses cluster-based sample splitting; standard errors and confidence intervals are computed with the cluster robust variance estimator from Chiang et al. (2021).

```{r}
#| label: sim-twoway-pliv-fit-summary

dml_pliv_obj$fit()
dml_pliv_obj$summary()
```

The estimate is close to the true $\theta_0 = 1$, and the reported standard errors and p-values are valid under two-way clustering.

## Part 2: One-Way Cluster Robust DML (Simulated Data)

For **one-way clustering**, correlation exists only along one dimension. We use the same DGP but set the second-dimension weights to zero so that only the first cluster variable matters. We then restrict the data backend to a single cluster variable.

### Simulate One-Way Cluster Data

```{r}
#| label: sim-oneway-ckms-generate

obj_dml_data_oneway <- make_pliv_multiway_cluster_CKMS2021(
  N, M, dim_X,
  omega_X = c(0.25, 0),
  omega_epsilon = c(0.25, 0),
  omega_v = c(0.25, 0),
  omega_V = c(0.25, 0)
)
```

### Use a Single Cluster Variable

```{r}
#| label: sim-oneway-set-cluster-cols

obj_dml_data_oneway$cluster_cols <- "cluster_var_i"
print(obj_dml_data_oneway)
```

### Fit One-Way Cluster Robust PLIV

With one-way clustering, the number of folds equals $K$ (e.g. 3), not $K^2$.

```{r}
#| label: sim-oneway-pliv-fit-summary

dml_pliv_oneway <- DoubleMLPLIV$new(obj_dml_data_oneway,
                                    ml_l, ml_m, ml_r,
                                    n_folds = 3)
dml_pliv_oneway$fit()
dml_pliv_oneway$summary()
```


## Part 3: Real-Data Application (BLP Automobile Data)

We replicate the consumer demand application in Chiang et al. (2021) using the **Berry, Levinsohn, and Pakes (1995)** automobile data from the **hdm** package. The outcome is demand (y), the treatment is log price, and we use instruments and covariates. Observations are clustered by **product** (model.id) and **market** (cdid); we compare **no clustering**, **one-way** (product or market), and **two-way** clustering.

### Load and Prepare BLP Data

```{r}
#| label: blp-load-prepare
#| warning: false

data(BLP, package = "hdm")
blp_data <- BLP$BLP
blp_data$price <- blp_data$price + 11.761
blp_data$log_p <- log(blp_data$price)
```

```{r}
#| label: blp-head-covariates

x_cols <- c("hpwt", "air", "mpd", "space")
head(blp_data[x_cols])
```

### Construct IV and Covariate Matrix

We use the same formula as in the DoubleML documentation to build the design matrix and instruments.

```{r}
#| label: blp-iv-and-model-matrix

iv_vars <- as.data.frame(hdm:::constructIV(blp_data$firm.id,
                                           blp_data$cdid,
                                           blp_data$id,
                                           blp_data[x_cols]))

formula <- formula(paste0(" ~ -1 + (hpwt + air + mpd + space)^2",
                         "+ I(hpwt^2)*(air + mpd + space)",
                         "+ I(air^2)*(hpwt + mpd + space)",
                         "+ I(mpd^2)*(hpwt + air + space)",
                         "+ I(space^2)*(hpwt + air + mpd)",
                         "+ I(space^2) + I(hpwt^3) + I(air^3) + I(mpd^3) + I(space^3)"))
data_transf <- data.frame(model.matrix(formula, blp_data))
```

```{r}
#| label: blp-assemble-dml-df

y_col <- "y"
d_col <- "log_p"
cluster_cols <- c("model.id", "cdid")
all_z_cols <- c("sum.other.hpwt", "sum.other.mpd", "sum.other.space")
z_col <- all_z_cols[1]

dml_df <- cbind(blp_data[c(y_col, d_col, cluster_cols)],
                data_transf,
                iv_vars[all_z_cols])
```

### Initialize DoubleMLClusterData for Real Data (via `R/DMLearner.R`)

We use **`doubleml_data_from_data_frame()`** from **RCausalML** [`R/DMLearner.R`](../../../R/DMLearner.R), which wraps **`DoubleML::double_ml_data_from_data_frame()`**. Passing **`cluster_cols`** returns a **`DoubleMLClusterData`** object.

```{r}
#| label: blp-doubleml-cluster-backend

dml_data <- doubleml_data_from_data_frame(
  dml_df,
  y_col = y_col,
  d_cols = d_col,
  z_cols = z_col,
  cluster_cols = cluster_cols,
  x_cols = names(data_transf)
)
stopifnot(!is.null(dml_data), inherits(dml_data, "DoubleMLClusterData"))
print(dml_data)
```

### Compare Specifications: Zero-, One-, and Two-Way Clustering

We fit the PLIV model under four specifications and store coefficients and standard errors.

```{r}
#| label: blp-coef-se-initialize

lasso <- lrn("regr.cv_glmnet", nfolds = 10, s = "lambda.min")
coef_df <- data.frame(matrix(NA_real_, ncol = 4, nrow = 1))
colnames(coef_df) <- c("zero_way", "one_way_product", "one_way_market", "two_way")
rownames(coef_df) <- all_z_cols[1]
se_df <- coef_df
n_rep <- 10
```

#### Two-way clustering (product and market)

```{r}
#| label: blp-fit-two-way-clustering

set.seed(1111)
dml_data$z_cols <- z_col
dml_data$cluster_cols <- c("model.id", "cdid")
dml_pliv <- DoubleMLPLIV$new(dml_data, lasso, lasso, lasso,
                             n_folds = 2, n_rep = n_rep)
dml_pliv$fit()
coef_df[1, 4] <- dml_pliv$coef
se_df[1, 4] <- dml_pliv$se
```

#### One-way clustering (product)

```{r}
#| label: blp-fit-oneway-product

set.seed(2222)
dml_data$cluster_cols <- "model.id"
dml_pliv <- DoubleMLPLIV$new(dml_data, lasso, lasso, lasso,
                             n_folds = 4, n_rep = n_rep)
dml_pliv$fit()
coef_df[1, 2] <- dml_pliv$coef
se_df[1, 2] <- dml_pliv$se
```

#### One-way clustering (market)

```{r}
#| label: blp-fit-oneway-market

set.seed(3333)
dml_data$cluster_cols <- "cdid"
dml_pliv <- DoubleMLPLIV$new(dml_data, lasso, lasso, lasso,
                             n_folds = 4, n_rep = n_rep)
dml_pliv$fit()
coef_df[1, 3] <- dml_pliv$coef
se_df[1, 3] <- dml_pliv$se
```

#### No clustering (standard DoubleMLData)

With **`cluster_cols = NULL`**, **`doubleml_data_from_data_frame()`** builds a standard **`DoubleMLData`** object (same **DMLearner.R** helper).

```{r}
#| label: blp-fit-zero-way-no-clustering

dml_data_no_cluster <- doubleml_data_from_data_frame(
  dml_df,
  y_col = y_col,
  d_cols = d_col,
  z_cols = z_col,
  x_cols = names(data_transf),
  cluster_cols = NULL
)
stopifnot(!is.null(dml_data_no_cluster), inherits(dml_data_no_cluster, "DoubleMLData"))
set.seed(4444)
dml_pliv <- DoubleMLPLIV$new(dml_data_no_cluster, lasso, lasso, lasso,
                             n_folds = 4, n_rep = n_rep)
dml_pliv$fit()
coef_df[1, 1] <- dml_pliv$coef
se_df[1, 1] <- dml_pliv$se
```

### Application Results

```{r}
#| label: blp-results-coefficients

coef_df
```

```{r}
#| label: blp-results-standard-errors

se_df
```

Interpretation: the **price coefficient** (effect of log price on demand) is similar across specifications, but **standard errors increase** as we account for clustering. Two-way clustering yields the largest standard errors because we allow correlation both within products and within markets. Inference that ignores clustering would be anti-conservative.


## Summary and Conclusion

**Cluster Robust DML** is the "production-ready" version of Double Machine Learning. Since most observational data in economics, social sciences, and environmental studies is clustered (by time, location, or group), **you should default to Cluster Robust DML** whenever a cluster identifier is available. It preserves the causal validity of DML while respecting the dependence structure of your data. This tutorial used **simulated data** from the PLIV multiway cluster DGP (Chiang et al. 2021) and **real data** (BLP automobile demand) to illustrate one-way and two-way cluster robust double machine learning in R, following the [DoubleML R Cluster Robust example](https://docs.doubleml.org/stable/examples/R_double_ml_multiway_cluster.html). **BLP data backends** were built with **RCausalML** **`doubleml_data_from_data_frame()`** from **`R/DMLearner.R`**. This tutorial covered:


| Step                   | Description                                                                               |
|-----------------------|-------------------------------------------------|
| **Clustered data**     | **`doubleml_data_from_data_frame(..., cluster_cols = ...)`** → `DoubleMLClusterData` (see **`R/DMLearner.R`**) |
| **Two-way clustering** | Sample splitting uses $K^2$ folds; variance estimator from Chiang et al. (2021)           |
| **One-way clustering** | Set `cluster_cols` to a single variable; $K$ folds                                        |
| **No clustering**      | **`doubleml_data_from_data_frame(..., cluster_cols = NULL)`** → `DoubleMLData`            |
| **Real data**          | BLP data: cluster by product and/or market; cluster robust SEs are larger than non-robust |


## Resources

-   Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile prices in market equilibrium. *Econometrica*, 63(4), 841–890. <https://doi.org/10.2307/2171802>
-   Cameron, A. C., Gelbach, J. B., & Miller, D. L. (2011). Robust inference with multiway clustering. *Journal of Business & Economic Statistics*, 29(2), 238–249. <https://doi.org/10.1198/jbes.2010.07136>

-   Chiang, H. D., Kato, K., Ma, Y., & Sasaki, Y. (2021). Multiway cluster robust double/debiased machine learning. *Journal of Business & Economic Statistics*. <https://doi.org/10.1080/07350015.2021.1895815>

-   Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W., & Robins, J. (2018). Double/debiased machine learning for treatment and structural parameters. *The Econometrics Journal*, 21(1), C1–C68. <https://doi.org/10.1111/ectj.12097>

-   [DoubleML R: Cluster Robust Double Machine Learning](https://docs.doubleml.org/stable/examples/R_double_ml_multiway_cluster.html)

-   [DoubleML User Guide](https://docs.doubleml.org/stable/guide/index.html)

-   RCausalML **`R/DMLearner.R`**: `doubleml_data_from_data_frame()`, `doubleml_data_from_matrix()`, `doubleml_plr`, `doubleml_pliv`, DiD helpers (`doubleml_did_*`), and related DoubleML bridges

-   [hdm package (high-dimensional models)](https://cran.r-project.org/package=hdm)