# 9. Boosted Multi-Arm Causal Model {.unnumbered}

A **Boosted Multi-Arm Causal Model** estimates heterogeneous treatment effects (HTE) when there are **multiple treatment arms** (e.g., `placebo`, `A`, `B`) using a practical, model-based approach:

- **Outcome models (T-learner style)**: fit one boosted model per treatment arm to learn $\mu_k(x) = \mathbb{E}[Y \mid X = x, W = k]$.
- **Treatment contrasts**: compute $\tau_{k-b}(x) = \mu_k(x) - \mu_b(x)$ relative to a chosen **baseline** arm $b$.
- **Propensity scores (optional, for diagnostics/weights)**: estimate $\Pr(W = k \mid X)$ with a multiclass probability forest.

In `RCausalML`, this is implemented by the R6 class `MultiArmCausalBoost` in `R/multi_arm_causal_boost.R`.

## Overview

This tutorial shows how to:

- Prepare a multi-arm dataset (simulated treatments) with **two outcomes** (`log_time`, `health_status`)
- **Train** `MultiArmCausalBoost` on both outcomes jointly (one booster per arm and per outcome)
- **Predict** multi-arm contrasts on the test set (`tau_hat_log_time`, `tau_hat_health_status`)
- Estimate **doubly robust (AIPW-style) average treatment effects** on the training sample (`ate_log_time`, `ate_health_status`)
- Plot **variable importance** for each outcome (XGBoost gain)
- Compute **kernel SHAP** for predicted CATE with `{kernelshap}` and visualize with `{shapviz}`
- (Optional) save/load a fitted model

## Scientific Terminology for Beginners

This glossary introduces key scientific terms used in this notebook with simple examples.

### 1) Multi-Arm Treatment

- **Explanation**: A study with more than two treatment options.
- **Example**: `placebo`, `A`, and `B` are three arms.

### 2) Baseline Arm

- **Explanation**: Reference treatment used for effect comparisons.
- **Example**: Compare each active arm against `placebo`.

### 3) Outcome Model

- **Explanation**: A model that predicts expected outcome within each treatment arm.
- **Example**: One boosted model estimates outcomes for arm `A`.

### 4) Treatment Contrast

- **Explanation**: Difference in predicted outcomes between two arms.
- **Example**: `tau_A-control(x) = mu_A(x) - mu_control(x)`.

### 5) Propensity Score

- **Explanation**: Probability of receiving each treatment arm given covariates.
- **Example**: A subject has probabilities (0.3, 0.5, 0.2) across three arms.

### 6) AIPW

- **Explanation**: A doubly robust estimator that combines outcome and propensity models.
- **Example**: ATE remains consistent if either model is correctly specified.

### 7) SHAP Values

- **Explanation**: Feature attributions explaining how variables affect predictions.
- **Example**: Age contributes +0.4 to the predicted treatment contrast.

### 8) Heterogeneous Treatment Effects

- **Explanation**: Effects vary across subgroups with different covariates.
- **Example**: Arm B helps severe cases more than mild cases.

### Load and Check Required Libraries

```{r load-packages}
packages <- c(
  "tidyverse",
  "survival",
  "RCausalML",
  "xgboost",
  "ranger",
  "ggplot2",
  "kernelshap",
  "shapviz",
  "patchwork"
)
```

### Install Missing Packages

```{r install-missing-packages}
#| warning: false
#| error: false
#new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r verify-installation}
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

## Load Required Libraries

```{r load-libraries}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

```{r load-multi-arm-causal-boost-impl}
#| warning: false
#| error: false
if (!exists("MultiArmCausalBoost")) {
  wd <- getwd()
  impl_path <- NULL

  for (i in 0:8) {
    candidate <- file.path(wd, "R", "multi_arm_causal_boost.R")
    if (file.exists(candidate)) {
      impl_path <- candidate
      break
    }
    parent <- dirname(wd)
    if (identical(parent, wd)) break
    wd <- parent
  }

  if (is.null(impl_path)) {
    stop("MultiArmCausalBoost not found, and could not locate 'R/multi_arm_causal_boost.R' by walking up from getwd().")
  }

  source(impl_path)
}
```

### Data: Lung dataset with simulated multi-arm treatment

We use `lung` from `{survival}` and simulate a 3-arm treatment:

- `placebo` (baseline)
- `A`
- `B`

We also create two outcomes:

- `log_time`: log survival time
- `health_status`: a simulated binary secondary outcome with treatment-dependent probabilities

```{r load-data}
#| warning: false
#| error: false

data(lung, package = "survival")

set.seed(123)
lung_data <- lung %>%
  dplyr::filter(!is.na(age), !is.na(sex), !is.na(ph.ecog), !is.na(time)) %>%
  dplyr::mutate(
    treatment = as.factor(sample(
      c("placebo", "A", "B"),
      n(),
      replace = TRUE,
      prob = c(0.4, 0.3, 0.3)
    )),
    health_status = rbinom(
      n(),
      1,
      prob = 0.3 + 0.10 * (treatment == "A") + 0.15 * (treatment == "B")
    ),
    log_time = log(time)
  )

# For a fast, reproducible tutorial render, keep a moderate sample size.
set.seed(123)
n_keep <- min(250, nrow(lung_data))
lung_data <- lung_data %>% dplyr::slice_sample(n = n_keep)

# Covariates, treatment, outcomes
X <- lung_data %>% dplyr::select(age, sex, ph.ecog) %>% as.matrix()
W <- relevel(lung_data$treatment, ref = "placebo")
Y <- lung_data %>% dplyr::select(log_time, health_status) %>% as.matrix()
colnames(Y) <- c("log_time", "health_status")

# Train/test split
set.seed(123)
n <- nrow(lung_data)
train_idx <- sample(seq_len(n), size = floor(0.8 * n))
test_idx  <- setdiff(seq_len(n), train_idx)

X_train <- X[train_idx, , drop = FALSE]
W_train <- W[train_idx]
Y_train <- Y[train_idx, , drop = FALSE]

X_test <- X[test_idx, , drop = FALSE]
W_test <- W[test_idx]
Y_test <- Y[test_idx, , drop = FALSE]
```

### Train `MultiArmCausalBoost` on `log_time` and `health_status`

`MultiArmCausalBoost` fits:

- one XGBoost model per treatment arm **and** per outcome (here: `log_time` and `health_status`)
- one multiclass propensity model $\Pr(W \mid X)$ via `ranger` (probability forest)

Key choices you control:

- **baseline**: the reference arm for contrasts (we use `placebo`)
- **xgboost parameters**: passed as `parameters = list(...)` (same objective for both outcomes here; binary `health_status` uses squared error for a smooth tutorial)
- **nrounds**: boosting rounds (kept moderate for tutorial runtime)

```{r fit-model}
#| warning: false
#| error: false

model <- MultiArmCausalBoost$new(
  parameters = list(
    eta = 0.10,
    max_depth = 3L,
    subsample = 0.8,
    colsample_bytree = 0.8,
    tree_method = "hist",
    objective = "reg:squarederror"
  ),
  baseline = "placebo"
)

model$fit(
  X = X_train,
  Y = Y_train,
  W = W_train,
  nrounds = 10L,
  verbose = 1L
)

model$summary()
```

## Predict contrasts on the test set (`tau_hat_*`)

The `predict()` method returns a list:

- `mu_hat`: potential outcomes array $[n, \text{arms}, \text{outcomes}]$
- `tau_hat`: contrasts array $[n, (\text{arms} - 1), \text{outcomes}]$ relative to `baseline`
- `propensity_score`: matrix $[n, \text{arms}]$

Below we store contrasts for each outcome as $n \times (\text{number of contrasts})$ matrices.

```{r predict-test}
#| warning: false
#| error: false

pred <- model$predict(X_test, drop = FALSE)

tau_hat_log_time <- pred$tau_hat[, , "log_time", drop = TRUE]
tau_hat_health_status <- pred$tau_hat[, , "health_status", drop = TRUE]
cn_tau <- dimnames(pred$tau_hat)[[2]]
if (is.null(dim(tau_hat_log_time))) {
  tau_hat_log_time <- matrix(tau_hat_log_time, ncol = 1L, dimnames = list(NULL, cn_tau))
} else {
  colnames(tau_hat_log_time) <- cn_tau
}
if (is.null(dim(tau_hat_health_status))) {
  tau_hat_health_status <- matrix(
    tau_hat_health_status,
    ncol = 1L,
    dimnames = list(NULL, cn_tau)
  )
} else {
  colnames(tau_hat_health_status) <- cn_tau
}

str(pred$mu_hat)
str(pred$tau_hat)
head(tau_hat_log_time)
head(tau_hat_health_status)
head(pred$propensity_score)
```

## Doubly robust average treatment effects (AIPW plug-in)

For each non-baseline arm $k$ and outcome $m$, we use the standard multi-arm doubly robust score (same spirit as AIPW), evaluated at the fitted outcome models $\hat\mu_a(X)$ and propensities $\hat e_a(X)$:

$$
\psi_i^{(k)} = \hat\mu_k(X_i) - \hat\mu_b(X_i)
+ \bigl(Y_{im} - \hat\mu_{W_i}(X_i)\bigr)\left(\frac{\mathbb{1}\{W_i=k\}}{\hat e_k(X_i)} - \frac{\mathbb{1}\{W_i=b\}}{\hat e_b(X_i)}\right),
$$

with baseline $b$ = `placebo`. We average $\psi_i^{(k)}$ over the **training** sample (where $Y$ is observed). This is a **plug-in** DR estimator (nuisances fit on the same data); for rigorous inference, sample splitting or cross-fitting is preferred.

```{r aipw-ate}
#| warning: false
#| error: false

multi_arm_dr_ate <- function(Y, W, mu_hat, ps, baseline) {
  W_char <- as.character(droplevels(W))
  arms <- dimnames(mu_hat)[[2]]
  outcomes <- dimnames(mu_hat)[[3]]
  n <- nrow(Y)
  M <- dim(mu_hat)[3]
  arm_idx <- match(W_char, arms)
  contrast_arms <- setdiff(arms, baseline)
  cn <- paste0(contrast_arms, " - ", baseline)
  out <- matrix(
    NA_real_,
    nrow = length(contrast_arms),
    ncol = M,
    dimnames = list(cn, outcomes)
  )
  idx_row <- seq_len(n)
  for (m in seq_len(M)) {
    mu_w <- mu_hat[cbind(idx_row, arm_idx, m)]
    for (j in seq_along(contrast_arms)) {
      k <- contrast_arms[j]
      mu_k <- mu_hat[, k, m]
      mu_b <- mu_hat[, baseline, m]
      e_k <- ps[, k]
      e_b <- ps[, baseline]
      I_k <- as.numeric(W_char == k)
      I_b <- as.numeric(W_char == baseline)
      psi <- (mu_k - mu_b) + (Y[, m] - mu_w) * (I_k / e_k - I_b / e_b)
      out[j, m] <- mean(psi)
    }
  }
  out
}

pred_train <- model$predict(X_train, drop = FALSE)
ate_tab <- multi_arm_dr_ate(
  Y_train,
  W_train,
  pred_train$mu_hat,
  pred_train$propensity_score,
  baseline = model$baseline
)
ate_log_time <- ate_tab[, "log_time"]
ate_health_status <- ate_tab[, "health_status"]

print(ate_tab)
```

### Summarize average (model-implied) contrasts on the test set

These are **not** doubly-robust estimates; they’re simply averages of the model’s predicted individual contrasts (use `ate_log_time` / `ate_health_status` above for DR averages on the training sample).

```{r avg-contrasts}
#| warning: false
#| error: false

tau_test <- pred$tau_hat

avg_tau <- apply(tau_test, c(2, 3), mean)
avg_tau_df <- as.data.frame(as.table(avg_tau))
colnames(avg_tau_df) <- c("contrast", "outcome", "avg_contrast")

avg_tau_df <- tibble::as_tibble(avg_tau_df)
avg_tau_df %>%
  dplyr::arrange(outcome, contrast) %>%
  print(n = Inf)
```

### Visualize heterogeneity in a contrast

As a simple check, plot a contrast against a key covariate.

```{r cate-vs-age}
#| fig.width: 7
#| fig.height: 4.5
#| warning: false
#| error: false

contrast_names <- dimnames(pred$tau_hat)[[2]]
outcome_names <- dimnames(pred$tau_hat)[[3]]

# pick the first available contrast/outcome for a compact demo
chosen_contrast <- contrast_names[1]
chosen_outcome  <- outcome_names[1]

tau_slice <- if (identical(chosen_outcome, "log_time")) {
  tau_hat_log_time
} else {
  tau_hat_health_status
}
plot_df <- tibble::tibble(
  age = X_test[, "age"],
  tau = tau_slice[, chosen_contrast]
)

ggplot(plot_df, aes(x = age, y = tau)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, color = "#4E79A7") +
  labs(
    title = paste0("Estimated contrast vs age: ", chosen_contrast, " (", chosen_outcome, ")"),
    x = "age",
    y = "Estimated treatment contrast"
  ) +
  theme_minimal()
```

## Variable importance: `log_time` and `health_status`

Because the model fits **one booster per (arm, outcome)**, gain-based importance is specific to an arm/outcome pair. Below: arm `A` with outcome index `1L` = `log_time`, then `2L` = `health_status` (`plot_importance()` takes a numeric outcome index).

```{r importance-log-time}
#| fig.width: 7
#| fig.height: 4.5
#| warning: false
#| error: false

model$plot_importance(arm = "A", outcome = 1L, top_n = 10L, metric = "Gain")
```

```{r importance-health-status}
#| fig.width: 7
#| fig.height: 4.5
#| warning: false
#| error: false

model$plot_importance(arm = "A", outcome = 2L, top_n = 10L, metric = "Gain")
```

### Access the underlying XGBoost model directly

If you need custom XGBoost tooling (e.g., advanced importance, SHAP from `xgboost`), you can retrieve a fitted booster:

```{r extract-booster}
#| warning: false
#| error: false

booster_A_log_time <- model$outcome_models[["A"]][["log_time"]]
booster_A_log_time
```

### Kernel SHAP for predicted CATE (`shapviz` + `kernelshap`)

We explain the **predicted contrast** $\hat\tau_k(X) = \hat\mu_k(X) - \hat\mu_{\text{placebo}}(X)$ as a function of $X$, separately for `log_time` and `health_status`. Use a small background set and a subsample of the test rows so kernel SHAP stays fast.

```{r kernel-shap-cate}
#| fig.width: 10
#| fig.height: 10
#| warning: false
#| error: false

library(kernelshap)
library(shapviz)
library(patchwork)

pred_A_vs_placebo_log <- function(object, newdata) {
  pr <- object$predict(as.matrix(newdata), drop = FALSE)
  pr$tau_hat[, "A - placebo", "log_time"]
}

pred_B_vs_placebo_log <- function(object, newdata) {
  pr <- object$predict(as.matrix(newdata), drop = FALSE)
  pr$tau_hat[, "B - placebo", "log_time"]
}

pred_A_vs_placebo_health <- function(object, newdata) {
  pr <- object$predict(as.matrix(newdata), drop = FALSE)
  pr$tau_hat[, "A - placebo", "health_status"]
}

pred_B_vs_placebo_health <- function(object, newdata) {
  pr <- object$predict(as.matrix(newdata), drop = FALSE)
  pr$tau_hat[, "B - placebo", "health_status"]
}

set.seed(42)
n_bg <- min(50L, nrow(X_train))
bg_idx <- sample.int(nrow(X_train), n_bg)
bg_small <- X_train[bg_idx, , drop = FALSE]

n_explain <- min(40L, nrow(X_test))
X_shap <- X_test[seq_len(n_explain), , drop = FALSE]

shap_A_log <- kernelshap(
  object = model,
  X = X_shap,
  bg_X = bg_small,
  pred_fun = pred_A_vs_placebo_log,
  verbose = FALSE
)

shap_B_log <- kernelshap(
  object = model,
  X = X_shap,
  bg_X = bg_small,
  pred_fun = pred_B_vs_placebo_log,
  verbose = FALSE
)

shap_A_health <- kernelshap(
  object = model,
  X = X_shap,
  bg_X = bg_small,
  pred_fun = pred_A_vs_placebo_health,
  verbose = FALSE
)

shap_B_health <- kernelshap(
  object = model,
  X = X_shap,
  bg_X = bg_small,
  pred_fun = pred_B_vs_placebo_health,
  verbose = FALSE
)

shp_A_log <- shapviz(shap_A_log)
shp_B_log <- shapviz(shap_B_log)
shp_A_health <- shapviz(shap_A_health)
shp_B_health <- shapviz(shap_B_health)

plt1 <- sv_importance(shp_A_log, kind = "both") + ggplot2::ggtitle("CATE SHAP — log_time: A vs placebo")
plt2 <- sv_importance(shp_B_log, kind = "both") + ggplot2::ggtitle("CATE SHAP — log_time: B vs placebo")
plt3 <- sv_importance(shp_A_health, kind = "both") + ggplot2::ggtitle("CATE SHAP — health_status: A vs placebo")
plt4 <- sv_importance(shp_B_health, kind = "both") + ggplot2::ggtitle("CATE SHAP — health_status: B vs placebo")

(plt1 | plt2) / (plt3 | plt4)
```

## Save and load a fitted model

`RCausalML` includes helpers to persist the fitted R6 object.

```{r save-load}
#| warning: false
#| error: false

tmp_path <- tempfile(fileext = ".rds")
save_multi_arm_causal_boost(model, tmp_path)

model2 <- load_multi_arm_causal_boost(tmp_path)
model2$summary()
```

## Notes and limitations

- Outcome models are **T-learner style** (separate boosters per arm); the **DR/AIPW averages** above reuse those fits plus the ranger propensities—they are not from a dedicated DR-learner objective.
- DR estimates use **plug-in nuisances** on the training set; prefer **cross-fitting** for formal inference.
- You need adequate sample size **per arm**, since each arm has its own outcome model(s).
- For multi-outcome $Y$, the implementation fits one booster per outcome per arm (no explicit outcome correlation modeling).
- Kernel SHAP explains **predicted** CATE $\hat\tau(X)$, not necessarily the structural causal effect unless identification holds.

## Summary

You now have a practical workflow for multi-arm causal effect modeling with boosting:

- fit `MultiArmCausalBoost` once on multi-outcome $Y$ (`log_time`, `health_status`)
- predict contrasts on the test set (`tau_hat_log_time`, `tau_hat_health_status`)
- summarize **doubly robust** average contrasts on the training sample (`ate_log_time`, `ate_health_status`)
- interpret drivers with per-outcome gain importance and kernel SHAP for CATE

## Resources

1.  Künzel, S. R., Sekhon, J. S., Bickel, P. J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. PNAS.
2.  `{xgboost}` documentation for boosted trees and model interpretation.
