# 6. Multi-arm/multi-outcome Causal Forest {.unnumbered}

A **Multi-Arm/Multi-Outcome Causal Forest** is an extension of the **Causal Forest** framework designed to estimate heterogeneous treatment effects in scenarios with **multiple treatment arms** (e.g., different interventions) and/or **multiple outcome variables** simultaneously. It builds on random forest principles to provide flexible, data-driven estimates of causal effects across diverse subgroups, leveraging shared information across treatments or outcomes to improve accuracy. This approach is particularly useful in complex experimental designs, such as clinical trials with multiple drugs or policy evaluations with various interventions affecting multiple metrics.

## Overview

A multi-arm/multi-outcome causal forest is an advanced A multi-arm/multi-outcome causal forest is an advanced machine learning method for estimating heterogeneous treatment effects (HTE) in settings with:

-   `Multiple treatments (multi-arm)`: More than two possible interventions or choices, not just "treatment" and "control".

-   `Multiple outcomes (multi-outcome)`: Several different outcome variables of interest, not just a single outcome.

Estimates causal effects (treatment effects) for multiple treatments (arms) or multiple outcomes in a single model, accounting for heterogeneity across covariates (e.g., patient characteristics). It answers questions like: "How does each treatment affect different outcomes across different subgroups?" Handles multiple treatment options (e.g., drug A, drug B, placebo) instead of a single treatment vs. control.

-   Use Cases:

  -   Clinical trials comparing multiple drugs on multiple health outcomes.
  -   Policy evaluation with several interventions and diverse impacts (e.g., education programs affecting test scores and attendance).
  -   Marketing studies analyzing multiple campaign strategies on various customer metrics.

## Scientific Terminology for Beginners

This glossary introduces key scientific terms used in this notebook with simple examples.

### 1) Treatment Arm

- **Explanation**: One treatment category in a multi-treatment study.
- **Example**: `control`, `drug A`, and `drug B` are three treatment arms.

### 2) Multi-Outcome

- **Explanation**: Modeling more than one outcome variable at the same time.
- **Example**: Estimating effects on both blood pressure and heart rate.

### 3) Counterfactual

- **Explanation**: The outcome that would have happened under a different treatment arm.
- **Example**: For a patient given `drug A`, the untreated outcome is counterfactual.

### 4) Contrast

- **Explanation**: Difference in effects between two treatment arms.
- **Example**: Effect of `drug B` minus effect of `control`.

### 5) Covariate Profile

- **Explanation**: The set of characteristics for one unit.
- **Example**: Age 45, female, high baseline risk.

### 6) Propensity Score

- **Explanation**: Probability of receiving each treatment arm given covariates.
- **Example**: A patient has 0.2 probability of arm A and 0.6 probability of arm B.

### 7) Heterogeneity

- **Explanation**: Treatment effects vary across units or subgroups.
- **Example**: Arm A works better for older patients than younger patients.

### 8) Joint Estimation

- **Explanation**: Estimating related outcomes/treatments together to share information.
- **Example**: A model uses both outcomes to improve precision for each outcome effect.

### How It Work

The Multi-Arm/Multi-Outcome Causal Forest builds on the **Generalized Random Forest (GRF)** framework, extending the Causal Forest algorithm (designed for single treatment/outcome) to handle multiple treatments and outcomes. Here’s a step-by-step explanation of its mechanics:

1.  Data Setup

  -   A dataset with:
  -   Covariates $X$ (e.g., patient age, gender).
  -   Treatment assignments $W$, where $W$ can take multiple values (e.g., 0 for control, 1 for treatment A, 2 for treatment B).
  -   Multiple outcome variables $Y_1, Y_2, \ldots, Y_k$ (e.g., different health metrics).

2.  Tree Construction

  -   `Splitting Rule`: Each tree in the forest splits the covariate space to maximize heterogeneity in treatment effects across all arms and outcomes. The splitting criterion optimizes a loss function that measures the quality of treatment effect estimates (e.g., variance of estimated effects).
  -   `Multi-Arm`: For multiple treatments, the tree estimates treatment effects for each arm relative to a baseline (e.g., control) or pairwise differences.
  -   `Multi-Outcome`: For multiple outcomes, the tree models correlations between outcomes, sharing information to improve estimation efficiency (e.g., using a joint loss function or covariance structure).
  -   `Honest Splitting*` Uses separate data for splitting and estimation within each tree to reduce bias (a hallmark of GRF).

3.  Treatment Effect Estimation

  -   For each leaf node in a tree, the algorithm estimates treatment effects for each treatment arm and outcome using local observations.
  -   `Multi-Arm`: Computes the effect of each treatment (e.g., $\tau_j(X) = E[Y | W=j, X] - E[Y | W=0, X]$) for treatment $j$ vs. control).
  -   `Multi-Outcome`: Estimates effects for each outcome (e.g., $\tau_{j,k}(X)$ for treatment $j$ on outcome $k$), often modeling outcome correlations to improve precision.
    -   Methods like local linear regression or moment-based estimation may be used within leaves to estimate effects.

4.  Forest Aggregation

  -   The forest averages predictions across many trees (typically hundreds or thousands) to produce robust estimates.
  -   For a given covariate $X$, the forest outputs a vector of treatment effect estimates for each treatment arm and outcome, e.g., $\hat{\tau}(X) = [\hat{\tau}_{1,1}(X), \hat{\tau}_{1,2}(X), \ldots, \hat{\tau}_{j,k}(X)]$.
  -   Variance estimates are also provided to assess uncertainty.

5.  Handling Complexity

  -   `Multi-Arm`: The forest balances comparisons across multiple treatments, reducing overfitting by sharing information across arms.
  -   `Multi-Outcome`: Leverages correlations between outcomes (e.g., via covariance weighting) to improve efficiency, especially when outcomes are related (e.g., systolic and diastolic blood pressure).
  -   `Nuisance Parameters`: Estimates propensity scores (probability of receiving each treatment) or outcome regressions to adjust for confounding, often using separate forests for these nuisance functions.

6.  Output:

  -   For each individual (defined by covariates $X$, the model provides:
        -   Estimated treatment effects for each treatment arm and outcome.
        -   Confidence intervals or standard errors for uncertainty.
    -   Visualizations or summaries of heterogeneous effects (e.g., which subgroups benefit most from a specific treatment).

### Key Differences from Standard Causal Forest

-   `Standard Causal Forest`: Handles one treatment (binary or continuous) and one outcome, estimating a single treatment effect per individual.
-   `Multi-Arm/Multi-Outcome Causal Forest`:
    -   Models multiple treatments (e.g., multiple drugs) and/or multiple outcomes (e.g., multiple health metrics).
    -   Shares information across arms/outcomes to improve estimation efficiency.
    -   Uses a joint objective function to optimize splits for all treatments and outcomes simultaneously.

### Advantages

-   Handles complex experiments with multiple treatments and outcomes.
-   Captures heterogeneity in treatment effects across subgroups.
-   Robust to non-linear relationships and high-dimensional covariates.
-   Improves efficiency by sharing information across arms/outcomes.

### Limitations

-   Requires sufficient sample size for each treatment arm and outcome.
-   Assumes unconfoundedness (no unmeasured confounders) unless combined with other methods (e.g., instrumental variables).
-   Computationally intensive for large datasets or many outcomes.
-   Interpretability can be challenging due to complex output (multiple effects per individual).

## Multi-arm/multi-outcome Causal Forestt in  R

This tutorial demonstrates how to implement a **multi-arm/multi-outcome causal forest** in R using the `{grf}` package. The example uses the `lung` dataset from the `{survival}` package, simulating a multi-arm treatment scenario with multiple outcomes.

### Load and Check Required Libraries


```{r}
#| label: load-packages
packages <- c('tidyverse', 
              'plyr',
              'RCausalML',
              'causaldata',
              'mlbench',
              'survival',
              'BART',
              'grf',
              'kernelshap',
              'shapviz')
```

### Install Missing Packages

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
#new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: verify-installation
# Verify installation
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: load-libraries
# Load packages with suppressed messages
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Check Loaded Packages

```{r}
#| label: check-loaded-packages
# Check loaded packages
cat("Successfully loaded packages:\n")
print(search()[grepl("package:", search())])
```

### Load and Prepare the Lung Dataset

The lung dataset includes survival data for lung cancer patients. Since lung lacks a treatment variable and multiple outcomes, we:

-   Simulates a multi-arm `treatment` variable with three levels: `placebo`, `A`, and `B` and Assigns probabilities (40% placebo, 30% A, 30% B) for each treatment.

-   Simulates a binary secondary outcome (e.g., `health_status`, 1 = yes, 0 = no) with treatment-dependent probabilities.

-   Use log-transformed survival time (`log_time`) as the primary outcome to ensure numerical stability

```{r}
#| label: load-data
#| warning: false
#| error: false

# Load and modify the lung data
data(lung, package = "survival")
lung_data <- lung %>%
  dplyr::filter(!is.na(age), !is.na(sex), !is.na(ph.ecog)) %>%
  dplyr::mutate(
    treatment = as.factor(sample(c("placebo", "A", "B"), n(), replace = TRUE, prob = c(0.4, 0.3, 0.3))),
    health_status = rbinom(n(), 1, prob = 0.3 + 0.1 * (treatment == "A") + 0.15 * (treatment == "B")),
    log_time = log(time)
  )

# Define covariates, treatment, and outcomes
X <- lung_data %>% select(age, sex, ph.ecog) %>% as.matrix()
W <- lung_data$treatment
Y <- lung_data %>% select(log_time, health_status) %>% as.matrix()
colnames(Y) <- c("log_time", "health_status")

# Split into training and test sets
set.seed(123)
n <- nrow(lung_data)
train_idx <- sample(1:n, size = 0.8 * n)
test_idx <- setdiff(1:n, train_idx)
X_train <- X[train_idx, ]
X_test <- X[test_idx, ]
W_train <- W[train_idx]
W_test <- W[test_idx]
Y_train <- Y[train_idx, ]
Y_test <- Y[test_idx, ]

```



### Train the Multi-arm/Multi-outcome Causal Forest

Since we have multiple treatment arms ("placebo", "A", "B") and multiple outcomes (log_time and health_status), and the grf package’s `multi_arm_causal_forest` function is designed for multiple treatments but a single outcome, we fit a separate model for each outcome.


```{r}
#| label: train-macf
#| warning: false
macf_log_time <- multi_arm_causal_forest(X = X_train,
                                         Y = Y_train[, "log_time"], 
                                         W = W_train)
macf_health_status <- multi_arm_causal_forest(X = X_train, 
                                              Y = Y_train[, "health_status"], 
                                              W = W_train)
```

### Predict on the test data set

```{r}
#| label: predict-test
# Predict treatment effects
# Predict treatment effects on the test set
tau_hat_log_time <- predict(macf_log_time, newdata = X_test)
tau_hat_health_status <- predict(macf_health_status, newdata = X_test)
```

### Calculate Doubly Robust Average Treatment Effects (AIPW)

We compute the average treatment effects (ATE) for each arm relative to "placebo" using the doubly robust Augmented Inverse Propensity Weighting (AIPW) method provided by {grf}.


```{r}
#| label: ate-aipw
# Calculate ATE for log_time
ate_log_time <- average_treatment_effect(macf_log_time)

# Calculate ATE for health_status
ate_health_status <- average_treatment_effect(macf_health_status)

# Display the results
print("Average Treatment Effects for log_time:")
print(ate_log_time)
print("Average Treatment Effects for health_status:")
print(ate_health_status)
```

### Plot Variable Importance

Finally, we plot the importance of each covariate (age, sex, ph.ecog) for each outcome using ggplot2.

#### log_time Outcome

```{r}
#| label: fig-var-imp-log-time
#| fig.width: 5.5
#| fig.height: 4.5
# Variable importance for log_time
var_imp_log_time <- variable_importance(macf_log_time)
var_imp_df_log_time <- data.frame(
  Variable = colnames(X),
  Importance = var_imp_log_time
)

ggplot(var_imp_df_log_time, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Variable Importance for log_time", x = "Variable", y = "Importance") +
  theme_minimal()

```

#### Health_status Outcome

```{r}
#| label: fig-var-imp-health-status
#| fig.width: 5.5
#| fig.height: 4.5
# Variable importance for health_status
var_imp_health_status <- variable_importance(macf_health_status)
var_imp_df_health_status <- data.frame(
  Variable = colnames(X),
  Importance = var_imp_health_status
)

ggplot(var_imp_df_health_status, aes(x = reorder(Variable, Importance), y = Importance)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Variable Importance for health_status", x = "Variable", y = "Importance") +
  theme_minimal()
```


### Compute Kernel SHAP values for CATE using shapviz

We use the `explain_cate()` function from the RCausalML package (see `R/shapviz_integration.R`) to compute SHAP values for the CATE predictions of each outcome. Here we show both `log_time` and `health_status` outcomes.

#### Step  “placebo” the baseline (highly recommended for interpretability)
Run this **before** refitting the forests (or right after you create `W_train`):

```{r}
#| label: relevel-refit-macf
# Re-level so placebo is the reference (baseline)
W_train <- relevel(as.factor(W_train), ref = "placebo")
W_test  <- relevel(as.factor(W_test),  ref = "placebo")   # optional but consistent

# Refit both forests with the new baseline
macf_log_time <- multi_arm_causal_forest(
  X = X_train,
  Y = Y_train[, "log_time"],
  W = W_train
)

macf_health_status <- multi_arm_causal_forest(
  X = X_train,
  Y = Y_train[, "health_status"],
  W = W_train
)
```

After this, the column names in `predict()$predictions` will be **"A - placebo"** and **"B - placebo"** (exactly what we want).

#### Check the new structure (quick diagnostic)

```{r}
#| label: check-predict-colnames
p <- predict(macf_log_time, newdata = head(X_test, 5))
print(colnames(p$predictions[, , 1]))  
```

#### Define clean prediction wrappers (one per contrast)

Add these functions right before the SHAP computation:

```{r}
#| label: pred-wrappers-shap
# Prediction wrapper for log_time outcome
pred_A_vs_placebo_log <- function(object, newdata) {
  p <- predict(object, newdata = newdata)
  p$predictions[, "A - placebo", 1]          # extracts the scalar vector we need
}

pred_B_vs_placebo_log <- function(object, newdata) {
  p <- predict(object, newdata = newdata)
  p$predictions[, "B - placebo", 1]
}

# Same wrappers for health_status outcome
pred_A_vs_placebo_health <- function(object, newdata) {
  p <- predict(object, newdata = newdata)
  p$predictions[, "A - placebo", 1]
}

pred_B_vs_placebo_health <- function(object, newdata) {
  p <- predict(object, newdata = newdata)
  p$predictions[, "B - placebo", 1]
}
```

#### Compute Kernel SHAP (the real replacement for `explain_cate()`)

```{r}
#| label: kernel-shap-compute
library(kernelshap)
library(shapviz)
library(future)

#  Parallel backend 
n_cores <- max(1L, parallel::detectCores() - 1L)
plan(multisession, workers = n_cores)

# Small background sample (50 rows) – makes Kernel SHAP fast and stable
set.seed(42)
bg_idx <- sample(nrow(X), min(50, nrow(X)))
bg_small <- X[bg_idx, , drop = FALSE]

# ── log_time outcome ───────────────────────────────────────────────
shap_A_log <- kernelshap(
  object   = macf_log_time,
  X        = X_test,                    # or X[1:500, ] if you want faster
  bg_X     = bg_small,
  pred_fun = pred_A_vs_placebo_log,
  verbose  = FALSE
)

shap_B_log <- kernelshap(
  object   = macf_log_time,
  X        = X_test,
  bg_X     = bg_small,
  pred_fun = pred_B_vs_placebo_log,
  verbose  = FALSE
)

# ── health_status outcome ──────────────────────────────────────────
shap_A_health <- kernelshap(
  object   = macf_health_status,
  X        = X_test,
  bg_X     = bg_small,
  pred_fun = pred_A_vs_placebo_health,
  verbose  = FALSE
)

shap_B_health <- kernelshap(
  object   = macf_health_status,
  X        = X_test,
  bg_X     = bg_small,
  pred_fun = pred_B_vs_placebo_health,
  verbose  = FALSE
)

# Reset workers 
plan(sequential)
```

### Exact Kernel SHAP values (preview)

```{r}
#| label: shap-preview-kable
#| results: asis
#| message: false
#| warning: false
library(knitr)

# Show a small preview so values appear in HTML
sv_A_log <- shapviz(shap_A_log)
cat("**SHAP matrix (log_time: A vs placebo)**\n\n")
print(knitr::kable(head(as.data.frame(sv_A_log$S), 10), digits = 4))
```

### Turn into shapviz objects and visualize (panel display)

```{r}
#| label: shapviz-objects

# Convert to shapviz objects
shp_A_log    <- shapviz(shap_A_log)
shp_B_log    <- shapviz(shap_B_log)
shp_A_health <- shapviz(shap_A_health)
shp_B_health <- shapviz(shap_B_health)
```

### Importance Plots

```{r}
#| label: fig-shap-importance-panel
#| fig.width: 10
#| fig.height: 10

library(patchwork) # For panel display

# Importance plots (beeswarm + bar) for each (treatment, outcome)
plt1 <- sv_importance(shp_A_log,    kind = "both") + ggtitle("log_time: A vs placebo")
plt2 <- sv_importance(shp_B_log,    kind = "both") + ggtitle("log_time: B vs placebo")
plt3 <- sv_importance(shp_A_health, kind = "both") + ggtitle("health_status: A vs placebo")
plt4 <- sv_importance(shp_B_health, kind = "both") + ggtitle("health_status: B vs placebo")

# Arrange plots in a 2x2 panel
(plt1 | plt2) / (plt3 | plt4)
```

### Dependence plots for each covariate

```{r}
#| label: shap-list-features
# List available feature names for each SHAP object
print("Available features in SHAP object (log_time):")
print(colnames(shp_A_log$X))
print("Available features in SHAP object (health_status):")
print(colnames(shp_A_health$X))

# For illustration, show dependence plots for the first three available features for BOTH log_time and health_status
available_feats_log <- colnames(shp_A_log$X)
available_feats_health <- colnames(shp_A_health$X)

# Logging available features for each outcome
print("Features (log_time):")
print(available_feats_log)
print("Features (health_status):")
print(available_feats_health)
```


```{r}
#| label: fig-shap-dependence
#| fig.width: 12
#| fig.height: 10
# Dependence plots for log_time outcome (A vs placebo)
plt_A_log1 <- sv_dependence(shp_A_log, v = available_feats_log[1]) + ggtitle(paste("log_time (A):", available_feats_log[1]))
plt_A_log2 <- sv_dependence(shp_A_log, v = available_feats_log[2]) + ggtitle(paste("log_time (A):", available_feats_log[2]))
plt_A_log3 <- sv_dependence(shp_A_log, v = available_feats_log[3]) + ggtitle(paste("log_time (A):", available_feats_log[3]))

# Dependence plots for health_status outcome (A vs placebo)
plt_A_health1 <- sv_dependence(shp_A_health, v = available_feats_health[1]) + ggtitle(paste("health_status (A):", available_feats_health[1]))
plt_A_health2 <- sv_dependence(shp_A_health, v = available_feats_health[2]) + ggtitle(paste("health_status (A):", available_feats_health[2]))
plt_A_health3 <- sv_dependence(shp_A_health, v = available_feats_health[3]) + ggtitle(paste("health_status (A):", available_feats_health[3]))

library(patchwork)
# Display all dependence plots in a 2x3 grid (top: log_time, bottom: health_status)
(plt_A_log1 | plt_A_log2 | plt_A_log3) /
(plt_A_health1 | plt_A_health2 | plt_A_health3)
```

## Summary and Conclusion


Multi-arm/multi-outcome causal forests are powerful tools for estimating heterogeneous treatment effects in complex settings with multiple treatments and outcomes. By leveraging the flexibility of random forests, they provide robust estimates that account for individual differences and correlations between outcomes. This approach is particularly useful in fields like healthcare, policy evaluation, and marketing, where understanding the nuanced effects of different interventions is crucial. This tutorial demonstrated how to implement a multi-arm/multi-outcome causal forest in R, using the `{RCausalML}` package intregated with with {grf} packages to analyze treatment effects on multiple outcomes simultaneously. This approach allows researchers to gain deeper insights into the effectiveness of various interventions across diverse populations and outcomes, ultimately leading to more informed decision-making. 

We also integrated SHAP values using the `kernelshap` and `shapviz` packages to interpret the heterogeneous treatment effects, providing a comprehensive framework for both estimation and interpretation in multi-arm/multi-outcome causal inference.

If you want a boosted (XGBoost-based) multi-arm approach that fits one outcome model per arm and then forms contrasts relative to a baseline, see `08-multi-arm-causal-boost-r.qmd`.

## Resources

1.  Wager, S., & Athey, S. (2018). Estimation and Inference of Heterogeneous Treatment Effects using Random Forests. JASA. 

2.  Athey, S., Tibshirani, J., & Wager, S. (2019). Generalized Random Forests. Ann. Statist. 

3.  Multi-arm/multi-outcome implementation available in grf R package
