# Uplift Trees — Regression (Abortion Data)

In this notebook we apply **uplift regression** methods from **RCausalML** to the **abortion** dataset from `{causaldata}`: a **continuous** outcome (logged gonorrhea rate) and a **binary** treatment (early repeal of abortion prohibition). The workflow parallels [03-uplift-trees-regression-r.qmd](03-uplift-trees-regression-r.qmd), but replaces simulated multi-arm data with this observational use case.


## Scientific Terminology for Beginners

This notebook uses policy data, so causal terminology is especially important.

- **Observational data**: Data where treatment is not randomly assigned by the analyst.
- **Confounding**: When other variables affect both treatment and outcome, potentially biasing effect estimates.
- **Identification assumptions**: Conditions needed to interpret estimates causally (for example, no unmeasured confounding).
- **CATE**: Subgroup-specific expected treatment effect, conditional on covariates $X$.
- **Policy effect heterogeneity**: Different policy impacts across demographic or socioeconomic groups.
- **Interpretability**: Ability to explain why the model predicts stronger or weaker effects for certain profiles.

### Simple example

Suppose we compare regions with and without an early policy change.

- Average logged outcome in treated-like subgroup is lower by 0.12 relative to comparable control subgroup.
- The model estimates subgroup CATE near **-0.12** on the log scale.

For a beginner, this means the policy is associated with a reduction in the outcome for that subgroup, under the modeling assumptions.

## Overview

Uplift modeling with **continuous outcomes** targets the **conditional average treatment effect (CATE)** $\tau(x) = \mathbb{E}[Y \mid T=1, X=x] - \mathbb{E}[Y \mid T=0, X=x]$. Here $Y$ is **lnr** (logged gonorrhea rate per 100,000 among 15–19 year olds) and $T$ is **repeal**. This is an observational panel-style dataset; we illustrate **heterogeneous effect estimation** and **ranking diagnostics** with the same tools as the synthetic regression notebook, without claiming causal identification beyond what your design assumptions allow.

## Implementation in R

For **continuous** outcomes, `uplift_rf_kl()` fits a **T-learner random forest** (`ranger`): separate outcome models under treatment and control, with $\hat{\tau}(x) = \hat{\mu}_1(x) - \hat{\mu}_0(x)$. The **abortion** data have only **one treated arm** vs control, so **`uplift_rf_multi()`** (multiple non-control arms) is not used here; see the synthetic regression notebook for that pattern.

**Interaction trees** (`interaction_tree()`) and **causal inference trees** (`causal_inference_tree()`) give **explicit partition rules** for mean outcome contrasts between arms. We evaluate **cumulative gain** (`plot_gain()`) on the holdout set using predicted CATE as the model score, and use **kernel SHAP** (`kernelshap`, `shapviz`) to summarize drivers of predicted uplift.

We also apply **filter-based feature selection** following [11-feature-selection-uplift.qmd](11-feature-selection-uplift.qmd) (Zhao et al. 2020): the **F-filter** (`get_importance()`, `method = "F"`) ranks covariates by F-tests for treatment–feature interaction in **linear models**, which is appropriate for continuous **lnr**. The **LR** and **KL** options in `get_importance()` are implemented for **binary** outcomes in this package; they are omitted here for that reason (see `?get_importance`).

### Setup

### Load and check required libraries

```{r setup-packages}
packages <- c(
  "tidyverse",
  "RCausalML",
  "causaldata",
  "rpart.plot",
  "shapviz",
  "kernelshap"
)
```

### Install missing packages (if needed)

```{r install-missing-packages}
#| warning: false
#| error: false
# new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
# if (length(new_packages)) install.packages(new_packages)
```

### Load required libraries

```{r load-required-libraries}
if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(".", quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Data

R package `{causaldata}` provides a collection of datasets for causal inference. We integrate {causaldata} with {RCausalML} to easily load datasets. In this tutorial, we use the **abortion** dataset, which contains state-year observations on gonorrhea rates among 15–19 year olds, with a binary indicator for early repeal of abortion prohibition. This dataset is commonly used to study the effects of abortion legalization on risky sexual behavior among teenagers, proxied by gonorrhea rates.

This dataset is commonly used to study the **causal effects of abortion legalization policies** (specifically, early repeal of abortion prohibitions in certain U.S. states) on **risky sexual behavior** among teenagers. Risky behavior is proxied by the **incidence of gonorrhea** (a sexually transmitted infection) among 15–19 year olds. The idea is that access to abortion might influence sexual decision-making or risk-taking behavior in this age group.

Here are the most important columns (based on the package documentation):

- **fip** — State FIPS code (numeric identifier for each U.S. state)
- **age** — Age in years (focused on 15–19 year olds)
- **race** — Race: 1 = white, 2 = black (or similar coding)
- **year** — Calendar year
- **t** — A rescaled time variable (often centered around policy change for DiD)
- **sex** — Sex: 1 = male, 2 = female
- **totpop** — Total population (state-level)
- **ir** — Incarcerated males per 100,000 (proxy for criminal justice environment)
- **crack** — Crack cocaine index (proxy for drug epidemic intensity)
- **alcohol** — Alcohol consumption per capita
- **income** — Real income per capita
- **ur** — State unemployment rate
- **poverty** — Poverty rate
- **repeal** — Treatment indicator: 1 = state had an **early repeal** of abortion prohibition (treatment group), 0 = otherwise (control group). This is the main policy/treatment variable.
- **pi** — Parental involvement law in effect (another policy control)
- **bf15** — Indicator: Is a black female in the 15–19 age group (useful for subgroup analysis)

```{r list-causaldata-datasets}
# causaldata integration
list_causaldata_datasets()
```

### Load Data

The **outcome variable** is typically **lnr** (logged gonorrhea rate per 100,000 population among 15–19 year olds), which measures the logged incidence of gonorrhea as a proxy for risky sexual behavior.

```{r load-abortion-data}
# Load abortion dataset (use $data for the raw dataframe; load_causaldata returns list with X, w, y, data, citation)
abortion <- load_causaldata("abortion")$data
# Quick look
head(abortion)
str(abortion)
summary(abortion$repeal)      # Treatment balance
table(abortion$repeal, abortion$race)  # Example crosstab
```

### Data Processing

```{r abortion-data-processing}
# Select relevant subset (focus on 15-19 year olds, complete cases)
abortion_clean <- abortion %>%
  dplyr::filter(age >= 15 & age <= 19) %>%
  na.omit()

# Define covariates (X), outcome (Y), treatment (W)
X <- abortion_clean[, c("age", "race", "sex", "totpop", "ir", "crack",
                        "alcohol", "income", "ur", "poverty")]

Y <- abortion_clean$lnr          # Logged gonorrhea rate per 100,000 (continuous)
W <- abortion_clean$repeal       # 1 = early repeal (treatment), 0 = otherwise

# Convert X to numeric matrix
X <- as.matrix(X)
colnames(X) <- c("age", "race", "sex", "totpop", "ir", "crack",
                 "alcohol", "income", "ur", "poverty")

# Remove near-zero variance columns (if any)
X <- X[, apply(X, 2, stats::var) > 1e-10, drop = FALSE]
cat("Covariates used:", colnames(X), "\n")
```

### Split Data into Training and Test Sets

We split the data into 80% training and 20% test sets using random sampling (fixed seed for reproducibility).

```{r train-test-split-abortion}
set.seed(111)
train_prop <- 0.8
n <- nrow(abortion_clean)
train_idx <- sample(seq_len(n), size = round(train_prop * n))
X_train <- X[train_idx, , drop = FALSE]
Y_train <- Y[train_idx]
W_train <- W[train_idx]
X_test <- X[-train_idx, , drop = FALSE]
Y_test <- Y[-train_idx]
W_test <- W[-train_idx]
cat("Training set size:", nrow(X_train), "\n")
cat("Test set size:", nrow(X_test), "\n")
cat("Column names of X_train:", colnames(X_train), "\n")
```

```{r abortion-x-names}
x_names <- colnames(X_train)
```

### Feature selection (F-filter; Zhao et al. 2020)

This mirrors the workflow in [11-feature-selection-uplift.qmd](11-feature-selection-uplift.qmd), adapted for **continuous** `lnr`. Rankings use **`get_importance(..., method = "F")`** on the **training set only** (no test leakage). Treatment is encoded as `control` / `treated` from `repeal`.

```{r abortion-fs-dataframe}
df_fs_train <- as.data.frame(X_train, stringsAsFactors = FALSE)
names(df_fs_train) <- x_names
df_fs_train$lnr <- Y_train
df_fs_train$treatment_group_key <- ifelse(W_train == 1, "treated", "control")
```

#### method = F (interaction F-tests; orders 1–3)

```{r abortion-filter-F}
f_imp_1 <- get_importance(
  df_fs_train,
  features = x_names,
  y_name = "lnr",
  method = "F",
  experiment_group_column = "treatment_group_key",
  control_group = "control",
  treatment_group = "treated",
  order = 1
)
head(f_imp_1)

f_imp_2 <- get_importance(
  df_fs_train,
  features = x_names,
  y_name = "lnr",
  method = "F",
  experiment_group_column = "treatment_group_key",
  control_group = "control",
  treatment_group = "treated",
  order = 2
)
head(f_imp_2)

f_imp_3 <- get_importance(
  df_fs_train,
  features = x_names,
  y_name = "lnr",
  method = "F",
  experiment_group_column = "treatment_group_key",
  control_group = "control",
  treatment_group = "treated",
  order = 3
)
head(f_imp_3)
```

#### Subsets for evaluation (top 3, top 6, all; capped by $p$)

```{r abortion-top-feature-sets}
ranked_f <- f_imp_1$feature
n_f <- length(ranked_f)
top_3_features <- ranked_f[seq_len(min(3L, n_f))]
top_6_features <- ranked_f[seq_len(min(6L, n_f))]
cat("Top 3 (F1):", paste(top_3_features, collapse = ", "), "\n")
cat("Top 6 (F1):", paste(top_6_features, collapse = ", "), "\n")
```

```{r abortion-plot-f-scores}
#| fig.width: 7
#| fig.height: 5
par(mar = c(4, 8, 2, 1))
n_bar <- min(20L, nrow(f_imp_1))
barplot(
  rev(head(f_imp_1$score, n_bar)),
  names.arg = rev(head(f_imp_1$feature, n_bar)),
  horiz = TRUE,
  las = 1,
  main = "F1 filter: top features (abortion, train)",
  col = "steelblue",
  border = NA
)
```

#### Performance vs feature subset: T-Learner, R-Learner, S-Learner

We compare **cumulative gain** and **AUUC** (mean of normalized cumulative gain) on the **test** set, following [11-feature-selection-uplift.qmd](11-feature-selection-uplift.qmd). Outcome column is **`lnr`**; `is_treated` is `W_test`.

```{r abortion-fs-test-idx}
is_treated <- as.integer(W_test)
y_test <- Y_test
n_test <- nrow(X_test)
```

```{r abortion-uplift-tlearner-fs}
tl <- TLearner(learner = "ranger", control_name = 0)
tl <- fit(tl, X_train, W_train, Y_train)
pred_all <- predict(tl, X_test)
if (is.matrix(pred_all)) pred_all <- pred_all[, 1]

tl_t3 <- TLearner(learner = "ranger", control_name = 0)
tl_t3 <- fit(tl_t3, X_train[, top_3_features, drop = FALSE], W_train, Y_train)
pred_t3 <- predict(tl_t3, X_test[, top_3_features, drop = FALSE])
if (is.matrix(pred_t3)) pred_t3 <- pred_t3[, 1]

tl_t6 <- TLearner(learner = "ranger", control_name = 0)
tl_t6 <- fit(tl_t6, X_train[, top_6_features, drop = FALSE], W_train, Y_train)
pred_t6 <- predict(tl_t6, X_test[, top_6_features, drop = FALSE])
if (is.matrix(pred_t6)) pred_t6 <- pred_t6[, 1]
```

```{r abortion-df-preds-uplift-fs}
model_cols_uplift <- c("All", "Top_3", "Top_6", "Random")
set.seed(42)
df_preds_uplift <- data.frame(
  All = pred_all,
  Top_3 = pred_t3,
  Top_6 = pred_t6,
  Random = stats::runif(n_test),
  is_treated = is_treated,
  lnr = y_test
)
```

```{r abortion-plot-gain-uplift-fs}
#| fig.width: 7
#| fig.height: 5
plot_gain(
  df_preds_uplift,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  model_cols = model_cols_uplift,
  main = "T-Learner (ranger): cumulative gain — abortion test"
)
```

```{r abortion-auuc-uplift-fs}
gain_uplift <- get_cumgain(
  df_preds_uplift,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  normalize = TRUE
)
auuc_uplift <- colMeans(gain_uplift[, model_cols_uplift, drop = FALSE])
print(auuc_uplift)
```

```{r abortion-rlearner-fs}
rl <- RLearner(learner = "ranger", control_name = 0, n_fold = 3)
rl <- fit(rl, X_train, W_train, Y_train)
pred_r_all <- predict(rl, X_test)
if (is.matrix(pred_r_all)) pred_r_all <- pred_r_all[, 1]

rl_t3 <- RLearner(learner = "ranger", control_name = 0, n_fold = 3)
rl_t3 <- fit(rl_t3, X_train[, top_3_features, drop = FALSE], W_train, Y_train)
pred_r_t3 <- predict(rl_t3, X_test[, top_3_features, drop = FALSE])
if (is.matrix(pred_r_t3)) pred_r_t3 <- pred_r_t3[, 1]

rl_t6 <- RLearner(learner = "ranger", control_name = 0, n_fold = 3)
rl_t6 <- fit(rl_t6, X_train[, top_6_features, drop = FALSE], W_train, Y_train)
pred_r_t6 <- predict(rl_t6, X_test[, top_6_features, drop = FALSE])
if (is.matrix(pred_r_t6)) pred_r_t6 <- pred_r_t6[, 1]
```

```{r abortion-rlearner-gain-auuc-fs}
set.seed(42)
df_preds_r <- data.frame(
  All = pred_r_all,
  Top_3 = pred_r_t3,
  Top_6 = pred_r_t6,
  Random = stats::runif(n_test),
  is_treated = is_treated,
  lnr = y_test
)
plot_gain(
  df_preds_r,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  model_cols = model_cols_uplift,
  main = "R-Learner: cumulative gain — abortion test"
)
gain_r <- get_cumgain(
  df_preds_r,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  normalize = TRUE
)
auuc_r <- colMeans(gain_r[, model_cols_uplift, drop = FALSE])
print(auuc_r)
```

```{r abortion-slearner-fs}
sl <- SLearner(learner = "ranger", control_name = 0)
sl <- fit(sl, X_train, W_train, Y_train)
pred_s_all <- predict(sl, X_test)
if (is.matrix(pred_s_all)) pred_s_all <- pred_s_all[, 1]

sl_t3 <- SLearner(learner = "ranger", control_name = 0)
sl_t3 <- fit(sl_t3, X_train[, top_3_features, drop = FALSE], W_train, Y_train)
pred_s_t3 <- predict(sl_t3, X_test[, top_3_features, drop = FALSE])
if (is.matrix(pred_s_t3)) pred_s_t3 <- pred_s_t3[, 1]

sl_t6 <- SLearner(learner = "ranger", control_name = 0)
sl_t6 <- fit(sl_t6, X_train[, top_6_features, drop = FALSE], W_train, Y_train)
pred_s_t6 <- predict(sl_t6, X_test[, top_6_features, drop = FALSE])
if (is.matrix(pred_s_t6)) pred_s_t6 <- pred_s_t6[, 1]
```

```{r abortion-slearner-gain-auuc-fs}
set.seed(42)
df_preds_s <- data.frame(
  All = pred_s_all,
  Top_3 = pred_s_t3,
  Top_6 = pred_s_t6,
  Random = stats::runif(n_test),
  is_treated = is_treated,
  lnr = y_test
)
plot_gain(
  df_preds_s,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  model_cols = model_cols_uplift,
  main = "S-Learner: cumulative gain — abortion test"
)
gain_s <- get_cumgain(
  df_preds_s,
  outcome_col = "lnr",
  treatment_col = "is_treated",
  normalize = TRUE
)
auuc_s <- colMeans(gain_s[, model_cols_uplift, drop = FALSE])
print(auuc_s)
```

```{r abortion-auuc-summary-fs}
tab <- rbind(
  data.frame(Learner = "T-Learner", model = names(auuc_uplift), AUUC = as.numeric(auuc_uplift)),
  data.frame(Learner = "R-Learner", model = names(auuc_r), AUUC = as.numeric(auuc_r)),
  data.frame(Learner = "S-Learner", model = names(auuc_s), AUUC = as.numeric(auuc_s))
)
knitr::kable(tab, digits = 4)
```

As in the classification feature-selection notebook, smaller feature sets can match or beat “all features” on AUUC depending on noise and hyperparameters.

Mean outcome and sample size by treatment arm:

```{r abortion-outcome-by-arm}
agg <- aggregate(
  lnr ~ repeal,
  data = data.frame(lnr = Y, repeal = W),
  FUN = function(z) c(mean = mean(z), sd = stats::sd(z), size = length(z))
)
agg <- do.call(data.frame, agg)
names(agg) <- c("repeal", "mean", "sd", "size")
agg <- rbind(
  agg,
  data.frame(
    repeal = NA,
    mean = mean(Y),
    sd = stats::sd(Y),
    size = length(Y)
  )
)
agg$repeal[is.na(agg$repeal)] <- "All"
print(agg)
```

### Binary treatment: T-learner random forest (`uplift_rf_kl`)

With a continuous outcome, `uplift_rf_kl()` fits **two** `ranger` models (treated vs control). We fit on the training set and evaluate uplift ranking on the test set.

```{r uplift-rf-binary-abortion}
uplift_bin_rf <- uplift_rf_kl(
  X_train, W_train, Y_train,
  n_trees = 100,
  min_node_size = 80,
  max_depth = 8
)
pred_tr <- predict(uplift_bin_rf, X_train)
pred_te <- predict(uplift_bin_rf, X_test)
cat(
  "Mean predicted CATE — train:", round(mean(pred_tr), 4),
  " test:", round(mean(pred_te), 4), "\n"
)
```

### Uplift (gain) curve on the test set

We pass the test-set **outcome**, **treatment**, and **predicted CATE** (as the model score column) to `plot_gain()`.

```{r abortion-auuc-metrics}
auuc_metrics <- data.frame(
  is_treated = W_test,
  outcome = Y_test,
  uplift_model = pred_te
)
head(auuc_metrics)
```

```{r abortion-plot-gain}
#| fig.width: 6
#| fig.height: 5
plot_gain(auuc_metrics, outcome_col = "outcome", treatment_col = "is_treated")
```

### Interaction tree

```{r abortion-it-tree}
it_fit <- interaction_tree(
  X_train, W_train, Y_train,
  min_node_size = 200,
  max_depth = 4
)
cat(uplift_tree_string(it_fit$tree, it_fit$X_names))
```

Leaf labels summarize **$\tau \approx \mathbb{E}[Y \mid T{=}1] - \mathbb{E}[Y \mid T{=}0]$** within each region.

```{r abortion-it-tree-plot}
#| fig.width: 8
#| fig.height: 8
uplift_tree_plot(
  it_fit,
  main = "Interaction tree — lnr vs repeal (abortion data, train)"
)
```

### Optional: causal inference tree (CIT)

```{r abortion-cit-tree}
#| fig.width: 8
#| fig.height: 8
cit_fit <- causal_inference_tree(
  X_train, W_train, Y_train,
  min_node_size = 200,
  max_depth = 4
)
uplift_tree_plot(
  cit_fit,
  main = "Causal inference tree (CIT) — abortion data (train)"
)
```

### Ranger variable importance (T-learner arms)

There is **no** `forest$trees` list for this object, so `uplift_forest_plot()` does not apply. **Variable importance** from `ranger` is a useful global summary.

```{r abortion-ranger-importance}
imp0 <- uplift_bin_rf$fit_0$variable.importance
imp1 <- uplift_bin_rf$fit_1$variable.importance
imp <- sort((imp0 + imp1) / 2, decreasing = TRUE)
print(round(head(imp, 8), 4))
```

### Distribution of predicted CATE (full sample)

```{r abortion-hist-cate}
uf_full <- uplift_rf_kl(
  X, W, Y,
  n_trees = 80,
  min_node_size = 40,
  max_depth = 8
)
pred_cate <- predict(uf_full, X)
```

```{r abortion-plot-cate-hist}
#| fig.width: 8
#| fig.height: 6
hist(
  pred_cate,
  breaks = 30,
  main = "T-learner forest: predicted CATE — abortion (full sample)",
  xlab = "Predicted treatment effect (lnr scale)",
  col = "steelblue",
  border = NA
)
abline(v = mean(pred_cate), lty = 2, col = "red")
legend(
  "topright",
  legend = sprintf("Mean = %.3f", mean(pred_cate)),
  lty = 2,
  col = "red",
  bty = "n"
)
```

### Kernel SHAP for predicted uplift (CATE)

We explain $\hat{\tau}(X)$ from **`uplift_bin_rf`** on a subset of the training data (full training rows can be slow).

```{r compute-shap-uplift-abortion}
set.seed(2026)

n_explain <- min(100L, nrow(X_train))
bg_size <- min(100L, nrow(X_train))
X_explain <- X_train[seq_len(n_explain), , drop = FALSE]
bg_idx <- sample.int(nrow(X_train), bg_size)
bg_X <- X_train[bg_idx, , drop = FALSE]

pred_fun <- function(object, newdata) {
  as.vector(predict(object, newdata))
}

library(future)
plan(multisession, workers = max(1L, parallel::detectCores() - 1L))

shap_ksh <- kernelshap(
  object = uplift_bin_rf,
  X = X_explain,
  bg_X = bg_X,
  pred_fun = pred_fun,
  bg_w = NULL,
  verbose = FALSE
)
plan(sequential)

shp_uplift <- shapviz(shap_ksh, X = X_explain)
```

```{r shap-importance-beeswarm-uplift-abortion}
#| fig.width: 7
#| fig.height: 5.5
sv_importance(shp_uplift, kind = "beeswarm")
sv_importance(shp_uplift, show_numbers = TRUE)
```

```{r shap-dependence-uplift-abortion}
#| fig.width: 8
#| fig.height: 6
important_vars <- x_names[seq_len(min(4L, length(x_names)))]
sv_dependence(shp_uplift, v = important_vars, share_y = TRUE)
```

```{r shap-dependence-single-uplift-abortion}
if (length(x_names) >= 2) {
  sv_dependence(shp_uplift, v = x_names[[2]], color_var = x_names[[1]])
}
```

```{r shap-waterfall-single-uplift-abortion}
#| fig.width: 7
#| fig.height: 4.5
preds_explain <- pred_fun(uplift_bin_rf, X_explain)
row_id_max <- which.max(preds_explain)

sv_waterfall(shp_uplift, row_id = row_id_max) +
  ggplot2::ggtitle("SHAP waterfall — highest estimated CATE in explain set")

sv_force(shp_uplift, row_id = row_id_max)
```

```{r shap-test-set-uplift-abortion}
n_te <- min(10L, nrow(X_test))
shap_test <- kernelshap(
  uplift_bin_rf,
  X = X_test[seq_len(n_te), , drop = FALSE],
  bg_X = bg_X,
  pred_fun = pred_fun,
  verbose = FALSE
)
shp_test <- shapviz(shap_test, X = X_test[seq_len(n_te), , drop = FALSE])
sv_importance(shp_test, kind = "beeswarm")
```

### Interpretation notes

- **Positive SHAP** pushes the predicted **CATE** up; **negative SHAP** pushes it down.
- Explanations target **heterogeneity in** $\hat{\tau}(X)$, not the raw outcome margin $\mathbb{E}[Y \mid X]$.
- **Kernel SHAP** is model-agnostic and uses T-learner predictions; standard errors and causal interpretation of $\tau(x)$ in this dataset depend on **identification** (e.g. confounding, panel structure), not on the ML estimator alone.

## Summary

On the **abortion** data, **continuous-outcome** uplift tools in RCausalML reduce to a **binary** T-learner forest (`uplift_rf_kl`), plus **interaction / CIT** trees for interpretable partitions and **gain / SHAP** for evaluation and explanation. **Feature selection** uses the **F-filter** from `get_importance()` (training data only), then compares T-, R-, and S-learners with top-3 / top-6 subsets vs all covariates on test **AUUC**, as in [11-feature-selection-uplift.qmd](11-feature-selection-uplift.qmd). For **multiple non-control arms**, follow [03-uplift-trees-regression-r.qmd](03-uplift-trees-regression-r.qmd).

## References

- [Uplift Trees Example with Synthetic Data (Python CausalML)](https://causalml.readthedocs.io/en/latest/examples/uplift_trees_with_synthetic_data.html)
- [Feature Selection for Uplift Trees (Python CausalML)](https://causalml.readthedocs.io/en/latest/examples/feature_selection.html)
- Zhao et al. (2020), [Feature Selection Methods for Uplift Modeling](https://arxiv.org/abs/2005.03447)
- Gutierrez and Gérardy (2017), [Causal inference and uplift modelling: A review of the literature](http://proceedings.mlr.press/v67/gutierrez17a/gutierrez17a.pdf)
