# RNN/LSTM-Based Causal Models {.unnumbered}

**Recurrent Neural Networks (RNNs)** and their gated variants process sequences step-by-step, maintaining a hidden state $h_t$ that accumulates temporal causal history:

$$h_t = f(W_h h_{t-1} + W_x x_t + b)$$

Three architectural extensions turn this sequential memory into explicit causal machinery:

- **Causal masking in LSTMs** → which variables are allowed to influence which (learned sparse adjacency)
- **Reverse-time attention** → which past timesteps and which variables mattered most (RETAIN)
- **Regime-aware conditioning** → how system dynamics shift under external interventions

## The Three Models

### 1) Causal LSTM

A standard LSTM augmented with a learnable causal adjacency mask $G \in [0,1]^{d \times d}$ that gates which variables are allowed to influence variable $i$:

$$\tilde{x}_t^{(i)} = G_{i,:} \odot x_t$$

The LSTM transitions for variable $i$ are then computed on $[h_{t-1}; \tilde{x}_t^{(i)}]$. The mask is soft (continuous relaxation via sigmoid) during training and can be hardened to $\{0,1\}$ at test time. An L1 sparsity penalty on off-diagonal entries encourages parsimonious graphs:

$$\mathcal{L}_{\text{sparse}} = \lambda \sum_{i \neq j} G_{ij}$$

### 2) RETAIN (Reverse Time Attention)

RETAIN (Choi et al., 2016) processes input sequences in reverse order and learns two attention channels simultaneously:

- **Temporal attention** $\alpha_t$ — which timestep contributes most
- **Variable attention** $\beta_t$ — which variable matters at each timestep

The context vector is:

$$c = \sum_{t=1}^{T} \alpha_t \cdot (\beta_t \odot v_t)$$

A practical attribution score combining both channels:

$$\text{Contribution}(j, t) = \alpha_t \cdot |\beta_{t,j}| \cdot |w_j|$$

RETAIN's dual-GRU architecture produces interpretable, attribution-based causal influence matrices that are especially useful when explaining model decisions to stakeholders.

### 3) Intervention-Aware RNN

To model changing data-generating regimes (policy changes, market shocks, clinical treatments), the input is augmented with learned regime embeddings and an explicit intervention indicator:

$$h_t = \text{LSTM}(h_{t-1},\ [x_t;\ r_t;\ I_t])$$

where $r_t$ is a learned regime embedding (from a GRU-based regime detector) and $I_t$ is an intervention indicator. A regime-conditioned causal weight matrix $\Psi_k \in \mathbb{R}^{d \times d}$ is maintained per regime $k$; the effective causal matrix is the regime-probability-weighted average.

## Implementation in R

We use `{RCausalML}`'s `rnn_causal_model()` function (and its convenience wrappers `causal_lstm_model()`, `retain_model()`, `intervention_rnn_model()`) to fit all three architectures on S&P 500 sector ETF daily log-return data (2018–2024).

### Load and Check Required Libraries

```{r}
#| label: packages-list
#| warning: false
packages <- c(
  'tidyverse',
  'plyr',
  'RCausalML',
  'torch',
  'tidyquant',
  'reshape2',
  'scales',
  'patchwork',
  'igraph',
  'zoo'
)
```

### Install Missing Packages

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# Install missing packages
#new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: verify-installation
#| warning: false
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: load-required-libraries
#| warning: false
# When rendering from package root, use local RCausalML
if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(".", quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

```{r}
#| label: setup
#| include: true
set.seed(42)
device_use <- NULL
if (requireNamespace("torch", quietly = TRUE)) {
  device_use <- if (torch::cuda_is_available()) "cuda" else "cpu"
  torch::torch_manual_seed(42)
}
cat(sprintf("Using device: %s\n", device_use))
```

## Data Loading and Preprocessing

This section downloads S&P 500 sector ETF prices (the same universe used in the companion notebooks), converts them to standardized log-returns, builds lagged input/output arrays for training, and derives a volatility-spike intervention proxy (top-25% rolling-variance days).

> **Note:** If online market data is unavailable, the notebook automatically falls back to synthetic demo data so all downstream cells remain runnable.

```{r}
#| label: data-download
#| warning: false

TICKERS <- c(
  "XLF" = "Financials",
  "XLE" = "Energy",
  "XLK" = "Technology",
  "XLV" = "Healthcare",
  "XLI" = "Industrials",
  "XLY" = "ConsumerDisc",
  "XLP" = "ConsumerStap",
  "XLU" = "Utilities"
)

fetch_close_prices <- function(tickers, start = "2018-01-01", end = "2024-01-01") {
  tryCatch({
    raw <- tidyquant::tq_get(names(tickers), get = "stock.prices",
                              from = start, to = end, complete_cases = TRUE)
    if (is.null(raw) || nrow(raw) == 0) return(NULL)
    raw |>
      dplyr::select(symbol, date, adjusted) |>
      tidyr::pivot_wider(names_from = symbol, values_from = adjusted) |>
      dplyr::arrange(date) |>
      tidyr::drop_na()
  }, error = function(e) NULL)
}

close_wide <- fetch_close_prices(TICKERS)

if (is.null(close_wide) || nrow(close_wide) < 30) {
  message("Warning: market data unavailable; using synthetic demo data.")
  set.seed(42)
  n_steps <- 1600L
  d_syn   <- length(TICKERS)
  A_syn   <- matrix(0, d_syn, d_syn)
  for (i in seq_len(d_syn)) {
    A_syn[i, i] <- 0.45
    if (i > 1)                 A_syn[i, i - 1] <- 0.25
    if (i > 2 && i %% 2 == 1) A_syn[i, i - 2] <- 0.15
  }
  x_syn <- matrix(0, n_steps, d_syn)
  x_syn[1, ] <- rnorm(d_syn)
  noise_syn  <- 0.15 * matrix(rnorm(n_steps * d_syn), n_steps, d_syn)
  for (t in seq(2L, n_steps)) x_syn[t, ] <- x_syn[t - 1L, ] %*% t(A_syn) + noise_syn[t, ]
  returns_df <- as.data.frame(x_syn)
  colnames(returns_df) <- unname(TICKERS)
  data_np <- scale(x_syn)
} else {
  price_mat <- as.matrix(close_wide[, names(TICKERS)])
  log_ret   <- log(price_mat[-1L, ] / price_mat[-nrow(price_mat), ])
  colnames(log_ret) <- unname(TICKERS[colnames(price_mat)])
  returns_df <- as.data.frame(log_ret)
  data_np    <- scale(log_ret)
}

VAR_NAMES <- colnames(data_np)
d         <- ncol(data_np)
Tt        <- nrow(data_np)

cat(sprintf("d=%d variables, T=%d\n", d, Tt))
cat("Variables:", paste(VAR_NAMES, collapse = ", "), "\n")
```

```{r}
#| label: build-lag-dataset
#| warning: false

LAG   <- 20L
AHEAD <- 1L

build_lag_dataset <- function(data, lag, ahead = 1L) {
  N     <- nrow(data) - lag - ahead + 1L
  x_seq <- array(NA_real_, dim = c(N, lag, ncol(data)))
  y_mat <- matrix(NA_real_, nrow = N, ncol = ncol(data))
  for (t in seq_len(N)) {
    x_seq[t, , ] <- data[t:(t + lag - 1L), , drop = FALSE]
    y_mat[t,  ]  <- data[t + lag + ahead - 1L, ]
  }
  list(x_seq = x_seq, y_mat = y_mat)
}

lag_data <- build_lag_dataset(data_np, LAG, AHEAD)
X_seq    <- lag_data$x_seq
Y_mat    <- lag_data$y_mat

split  <- floor(0.8 * nrow(X_seq))
X_tr   <- X_seq[seq_len(split), , ]
X_val  <- X_seq[(split + 1L):nrow(X_seq), , ]
Y_tr   <- Y_mat[seq_len(split), ]
Y_val  <- Y_mat[(split + 1L):nrow(Y_mat), ]

cat(sprintf("Train: (%d, %d, %d)  Val: (%d, %d, %d)\n",
            dim(X_tr)[1], dim(X_tr)[2], dim(X_tr)[3],
            dim(X_val)[1], dim(X_val)[2], dim(X_val)[3]))
```

```{r}
#| label: build-intervention-vector
#| warning: false

# Intervention proxy: top-25% rolling-variance days (volatility spikes)
roll_sd <- zoo::rollapply(data_np, width = 10, FUN = sd, by.column = TRUE,
                          fill = NA, align = "right")
if (!requireNamespace("zoo", quietly = TRUE)) {
  roll_sd <- matrix(NA_real_, nrow(data_np), d)
  for (j in seq_len(d)) {
    for (t in seq(10L, nrow(data_np)))
      roll_sd[t, j] <- sd(data_np[(t - 9L):t, j])
  }
}
global_sd  <- rowMeans(roll_sd, na.rm = TRUE)
global_sd[is.na(global_sd)] <- 0
threshold  <- quantile(global_sd, 0.75)
I_raw      <- as.numeric(global_sd > threshold)

cat(sprintf("Intervention rate: %.1f%%\n", 100 * mean(I_raw)))
cat(sprintf("Intervention vector length: %d\n", length(I_raw)))
```

## Model Architectures Overview

In this section, we describe the three RNN/LSTM-based causal neural network models implemented in `{RCausalML}` and fitted using `rnn_causal_model()`. Each architecture is designed for multivariate time-series causal inference.

### CausalLSTM Architecture

The R implementation of `CausalLSTM` in `{RCausalML}`:

1. **LearnableCausalMask**: a $d \times d$ sigmoid-parameterized adjacency matrix with diagonal fixed to 1 (self-causality) and an L1 sparsity penalty on off-diagonal entries.
2. **Per-variable LSTMs**: each of the $d$ variables has its own `nn_lstm` fed the masked input $G_{i,:} \odot x_t$.
3. **Output heads**: two-layer MLP (hidden→hidden/2→1) per variable producing one-step-ahead forecasts.
4. **Sparsity loss**: added to MSE during training, promoting parsimonious causal graphs.
5. **Causal matrix**: the soft mask $G$ (or its hardened version) extracted at evaluation time.

### RETAIN Architecture

The R implementation of `RETAIN` in `{RCausalML}`:

1. **Embedding layer**: linear projection of each timestep's features to `hidden_alpha` dimensions.
2. **Reverse-time processing**: input embedded sequence is flipped in time before GRU processing.
3. **Alpha GRU + head**: GRU over reversed embeddings → scalar softmax attention weights $\alpha_t$ (one per timestep).
4. **Beta GRU + head**: second GRU → tanh-transformed feature weights $\beta_t \in \mathbb{R}^d$ (one vector per timestep).
5. **Context aggregation**: $c = \sum_t \alpha_t \cdot (\beta_t \odot x_t)$ → output head.
6. **Causal matrix**: aggregated $|W| \cdot |\beta|$ attribution averaged over validation batches.

### Intervention-Aware RNN Architecture

The R implementation of `InterventionAwareRNN` in `{RCausalML}`:

1. **RegimeDetector**: GRU processes input sequence → softmax over $K$ regime logits.
2. **Regime embedding**: linear projection of regime probabilities → dense `regime_dim`-vector, repeated along sequence length.
3. **Intervention projection**: scalar intervention indicator $I_t$ → dense `regime_dim`-vector per timestep.
4. **Augmented LSTM**: input is $[x_t; r_t; I_t] \in \mathbb{R}^{d + 2 \cdot \text{regime\_dim}}$ at each step.
5. **Prediction head**: GELU-activated MLP from last LSTM state → $d$-vector forecast.
6. **Causal matrix**: regime-probability-softmax-weighted average of learnable $\Psi_k$ matrices.

```
Input x_t (d variables)
      │
      ├──────────────────────────────────────────────────┐
      │                                                  │
      ▼                                                  ▼
RegimeDetector (GRU)                        Intervention Indicator I_t
      │                                                  │
      ▼                                                  ▼
regime_embed (Linear)                      interv_proj (Linear)
      │                                                  │
      └────────────────────┬─────────────────────────────┘
                           │
                           ▼
              Concatenate [x_t; r_t; I_t]
                           │
                           ▼
                    LSTM Encoder
                           │
                           ▼
                  Prediction Head (GELU MLP)
                           │
                           ▼
                 Forecast ŷ (d variables)
```

## Fit All Three Models

`rnn_causal_model()` is the single entry point that fits CausalLSTM, RETAIN, and the Intervention-Aware RNN in sequence. Internally it:

- Builds windowed `(lag, d)` input/output tensors and an `(lag,)` intervention window per sample.
- Trains each model with AdamW + cosine LR schedule + early stopping.
- Adds the sparsity loss $\mathcal{L}_{\text{sparse}}$ for CausalLSTM.
- Extracts the $d \times d$ causal matrix from each trained model.

```{r}
#| label: fit-all-models
#| warning: false

cat("Fitting CausalLSTM, RETAIN, and Intervention-Aware RNN ...\n")

rnn_fit <- rnn_causal_model(
  data             = data_np,
  lag              = LAG,
  models           = c("causal_lstm", "retain", "intervention_rnn"),
  intervention     = I_raw,
  hidden           = 64L,
  n_layers         = 2L,
  n_regimes        = 3L,
  regime_dim       = 8L,
  hidden_alpha     = 32L,
  hidden_beta      = 32L,
  dropout          = 0.2,
  lam_sparse       = 0.005,
  epochs           = 60L,
  lr               = 3e-4,
  patience         = 15L,
  batch_size       = 64L,
  val_split        = 0.2,
  device           = device_use,
  verbose          = TRUE
)

cat("\nTraining complete.\n")
cat("Validation MSE:\n")
print(round(rnn_fit$val_mse, 6))
```

### Validation MSE Summary Table

```{r}
#| label: val-mse-table
#| warning: false

val_mse_df <- data.frame(
  Model   = c("CausalLSTM", "RETAIN", "InterventionRNN"),
  Val_MSE = unname(rnn_fit$val_mse)
) |> dplyr::arrange(Val_MSE)

print(val_mse_df)
```

## Training Loss Curves

```{r}
#| label: plot-training-loss
#| warning: false

hist_list <- rnn_fit$histories

make_loss_df <- function(nm) {
  h <- hist_list[[nm]]
  if (is.null(h)) return(NULL)
  n_ep <- sum(h$val != 0)
  data.frame(
    epoch = seq_len(n_ep),
    val   = h$val[seq_len(n_ep)],
    train = h$train[seq_len(n_ep)],
    model = nm
  )
}

loss_df <- dplyr::bind_rows(lapply(names(hist_list), make_loss_df))
loss_df$model <- factor(loss_df$model,
                         levels = c("causal_lstm", "retain", "intervention_rnn"),
                         labels = c("CausalLSTM", "RETAIN", "InterventionRNN"))

ggplot(loss_df, aes(x = epoch, y = val, colour = model)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(
    "CausalLSTM"     = "#1b9e77",
    "RETAIN"         = "#d95f02",
    "InterventionRNN"= "#7570b3"
  )) +
  labs(
    title  = "Validation Loss Curves — RNN/LSTM Causal Models",
    x      = "Epoch",
    y      = "MSE",
    colour = NULL
  ) +
  theme_minimal()
```

### Training vs. Validation Loss per Model

```{r}
#| label: plot-train-val-loss
#| warning: false
#| fig-width: 12
#| fig-height: 4

plot_model_loss <- function(nm, label, colour) {
  h <- hist_list[[nm]]
  if (is.null(h)) return(NULL)
  n_ep <- sum(h$val != 0)
  df <- data.frame(
    epoch = seq_len(n_ep),
    Train = h$train[seq_len(n_ep)],
    Val   = h$val[seq_len(n_ep)]
  ) |> tidyr::pivot_longer(-epoch, names_to = "split", values_to = "loss")
  ggplot(df, aes(x = epoch, y = loss, colour = split)) +
    geom_line(linewidth = 0.85) +
    scale_colour_manual(values = c("Train" = colour, "Val" = "grey40")) +
    labs(title = label, x = "Epoch", y = "MSE", colour = NULL) +
    theme_minimal(base_size = 9)
}

p1 <- plot_model_loss("causal_lstm",       "CausalLSTM",      "#1b9e77")
p2 <- plot_model_loss("retain",            "RETAIN",          "#d95f02")
p3 <- plot_model_loss("intervention_rnn",  "InterventionRNN", "#7570b3")

patchwork::wrap_plots(p1, p2, p3, ncol = 3)
```

## Causal Matrix Visualization

For each model, the learned causal matrix $C \in \mathbb{R}^{d \times d}$ encodes variable-to-variable influence:

- **CausalLSTM**: the soft sigmoid mask $G$, where $G_{ij}$ encodes the learned strength of the causal link from variable $j$ to variable $i$.
- **RETAIN**: aggregated output-weight-scaled beta attribution $|W| \cdot |\bar\beta|$, averaged over validation batches.
- **InterventionRNN**: softmax-normalized average of learnable regime-conditioned causal weight matrices $\Psi_k$.

```{r}
#| label: extract-causal-matrices
#| warning: false

C_lstm   <- causal_matrix_rnn(rnn_fit, model = "causal_lstm")
C_retain <- causal_matrix_rnn(rnn_fit, model = "retain")
C_interv <- causal_matrix_rnn(rnn_fit, model = "intervention_rnn")

cat("CausalLSTM causal matrix (first 3 rows):\n")
print(round(C_lstm[1:3, ], 4))
cat("\nDiagonal (all 1 — guaranteed by mask design):", round(diag(C_lstm), 3), "\n")
```

### Heatmap Comparison

```{r}
#| label: heatmap-comparison
#| warning: false
#| fig-width: 15
#| fig-height: 5

plot_causal_heatmap <- function(C, title, var_names) {
  A <- as.matrix(C)
  diag(A) <- NA_real_
  rownames(A) <- var_names
  colnames(A) <- var_names
  df <- reshape2::melt(A, na.rm = TRUE)
  colnames(df) <- c("Target", "Source", "Weight")
  df$Target <- factor(df$Target, levels = rev(var_names))
  df$Source <- factor(df$Source, levels = var_names)
  ggplot(df, aes(x = Source, y = Target, fill = Weight)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = round(Weight, 2)), size = 2.5, colour = "black") +
    scale_fill_gradient(low = "white", high = "#2C7BB6",
                        na.value = "grey90", name = "Weight") +
    labs(title = title, x = "Source (cause)", y = "Target (effect)") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

p_lstm   <- plot_causal_heatmap(C_lstm,   "CausalLSTM — Learned Mask",          VAR_NAMES)
p_retain <- plot_causal_heatmap(C_retain, "RETAIN — Beta Attribution",           VAR_NAMES)
p_interv <- plot_causal_heatmap(C_interv, "InterventionRNN — Regime-Avg Matrix", VAR_NAMES)

patchwork::wrap_plots(p_lstm, p_retain, p_interv, ncol = 3)
```

### Causal Graph Network Diagrams

```{r}
#| label: network-diagram-comparison
#| warning: false

draw_rnn_igraph <- function(C, title, var_names, threshold = 0.12,
                             node_colour = "steelblue") {
  M     <- as.matrix(C)
  M[is.na(M)] <- 0.0
  A_bin <- (M > threshold) * 1L
  diag(A_bin) <- 0L
  g <- igraph::graph_from_adjacency_matrix(
    A_bin, mode = "directed", weighted = NULL, diag = FALSE
  )
  igraph::V(g)$name        <- var_names
  igraph::V(g)$color       <- node_colour
  igraph::V(g)$label       <- var_names
  igraph::V(g)$size        <- 28
  igraph::V(g)$label.color <- "white"
  igraph::V(g)$label.cex   <- 0.8
  igraph::E(g)$color       <- "grey50"
  igraph::E(g)$arrow.size  <- 0.6

  old_par <- par(mar = c(0, 0, 2, 0))
  plot(g,
       layout = igraph::layout_in_circle(g),
       main   = title,
       vertex.frame.color = "white")
  par(old_par)
}

par(mfrow = c(1, 3))
draw_rnn_igraph(C_lstm,   "CausalLSTM (threshold > 0.12)",     VAR_NAMES, 0.12, "steelblue")
draw_rnn_igraph(C_retain, "RETAIN (threshold > 0.12)",         VAR_NAMES, 0.12, "darkorange")
draw_rnn_igraph(C_interv, "InterventionRNN (threshold > 0.12)",VAR_NAMES, 0.12, "mediumpurple")
par(mfrow = c(1, 1))
```

## Graph Statistics

```{r}
#| label: graph-statistics
#| warning: false

graph_stats_rnn <- function(C, threshold = 0.12, name = "") {
  M     <- as.matrix(C)
  M[is.na(M)] <- 0.0
  A_bin <- (M > threshold) * 1L
  diag(A_bin) <- 0L
  g    <- igraph::graph_from_adjacency_matrix(
    A_bin, mode = "directed", weighted = NULL, diag = FALSE
  )
  d_n  <- igraph::gorder(g)
  e_n  <- igraph::gsize(g)
  dens <- if (d_n > 1) e_n / (d_n * (d_n - 1)) else 0.0
  data.frame(
    model          = name,
    nodes          = d_n,
    edges          = e_n,
    density        = round(dens, 4),
    is_dag         = igraph::is_dag(g),
    max_in_degree  = if (d_n > 0) max(igraph::degree(g, mode = "in"))  else 0L,
    max_out_degree = if (d_n > 0) max(igraph::degree(g, mode = "out")) else 0L
  )
}

stats_all <- do.call(rbind, list(
  graph_stats_rnn(C_lstm,   0.12, "CausalLSTM"),
  graph_stats_rnn(C_retain, 0.12, "RETAIN"),
  graph_stats_rnn(C_interv, 0.12, "InterventionRNN")
))
rownames(stats_all) <- NULL

cat("Graph statistics (threshold = 0.12):\n")
print(stats_all)
```

## Model Validation: Prediction Error on Held-Out Data

```{r}
#| label: validation-mse
#| warning: false

pred_lstm   <- predict(rnn_fit, model = "causal_lstm",      newdata = X_val)
pred_retain <- predict(rnn_fit, model = "retain",           newdata = X_val)
pred_interv <- predict(rnn_fit, model = "intervention_rnn", newdata = X_val,
                        intervention = matrix(0.0, nrow(X_val), LAG))  # zero-intervention baseline

mse_lstm   <- mean((pred_lstm   - Y_val)^2)
mse_retain <- mean((pred_retain - Y_val)^2)
mse_interv <- mean((pred_interv - Y_val)^2)

val_df <- data.frame(
  model   = c("CausalLSTM", "RETAIN", "InterventionRNN"),
  val_mse = c(mse_lstm, mse_retain, mse_interv)
)

cat("Validation MSE (next-step reconstruction):\n")
print(val_df)

ggplot(val_df, aes(x = reorder(model, val_mse), y = val_mse, fill = model)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_manual(values = c(
    "CausalLSTM"      = "#1b9e77",
    "RETAIN"          = "#d95f02",
    "InterventionRNN" = "#7570b3"
  )) +
  coord_flip() +
  labs(
    title = "Validation MSE by Model",
    x     = NULL,
    y     = "Mean Squared Error"
  ) +
  theme_minimal()
```

### Per-Variable Prediction Error

```{r}
#| label: per-variable-mse
#| warning: false

per_var_mse <- data.frame(
  variable         = VAR_NAMES,
  CausalLSTM       = colMeans((pred_lstm   - Y_val)^2),
  RETAIN           = colMeans((pred_retain - Y_val)^2),
  InterventionRNN  = colMeans((pred_interv - Y_val)^2)
)

cat("Per-variable MSE:\n")
print(data.frame(
  variable = per_var_mse$variable,
  round(per_var_mse[, -1], 5)
))

pv_long <- tidyr::pivot_longer(per_var_mse, -variable,
                                names_to = "model", values_to = "mse")

ggplot(pv_long, aes(x = variable, y = mse, fill = model)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(
    "CausalLSTM"     = "#1b9e77",
    "RETAIN"         = "#d95f02",
    "InterventionRNN"= "#7570b3"
  )) +
  labs(
    title = "Per-Variable Validation MSE",
    x     = "Sector",
    y     = "MSE",
    fill  = "Model"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## RETAIN Attribution Analysis

RETAIN's dual-channel attention produces interpretable attribution scores. We extract `alpha` (temporal importance) and `beta` (variable importance) from the validation set and visualise them.

```{r}
#| label: retain-attribution
#| warning: false
#| fig-width: 12
#| fig-height: 5

retain_mod <- rnn_fit$models[["retain"]]
retain_mod$eval()

dev_use <- if (is.null(device_use)) "cpu" else device_use
n_plot  <- min(200L, nrow(X_val))
x_sub   <- torch::torch_tensor(X_val[seq_len(n_plot), , ],
                                dtype  = torch::torch_float32(),
                                device = dev_use)

retain_out <- torch::with_no_grad({ retain_mod$forward(x_sub) })

# Alpha: temporal attention (n_plot x LAG)
alpha_mat <- as.matrix(retain_out$alpha$cpu())
alpha_mean <- colMeans(alpha_mat)

# Beta: variable attention averaged over time (n_plot x d)
beta_3d  <- retain_out$beta$abs()$cpu()
beta_mat <- as.matrix(beta_3d$mean(dim = 2L))  # (n_plot, d)
beta_mean <- colMeans(beta_mat)

# --- Plot 1: Mean temporal attention across validation set ---
alpha_df <- data.frame(
  lag    = seq_len(LAG),
  weight = alpha_mean
)

p_alpha <- ggplot(alpha_df, aes(x = lag, y = weight)) +
  geom_col(fill = "#d95f02", alpha = 0.8) +
  labs(
    title = "RETAIN: Mean Temporal Attention (α)",
    x     = "Lag (steps back from t)",
    y     = "Mean α weight"
  ) +
  theme_minimal()

# --- Plot 2: Mean variable attention ---
beta_df <- data.frame(
  variable = VAR_NAMES,
  weight   = beta_mean
)

p_beta <- ggplot(beta_df, aes(x = reorder(variable, -weight), y = weight, fill = variable)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "RETAIN: Mean Variable Attention (|β|)",
    x     = "Variable",
    y     = "Mean |β| weight"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

patchwork::wrap_plots(p_alpha, p_beta, ncol = 2)
```

## Intervention Effect Analysis

For the Intervention-Aware RNN we compare predictions under zero-intervention vs. the observed intervention signal on the validation set.

```{r}
#| label: intervention-effect
#| warning: false
#| fig-width: 10
#| fig-height: 4

interv_mod <- rnn_fit$models[["intervention_rnn"]]
interv_mod$eval()

dev_use <- if (is.null(device_use)) "cpu" else device_use
n_plot  <- min(200L, nrow(X_val))
x_sub_t <- torch::torch_tensor(X_val[seq_len(n_plot), , ],
                                dtype   = torch::torch_float32(),
                                device  = dev_use)

# Reconstruct the intervention window for validation samples using X_val rows
n_full <- nrow(X_val)
# X_val rows come from positions (split+1):(N) of X_seq, which map to time
# indices (split+1):(N) in the lagged dataset, i.e. original rows (split+1+LAG-1)...
I_val_win <- matrix(0.0, n_full, LAG)
offset <- split  # first validation sample starts at row split+1 in X_seq
for (t in seq_len(n_full)) {
  ts   <- offset + t            # position in X_seq (1-indexed)
  i_st <- ts                    # original series index for the lag window start
  i_en <- ts + LAG - 1L
  if (i_en <= length(I_raw))
    I_val_win[t, ] <- I_raw[i_st:i_en]
}
I_val_win <- pmin(pmax(I_val_win, 0), 1)

i_obs_t  <- torch::torch_tensor(I_val_win[seq_len(n_plot), ],
                                 dtype  = torch::torch_float32(),
                                 device = dev_use)
i_zero_t <- torch::torch_zeros_like(i_obs_t)

out_obs  <- torch::with_no_grad({ interv_mod$forward(x_sub_t, interv = i_obs_t)  })
out_zero <- torch::with_no_grad({ interv_mod$forward(x_sub_t, interv = i_zero_t) })

pred_obs  <- as.matrix(out_obs$pred$cpu())
pred_zero <- as.matrix(out_zero$pred$cpu())

# Average absolute prediction difference per variable
diff_df <- data.frame(
  variable = VAR_NAMES,
  avg_diff = colMeans(abs(pred_obs - pred_zero))
)

ggplot(diff_df, aes(x = reorder(variable, -avg_diff), y = avg_diff, fill = variable)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Intervention Effect: Mean |Prediction Change| under Observed vs. Zero Intervention",
    x     = "Sector variable",
    y     = "Mean |Δ prediction|"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Interpretation Notes

- **CausalLSTM** yields an explicit, sparse, directed adjacency matrix through a learnable structural mask — the closest to a formal causal graph among the three.
- **RETAIN** gives time- and variable-level attribution, making it the most interpretable architecture for explaining decisions to domain experts.
- **Intervention-Aware RNN** can adapt dynamics when volatility/shock regimes change and explicitly quantifies how an intervention changes predictions.

In practice, combining these models is often powerful:

1. Use **RETAIN** for attribution-based explanation.
2. Use **CausalLSTM** for enforcing structural sparsity in the causal graph.
3. Use **Intervention-Aware RNN** when regime shifts or external shocks are known to occur.

## Summary and Conclusions

In this notebook, we explored causal modelling approaches using temporal deep learning models on multivariate S&P 500 sector ETF data. We compared three architectures — **CausalLSTM**, **RETAIN**, and **Intervention-Aware RNN** — on predictive accuracy and causal interpretability.

**Key Findings:**

- All three models achieved comparable validation MSE, demonstrating that causal structural constraints need not hurt predictive performance.
- **CausalLSTM** provided a sparse, interpretable adjacency matrix that can reveal directed causal relationships between variables through the learned mask.
- **RETAIN** generated meaningful dual-channel attribution scores ($\alpha$ temporal, $\beta$ variable), making it well-suited for interpretation-first workflows.
- The **Intervention-Aware RNN** effectively captured regime changes, with distinct prediction responses under observed vs. zero intervention, especially useful in datasets with known exogenous shocks.

**Conclusions and Recommendations:**

- Causality-aware neural architectures (like CausalLSTM) enhance both transparency and scientific value in longitudinal modelling tasks.
- Attribution-based models (like RETAIN) are recommended when explaining predictions to stakeholders is a priority.
- Intervention-Aware RNNs are most useful when domain knowledge provides a binary or continuous treatment/shock signal.
- In practice, deploying all three in combination leverages their complementary strengths.
- Further work could include benchmarking on richer synthetic or real-world datasets, bootstrap uncertainty on causal matrices, and integration with formal causal discovery pipelines.

## Resources

### Primary References

- **CausalLSTM (2025):** [Causality-Aware Recurrent Neural Networks for Time Series Analysis](https://pmc.ncbi.nlm.nih.gov/articles/PMC11876796/) — structural mask on LSTM inputs for sparse causal discovery.
- **RETAIN (2016):** [An Interpretable Predictive Model for Healthcare Using Reverse Time Attention Mechanism](https://arxiv.org/abs/1608.05745) — Choi et al.; dual GRU-based attention for interpretable sequential modelling.
- **Intervention-Aware RNNs:** [Detecting and Adapting to Regime Change](https://arxiv.org/abs/2010.08918) — regime-conditioned recurrent networks for non-stationary sequences.

### Background Texts

- **Causality (Judea Pearl):** [Causality: Models, Reasoning and Inference](https://bayes.cs.ucla.edu/BOOK-2K/) — definitive reference for structural causal models and Pearl's Ladder of Causation.
- **Elements of Causal Inference (Peters, Janzing & Schölkopf):** [MIT Press](https://mitpress.mit.edu/9780262037310/elements-of-causal-inference/) — rigorous treatment of SCMs and identifiability.
- **The Book of Why (J. Pearl & D. Mackenzie)** — accessible introduction to causal reasoning.

### Software and Datasets

- **R torch:** <https://torch.mlverse.org/> — PyTorch bindings for R used by `{RCausalML}`.
- **tidyquant:** <https://business-science.github.io/tidyquant/> — tidy finance data retrieval.
- **{RCausalML}:** `rnn_causal_model()`, `causal_matrix_rnn()`, `predict.rnn_causal_model()` — main API for this notebook.

### Companion Notebooks in This Series

- `05-04-01-DeepCausalML-timeseries-neural-granger-cusuality-r.qmd` — Neural Granger Causality (cMLP, cLSTM, SRU, NRI) on the same sector ETF universe.
- `05-04-02-DeepCausalML-timeseries-structural-causal-model-SMC-r.qmd` — Deep Structural Causal Models: DeepSCM, DECI, DynoTEARS on sector ETF data.
- `05-04-03-DeepCausalML-attension-transformer-r.qmd` — Attention-Based/Transformer Causal Models: TCDF, CausalTransformer, TFT on sector ETF data.
- `05-03-05-DeepCausalML-causal-structural-learning-regularization-CASTLE-r.qmd` — CASTLE: simultaneous supervised prediction and causal graph discovery.

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Hidden state $h_t$ | An internal vector that the RNN updates at each step, carrying "memory" of prior inputs. | Like a running mental note of everything you've read in a book so far. |
| LSTM gates | Learned sigmoid-activated gates (forget, input, output) that control what information to keep or discard. | A spam filter deciding which emails to delete vs. archive vs. read. |
| Causal mask $G$ | A matrix that zeroes out inputs from variables that should not causally influence a target variable. | A teacher blocking a student from looking at irrelevant notes during a test. |
| Sparsity penalty | A regularisation term that penalises non-zero off-diagonal mask entries, encouraging simpler graphs. | A penalty for using too many ingredients in a recipe. |
| Reverse-time attention | Attention applied to the sequence in reverse chronological order, so recent events are processed first. | Reading a mystery novel from the last chapter backwards to find the most recent clues first. |
| Alpha attention $\alpha_t$ | Scalar weight per timestep, indicating how important that timestep was for the prediction. | Highlighting which day in a week was most predictive of the weekend outcome. |
| Beta attention $\beta_t$ | Per-feature vector weight at each timestep, indicating which variable mattered at that moment. | At each highlighted day, noting which specific event (weather, news, etc.) drove the outcome. |
| Regime | A latent system state (e.g., bull vs. bear market) that governs the data-generating process. | A traffic light: red (stop), yellow (caution), green (go) — different rules per state. |
| Intervention indicator $I_t$ | A binary or continuous signal marking that an external event occurred at timestep $t$. | A flag indicating that a new drug was administered on a given day in a clinical trial. |
| Regime-conditioned causal matrix | A separate $d \times d$ causal weight matrix per latent regime, averaged by regime probability. | Different influence tables for different market conditions — bull vs. bear. |
| Early stopping | Halting training when validation loss stops improving, preventing overfitting. | Stopping practice when further rehearsal no longer improves performance. |
| AdamW | Adam optimizer with decoupled weight decay regularisation. | Gradient descent that adapts learning rates per parameter and penalises large weights. |
| Cosine annealing | A learning-rate schedule that decays the LR following a cosine curve. | Gradually slowing down training as you approach convergence. |
| Lag / window | Number of past time steps used as input to the model. | Using the last 20 trading days to predict the next day's return. |
