# Neural Granger Causality Models {.unnumbered}

**Neural Granger Causality** extends the classical Granger framework into the nonlinear regime by replacing linear coefficient blocks with neural networks while preserving the sparse input-selection logic of group LASSO. The result is a family of models that discover *who causes whom* in a multivariate time series without assuming linearity.

> If a neural network trained to predict $Y_t$ assigns zero (or near-zero) weight to all lags of $X$, then $X$ does not Granger-cause $Y$ — regardless of how complex the network is.

## Overview

Classical Granger causality tests whether past values of $X$ significantly improve the forecast of $Y$ beyond what $Y$'s own history provides. For lag order $p$ this is the vector autoregression (VAR):

$$
Y_t = \sum_{k=1}^{p} A_k Y_{t-k} + \epsilon_t
$$

If the coefficient block connecting $X_{t-k} \rightarrow Y_t$ is nonzero, then $X$ Granger-causes $Y$. The method is powerful but **inherently linear** and can miss nonlinear dynamics.

**Neural Granger Causality** solves this by replacing the fixed coefficient $A_k$ with a learned nonlinear function, while enforcing structured sparsity — specifically **group LASSO** — on the input layer weights. A zero group norm for all lags of variable $X$ means $X$ is excluded from predicting $Y$.

## The Four Models

### 1) cMLP — Component-wise MLP

A separate MLP is trained for each target variable $Y_i$:

$$
\hat{Y}_{i,t} = \text{MLP}_i\!\left(Y_{1,t-1:t-p},\ldots,Y_{d,t-1:t-p}\right)
$$

Training objective:

$$
\mathcal{L} = \underbrace{\text{MSE}}_{\text{prediction}} + \lambda \underbrace{\sum_j \lVert W_j^{(1)} \rVert_F}_{\text{group LASSO on input weights}}
$$

Sparsity on the first-layer weight block $W_j^{(1)}$ (columns corresponding to all lags of variable $j$) causes entire input variables to be selected or zeroed out, making the inferred causal graph readable directly from those norms.

### 2) cLSTM — Component-wise LSTM

One LSTM per target variable. Better for sequential / long-lag dynamics:

$$
h_t = \text{LSTM}_i(Y_{1:d,\ t-p:t-1}), \qquad \hat{Y}_{i,t} = W_{\text{out}} h_t
$$

Group LASSO is applied to the LSTM input-weight matrix $W_{ih}$, grouping all gates for each source variable $j$ together.

### 3) ECONOMY-SRU — Structurally Constrained Recurrent Unit

A simplified GRU with an explicit learnable **causal mask**:

$$
G \in [0,1]^{d \times d}, \qquad G_{ij} = \sigma(\theta_{ij})
$$

The mask gates each variable's contribution to every other variable's GRU update. Sparsity is induced by an $\ell_1$ penalty $\lambda \sum_{ij} G_{ij}$ on the mask, driving weak links to zero.

### 4) NRI — Neural Relational Inference

NRI (Kipf et al., 2018) is a graph-VAE approach that simultaneously learns edge types and dynamics:

- **Encoder** infers latent edge distributions $q(z \mid X)$ via a message-passing network
- **Decoder** predicts trajectories from sampled edges

$$
\mathcal{L} = \mathbb{E}_{q(z\mid X)}[\log p(X\mid z)] - \mathrm{KL}[q(z\mid X)\|p(z)]
$$

The posterior probability of a *causal edge* between each variable pair forms the inferred adjacency matrix.

## Implementation in R

We use {RCausalML}'s `neural_granger_ml()` wrapper — which packages all four models (cMLP, cLSTM, EconomySRU, NRI) — trained on **daily log returns of S&P 500 sector ETFs** (2018–2023).

### Load and Check Required Libraries

```{r}
#| label: packages-list
#| warning: false
packages <- c(
  'tidyverse',
  'plyr',
  'RCausalML',
  'torch',
  'tidyquant',
  'reshape2',
  'scales'
)
```

### Install Missing Packages

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# Install missing packages
#new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: verify-installation
#| warning: false
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: load-required-libraries
#| warning: false
# When rendering from package root, use local RCausalML (so neural_granger_ml fixes are used)
if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(".", quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

```{r}
#| label: setup
#| include: true
set.seed(42)
if (requireNamespace("torch", quietly = TRUE)) {
  torch_manual_seed(42)
  device_use <- if (torch::cuda_is_available()) "cuda" else "cpu"
  cat(sprintf("Using device: %s\n", device_use))
} else {
  device_use <- "cpu"
  cat("torch not available; some models will be skipped.\n")
}
```

## Data: S&P 500 Sector ETF Daily Log Returns

We fetch daily adjusted closing prices for eight major S&P 500 sector ETFs using `tidyquant`, compute daily log returns, standardise them (zero mean, unit variance), and store them in a matrix `data_np` with columns named after each sector.

The sector ETFs are:

| Ticker | Sector |
|--------|--------|
| XLF | Financials |
| XLE | Energy |
| XLK | Technology |
| XLV | Healthcare |
| XLI | Industrials |
| XLY | Consumer Discretionary |
| XLP | Consumer Staples |
| XLU | Utilities |

```{r}
#| label: data-loading
#| warning: false
#| message: false

TICKERS <- c(
  XLF = "Financials",
  XLE = "Energy",
  XLK = "Technology",
  XLV = "Healthcare",
  XLI = "Industrials",
  XLY = "ConsumerDisc",
  XLP = "ConsumerStap",
  XLU = "Utilities"
)

# Download adjusted closing prices via tidyquant
raw_prices <- tq_get(
  names(TICKERS),
  from = "2018-01-01",
  to   = "2024-01-01",
  get  = "stock.prices"
) |>
  select(symbol, date, adjusted) |>
  pivot_wider(names_from = symbol, values_from = adjusted) |>
  arrange(date)

# Daily log returns; drop the first row (NA from lag)
returns_mat <- raw_prices |>
  select(-date) |>
  mutate(across(everything(), ~ log(. / lag(.)))) |>
  slice(-1)

# Rename columns to sector names (match TICKERS order)
colnames(returns_mat) <- TICKERS[colnames(returns_mat)]

cat(sprintf("Shape: %d rows x %d columns\n", nrow(returns_mat), ncol(returns_mat)))
print(round(sapply(returns_mat, function(x) c(
  mean = mean(x, na.rm = TRUE),
  sd   = sd(x, na.rm = TRUE),
  min  = min(x, na.rm = TRUE),
  max  = max(x, na.rm = TRUE)
)), 4))

VAR_NAMES <- colnames(returns_mat)
T_obs <- nrow(returns_mat)
d <- ncol(returns_mat)
cat(sprintf("\nVariables (%d): %s\n", d, paste(VAR_NAMES, collapse = ", ")))
cat(sprintf("Time steps: %d\n", T_obs))
```

### Standardise the Returns

```{r}
#| label: standardise
#| warning: false
# Standardise to zero mean / unit variance (as float32 equivalent in R)
col_means <- colMeans(returns_mat, na.rm = TRUE)
col_sds   <- apply(returns_mat, 2, sd, na.rm = TRUE)
col_sds[col_sds == 0] <- 1

data_np <- scale(as.matrix(returns_mat), center = col_means, scale = col_sds)
storage.mode(data_np) <- "double"

cat("Standardised data: first 5 rows\n")
print(round(head(data_np, 5), 4))
```

## Data Preparation: Lag Dataset

For neural Granger causality we frame the multivariate time series as a **supervised regression problem**: given all variables over a lag window of $p$ steps, predict the next values. Sparsity in the first-layer weights then reveals the causal structure.

```{r}
#| label: lag-dataset
#| warning: false

LAG <- 5L  # Trading week

build_lag_dataset <- function(data, lag) {
  n <- nrow(data) - lag
  d <- ncol(data)
  x_flat <- matrix(0, nrow = n, ncol = d * lag)
  x_seq  <- array(0,  dim = c(n, lag, d))
  y      <- matrix(0, nrow = n, ncol = d)
  for (t in seq_len(n)) {
    window <- data[t:(t + lag - 1), , drop = FALSE]  # (lag, d)
    x_flat[t, ] <- as.vector(t(window))              # row-major flatten: (d*lag,)
    x_seq[t, , ] <- window
    y[t, ] <- data[t + lag, ]
  }
  colnames(y) <- colnames(data)
  list(x_flat = x_flat, x_seq = x_seq, y = y)
}

prep      <- build_lag_dataset(data_np, LAG)
n_total   <- nrow(prep$y)
split_idx <- floor(0.8 * n_total)

x_flat_tr  <- prep$x_flat[seq_len(split_idx), , drop = FALSE]
x_flat_val <- prep$x_flat[(split_idx + 1L):n_total, , drop = FALSE]
x_seq_tr   <- prep$x_seq[seq_len(split_idx), , , drop = FALSE]
x_seq_val  <- prep$x_seq[(split_idx + 1L):n_total, , , drop = FALSE]
y_tr       <- prep$y[seq_len(split_idx), , drop = FALSE]
y_val      <- prep$y[(split_idx + 1L):n_total, , drop = FALSE]

cat(sprintf(
  "Train: x_flat %s | Val: x_flat %s\n",
  paste(dim(x_flat_tr),  collapse = " x "),
  paste(dim(x_flat_val), collapse = " x ")
))
```

## Fit All Four Neural Granger Models

{RCausalML}'s `neural_granger_ml()` trains **cMLP**, **cLSTM**, **EconomySRU**, and **NRI** in a single call. Under the hood each model is a `torch` `nn_module`; training uses Adam with gradient clipping and group-LASSO / sparsity regularisation.

```{r}
#| label: train-models
#| warning: false

EPOCHS     <- 60L
LAM        <- 0.005
HIDDEN     <- 32L
LR         <- 5e-4
BATCH_SIZE <- 32L

cat("Fitting Neural Granger Causality models via RCausalML::neural_granger_ml() ...\n")

ngc_fit <- neural_granger_ml(
  data       = data_np,
  lag        = LAG,
  models     = c("cmlp", "clstm", "economysru", "nri"),
  hidden     = HIDDEN,
  lam        = LAM,
  epochs     = EPOCHS,
  batch_size = BATCH_SIZE,
  lr         = LR,
  val_split  = 0.2,
  n_edge_types = 2L,
  temp       = 0.5,
  device     = device_use,
  verbose    = TRUE
)

cat("\nTraining complete. Validation MSE per model:\n")
val_mse_df <- data.frame(
  Model   = names(ngc_fit$val_mse),
  Val_MSE = unlist(ngc_fit$val_mse),
  Val_RMSE = sqrt(unlist(ngc_fit$val_mse)),
  row.names = NULL
)
val_mse_df <- val_mse_df[order(val_mse_df$Val_MSE), ]
print(round(val_mse_df, 6))
```

### Training Curves

```{r}
#| label: training-curves
#| warning: false
#| fig.width: 12
#| fig.height: 4

model_labels <- c(cmlp = "cMLP", clstm = "cLSTM", economysru = "ECONOMY-SRU", nri = "NRI")

loss_list <- lapply(names(ngc_fit$histories), function(m) {
  h <- ngc_fit$histories[[m]]
  data.frame(
    model  = model_labels[m],
    epoch  = seq_along(h$train_loss),
    train  = h$train_loss,
    val    = h$val_loss
  )
})
loss_df <- bind_rows(loss_list)

loss_long <- loss_df |>
  pivot_longer(cols = c(train, val), names_to = "split", values_to = "loss") |>
  mutate(split = factor(split, levels = c("train", "val"), labels = c("Train", "Validation")))

ggplot(loss_long, aes(x = epoch, y = loss, color = split)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ model, nrow = 1, scales = "free_y") +
  scale_color_manual(values = c(Train = "#1b9e77", Validation = "#d95f02")) +
  labs(
    title  = "Neural Granger Causality — Training Curves (S&P 500 Sector ETFs)",
    x      = "Epoch",
    y      = "MSE Loss",
    color  = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

## Inferred Causal Matrices

Each model produces a $d \times d$ matrix where entry $(i, j)$ encodes the **strength of influence from source variable $j$ to target variable $i$**:

- **cMLP / cLSTM**: Frobenius norm of first-layer weight block for variable $j$
- **EconomySRU**: learned sigmoid gate value $G_{ij}$
- **NRI**: posterior probability of a causal edge from $j$ to $i$

```{r}
#| label: causal-matrices
#| warning: false
#| fig.width: 11
#| fig.height: 10

plot_causal_heatmap <- function(mat, title, var_names) {
  mat_df <- as.data.frame(mat)
  colnames(mat_df) <- var_names
  rownames(mat_df) <- var_names
  mat_long <- mat_df |>
    rownames_to_column("Target") |>
    pivot_longer(-Target, names_to = "Source", values_to = "Strength") |>
    mutate(
      Target = factor(Target, levels = rev(var_names)),
      Source = factor(Source, levels = var_names)
    )
  ggplot(mat_long, aes(x = Source, y = Target, fill = Strength)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradient(low = "white", high = "#6a0dad", name = "Strength") +
    labs(title = title, x = "Source variable", y = "Target variable") +
    theme_minimal() +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, size = 9),
      axis.text.y  = element_text(size = 9),
      plot.title   = element_text(face = "bold", size = 10)
    )
}

p_cmlp <- plot_causal_heatmap(
  ngc_fit$causal_matrices$cmlp, "cMLP inferred causal strengths", VAR_NAMES)
p_clstm <- plot_causal_heatmap(
  ngc_fit$causal_matrices$clstm, "cLSTM inferred causal strengths", VAR_NAMES)
p_sru <- plot_causal_heatmap(
  ngc_fit$causal_matrices$economysru, "ECONOMY-SRU inferred causal strengths", VAR_NAMES)
p_nri <- plot_causal_heatmap(
  ngc_fit$causal_matrices$nri, "NRI inferred causal strengths", VAR_NAMES)

library(patchwork)
(p_cmlp | p_clstm) / (p_sru | p_nri) +
  plot_annotation(
    title   = "Neural Granger Causality — Inferred Causal Strengths (S&P 500 Sector ETFs)",
    caption = "Row = target variable, Column = source variable. Higher = stronger inferred causal link.",
    theme   = theme(plot.title = element_text(face = "bold", size = 12))
  )
```

## Validation: Model Performance Comparison

```{r}
#| label: performance-bar
#| warning: false
#| fig.width: 7
#| fig.height: 4

val_mse_plot <- data.frame(
  Model   = model_labels[names(ngc_fit$val_mse)],
  Val_MSE = unlist(ngc_fit$val_mse)
)

ggplot(val_mse_plot, aes(x = reorder(Model, Val_MSE), y = Val_MSE, fill = Model)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title  = "Validation MSE by Model (S&P 500 Sector ETFs)",
    x      = NULL,
    y      = "Validation MSE"
  ) +
  coord_flip() +
  theme_minimal()
```

## Consensus Causal Graph

Average the causal matrices across all four models (after min-max normalisation) to derive a **consensus strength** for each directed variable pair.

```{r}
#| label: consensus-graph
#| warning: false
#| fig.width: 9
#| fig.height: 5

minmax <- function(m) {
  rng <- range(m)
  if (diff(rng) < 1e-10) return(matrix(0, nrow(m), ncol(m)))
  (m - rng[1]) / diff(rng)
}

mats_norm <- lapply(ngc_fit$causal_matrices, minmax)
consensus <- Reduce("+", mats_norm) / length(mats_norm)
rownames(consensus) <- VAR_NAMES
colnames(consensus) <- VAR_NAMES
diag(consensus) <- 0

cat("Consensus causal matrix (min-max normalised, diagonal zeroed):\n")
print(round(consensus, 3))

# Visualise as a heatmap
consensus_long <- as.data.frame(consensus) |>
  rownames_to_column("Target") |>
  pivot_longer(-Target, names_to = "Source", values_to = "Strength") |>
  mutate(
    Target = factor(Target, levels = rev(VAR_NAMES)),
    Source = factor(Source, levels = VAR_NAMES)
  )

ggplot(consensus_long, aes(x = Source, y = Target, fill = Strength)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = ifelse(Strength > 0.3, sprintf("%.2f", Strength), "")),
            size = 3, color = "white") +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b", name = "Consensus\nStrength") +
  labs(
    title   = "Consensus Causal Strengths across cMLP / cLSTM / ECONOMY-SRU / NRI",
    subtitle = "Values > 0.30 labelled. Diagonal zeroed.",
    x       = "Source variable",
    y       = "Target variable"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Notes and Extensions

-   This is a **pedagogical** demonstration on real but noisy financial data. Inferred causal links should be treated as **exploratory** — financial returns are weakly causal and heavily confounded.
-   You can tune `LAG`, `LAM`, `HIDDEN`, and `EPOCHS` to trade off fit vs. sparsity. Increasing `LAM` produces sparser, more interpretable graphs; decreasing it may reveal weaker relationships.
-   For stricter graph sparsity with NRI, consider annealing the Gumbel-Softmax temperature and adding explicit edge-sparsity priors.
-   A rigorous causal interpretation requires domain knowledge, interventional data, or additional structural assumptions (e.g., acyclicity constraints).

## Summary and Conclusions

-   **Classical Granger causality** tests linear predictive improvement but misses nonlinear relationships — Neural Granger Causality addresses this by replacing coefficient blocks with neural networks.
-   **Four architectures** cover the main design choices: feed-forward per-variable heads (**cMLP**), recurrent per-variable heads (**cLSTM**), structured gated recurrence (**EconomySRU**), and amortised graph inference (**NRI**).
-   **Structured sparsity** (group LASSO, $\ell_1$ mask penalty, or KL prior) is the common thread — without it, neural networks spread weight everywhere and no causal graph can be read off.
-   **{RCausalML}** exposes all four models through a single `neural_granger_ml()` call; the returned object contains trained models, per-epoch histories, validation MSEs, and $d \times d$ causal matrices ready for downstream analysis.
-   **On S&P 500 sector ETFs** (2018–2023), cLSTM and ECONOMY-SRU typically achieve slightly lower validation MSE than cMLP; NRI behaves differently because it optimises an ELBO rather than pure MSE.
-   **Consensus graph** (average of min-max normalised matrices) can highlight cross-model stable links; strong consensus pairs across all four models are the most reliable candidates for further investigation.

## Resources

### Neural Granger Causality — Primary References

-   **Tank et al. (2018) — cMLP / cLSTM:** [Neural Granger Causality for Nonlinear Time Series](https://arxiv.org/abs/1802.05842) — introduces the component-wise MLP and LSTM with group-LASSO input regularisation.
-   **Kipf et al. (2018) — NRI:** [Neural Relational Inference for Interacting Systems](https://arxiv.org/abs/1802.04687) — graph-VAE formulation with encoder/decoder message-passing.

### Granger Causality Background

-   **Granger (1969):** Investigating Causal Relations by Econometric Models and Cross-spectral Methods — the original formulation.
-   **Shojaie & Fox (2022):** [Granger Causality: A Review and Recent Advances](https://www.annualreview.org/doi/10.1146/annurev-statistics-040120-010930) — comprehensive modern review.

### Software Stack (this tutorial)

-   **torch for R:** <https://torch.mlverse.org/>
-   **tidyquant:** <https://business-science.github.io/tidyquant/>
-   **RCausalML:** `neural_granger_ml()` in `R/causalDeepNet.R`
-   **patchwork:** <https://patchwork.data-imaginist.com/>

### Companion Material

-   `05-03-05-DeepCausalML-causal-structural-learning-regularization-CASTLE-r.qmd` — CASTLE uses a similar group-sparsity idea for structure learning in a supervised setting; compare the two approaches.
-   `05-04-00-DeepCuasalML-timesseries-introduction-r.qmd` — overview of all deep causal time series methods in this series.

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Granger causality | X causes Y if knowing X's past improves forecasting Y. | Past Energy ETF returns help predict Financials ETF returns. |
| Group LASSO | Regularisation that zeros out entire groups of weights at once. | All five lagged inputs of Energy are set to zero if Energy doesn't Granger-cause the target. |
| cMLP | Separate MLP per target variable with group-sparse input layer. | Eight MLPs, one per sector ETF, each reading all other sectors' lagged returns. |
| cLSTM | Separate LSTM per target with sparse input-weight Granger structure. | Same as cMLP but with memory for sequential dependencies. |
| EconomySRU | GRU with a learnable causal gate matrix $G_{ij}$. | The model learns which sector pairs actively influence each other's hidden state. |
| NRI | Graph-VAE that jointly infers edge types and dynamics. | A probabilistic model that puts a probability on each "Energy → Financials" edge. |
| Causal matrix | $d \times d$ matrix of inferred directed influence strengths. | Row = target, column = source; entry $(i,j)$ indicates how strongly $j$ Granger-causes $i$. |
| Consensus graph | Average normalised causal matrix across models. | If all four models agree on a strong link, it is more likely to be a real relationship. |
