# 6. Domain Adaptation Learner (DA-Learner)

Thisssss notebook demonstrates the **Domain Adaptation Learner (DA-Learner)**, a meta-learner for estimating Conditional Average Treatment Effects (CATE) in the presence of significant domain shift between treatment and control groups. The DA-Learner uses propensity-score-based weighting to adapt the outcome models, making it robust to selection bias and distributional differences. We wwill use RCausalML's `DomainAdaptationLearner` to implement this method on a real dataset, showcasing its advantages over standard meta-learners in scenarios with strong domain shift.          

## Overview

**Domain Adaptation Learner** (also called **DA-Learner** or `DomainAdaptationLearner` in the EconML library) is a meta-learner designed for estimating the **Conditional Average Treatment Effect (CATE)**, denoted as $\tau(x) = E[Y(1) - Y(0) \mid X = x]$, from observational data.

It is particularly useful when the probability distributions of the outcome under treatment $P(Y^1 \mid X)$ and under control $P(Y^0 \mid X)$ differ substantially (i.e., there is a significant **domain shift** or **selection bias** between the treatment and control groups). Standard meta-learners like the plain T-Learner can produce biased estimates in such cases because the outcome models trained on one group do not generalize well to the other group.

It can be viewed as a **domain-adapted variant of the T-Learner** (or sometimes described in literature as a weighted extension related to the X-Learner) that uses **propensity-score-based sample weighting** to align the training distributions. This makes the outcome models more robust when predicting counterfactuals across domains.

### How It Works 

The algorithm follows these main steps:

1. **Estimate the propensity score** $g(x) = P(T=1 \mid X=x)$ (the probability of receiving treatment given covariates $X$). This is usually done with a separate classifier (e.g., logistic regression, random forest, etc.).

2. **Train two outcome models with domain-adaptive weighting** (this is the key "domain adaptation" part):
   - Control outcome model $\hat{\mu}_0$ (predicts $Y^0$) is trained **only on control samples** ($T=0$), but each control sample is re-weighted by $\frac{g(X^0)}{1 - g(X^0)}$.  
     This weighting makes the control data distribution look more like the treatment-group distribution.
   - Treatment outcome model $\hat{\mu}_1$ (predicts $Y^1$) is trained **only on treated samples** ($T=1$), but each treated sample is re-weighted by $\frac{1 - g(X^1)}{g(X^1)}$.  
     This makes the treated data distribution look more like the control-group distribution.

   Any machine-learning regressor can be used for $M_1$ and $M_2$ (e.g., RandomForest, LightGBM, neural nets).

3. **Compute imputed treatment-effect residuals** (pseudo-effects) for each group:
   - For treated units: $\hat{D}^1 = Y^1 - \hat{\mu}_0(X^1)$  
     (actual outcome minus the *adapted* prediction of what the outcome would have been under control).
   - For control units: $\hat{D}^0 = \hat{\mu}_1(X^0) - Y^0$  
     (the *adapted* prediction of what the outcome would have been under treatment minus the actual outcome).

4. **Train a final CATE model** on the combined residuals:
   - Concatenate the two sets of residuals and features: $\hat{D}^0$ (from controls) and $\hat{D}^1$ (from treated), with their respective $X^0$ and $X^1$.
   - Fit a model $M_3$ that predicts the residuals as a function of $X$. This final model directly gives $\hat{\tau}(x)$.

### Mathematical Formulation

The full procedure is expressed as:

$$
\begin{split}
\hat{\mu}_0 &= M_1\left(Y^0 \sim X^0,\ \text{weights}=\frac{g(X^0)}{1-g(X^0)}\right) \\
\hat{\mu}_1 &= M_2\left(Y^1 \sim X^1,\ \text{weights}=\frac{1-g(X^1)}{g(X^1)}\right) \\
\hat{D}^1 &= Y^1 - \hat{\mu}_0(X^1) \\
\hat{D}^0 &= \hat{\mu}_1(X^0) - Y^0 \\
\hat{\tau} &= M_3(\hat{D}^0 \mid \hat{D}^1 \sim X^0 \mid X^1)
\end{split}
$$

where:

- $M_1, M_2, M_3$ = arbitrary ML models,
- $g(x)$ = propensity score,
- $|$ denotes dataset concatenation (training on the combined data).

### Why the Weighting Works (Intuition)

The weights come from **importance weighting** in domain adaptation:

- To predict the counterfactual $Y^0$ for *treated* individuals, you want the control training data to mimic the treated population → multiply control samples by $\frac{P(\text{treated})}{P(\text{control})} = \frac{g}{1-g}$.
- Symmetrically for the treatment model.

This reduces bias when the support or distribution of $X$ differs between groups, which is common in real-world observational data.

## Implementation in in R

RCausalML provides an implementation of the DA-Learner through the `DomainAdaptationLearner` class. Below is an example of how to use it:


### Setup

#### Define required packages

```{r}
#| label: packages-list
#| warning: false
packages <- c(
  "tidyverse",
  "plyr",
  "RCausalML",
  "causaldata",
  "mlbench",
  "xgboost",
  "kernelshap",
  "shapviz",
  "future",
  "mlr3",
  "mlr3learners",
  "nnet",
  "gridExtra",
  "survival"
)
```

### Install Missing Packages

#### Optional install commands (commented out)

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
# if (length(new_packages)) install.packages(new_packages)
# remotes::install_github("mlr-org/mlr3learners")
```

### Verify Installation

#### Check package namespaces

```{r}
#| label: verify-installation
#| warning: false
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

#### Attach libraries and local RCausalML

```{r}
#| label: load-required-libraries
#| warning: false

if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(".", quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Data

Synthetic data with **confounded treatment** (propensity depends on $X_1$ via a logistic link) and an outcome that depends on treatment and $X_1$, matching the usual Python sketch:

$T \sim \mathrm{Bernoulli}(\mathrm{expit}(X_1))$, $y = (1 + 0.5 X_1) T + X_1 + \varepsilon$.

#### Simulate confounded synthetic data

```{r}
#| label: synthetic-dgp-da-learner
#| warning: false

set.seed(123)
n <- 1000L
p <- 5L
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
colnames(X) <- paste0("X", seq_len(p))
# Treatment w corresponds to Python T; avoid naming it `T` in R (conflicts with TRUE).
w <- rbinom(n, 1L, prob = stats::plogis(X[, 1L]))
y <- (1 + 0.5 * X[, 1L]) * w + X[, 1L] + stats::rnorm(n)
```

Dimensions: `r nrow(X)` observations, `r ncol(X)` covariates; treatment rate `r mean(w)`.

For this DGP the **conditional treatment effect** is linear in the first covariate only:

$$
\tau(x) = \mathbb{E}[Y(1) - Y(0) \mid X = x] = 1 + 0.5 x_1.
$$

We use that closed form for **validation** (holdout RMSE / correlation). The **population** average effect is $\mathbb{E}[\tau(X)] = 1$ when $X_1 \sim \mathcal{N}(0,1)$.

### Fit `DomainAdaptationLearner`

The [EconML](https://econml.azurewebsites.net/) `DomainAdaptationLearner` takes three estimator choices. **RCausalML** uses the same stages but names the *constructor* arguments after the **base learner string** you pass in (not arbitrary sklearn objects):

| EconML argument | Role | RCausalML argument |
|-----------------|------|-------------------|
| `models` | Weighted outcome models $\hat\mu_0$, $\hat\mu_1$ on each arm (must support sample weights) | `learner` (`"lm"`, `"ranger"`, `"glmnet"`, `"xgb"`) |
| `final_models` | One regressor per non-control treatment: fits pseudo–treatment effects $\hat D$ on $X$ to get $\hat\tau(x)$ | `final_learner` (same options; default copies `learner`) |
| `propensity_model` | Classifier for $P(T=\text{arm}\mid X)$ vs control on the pooled $\{\text{control},\text{arm}\}$ sample | `propensity_learner` (`"glm"` or `"glmnet"`) |

After `fit()`, the fitted objects are stored as **`dal$final_models`** (list, one element per treatment level vs `control_name`) and **`dal$propensity_models`** (list, parallel: arm-vs-control propensity fit used for weights in that arm’s loop).

#### Fit DA-Learner and predict CATE

```{r}
#| label: fit-da-learner
#| warning: false

# EconML: models=..., final_models=..., propensity_model=...
dal <- DomainAdaptationLearner(
  learner = "lm",             # weighted outcome models (EconML `models`)
  final_learner = "lm",       # pseudo-effects on X -> CATE (EconML `final_models`)
  propensity_learner = "glm", # P(arm|X) vs control (EconML `propensity_model`)
  control_name = 0
)
dal <- fit(dal, X, w, y)
tau_hat <- predict(dal, X)
head(tau_hat, 10L)
```

#### Summarize `final_models` and `propensity_models`

```{r}
#| label: fitted-components-da
#| warning: false

cat("Treatment arms (non-control):", paste(dal$t_groups, collapse = ", "), "\n\n")

cat("final_models — estimators for pseudo-treatment effects (one list slot per arm):\n")
for (i in seq_along(dal$final_models)) {
  fm <- dal$final_models[[i]]
  cat("  arm ", dal$t_groups[i], ": type = ", fm$type, "\n", sep = "")
  if (identical(fm$type, "lm")) print(stats::coef(fm$model))
}

cat("\npropensity_models — fitted propensity for each arm vs control:\n")
for (i in seq_along(dal$propensity_models)) {
  pm <- dal$propensity_models[[i]]
  cat("  arm ", dal$t_groups[i], " vs control: type = ", pm$type, "\n", sep = "")
  if (identical(pm$type, "glm")) {
    print(stats::coef(pm$model))
  } else if (identical(pm$type, "glmnet")) {
    print(as.matrix(glmnet::coef(pm$model, s = "lambda.1se")))
  }
}
```

### Average treatment effect (`estimate_ate`)

#### Point and interval ATE estimates

```{r}
#| label: estimate-ate-da
#| warning: false

ate_dal <- estimate_ate(dal, X, w, y, pretrain = TRUE, return_ci = TRUE, bootstrap_ci = FALSE)
print(ate_dal)
cat(
  "\nPopulation ATE (theory): 1.0\n",
  "Sample mean true tau(X): ", mean(1 + 0.5 * X[, 1L]), "\n",
  sep = ""
)
```

### Holdout validation vs known $\tau(x)$

Split the data, **fit only on the training half**, predict on the test set, and compare to $1 + 0.5 X_1$. We also fit a plain **T-Learner** (`TLearner` in the same file) for contrast when selection bias is strong.

#### Holdout metrics: DA-Learner vs T-Learner

```{r}
#| label: validate-da-holdout
#| warning: false

set.seed(456)
idx <- sample.int(n, n)
n_te <- floor(0.3 * n)
te_idx <- idx[seq_len(n_te)]
tr_idx <- idx[-seq_len(n_te)]

true_tau <- function(Xm) 1 + 0.5 * Xm[, 1L]

X_tr <- X[tr_idx, , drop = FALSE]
w_tr <- w[tr_idx]
y_tr <- y[tr_idx]
X_te <- X[te_idx, , drop = FALSE]
tau_te <- true_tau(X_te)

dal_v <- DomainAdaptationLearner(
  learner = "lm",             # weighted outcome models (EconML `models`)
  final_learner = "lm",       # pseudo-effect on X (EconML `final_models`)
  propensity_learner = "glm", # propensity ê(x) (EconML `propensity_model`)
  control_name = 0
)
dal_v <- fit(dal_v, X_tr, w_tr, y_tr)
pred_da <- as.numeric(predict(dal_v, X_te))

tl_v <- TLearner(learner = "lm", control_name = 0)
tl_v <- fit(tl_v, X_tr, w_tr, y_tr)
pred_tl <- as.numeric(predict(tl_v, X_te))

rmse <- function(a, b) sqrt(mean((a - b)^2))
mae <- function(a, b) mean(abs(a - b))

val_tbl <- tibble::tibble(
  learner = c("DomainAdaptationLearner", "TLearner"),
  RMSE = c(rmse(pred_da, tau_te), rmse(pred_tl, tau_te)),
  MAE = c(mae(pred_da, tau_te), mae(pred_tl, tau_te)),
  cor_true = c(cor(pred_da, tau_te), cor(pred_tl, tau_te))
)
print(val_tbl)
```

#### Scatter: predicted vs true CATE on holdout

```{r}
#| label: fig-cate-scatter-da
#| warning: false
#| fig-cap: "Holdout predicted CATE vs true $\\tau(x) = 1 + 0.5 X_1$."
#| fig-width: 7
#| fig-height: 4

viz <- tibble::tibble(
  true_tau = tau_te,
  DA_Learner = pred_da,
  T_Learner = pred_tl
) |>
  tidyr::pivot_longer(-true_tau, names_to = "method", values_to = "predicted")

ggplot2::ggplot(viz, ggplot2::aes(x = true_tau, y = predicted, color = method)) +
  ggplot2::geom_point(alpha = 0.35, size = 1.2) +
  ggplot2::geom_abline(intercept = 0, slope = 1, linetype = 2, color = "gray40") +
  ggplot2::coord_fixed() +
  ggplot2::theme_bw() +
  ggplot2::labs(
    x = expression("True " * tau * (x)),
    y = expression("Predicted " * hat(tau) * (x)),
    color = NULL
  )
```

### SHAP for predicted CATE (synthetic DGP, training subsample)

As in the binary-treatment meta-learner notebook (**Part C** there), we attribute **predicted** $\hat{\tau}(X)$ from each fitted meta-learner to covariates using **permutation SHAP** (`explain_cate(..., use_permshap = TRUE)`). Background and explanation rows are drawn **only from the training split** (`X_tr`) defined in the holdout section above. The **X-Learner** is fit **without** a fixed propensity vector, so `predict()` evaluates learned $e(X)$ on any `newdata`, which **kernelshap** requires when features are perturbed. Parallel workers use the same **future** / `parallel` / `parallel_args` pattern as that notebook, with **ranger** included in `parallel_args$packages` for the **DA-, S-, and T-learners**, and **xgboost** / **glmnet** for the **X-Learner** (stage-2 surfaces and glmnet propensity).

#### Permutation SHAP and mean $\|\mathrm{SHAP}\|$ by learner

```{r da-shap-cate-meta}
#| warning: false
#| fig.width: 10
#| fig.height: 9
fast_run <- Sys.getenv("CAUSALML_FAST_RENDER", "TRUE") == "TRUE"
has_ks_da <- requireNamespace("kernelshap", quietly = TRUE)
has_sv_da <- requireNamespace("shapviz", quietly = TRUE)
shp_list_da <- NULL
X_explain_da <- NULL
if (!has_ks_da || !has_sv_da) {
  message("Skipping SHAP: install.packages(c(\"kernelshap\", \"shapviz\")) to run this section.")
} else if (!exists("X_tr") || nrow(X_tr) < 30L) {
  message("Skipping SHAP: run the holdout chunk first so X_tr is defined with enough rows.")
} else {
  suppressPackageStartupMessages({
    library(kernelshap)
    library(shapviz)
  })
  num_trees_shap <- if (fast_run) 80L else 200L
  nrounds_x_shap <- if (fast_run) 40L else 80L
  learner_da_shap <- fit(
    DomainAdaptationLearner(
      learner = "ranger",
      final_learner = "ranger",
      propensity_learner = "glm",
      control_name = 0
    ),
    X_tr, w_tr, y_tr,
    num.trees = num_trees_shap
  )
  learner_s_shap <- fit(
    SLearner(learner = "ranger", control_name = 0),
    X_tr, w_tr, y_tr,
    num.trees = num_trees_shap
  )
  learner_t_shap <- fit(
    TLearner(learner = "ranger", control_name = 0),
    X_tr, w_tr, y_tr,
    num.trees = num_trees_shap
  )
  learner_x_shap <- fit(
    XLearner(learner = "xgb", control_name = 0),
    X_tr, w_tr, y_tr,
    nrounds = nrounds_x_shap
  )

  set.seed(4242)
  n_explain_da <- if (fast_run) 80L else 200L
  bg_size_da <- if (fast_run) 100L else min(300L, nrow(X_tr))
  n_explain_da <- min(n_explain_da, nrow(X_tr))
  bg_size_da <- min(bg_size_da, nrow(X_tr))
  idx_ex_da <- sample.int(nrow(X_tr), n_explain_da)
  bg_idx_da <- sample.int(nrow(X_tr), bg_size_da)
  X_explain_da <- as.data.frame(X_tr[idx_ex_da, , drop = FALSE])
  bg_X_da <- as.data.frame(X_tr[bg_idx_da, , drop = FALSE])

  has_future_da <- requireNamespace("future", quietly = TRUE)
  n_workers_shap_da <- if (has_future_da) {
    min(4L, future::availableCores(omit = 1L))
  } else {
    1L
  }
  use_parallel_shap_da <- isTRUE(n_workers_shap_da > 1L)
  if (use_parallel_shap_da) {
    shap_plan_old_da <- future::plan()
    on.exit(future::plan(shap_plan_old_da), add = TRUE)
    future::plan(future::multisession, workers = n_workers_shap_da)
  }

  meta_da_shap <- list(
    "DA (ranger)" = learner_da_shap,
    "S (ranger)" = learner_s_shap,
    "T (ranger)" = learner_t_shap,
    "X (xgb, learned e(X))" = learner_x_shap
  )

  shap_parallel_args_da <- if (use_parallel_shap_da) {
    list(packages = c("RCausalML", "ranger", "xgboost", "glmnet"))
  } else {
    NULL
  }

  shap_list_da <- vector("list", length(meta_da_shap))
  names(shap_list_da) <- names(meta_da_shap)
  for (i in seq_along(meta_da_shap)) {
    shap_list_da[[i]] <- explain_cate(
      meta_da_shap[[i]],
      X = X_explain_da,
      bg_X = bg_X_da,
      use_permshap = TRUE,
      verbose = FALSE,
      parallel = use_parallel_shap_da,
      parallel_args = shap_parallel_args_da
    )
  }

  shp_list_da <- lapply(shap_list_da, function(ks) shapviz::shapviz(ks, X = X_explain_da))

  imp_rows_da <- lapply(names(shap_list_da), function(lab) {
    ks <- shap_list_da[[lab]]
    S_mat <- as.matrix(ks$S)
    data.frame(
      feature = colnames(ks$X),
      mean_abs_shap = colMeans(abs(S_mat)),
      learner = lab,
      stringsAsFactors = FALSE
    )
  })
  imp_df_da <- dplyr::bind_rows(imp_rows_da)

  p_imp_da <- ggplot2::ggplot(imp_df_da, ggplot2::aes(x = reorder(feature, mean_abs_shap), y = mean_abs_shap)) +
    ggplot2::geom_col(fill = "steelblue", width = 0.75) +
    ggplot2::facet_wrap(~learner, scales = "free_y", ncol = 2) +
    ggplot2::coord_flip() +
    ggplot2::labs(
      title = "Synthetic DGP: mean |SHAP| for predicted CATE by meta-learner (train subsample)",
      x = NULL,
      y = "Mean |SHAP|"
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(strip.text = ggplot2::element_text(face = "bold"))
  print(p_imp_da)
}
```

### Visualize SHAP with beeswarm and waterfall plots (T-Learner with ranger base)

#### Beeswarm for each learner; waterfall for highest predicted CATE (T-Learner)

```{r da-shap-beeswarm-waterfall}
#| warning: false
#| fig.width: 9
#| fig.height: 11
if (!is.null(shp_list_da) && length(shp_list_da) > 0) {
  for (lab in names(shp_list_da)) {
    print(
      shapviz::sv_importance(shp_list_da[[lab]], kind = "beeswarm") +
        ggplot2::ggtitle(lab)
    )
  }
  pred_t_da <- as.numeric(predict(learner_t_shap, as.matrix(X_explain_da)))
  row_hi_da <- which.max(pred_t_da)
  print(
    shapviz::sv_waterfall(shp_list_da[["T (ranger)"]], row_id = row_hi_da) +
      ggplot2::ggtitle(
        "Waterfall SHAP — T-Learner (ranger), highest predicted CATE in explain set (synthetic DGP)"
      )
  )
}
```

## Summary and Conclusion

This notebook introduced the **Domain Adaptation Learner (DA-Learner)** for estimating **conditional average treatment effects (CATE)** when treatment and control groups can differ systematically in covariate distributions (**selection bias** / **domain shift**). We mapped **EconML**’s `models`, `final_models`, and `propensity_model` to **RCausalML**’s `learner`, `final_learner`, and `propensity_learner`, and showed how **`fit()`** stores fitted **`final_models`** and **`propensity_models`** for inspection.

On a **synthetic confounded DGP** with closed-form $\tau(x) = 1 + 0.5 x_1$, we **fit** `DomainAdaptationLearner`, reported **ATE** via **`estimate_ate()`**, and **validated** predictions on a **holdout** split against the true CATE (RMSE, MAE, correlation), alongside a plain **T-Learner** for contrast. **Permutation SHAP** via **`explain_cate(..., use_permshap = TRUE)`** on a **training subsample** illustrated which covariates drive **predicted** $\hat{\tau}(X)$ for **DA-, S-, T-, and X-learners** (with **parallel** **future** workers and **`parallel_args$packages`** where applicable). SHAP values summarize the **fitted surfaces**, not additional causal identification beyond the assumptions already encoded in the estimators.

## Resources

- [EconML: `DomainAdaptationLearner`](https://econml.azurewebsites.net/) — Python reference implementation and API (models, final_models, propensity_model).
- [CausalML: meta-learners with synthetic data](https://causalml.readthedocs.io/en/latest/examples/meta_learners_with_synthetic_data.html) — related meta-learner workflows and benchmarks.
- [kernelshap](https://CRAN.R-project.org/package=kernelshap) — Kernel and permutation SHAP used by **`RCausalML::explain_cate()`**.
- [shapviz](https://CRAN.R-project.org/package=shapviz) — Visualization (`sv_importance`, `sv_waterfall`, beeswarm plots).
- Künzel et al. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *PNAS*.
- Nie & Wager (2018). Quasi-oracle estimation of heterogeneous treatment effects. *Biometrika*.

## Scientific Terminology for Beginners

This quick glossary explains the core scientific terms used in this notebook in plain language.

### 1) Potential Outcomes

- **Meaning**: For each person, imagine two possible worlds: one where they receive treatment and one where they do not.
- **Example**: If a patient gets a drug, one potential outcome is their recovery with the drug, and another potential outcome is their recovery without the drug.

### 2) Treatment Effect

- **Meaning**: The difference between outcome with treatment and outcome without treatment.
- **Example**: If expected blood pressure decreases by 8 points with treatment and 3 points without treatment, the treatment effect is 5 points.

### 3) CATE (Conditional Average Treatment Effect)

- **Meaning**: The average treatment effect for a specific subgroup with features \(X\) (for example, age group or risk profile).
- **Example**: The treatment may work better for younger patients than older patients; each subgroup has its own CATE.

### 4) Confounding

- **Meaning**: A third factor affects both treatment assignment and the outcome, creating misleading associations.
- **Example**: Sicker patients may be more likely to receive treatment and also more likely to have poor outcomes, even if treatment helps.

### 5) Propensity Score

- **Meaning**: The probability of receiving treatment given covariates \(X\): \(P(T=1\mid X)\).
- **Example**: If two patients have similar age, health history, and baseline risk, they should have similar propensity scores.

### 6) Counterfactual

- **Meaning**: The unobserved outcome for the treatment status the person did not receive.
- **Example**: For a treated person, we observe treatment outcome but not what would have happened under control; that missing value is counterfactual.

### 7) Overlap (Positivity)

- **Meaning**: For each covariate profile, there is a non-zero chance of both treatment and control.
- **Example**: If a subgroup always receives treatment and never control, causal comparison for that subgroup is unreliable.

### 8) Nuisance Models

- **Meaning**: Helper models (like outcome models or propensity models) used to estimate the final causal quantity.
- **Example**: In DR- or R-learning, we first model \(E[Y\mid X,T]\) and \(P(T\mid X)\), then combine them to estimate CATE.

### 9) Cross-Fitting

- **Meaning**: A sample-splitting strategy where one part of data fits nuisance models and another part estimates treatment effects.
- **Example**: Train propensity/outcome models on fold A and estimate pseudo-outcomes on fold B, then rotate folds.

### 10) Doubly Robust Estimation

- **Meaning**: An estimator is consistent if either the outcome model or the propensity model is correctly specified.
- **Example**: Even if the propensity model is imperfect, a correctly specified outcome model can still yield valid effect estimates (and vice versa).

