# 4. Meta Learner with Continuous Treatments {#sec-meta-learners-continuous .unnumbered}

Continuous treatments (e.g., dosage levels) often call for a **dose-response function** or a well-defined contrast in $D$, not only a binary CATE. This tutorial covers **RCausalML**’s **`r_learner_continuous()`** for **Conditional Average Dose Response (CADR)** on a dose grid $\tau(d,x)$, and—new in the package—**S-, T-, X-, R-, and DR-learners** with **`treatment_type = "continuous"`** for flexible single-vector summaries when $D$ is continuous.

## Overview {#sec-continuous-overview}

The notebook has two parts:

- **Part A — Synthetic data:** We study **continuous** $D$ with a known DGP: CADR $\tau(d,x) = \frac{\partial}{\partial d}\mathbb{E}[Y(d)|X=x]$, generalized R-learner (`r_learner_continuous()`), SHAP on a surrogate, and full diagnostics (overlap, balance, calibration).
- **Part B — NMES use case:** We use **`nmes_data`** from **`{causaldrf}`** (1987 NMES): **outcome** `TOTALEXP`, **continuous treatment** `packyears`, and the same covariate set as below. We fit **`r_learner_continuous()`** for the CADR-style curve, then fit **S-, T-, X-, R-, and DR-learners** in two ways: (1) **native continuous** treatment via `treatment_type = "continuous"` (S fits $\mu(X,D)$ and contrasts default quartiles of $D$; T/X split at the sample median of $D$; R uses Robinson-style nuisances; DR uses a continuous DR pseudo-outcome with cross-fitting); (2) **binary** high vs. low pack-years (median split) for a direct discrete contrast and for **permutation SHAP** (`explain_cate()`), which we keep on the binary fits for render time. We also compare **double machine learning** (**`LinearDML`**, **`SparseLinearDML`**, **`CausalForestDML`**) on the same NMES subsample, report **ATE**s and simple **holdout** nuisance metrics, and add **SHAP** for the DML predicted CATE surfaces.

We implement a **generalized R-learner** using cross-fitted outcome and treatment models, residualization, and local linear regression over a dose grid, plus **SHAP** (forest surrogate + permutation SHAP) in Part A.

## Implementation in R {#sec-continuous-implementation}

In R, we use **RCausalML** for `r_learner_continuous()`, **`{causaldrf}`** for **`nmes_data`**, `mlr3` for several diagnostics, **`kernelshap`** and **`shapviz`** for SHAP in Part A (optional), and `ggplot2` for visualization.

### Check and Install Required R Packages {#sec-continuous-packages-list}

Following R packages are required to run this notebook. If any of these packages are not installed, you can install them using the code below:

`grf`, `DoubleML`, `ggplot2`, `dplyr`, `mlr3`, `mlr3learners`, `nnet`, `kernelshap`, `shapviz`, `causaldrf`, `knitr`

```{r}
#| label: lst-packages-vector
#| lst-cap: "Required R package names used throughout the notebook."
packages <- c('tidyverse', 
              'plyr',
              #'data.table',
              'RCausalML',
              'DoubleML',
              'causaldrf',
              'grf',
              'mlr3',
              'mlr3learners',
              'nnet',
              'gridExtra',
              'kernelshap',
              'shapviz',
              'knitr'
         )
```

### Install Missing Packages {#sec-continuous-install-packages}

```{r}
#| label: lst-install-missing-packages
#| lst-cap: "Optional commands to install missing CRAN/GitHub dependencies (commented by default)."
#| warning: false
#| error: false
# Install missing packages
# new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
# if(length(new_packages)) install.packages(new_packages)
# remotes::install_github("mlr-org/mlr3learners")
```

### Verify Installation {#sec-continuous-verify-packages}

```{r}
#| label: lst-verify-package-installation
#| lst-cap: "Check that each required package namespace is available."
# Verify installation
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries {#sec-continuous-load-libraries}

```{r}
#| label: load-required-libraries
#| warning: false

if (requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(pkg_root, quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Check Loaded Packages {#sec-continuous-check-search-path}

```{r}
#| label: lst-check-loaded-packages
#| lst-cap: "Confirm which package environments are attached on the search path."
# Check loaded packages
cat("Successfully loaded packages:\n")
print(search()[grepl("package:", search())])
```

## Part A: Meta-learners with synthetic data {#sec-continuous-part-a}

We simulate a **continuous** treatment and a smooth dose–response surface so we can compare estimates to a known benchmark.

### Data Generation for Continuous Treatment {#sec-continuous-synthetic-dgp}

```{r}
#| label: lst-generate-continuous-treatment-data
#| lst-cap: "Simulate continuous treatment, covariates, and outcome from a known dose-response DGP."
# Continuous Treatment Generator (Dose-Response)
generate_continuous_treatment_data <- function(n = 5000) {
  X <- matrix(rnorm(n * 10), nrow = n)
  colnames(X) <- paste0("X", 1:10)
  D <- pmax(0, rnorm(n, mean = 2 + 1.5*X[,1] - X[,2], sd = 1.2))
  mu_0 <- 2*X[,1] - 1.5*X[,2] + X[,4]^2
  Y <- mu_0 + 0.8*D + 0.5*X[,1]*D - 0.3*X[,2]*D + 0.05*D^2 + rnorm(n, sd = 1.5)
  data.frame(Y, D, as.data.frame(X))
}
df_cont <- generate_continuous_treatment_data(5000)
```

### Estimand: Conditional Average Dose Response (CADR) {#sec-continuous-cadr-estimand}

For continuous treatment $D$ (e.g., dosage), we estimate:

$$\tau(d, x) = \frac{\partial}{\partial d} \mathbb{E}[Y(d) \mid X = x]$$

This is the **marginal effect** of an infinitesimal dose increase at level $d$ for units with covariates $x$.

### Generalized R-Learner for Continuous Treatments {#sec-continuous-r-learner}

Extends Robinson transformation to continuous treatments using **kernel smoothing**. The implementation lives in **RCausalML** as `r_learner_continuous()` (cross-fitted `ranger` nuisances and local weighted `lm`, matching this notebook’s previous `mlr3` recipe):

```{r}
#| label: lst-r-learner-continuous-fit
#| lst-cap: "Fit generalized R-learner on synthetic data and compute population-average curve vs. truth."
set.seed(123)
r_cont <- r_learner_continuous(
  df_cont,
  covariates = paste0("X", 1:10),
  dose_grid = seq(0, 6, by = 0.5),
  learner = "ranger",
  n_fold = 5L,
  seed = 123L,
  num.trees = 800L
)
avg_tau <- rowMeans(r_cont$tau_grid)
dose_grid <- r_cont$dose_grid
tau_true_avg <- 0.8 + 0.1 * dose_grid
```

### Average Dose-Response Curve {#sec-continuous-avg-dose-response}

```{r}
#| label: fig-average-dose-response
#| fig-cap: "Average estimated dose-response (R-learner) versus the true population curve over the dose grid."
#| fig.width: 7
#| fig.height: 5
ggplot(data.frame(dose = dose_grid, 
                  tau_hat = avg_tau,
                  tau_true = tau_true_avg),
       aes(x = dose)) +
  geom_line(aes(y = tau_hat, color = "Estimated"), size = 1.2) +
  geom_line(aes(y = tau_true, color = "True"), linetype = "dashed", size = 1.2) +
  labs(title = "Average Dose-Response Function",
       subtitle = "R-Learner estimate vs. ground truth (population average)",
       x = "Dosage Level (D)", 
       y = "Marginal Effect ∂E[Y(d)|X]/∂d",
       color = "Curve") +
  scale_color_manual(values = c("Estimated" = "#56B4E9", "True" = "darkred")) +
  theme_minimal(base_size = 13)
```

### Heterogeneous Dose-Response: Effect Modification {#sec-continuous-heterogeneous-dose}

Visualize how dose-response varies by key covariate (X1):

```{r}
#| label: fig-heterogeneous-dose-response-strata
#| fig-cap: "Dose-response by low vs. high X1 (estimated vs. true), illustrating effect modification."
#| fig.width: 7
#| fig.height: 5
x1_low_idx <- which(df_cont$X1 < quantile(df_cont$X1, 0.25))
x1_high_idx <- which(df_cont$X1 > quantile(df_cont$X1, 0.75))
tau_low <- rowMeans(r_cont$tau_grid[, x1_low_idx])
tau_high <- rowMeans(r_cont$tau_grid[, x1_high_idx])
tau_true_low <- 0.465 + 0.1 * dose_grid
tau_true_high <- 1.135 + 0.1 * dose_grid

df_strata <- data.frame(
  dose = rep(dose_grid, 4),
  tau = c(tau_low, tau_high, tau_true_low, tau_true_high),
  stratum = rep(c("X1 Low (Est)", "X1 High (Est)", 
                  "X1 Low (True)", "X1 High (True)"), 
                each = length(dose_grid)),
  type = rep(c("Estimated", "Estimated", "True", "True"), each = length(dose_grid))
)

ggplot(df_strata, aes(x = dose, y = tau, color = stratum, linetype = type)) +
  geom_line(size = 1.1) +
  labs(title = "Heterogeneous Dose-Response by Covariate Strata",
       subtitle = "Effect modification: High X1 units benefit more from increased dosage",
       x = "Dosage Level (D)", 
       y = "Marginal Effect ∂E[Y(d)|X]/∂d",
       color = "Stratum", linetype = "Estimate Type") +
  scale_color_manual(values = c(
    "X1 Low (Est)" = "#E69F00", "X1 High (Est)" = "#009E73",
    "X1 Low (True)" = "#E69F00", "X1 High (True)" = "#009E73"
  )) +
  theme_minimal(base_size = 13)
```

### Native S/T/X/R/DR with `treatment_type = "continuous"` (synthetic) {#sec-continuous-native-meta-api}

**RCausalML** now accepts **continuous** `treatment` for `SLearner`, `TLearner`, `XLearner`, `RLearner`, and `DRLearner` via `treatment_type = "continuous"`. This is **not** the same object as **`r_learner_continuous()`** (which returns a **dose grid** $\times$ units surface for CADR): here you get a **single** $\hat{\tau}(X)$ vector per specification (e.g., S-learner contrast $\mu(X,d_1)-\mu(X,d_0)$ at training quartiles of $D$; T/X high-vs-low dose at the median; R-learner Robinson partial effect; DR continuous pseudo-outcome). The chunk below is a minimal sanity check on the same simulated data as above.

```{r}
#| label: lst-native-continuous-meta-synthetic
#| lst-cap: "Fit S/T/X/R/DR with treatment_type = continuous on synthetic D (small ranger forests)."
#| warning: false
X_syn <- as.matrix(df_cont[, paste0("X", 1:10)])
D_syn <- df_cont$D
Y_syn <- df_cont$Y
set.seed(11L)
nt_syn <- 200L
sl_syn_c <- fit(SLearner(learner = "ranger", treatment_type = "continuous"),
                X_syn, D_syn, Y_syn, num.trees = nt_syn)
tl_syn_c <- fit(TLearner(learner = "ranger", treatment_type = "continuous"),
                X_syn, D_syn, Y_syn, num.trees = nt_syn)
xl_syn_c <- fit(XLearner(learner = "ranger", treatment_type = "continuous"),
                X_syn, D_syn, Y_syn, num.trees = nt_syn)
rl_syn_c <- fit(RLearner(learner = "ranger", treatment_type = "continuous", n_fold = 5L),
                X_syn, D_syn, Y_syn, num.trees = nt_syn)
dr_syn_c <- fit(DRLearner(learner = "ranger", treatment_type = "continuous", n_fold = 3L),
                X_syn, D_syn, Y_syn, num.trees = nt_syn)
te_mean_native <- c(
  S  = mean(predict(sl_syn_c, X_syn)),
  T  = mean(predict(tl_syn_c, X_syn)),
  X  = mean(predict(xl_syn_c, X_syn)),
  R  = mean(predict(rl_syn_c, X_syn)),
  DR = mean(predict(dr_syn_c, X_syn))
)
print(round(te_mean_native, 3))
```

### Diagnostics & Validation Protocol {#sec-continuous-diagnostics}

Diagnostics help ensure that dose-response estimates are not driven by lack of overlap, covariate imbalance, or miscalibration. Below we implement overlap checks, balance assessment before/after weighting, and a CATE calibration plot.

#### Overlap Diagnostics {#sec-continuous-overlap}

For continuous treatments, **overlap** means that the conditional support of $D$ given $X$ has sufficient density across the dose range we care about. We check the marginal distribution of $D$ and the distribution of the conditional expectation $E[D|X]$ (estimated via a regression of $D$ on $X$) to assess common support.

```{r}
#| label: fig-overlap-diagnostics-synthetic
#| fig-cap: "Marginal dose density and distribution of estimated E[D|X] by X1 quartile (synthetic data)."
#| fig.width: 10
#| fig.height: 5
# Estimate E[D|X] for overlap diagnostics
task_d <- TaskRegr$new(id = "d_overlap",
                       backend = df_cont[, c("D", paste0("X", 1:10))],
                       target = "D")
learner_d <- lrn("regr.ranger", num.trees = 500)
learner_d$train(task_d)
e_hat_overlap <- learner_d$predict_newdata(df_cont[, paste0("X", 1:10)])$response

df_overlap <- data.frame(
  D = df_cont$D,
  e_hat = e_hat_overlap,
  X1_quartile = cut(df_cont$X1, breaks = quantile(df_cont$X1, 0:4/4),
                    labels = c("X1 Q1", "X1 Q2", "X1 Q3", "X1 Q4"), include.lowest = TRUE)
)

p1 <- ggplot(df_overlap, aes(x = D)) +
  geom_density(fill = "#56B4E9", alpha = 0.6) +
  labs(title = "Overlap: Marginal Distribution of Dose (D)",
       x = "Dosage (D)", y = "Density") +
  theme_minimal(base_size = 12)

p2 <- ggplot(df_overlap, aes(x = e_hat, fill = X1_quartile)) +
  geom_density(alpha = 0.5, position = "identity") +
  labs(title = "Overlap: Distribution of E[D|X] by X1 Quartile",
       x = "E[D|X] (estimated)", y = "Density", fill = "X1 stratum") +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

Interpretation: substantial overlap across covariate strata (e.g., E[D|X] distributions by X1 quartile) supports estimation of dose-response over the observed dose range; narrow or separated supports suggest limited overlap and cautious interpretation.

#### Balance Assessment (Before/After Weighting) {#sec-continuous-balance}

We stratify units by dose quartiles and assess **standardized mean differences (SMD)** of covariates across strata. We then compute **stabilized inverse probability of treatment weights** (IPTW) for continuous $D$ using a normal approximation: $w_i = f(D_i) / f(D_i|X_i)$, with $f(D|X)$ from $D|X \sim N(\hat{e}(X), \hat{\sigma}^2)$, and reassess balance after weighting.

```{r}
#| label: fig-balance-smd-synthetic
#| fig-cap: "Standardized mean differences (Q4 vs. Q1 dose) before and after stabilized IPTW weighting."
#| fig.width: 7
#| fig.height: 5
# Dose strata (quartiles)
df_cont$dose_stratum <- cut(df_cont$D,
  breaks = quantile(df_cont$D, 0:4/4),
  labels = c("Q1", "Q2", "Q3", "Q4"), include.lowest = TRUE)

# SMD before weighting: high (Q4) vs low (Q1) dose
covs <- paste0("X", 1:10)
low <- df_cont[df_cont$dose_stratum == "Q1", covs]
high <- df_cont[df_cont$dose_stratum == "Q4", covs]
smd_before <- (colMeans(high) - colMeans(low)) / sqrt((apply(high, 2, sd)^2 + apply(low, 2, sd)^2) / 2)

# Weights: stabilized IPTW for continuous D (normal approximation)
sigma_d <- sd(df_cont$D - e_hat_overlap)
f_D <- density(df_cont$D, from = min(df_cont$D), to = max(df_cont$D), n = 512)
f_D_at_D <- approx(f_D$x, f_D$y, xout = df_cont$D, rule = 2)$y
f_D_given_X <- dnorm(df_cont$D, mean = e_hat_overlap, sd = pmax(sigma_d, 0.1))
sw <- f_D_at_D / pmax(f_D_given_X, 1e-6)
sw <- pmin(sw, quantile(sw, 0.99))

# Weighted SMD (Q4 vs Q1)
df_cont$sw <- sw
w_low <- df_cont[df_cont$dose_stratum == "Q1", ]
w_high <- df_cont[df_cont$dose_stratum == "Q4", ]
for (v in covs) {
  m_low <- weighted.mean(w_low[[v]], w_low$sw)
  m_high <- weighted.mean(w_high[[v]], w_high$sw)
  s_low <- sqrt(weighted.mean((w_low[[v]] - m_low)^2, w_low$sw))
  s_high <- sqrt(weighted.mean((w_high[[v]] - m_high)^2, w_high$sw))
  smd_after_v <- (m_high - m_low) / sqrt((s_low^2 + s_high^2) / 2)
  if (v == covs[1]) smd_after <- setNames(numeric(length(covs)), covs)
  smd_after[v] <- smd_after_v
}

balance_df <- data.frame(
  covariate = rep(covs, 2),
  SMD = c(smd_before, smd_after),
  when = rep(c("Before weighting", "After weighting"), each = length(covs))
)

ggplot(balance_df, aes(x = covariate, y = SMD, fill = when)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = c(-0.1, 0.1), linetype = "dashed", color = "gray40") +
  labs(title = "Balance Assessment: Standardized Mean Difference (Q4 vs Q1 dose)",
       subtitle = "Covariates should be balanced (|SMD| < 0.1) after weighting",
       x = "Covariate", y = "SMD", fill = "") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### CATE Calibration Plot {#sec-continuous-cate-calibration}

We bin units by their **predicted marginal effect** at a reference dose (median dose) and plot mean predicted effect vs. mean observed outcome (residualized by baseline) in each bin. A well-calibrated estimator should show a positive alignment between predicted effect and realized outcomes.

```{r}
#| label: fig-cate-calibration-synthetic
#| fig-cap: "Calibration of predicted marginal effect at median dose vs. mean residualized outcome by quintile."
#| fig.width: 7
#| fig.height: 5
# Reference dose: median
ref_dose_idx <- which.min(abs(r_cont$dose_grid - median(df_cont$D)))
tau_at_ref <- r_cont$tau_grid[ref_dose_idx, ]

# Residualized outcome (Y - mu_hat) as proxy for treatment-driven variation
task_y <- TaskRegr$new(id = "y_cal", backend = df_cont[, c("Y", "D", paste0("X", 1:10))], target = "Y")
learner_y <- lrn("regr.ranger", num.trees = 500)
learner_y$train(task_y)
mu_hat_cal <- learner_y$predict_newdata(df_cont[, c("D", paste0("X", 1:10))])$response
Y_resid_cal <- df_cont$Y - mu_hat_cal

# Quintiles of predicted tau at reference dose (rank-based bins to handle ties)
df_cont$tau_hat_ref <- tau_at_ref
df_cont$Y_resid <- Y_resid_cal
n_bins <- 5L
r <- rank(tau_at_ref, ties.method = "first")
df_cont$tau_bin <- cut(r, breaks = seq(1, nrow(df_cont) + 1, length.out = n_bins + 1), labels = 1:n_bins, include.lowest = TRUE, right = FALSE)

cal_df <- df_cont %>%
  group_by(tau_bin) %>%
  summarise(mean_tau_hat = mean(tau_hat_ref), mean_Y_resid = mean(Y_resid), .groups = "drop")

ggplot(cal_df, aes(x = mean_tau_hat, y = mean_Y_resid)) +
  geom_point(size = 4, color = "#009E73") +
  geom_smooth(method = "lm", se = TRUE, color = "#56B4E9", linetype = "dashed") +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "gray40") +
  labs(title = "CATE Calibration Plot (at median dose)",
       subtitle = "Mean predicted marginal effect vs mean residualized outcome by quintile",
       x = "Mean predicted τ(d,x) in bin", y = "Mean residualized outcome in bin") +
  theme_minimal(base_size = 13)
```

### Decision Framework: Which Meta-Learner to Use? {#sec-continuous-decision-framework}

For **continuous treatments**, the choice of approach depends on data structure and goals:

| Criterion | Recommendation |
|-----------|----------------|
| **Target** | Full curve $\tau(d,x)$ on a grid → `r_learner_continuous()`. Single-vector heterogeneous “effects” with continuous $D$ in **RCausalML** → `SLearner` / `TLearner` / `XLearner` / `RLearner` / `DRLearner` with `treatment_type = "continuous"` (see help text for each estimand). |
| **Overlap** | Ensure sufficient density of $D$ across $X$; trim or restrict to common support if needed. |
| **Curvature** | If effect is linear in $d$ → local linear R-Learner or parametric DR model; if nonlinear → kernel smoothing, splines, or generalized random forests. |
| **Bandwidth** | Use cross-validation to choose bandwidth for kernel weights; Silverman-type rules are a starting point. |
| **Alternatives** | **Generalized random forests** (`grf`) for continuous treatment; **double ML** with series expansion in $d$; **causal BART** with continuous treatment. |

Use the **R-Learner** when you want a debiased, flexible estimator that leverages ML for nuisances and residualization; use **GRF** or **BART** when you prefer a single forest that directly targets the dose-response.


## Part B: Use case — NMES (`nmes_data` from `{causaldrf}`) {#sec-continuous-part-b-nmes}

**`nmes_data`** comes from the **1987 National Medical Expenditure Survey (NMES)** and is bundled in **`{causaldrf}`** (`data(nmes_data, package = "causaldrf")`). It is widely used for **continuous treatment** / dose-response methods (e.g., Imai & van Dyk, 2004). The extract has **9,708** observations and **12** variables (complete cases in the packaged version).

| Role | Variable | Description |
|------|-----------|-------------|
| **Outcome** | `TOTALEXP` | Total annual medical expenditures (USD) |
| **Treatment** | `packyears` | Pack-years: (cigarettes per day / 20) × years smoked |
| **Covariates** | `AGESMOKE` | Age when started smoking |
| | `LASTAGE` | Age at last interview |
| | `MALE` | Sex (indicator) |
| | `RACE3` | Race (coded) |
| | `educate` | Education |
| | `marital` | Marital status |
| | `SREGION` | Census region |
| | `POVSTALB` | Poverty status |
| | `beltuse` | Seat belt use |
| | `HSQACCWT` | Sampling weight (include as covariate here for simplicity; full survey inference would use design-based weights) |

Inference is only as credible as **unconfoundedness** and **overlap** over the support of `packyears`.

### Explore treatment and outcome {#sec-nmes-explore}

```{r}
#| label: fig-nmes-explore-scatter
#| fig-cap: "NMES: total medical expenditures vs. pack-years with a LOESS smooth and data structure summary."
#| fig.width: 8
#| fig.height: 5
data(nmes_data, package = "causaldrf")
str(nmes_data, give.attr = FALSE)

ggplot(nmes_data, aes(x = packyears, y = TOTALEXP)) +
  geom_point(alpha = 0.06, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9, color = "#D55E00") +
  labs(
    title = "NMES: total medical expenditures vs pack-years",
    x = "Pack-years (continuous treatment)",
    y = "Total annual medical expenditures ($)"
  ) +
  theme_minimal(base_size = 12)
```

### Prepare analysis data and fit generalized R-learner {#sec-nmes-rlearner-prep}

We map **`TOTALEXP` $\to$ `Y`**, **`packyears` $\to$ `D`**, and use the covariates above. The **dose grid** spans the 5th–95th percentiles of `packyears` to limit extrapolation.

```{r}
#| label: lst-nmes-rlearner-continuous-fit
#| lst-cap: "Prepare NMES analysis data and fit r_learner_continuous on pack-years with a restricted dose grid."
#| warning: false
covs_nmes <- c(
  "AGESMOKE", "LASTAGE", "MALE", "RACE3", "beltuse", "educate",
  "marital", "SREGION", "POVSTALB", "HSQACCWT"
)
df_nmes <- data.frame(
  Y = nmes_data$TOTALEXP,
  D = nmes_data$packyears,
  nmes_data[, covs_nmes, drop = FALSE],
  check.names = FALSE
)
df_nmes <- df_nmes[stats::complete.cases(df_nmes), , drop = FALSE]
# Numeric coding for all adjustment variables (avoids GLM propensity errors in DR-learner CV when
# factor levels differ across folds; ranger also treats these as continuous splits)
for (vn in covs_nmes) {
  df_nmes[[vn]] <- suppressWarnings(as.numeric(as.character(df_nmes[[vn]])))
}

dose_lo <- stats::quantile(df_nmes$D, 0.05)
dose_hi <- stats::quantile(df_nmes$D, 0.95)
dose_grid_nmes <- seq(dose_lo, dose_hi, length.out = 14L)

set.seed(789L)
r_nmes <- r_learner_continuous(
  df_nmes,
  outcome = "Y",
  treatment = "D",
  covariates = covs_nmes,
  dose_grid = dose_grid_nmes,
  learner = "ranger",
  n_fold = 5L,
  seed = 789L,
  num.trees = 400L
)
avg_tau_nmes <- rowMeans(r_nmes$tau_grid)
dose_grid_fit <- r_nmes$dose_grid
```

### Average estimated local slope vs pack-years (NMES) {#sec-nmes-avg-tau-plot}

```{r}
#| label: fig-nmes-avg-tau
#| fig-cap: "Population-average estimated local slope (row-mean τ̂(d)) over pack-years with a rug of observed D."
#| fig.width: 7
#| fig.height: 5
ggplot(
  data.frame(dose = dose_grid_fit, tau_hat = avg_tau_nmes),
  aes(x = dose, y = tau_hat)
) +
  geom_line(linewidth = 1.1, color = "#0072B2") +
  geom_rug(data = data.frame(D = df_nmes$D), aes(x = D), inherit.aes = FALSE, alpha = 0.12) +
  labs(
    title = "NMES: population-average estimated local slope (generalized R-learner)",
    subtitle = "Row-mean τ̂(d) over units; dose grid = 5th–95th percentile of pack-years",
    x = "Pack-years (d)",
    y = "Row-mean estimated local slope (averaged over units)"
  ) +
  theme_minimal(base_size = 13)
```

### Quick overlap check (continuous treatment) {#sec-nmes-overlap}

We plot the **marginal distribution of** `packyears` and the distribution of **$\hat{E}[D \mid X]$** (random forest) by sex, as a simple overlap diagnostic.

```{r}
#| label: fig-nmes-overlap-diagnostics
#| fig-cap: "NMES overlap: marginal density of pack-years and estimated E[D|X] by sex."
#| fig.width: 8
#| fig.height: 4
task_dn <- TaskRegr$new(
  id = "d_nmes",
  backend = df_nmes[, c("D", covs_nmes)],
  target = "D"
)
lrn_dn <- lrn("regr.ranger", num.trees = 400L)
lrn_dn$train(task_dn)
e_hat_nmes <- lrn_dn$predict_newdata(df_nmes[, covs_nmes])$response

p_d <- ggplot(df_nmes, aes(x = D)) +
  geom_density(fill = "#56B4E9", alpha = 0.65) +
  labs(title = "Overlap: marginal density of pack-years",
       x = "Pack-years (D)", y = "Density") +
  theme_minimal(base_size = 12)

p_e <- ggplot(
  data.frame(e_hat = e_hat_nmes, male = factor(df_nmes$MALE, labels = c("Female", "Male"))),
  aes(x = e_hat, fill = male)
) +
  geom_density(alpha = 0.45, position = "identity") +
  labs(title = "Distribution of estimated E[D|X] by sex",
       x = "Estimated E[packyears|X]", y = "Density", fill = NULL) +
  theme_minimal(base_size = 12)

gridExtra::grid.arrange(p_d, p_e, ncol = 2)
```

### Meta-learners on NMES: continuous `packyears` and binary median split {#sec-nmes-binary-meta}

For compute time when rendering, we fit meta-learners on a **random subsample** of up to **3,000** rows (the **`r_learner_continuous()`** fit above still uses **all** complete cases).

**1. Native continuous treatment.** Pass **`packyears`** as `treatment` with `treatment_type = "continuous"`. **S-Learner** contrasts predicted outcome at the **25th vs. 75th percentile** of training $D$ (override with `dose_values = c(d0, d1)`). **T- / X-Learner** use the **sample median** of $D$ as high-vs-low split (override with `dose_split`). **R-Learner** uses **Robinson-style** $\mathbb{E}[Y|X]$ and $\mathbb{E}[D|X]$ nuisances. **DR-Learner** uses a **continuous** cross-fitted pseudo-outcome (see package documentation). Covariates **exclude** raw `packyears` so dose enters only through `treatment`.

**2. Binary median split.** For a classic **ATE** of high vs. low exposure and for **SHAP** below, we also fit **binary** learners on

$$W_i = \mathbf{1}\{\texttt{packyears}_i > \mathrm{median}(\texttt{packyears})\}.$$

We still **do not** put raw `packyears` in $X$ when using $W$ (overlap).

```{r}
#| label: lst-nmes-binary-meta-prep
#| lst-cap: "Subsample NMES for meta-learners; keep continuous D and binary W = 1{D > median(D)}."
X_nmes <- as.matrix(df_nmes[, covs_nmes, drop = FALSE])
colnames(X_nmes) <- covs_nmes
y_nmes <- df_nmes$Y
w_high <- as.integer(df_nmes$D > stats::median(df_nmes$D))
stopifnot(all(w_high %in% c(0L, 1L)))
# Subsample for S/T/X/R/DR + SHAP (full N makes DR + five permshap runs very slow to render)
set.seed(4242L)
n_sub_meta <- min(3000L, nrow(X_nmes))
idx_sub <- sample.int(nrow(X_nmes), n_sub_meta)
X_meta <- X_nmes[idx_sub, , drop = FALSE]
y_meta <- y_nmes[idx_sub]
d_meta <- df_nmes$D[idx_sub]
w_meta <- w_high[idx_sub]
```

```{r}
#| label: lst-nmes-meta-learners-continuous-fit
#| lst-cap: "Fit S/T/X/R/DR with treatment_type = continuous on raw pack-years (NMES subsample)."
#| warning: false
set.seed(12024L)
ctrl <- 0L
nt_m <- 280L
sl_nmes_c <- SLearner(learner = "ranger", control_name = ctrl, treatment_type = "continuous")
tl_nmes_c <- TLearner(learner = "ranger", control_name = ctrl, treatment_type = "continuous")
xl_nmes_c <- XLearner(learner = "ranger", control_name = ctrl, treatment_type = "continuous")
rl_nmes_c <- RLearner(learner = "ranger", control_name = ctrl, n_fold = 5L, treatment_type = "continuous")
dr_nmes_c <- DRLearner(learner = "ranger", control_name = ctrl, n_fold = 3L, treatment_type = "continuous")

sl_nmes_c <- fit(sl_nmes_c, X_meta, d_meta, y_meta, num.trees = nt_m)
tl_nmes_c <- fit(tl_nmes_c, X_meta, d_meta, y_meta, num.trees = nt_m)
xl_nmes_c <- fit(xl_nmes_c, X_meta, d_meta, y_meta, num.trees = nt_m)
rl_nmes_c <- fit(rl_nmes_c, X_meta, d_meta, y_meta, num.trees = nt_m)
dr_nmes_c <- fit(dr_nmes_c, X_meta, d_meta, y_meta, num.trees = nt_m)
```

```{r}
#| label: tbl-nmes-meta-continuous-summary
#| tbl-cap: "NMES subsample: mean predicted score by learner with continuous packyears (native API); estimands differ by learner."
#| echo: true
mean_te <- function(fit_obj, X) mean(as.numeric(predict(fit_obj, X)), na.rm = TRUE)
ate_c_scalar <- function(fit_obj) {
  a <- estimate_ate(fit_obj, X_meta, d_meta, y_meta, return_ci = FALSE, pretrain = TRUE)$ate
  as.numeric(a)[1L]
}
tbl_cont <- data.frame(
  learner = c("S", "T", "X", "R", "DR"),
  mean_predicted_tau = c(
    mean_te(sl_nmes_c, X_meta),
    mean_te(tl_nmes_c, X_meta),
    mean_te(xl_nmes_c, X_meta),
    mean_te(rl_nmes_c, X_meta),
    mean_te(dr_nmes_c, X_meta)
  ),
  ate_plugin = c(
    ate_c_scalar(sl_nmes_c),
    ate_c_scalar(tl_nmes_c),
    ate_c_scalar(xl_nmes_c),
    ate_c_scalar(rl_nmes_c),
    ate_c_scalar(dr_nmes_c)
  )
)
knitr::kable(tbl_cont, digits = 2)
```

The **mean predicted** column averages $\hat{\tau}(X)$ from `predict()`; **`ate_plugin`** uses `estimate_ate()`’s internal variance formulas (continuous rows use high-vs-low-$D$ logic where applicable). Interpret **S** as a **dose contrast** at training quartiles, **T/X** as **above-median vs. below**, **R** as a **Robinson-type** effect, and **DR** as the **continuous DR** CATE surface aggregated to a mean—comparability across columns is limited.

```{r}
#| label: lst-nmes-meta-learners-binary-fit
#| lst-cap: "Fit S-, T-, X-, R-, and DR-learners (ranger) on binary W = high vs. low pack-years."
#| warning: false
set.seed(12025L)
sl_nmes <- SLearner(learner = "ranger", control_name = ctrl)
tl_nmes <- TLearner(learner = "ranger", control_name = ctrl)
xl_nmes <- XLearner(learner = "ranger", control_name = ctrl)
rl_nmes_bin <- RLearner(learner = "ranger", control_name = ctrl, n_fold = 5L)
dr_nmes <- DRLearner(learner = "ranger", control_name = ctrl, n_fold = 3L)

sl_nmes <- fit(sl_nmes, X_meta, w_meta, y_meta, num.trees = nt_m)
tl_nmes <- fit(tl_nmes, X_meta, w_meta, y_meta, num.trees = nt_m)
xl_nmes <- fit(xl_nmes, X_meta, w_meta, y_meta, num.trees = nt_m)
rl_nmes_bin <- fit(rl_nmes_bin, X_meta, w_meta, y_meta, num.trees = nt_m)
dr_nmes <- fit(dr_nmes, X_meta, w_meta, y_meta, num.trees = nt_m)
```

```{r}
#| label: tbl-nmes-meta-ate
#| tbl-cap: "Binary W: sample ATE of high vs. low pack-years on TOTALEXP (USD) by meta-learner (n ≤ 3,000 subsample)."
#| echo: true
ate_scalar <- function(fit_obj) {
  a <- estimate_ate(fit_obj, X_meta, w_meta, y_meta, return_ci = FALSE, pretrain = TRUE)$ate
  as.numeric(a)[1L]
}
ate_tbl <- data.frame(
  learner = c("S", "T", "X", "R", "DR"),
  ate = c(
    ate_scalar(sl_nmes),
    ate_scalar(tl_nmes),
    ate_scalar(xl_nmes),
    ate_scalar(rl_nmes_bin),
    ate_scalar(dr_nmes)
  )
)
knitr::kable(ate_tbl, digits = 2)
```

### Use case: Double machine learning on NMES (DML) {#sec-nmes-dml-use-case}

We reuse the **same subsample** (`X_meta`, `d_meta` = `packyears`, `y_meta` = `TOTALEXP`) and fit **partially linear DML** with **ranger** nuisances: **`LinearDML`** (OLS final stage on residual interactions), **`SparseLinearDML`** (Lasso final stage), and **`CausalForestDML`** (**`grf::causal_forest`** on residuals). All three target **$\hat{\theta}(X)$** in $\tilde{Y} \approx \tilde{D}\cdot\theta(X)$ with continuous $\tilde{D}$ (see `R/DMLearner.R`). There is **no ground-truth CATE** on NMES, so we compare **(i)** **ATE** with a large-sample normal CI, **(ii)** **holdout RMSE** of the cross-fitted **$\hat{\mathbb{E}}[Y\mid X]$** on a 30% test split (models refit on 70% train only), and **(iii)** **Spearman** correlation between predicted $\hat{\theta}(X)$ on the full subsample (agreement of **rankings** across methods).

```{r}
#| label: lst-nmes-dml-fit-compare
#| tbl-cap: "NMES subsample: Double ML variants — ATE, mean |CATE|, and pairwise Spearman correlation of predicted CATE."
#| warning: false
set.seed(2026L)
nt_dml <- 220L
n_f_dml <- 4L
if (!requireNamespace("grf", quietly = TRUE)) {
  dml_lin_nmes <- dml_sparse_nmes <- dml_cf_nmes <- NULL
  message("Skipping DML comparison: install.packages(\"grf\") for CausalForestDML.")
} else {
  dml_lin_nmes <- LinearDML(
    model_y = "ranger", model_t = "ranger",
    n_fold = n_f_dml, seed = 20261L
  )
  dml_sparse_nmes <- SparseLinearDML(
    model_y = "ranger", model_t = "ranger",
    alpha = 0.15, n_fold = n_f_dml, seed = 20262L
  )
  dml_cf_nmes <- CausalForestDML(
    model_y = "ranger", model_t = "ranger",
    n_fold = n_f_dml, seed = 20263L,
    num_trees = 800L, min_node_size = 8L
  )
  dml_lin_nmes <- fit(dml_lin_nmes, X_meta, d_meta, y_meta, num.trees = nt_dml)
  dml_sparse_nmes <- fit(dml_sparse_nmes, X_meta, d_meta, y_meta, num.trees = nt_dml)
  # CausalForestDML: grf::causal_forest already receives num.trees from the constructor; passing num.trees
  # through ... would duplicate the argument — omit here (ranger nuisances use their defaults).
  dml_cf_nmes <- fit(dml_cf_nmes, X_meta, d_meta, y_meta)

  ate_lin <- estimate_ate(dml_lin_nmes, X_meta, d_meta, y_meta, return_ci = TRUE, pretrain = TRUE)
  ate_spr <- estimate_ate(dml_sparse_nmes, X_meta, d_meta, y_meta, return_ci = TRUE, pretrain = TRUE)
  ate_cf <- estimate_ate(dml_cf_nmes, X_meta, d_meta, y_meta, return_ci = TRUE, pretrain = TRUE)

  tau_l <- as.numeric(predict(dml_lin_nmes, X_meta))
  tau_s <- as.numeric(predict(dml_sparse_nmes, X_meta))
  tau_c <- as.numeric(predict(dml_cf_nmes, X_meta))
  cors <- matrix(NA_real_, 3L, 3L, dimnames = list(c("LinearDML", "SparseLinearDML", "CausalForestDML"),
                                                    c("LinearDML", "SparseLinearDML", "CausalForestDML")))
  M <- cbind(tau_l, tau_s, tau_c)
  for (i in 1:3) for (j in 1:3) cors[i, j] <- suppressWarnings(stats::cor(M[, i], M[, j], method = "spearman", use = "complete.obs"))

  tbl_dml_main <- data.frame(
    method = c("LinearDML", "SparseLinearDML", "CausalForestDML"),
    ATE = c(as.numeric(ate_lin$ate)[1L], as.numeric(ate_spr$ate)[1L], as.numeric(ate_cf$ate)[1L]),
    CI_low = c(as.numeric(ate_lin$ate_lb)[1L], as.numeric(ate_spr$ate_lb)[1L], as.numeric(ate_cf$ate_lb)[1L]),
    CI_high = c(as.numeric(ate_lin$ate_ub)[1L], as.numeric(ate_spr$ate_ub)[1L], as.numeric(ate_cf$ate_ub)[1L]),
    mean_abs_CATE = c(mean(abs(tau_l)), mean(abs(tau_s)), mean(abs(tau_c))),
    stringsAsFactors = FALSE
  )

  n_te <- max(30L, floor(0.3 * nrow(X_meta)))
  ix_te <- sample.int(nrow(X_meta), n_te)
  ix_tr <- setdiff(seq_len(nrow(X_meta)), ix_te)
  X_tr <- X_meta[ix_tr, , drop = FALSE]
  X_te <- X_meta[ix_te, , drop = FALSE]
  d_tr <- d_meta[ix_tr]
  d_te <- d_meta[ix_te]
  y_tr <- y_meta[ix_tr]
  y_te <- y_meta[ix_te]

  rmse_yhat <- function(est, pass_nt = TRUE) {
    est_tr <- if (pass_nt) {
      fit(est, X_tr, d_tr, y_tr, num.trees = nt_dml)
    } else {
      fit(est, X_tr, d_tr, y_tr)
    }
    comp <- predict(est_tr, X_te, return_components = TRUE, verbose = FALSE)
    sqrt(mean((y_te - comp$yhat_c)^2, na.rm = TRUE))
  }
  # Refit fresh constructors so train-fold fits do not overwrite full-sample objects used for SHAP below
  tbl_dml_main$RMSE_EY_holdout <- c(
    rmse_yhat(LinearDML(model_y = "ranger", model_t = "ranger", n_fold = n_f_dml, seed = 3031L), TRUE),
    rmse_yhat(SparseLinearDML(model_y = "ranger", model_t = "ranger", alpha = 0.15, n_fold = n_f_dml, seed = 3032L), TRUE),
    rmse_yhat(CausalForestDML(model_y = "ranger", model_t = "ranger", n_fold = n_f_dml, seed = 3033L,
                              num_trees = 800L, min_node_size = 8L), FALSE)
  )

  knitr::kable(tbl_dml_main, digits = 2)
  cat("\nSpearman correlation of predicted CATE (full subsample):\n")
  knitr::kable(cors, digits = 3)
}
```

**Interpretation:** **ATE** rows aggregate $\hat{\theta}(X)$; intervals use the notebook’s default **large-sample** standard error for the mean CATE (not debiased DML inference). **RMSE_EY_holdout** measures how well the **cross-fitted outcome nuisance** predicts **$Y$** out of sample—higher overlap and smoother $\mathbb{E}[Y\mid X]$ often yield lower RMSE. **Spearman** shows whether **heterogeneity rankings** agree across linear, sparse, and forest final stages.

### SHAP for predicted CATE (DML, NMES) {#sec-nmes-shap-dml}

As in [§ NMES SHAP for meta-learners](#sec-nmes-shap-cate), we attribute **predicted** $\hat{\theta}(X)$ (not necessarily causal structure) using **`explain_cate(..., use_permshap = TRUE)`** and **`shapviz`**, on the **full-subsample** DML fits above.

```{r}
#| label: fig-nmes-shap-dml-importance
#| fig-cap: "Mean absolute SHAP values for predicted CATE: LinearDML, SparseLinearDML, and CausalForestDML (NMES subsample)."
#| warning: false
#| fig.width: 11
#| fig.height: 5
shp_nmes_dml_list <- NULL
has_ks_dml <- requireNamespace("kernelshap", quietly = TRUE)
has_sv_dml <- requireNamespace("shapviz", quietly = TRUE)
if (!is.null(dml_lin_nmes) && has_ks_dml && has_sv_dml) {
  set.seed(3131L)
  n_explain_d <- min(20L, nrow(X_meta))
  n_bg_d <- min(100L, nrow(X_meta))
  ix_ed <- sample.int(nrow(X_meta), n_explain_d)
  ix_bd <- sample.int(nrow(X_meta), n_bg_d)
  X_explain_dml <- as.data.frame(X_meta[ix_ed, , drop = FALSE])
  bg_X_dml <- as.data.frame(X_meta[ix_bd, , drop = FALSE])

  dml_for_shap <- list(
    "LinearDML" = dml_lin_nmes,
    "SparseLinearDML" = dml_sparse_nmes,
    "CausalForestDML" = dml_cf_nmes
  )
  shap_dml_raw <- vector("list", length(dml_for_shap))
  names(shap_dml_raw) <- names(dml_for_shap)
  for (i in seq_along(dml_for_shap)) {
    shap_dml_raw[[i]] <- explain_cate(
      dml_for_shap[[i]],
      X = X_explain_dml,
      bg_X = bg_X_dml,
      use_permshap = TRUE,
      verbose = FALSE,
      exact = FALSE,
      max_iter = 8L,
      tol = 0.12
    )
  }
  shp_nmes_dml_list <- lapply(shap_dml_raw, function(ks) shapviz(ks, X = X_explain_dml))

  imp_dml <- dplyr::bind_rows(lapply(names(shp_nmes_dml_list), function(lab) {
    ks_obj <- shap_dml_raw[[lab]]
    S_mat <- as.matrix(ks_obj$S)
    data.frame(
      feature = colnames(ks_obj$X),
      mean_abs_shap = colMeans(abs(S_mat)),
      learner = lab,
      stringsAsFactors = FALSE
    )
  }))

  print(
    ggplot2::ggplot(imp_dml, ggplot2::aes(x = stats::reorder(factor(feature), mean_abs_shap), y = mean_abs_shap)) +
      ggplot2::geom_col(fill = "#8F4CB3", width = 0.72) +
      ggplot2::facet_wrap(~learner, scales = "free_y", ncol = 3L) +
      ggplot2::coord_flip() +
      ggplot2::labs(
        title = "NMES DML: mean |SHAP| for predicted CATE",
        subtitle = "Continuous packyears; permutation SHAP on a small explain set (predicted θ̂(X), not causal guarantees)",
        x = NULL,
        y = "Mean |SHAP|"
      ) +
      ggplot2::theme_bw() +
      ggplot2::theme(strip.text = ggplot2::element_text(face = "bold"))
  )
} else if (!is.null(dml_lin_nmes)) {
  message("Skipping DML SHAP: install.packages(c(\"kernelshap\", \"shapviz\"))")
}
```

```{r}
#| label: fig-nmes-shap-dml-beeswarm
#| fig-cap: "Beeswarm SHAP for LinearDML predicted CATE on the NMES explain subset."
#| warning: false
#| fig.width: 8
#| fig.height: 6
if (!is.null(shp_nmes_dml_list) && length(shp_nmes_dml_list) > 0) {
  print(
    shapviz::sv_importance(shp_nmes_dml_list[["LinearDML"]], kind = "beeswarm") +
      ggplot2::ggtitle("LinearDML — beeswarm SHAP (NMES)")
  )
}
```

### SHAP interpretability for predicted CATE (NMES) {#sec-nmes-shap-cate}

[SHAP](https://github.com/shap/shap) attributes **predicted** $\hat{\tau}(X)$ to covariates. We use **`RCausalML::explain_cate()`** with **`use_permshap = TRUE`**, smaller explain/background samples, and relaxed **`max_iter`** / **`tol`** so the document renders in reasonable time. The **X-Learner** is fit **without** a fixed propensity vector so `predict()` evaluates learned $e(X)$ on any `newdata`, as required when `kernelshap` perturbs features.

We run SHAP for **S-, T-, X-, and R-learners** only: **DR-learner** predictions require multiple cross-fitted surfaces per call, so permutation SHAP at this sample size is prohibitively slow; the **DR row remains in the binary ATE table** above. (You can run `explain_cate()` on **`tl_nmes_c`** or other **continuous** fits similarly if needed; we keep SHAP on **binary** fits here for cost and a single estimand.)

```{r}
#| label: fig-nmes-shap-meta-importance
#| fig-cap: "Mean absolute SHAP values for predicted binary CATE by S/T/X/R meta-learner (NMES subsample)."
#| warning: false
#| fig.width: 11
#| fig.height: 10
has_ks_n <- requireNamespace("kernelshap", quietly = TRUE)
has_sv_n <- requireNamespace("shapviz", quietly = TRUE)
shp_nmes_list <- NULL
X_explain_nmes <- NULL
if (!has_ks_n || !has_sv_n) {
  message("Skipping NMES SHAP: install.packages(c(\"kernelshap\", \"shapviz\"))")
} else {
  suppressPackageStartupMessages({
    library(kernelshap)
    library(shapviz)
  })
  set.seed(3030L)
  n_explain_n <- min(24L, nrow(X_meta))
  n_bg_n <- min(120L, nrow(X_meta))
  ix_e <- sample.int(nrow(X_meta), n_explain_n)
  ix_b <- sample.int(nrow(X_meta), n_bg_n)
  X_explain_nmes <- as.data.frame(X_meta[ix_e, , drop = FALSE])
  bg_X_nmes <- as.data.frame(X_meta[ix_b, , drop = FALSE])

  meta_nmes_for_shap <- list(
    "S (ranger)" = sl_nmes,
    "T (ranger)" = tl_nmes,
    "X (ranger)" = xl_nmes,
    "R (ranger)" = rl_nmes_bin
  )
  shap_nmes_raw <- vector("list", length(meta_nmes_for_shap))
  names(shap_nmes_raw) <- names(meta_nmes_for_shap)
  for (i in seq_along(meta_nmes_for_shap)) {
    shap_nmes_raw[[i]] <- explain_cate(
      meta_nmes_for_shap[[i]],
      X = X_explain_nmes,
      bg_X = bg_X_nmes,
      use_permshap = TRUE,
      verbose = FALSE,
      exact = FALSE,
      max_iter = 8L,
      tol = 0.12
    )
  }
  shp_nmes_list <- lapply(shap_nmes_raw, function(ks) shapviz(ks, X = X_explain_nmes))

  imp_nmes <- dplyr::bind_rows(lapply(names(shp_nmes_list), function(lab) {
    ks_obj <- shap_nmes_raw[[lab]]
    S_mat <- as.matrix(ks_obj$S)
    data.frame(
      feature = colnames(ks_obj$X),
      mean_abs_shap = colMeans(abs(S_mat)),
      learner = lab,
      stringsAsFactors = FALSE
    )
  }))

  print(
    ggplot2::ggplot(imp_nmes, ggplot2::aes(x = stats::reorder(factor(feature), mean_abs_shap), y = mean_abs_shap)) +
      ggplot2::geom_col(fill = "steelblue", width = 0.72) +
      ggplot2::facet_wrap(~learner, scales = "free_y", ncol = 2) +
      ggplot2::coord_flip() +
      ggplot2::labs(
        title = "NMES: mean |SHAP| for predicted CATE by meta-learner (ranger)",
        subtitle = "Binary W = 1{packyears > median}; S/T/X/R only (DR omitted for SHAP cost); predicted CATE, not necessarily causal",
        x = NULL,
        y = "Mean |SHAP|"
      ) +
      ggplot2::theme_bw() +
      ggplot2::theme(strip.text = ggplot2::element_text(face = "bold"))
  )
}
```


### SHAP beeswarm and waterfall for the T-Learner surrogate {#sec-nmes-shap-beeswarm}

```{r}
#| label: fig-nmes-shap-beeswarm-waterfall
#| fig-cap: "Beeswarm SHAP for T- and X-learners and a waterfall for the highest predicted T-learner CATE in the explain sample."
#| warning: false
#| fig.width: 9
#| fig.height: 11
if (!is.null(shp_nmes_list) && length(shp_nmes_list) > 0) {
  for (lab in c("T (ranger)", "X (ranger)")) {
    print(shapviz::sv_importance(shp_nmes_list[[lab]], kind = "beeswarm") + ggplot2::ggtitle(lab))
  }
  pred_t_n <- as.numeric(predict(tl_nmes, as.matrix(X_explain_nmes)))
  row_top <- which.max(pred_t_n)
  print(
    shapviz::sv_waterfall(shp_nmes_list[["T (ranger)"]], row_id = row_top) +
      ggplot2::ggtitle("Waterfall SHAP — T-Learner (ranger), highest predicted CATE in explain sample")
  )
}
```

## Summary and Conclusions {#sec-continuous-summary}

This notebook demonstrated meta-learners for continuous treatments (**Part A**: synthetic $D$ with known CADR; **Part B**: NMES **`nmes_data`** with `TOTALEXP`, `packyears`, and covariates from **`{causaldrf}`**). Key takeaways:

- **CADR**: For continuous treatment $D$, a natural target is $\tau(d,x) = \frac{\partial}{\partial d}\mathbb{E}[Y(d)|X=x]$, the marginal effect of dose at $d$ given $x$, estimated on a **dose grid** by **`r_learner_continuous()`**.
- **Native continuous meta-learners**: **`SLearner`**, **`TLearner`**, **`XLearner`**, **`RLearner`**, and **`DRLearner`** accept **`treatment_type = "continuous"`** and numeric **`treatment`** (e.g. `packyears`); estimands differ (quartile contrast, median split, Robinson, continuous DR, etc.—see **Details** in each function’s documentation).
- **R-Learner for continuous (grid)**: Cross-fit outcome and treatment (E[D|X]) models, form residuals, then use local linear regression with kernel weights over a dose grid to estimate $\tau(d,x)$ (or population averages).
- **Bandwidth**: Silverman-type rule used here; in practice choose via cross-validation. Local linear regression reduces boundary bias vs. kernel averaging.
- **Heterogeneity**: Stratifying by covariates (e.g., X1) shows how dose-response differs across subgroups; the R-Learner output can be used for subgroup-specific curves.
- **SHAP**: Because `r_learner_continuous()` returns a dose $\times$ unit surface, we explain **variation in** $\hat{\tau}(d_{\mathrm{ref}}, x)$ across units by permutation SHAP on a **ranger surrogate** of those slopes at a fixed reference dose. Interpret as attribution for **model-estimated** heterogeneity, not as causal Shapley effects for the true DGP.
- **Diagnostics**: Overlap (distribution of $D$ and $E[D|X]$), balance (SMD before/after stabilized IPTW), and CATE calibration (predicted vs residualized outcome by quintile) support the validity of dose-response estimates.
- **Decision framework**: Use the R-Learner for debiased, flexible dose-response estimation; consider GRF or causal BART when a single forest or Bayesian approach is preferred. Always check overlap and balance.
- **NMES (Part B)**: Continuous **`r_learner_continuous()`** curve for **`packyears`**; **S/T/X/R/DR** with **`treatment_type = "continuous"`** on raw **`packyears`** (native API); **binary** median-split **S/T/X/R/DR** for discrete ATE and **SHAP**; **`LinearDML` / `SparseLinearDML` / `CausalForestDML`** on the same subsample with **ATE**, **holdout RMSE** for $\hat{\mathbb{E}}[Y\mid X]$, **Spearman** agreement of predicted CATE, and **permutation SHAP** for $\hat{\theta}(X)$; **`ranger`** in nuisances where used. Interpret under **unconfoundedness**; use survey weights properly in real reporting.

For a **full dose–response surface** on a grid, use **`r_learner_continuous()`** or methods such as **GRF**. For a **single vector** of heterogeneous effects with continuous $D$, use **`treatment_type = "continuous"`** on the standard meta-learners and read each learner’s estimand in the package help. Validate bandwidth and overlap in $D$ given $X$; use SHAP on an explicit target (here, surrogate predictions at $d_{\mathrm{ref}}$ or binary $\hat{\tau}(X)$) when reporting interpretability.

## Resources {#sec-continuous-resources}

1. Nie & Wager (2021). *Biometrika*: Quasi-oracle estimation (R-Learner theory)  
2. Kennedy et al. (2017). *JASA*: Nonparametric causal effects of continuous treatments  
3. Athey & Wager (2019). *arXiv*: Generalized random forests  
4. Chernozhukov et al. (2018). *Econometrica*: Double/debiased ML  
5. Imai, K., & van Dyk, D. A. (2004). Causal inference with general treatment regimes. *Journal of the American Statistical Association*, 99(467), 854–866. (NMES / continuous treatment)  
6. `{causaldrf}`: [https://CRAN.R-project.org/package=causaldrf](https://CRAN.R-project.org/package=causaldrf) — `data("nmes_data", package = "causaldrf")`

## Scientific Terminology for Beginners

This quick glossary explains the core scientific terms used in this notebook in plain language.

### 1) Potential Outcomes

- **Meaning**: For each person, imagine two possible worlds: one where they receive treatment and one where they do not.
- **Example**: If a patient gets a drug, one potential outcome is their recovery with the drug, and another potential outcome is their recovery without the drug.

### 2) Treatment Effect

- **Meaning**: The difference between outcome with treatment and outcome without treatment.
- **Example**: If expected blood pressure decreases by 8 points with treatment and 3 points without treatment, the treatment effect is 5 points.

### 3) CATE (Conditional Average Treatment Effect)

- **Meaning**: The average treatment effect for a specific subgroup with features \(X\) (for example, age group or risk profile).
- **Example**: The treatment may work better for younger patients than older patients; each subgroup has its own CATE.

### 4) Confounding

- **Meaning**: A third factor affects both treatment assignment and the outcome, creating misleading associations.
- **Example**: Sicker patients may be more likely to receive treatment and also more likely to have poor outcomes, even if treatment helps.

### 5) Propensity Score

- **Meaning**: The probability of receiving treatment given covariates \(X\): \(P(T=1\mid X)\).
- **Example**: If two patients have similar age, health history, and baseline risk, they should have similar propensity scores.

### 6) Counterfactual

- **Meaning**: The unobserved outcome for the treatment status the person did not receive.
- **Example**: For a treated person, we observe treatment outcome but not what would have happened under control; that missing value is counterfactual.

### 7) Overlap (Positivity)

- **Meaning**: For each covariate profile, there is a non-zero chance of both treatment and control.
- **Example**: If a subgroup always receives treatment and never control, causal comparison for that subgroup is unreliable.

### 8) Nuisance Models

- **Meaning**: Helper models (like outcome models or propensity models) used to estimate the final causal quantity.
- **Example**: In DR- or R-learning, we first model \(E[Y\mid X,T]\) and \(P(T\mid X)\), then combine them to estimate CATE.

### 9) Cross-Fitting

- **Meaning**: A sample-splitting strategy where one part of data fits nuisance models and another part estimates treatment effects.
- **Example**: Train propensity/outcome models on fold A and estimate pseudo-outcomes on fold B, then rotate folds.

### 10) Doubly Robust Estimation

- **Meaning**: An estimator is consistent if either the outcome model or the propensity model is correctly specified.
- **Example**: Even if the propensity model is imperfect, a correctly specified outcome model can still yield valid effect estimates (and vice versa).

