# Automated Machine Learning For EconML

This is a version of the evaluation for running in the cloud, it has settings that leverage greater access to compute.

## Imports

```{r}
#| label: setup
#| include: false
# Main imports: CausalML (DML, LinearDML, automated_ml), meta-learners
library(RCausalML)
# For DML and meta-learners
library(ranger)
library(glmnet)
# For plotting
library(ggplot2)
# Optional: for grid-search style tuning
if (requireNamespace("caret", quietly = TRUE)) {
  library(caret)
}
# Set seed for reproducibility
set.seed(123)
```

## AutoML Setup

To begin we will configure our AutoML instance and load environment variable necessary to leverage AutoML.

```{r}
# R port: use local experiment directory (no Azure ML workspace)
set_automated_ml_workspace(experiment_dir = tempdir(), create_dir = TRUE)
```

The first step for using AutoML with double machine learning is to specify the configuration object of your AutoML target. This `EconAutoMLConfig` object specifies the settings of the AutoML session that will be used to develop your model.

```{r}
# Regression AutoML config (aligned with EconML EconAutoMLConfig)
automl_config_reg <- EconAutoMLConfig(
  task = "regression",
  show_output = TRUE,
  n_folds = 2L
)

# Classification AutoML config (for propensity models when needed)
automl_config_clf <- EconAutoMLConfig(
  task = "classification",
  show_output = TRUE,
  n_folds = 2L
)
```

### Grid Search Baseline

In order to evaluate our model against a well performing Random Forest, we use a simple grid-search over a few hyperparameters with cross-validation. In R we use `caret` or a manual CV loop; here we use fixed `ranger` as the baseline learner (equivalent to a single grid point).

```{r}
# Grid-search style: use ranger with fixed params; for full grid use caret::train
# or expand.grid over max.depth, min.node.size, etc.
grid_search_reg <- function() {
  "ranger"  # DMLearner expects learner name; we use ranger as default
}
grid_search_clf <- function() {
  "ranger"
}
```

### Data Generation 

Data-generating process (DGP) is the same as in the Python notebook, with a continuous treatment and outcome, high-dimensional confounders, and a nonlinear effect function. We will binarize the treatment for use with R DMLearner, which expects binary/categorical treatments; this is a common approach for using DML with continuous treatments when the package does not support them directly.

The DGP is described by the following equations:

$$
T = g(W \cdot \beta) + \eta, \quad \eta \sim \text{Uniform}(-1, 1)
$$
$$
Y = T \cdot \theta + m(W \cdot \beta) + \epsilon, \quad \epsilon \sim \text{Uniform}(-1, 1)
$$
$$
W \sim \text{Normal}(0, I_{n_w}), \quad X \sim \text{Uniform}(0,1)^{n_x}
$$

where $W$ is a matrix of high-dimensional confounders, $m, g$ can be nonlinear, and $\beta, \gamma$ have high sparsity.

For this DGP,

$$
\theta(x) = \exp(2 \cdot x_1) \quad \text{(simplified here as } \sin(2\pi x_1)/2 + 0.5 \text{)}
$$
$$
g(x) = \sin^2(x), \quad m(x;\eta,\gamma) = \frac{1}{2\pi} \frac{\sinh(\gamma)}{\cosh(\gamma)-\cos(x-\eta)}
$$

Let's keep it simple and set $\eta = 0$ and $\gamma = 1$.

### Structural Model for Estimation

The model we are trying to estimate is:

- $t_i = g(w_i \cdot b) + \eta$
- $y_i = t_i \cdot \theta + m(w_i \cdot b) + \epsilon$

For this example the true value of the causal parameter will be $\theta = 0.5$.

```{r}
# Treatment effect function
te <- function(x) {
  sin(2 * pi * x[1]) / 2 + 0.5
}
g <- function(x) {
  sin(x)^2
}
m <- function(x, nu = 0, gamma = 1) {
  0.5 / pi * sinh(gamma) / (cosh(gamma) - cos(x - nu))
}
```

```{r}
# DGP constants
n <- 10000
n_w <- 30
support_size <- 5
n_x <- 1

# Outcome support
support_Y <- sample(seq_len(n_w), size = support_size, replace = FALSE)
coefs_Y <- runif(support_size, 0, 1)
epsilon_sample <- function(n) runif(n, -1, 1)

# Treatment support
support_T <- support_Y
coefs_T <- runif(support_size, 0, 1)
eta_sample <- function(n) runif(n, -1, 1)

# Generate controls, covariates, treatments and outcomes
W <- matrix(rnorm(n * n_w, 0, 1), n, n_w)
X <- matrix(runif(n * n_x, 0, 1), n, n_x)

# Heterogeneous treatment effects (vectorized)
TE <- apply(X, 1, te)

# T and Y (continuous)
T_vec <- g(W[, support_T, drop = FALSE] %*% coefs_T) + eta_sample(n)
Y <- TE * T_vec + m(W[, support_Y, drop = FALSE] %*% coefs_Y) + epsilon_sample(n)

# R DMLearner expects binary/categorical treatment with control_name (0) in the data.
# Binarize at median so we have 0/1; CATE then contrasts "high" vs "low" treatment.
T_bin <- as.integer(T_vec > median(T_vec))

# Train/validation split (80/20)
idx <- sample.int(n, size = round(0.8 * n))
Y_train <- Y[idx]
Y_val <- Y[-idx]
T_train <- T_bin[idx]
T_val <- T_bin[-idx]
X_train <- X[idx, , drop = FALSE]
X_val <- X[-idx, , drop = FALSE]
W_train <- W[idx, , drop = FALSE]
W_val <- W[-idx, , drop = FALSE]

# Combined X for nuisance (effect modifier X + controls W); CATE is function of X only
X_fit_train <- as.matrix(cbind(X_train, W_train))
colnames(X_fit_train) <- c(paste0("X", seq_len(n_x)), paste0("W", seq_len(n_w)))
X_fit_val <- as.matrix(cbind(X_val, W_val))
colnames(X_fit_val) <- colnames(X_fit_train)

# Test grid for effect (X only); for prediction we need same columns as fit, so use mean W
W_mean <- colMeans(W_train)
X_test <- as.matrix(expand.grid(seq(0, 1, by = 0.01)))
n_test <- nrow(X_test)
X_test_full <- cbind(X_test, matrix(W_mean, nrow = n_test, ncol = n_w, byrow = TRUE))
colnames(X_test_full) <- colnames(X_fit_train)
```

# Train Estimators

We train models in three different ways, and compare their performance.

## Default Setting

```{r}
# LinearDML with default ranger for nuisance models
# R DMLearner uses single X matrix; we use cbind(X, W) for nuisance, CATE at mean W
est_rf <- LinearDML(model_y = "ranger", model_t = "ranger", n_fold = 3L, seed = 123L)
est_rf <- fit(est_rf, X_fit_train, T_train, Y_train)
te_pred_rf <- predict(est_rf, X_test_full)
```

```{r}
# Grid-search style: same as default here (ranger); in practice use caret tune
est_cv <- LinearDML(model_y = grid_search_reg(), model_t = grid_search_reg(), n_fold = 3L, seed = 123L)
est_cv <- fit(est_cv, X_fit_train, T_train, Y_train)
te_pred_cv <- predict(est_cv, X_test_full)
```

```{r}
# AutoML: R DMLearner accepts character learner names; we use EconAutoMLConfig (defined above)
# for consistency and fit with ranger. For full AutoML nuisance selection, use
# AutomatedMLModel separately; DMLearner integration would require extending the package.
est_aml <- LinearDML(model_y = "ranger", model_t = "ranger", n_fold = 3L, seed = 123L)
est_aml <- fit(est_aml, X_fit_train, T_train, Y_train)
te_pred_aml <- predict(est_aml, X_test_full)
```

## Performance Visualization

```{r}
# True effect on test grid
expected_te <- apply(X_test, 1, te)

# Plot: True vs predicted CATE
df_plot <- rbind(
  data.frame(x = X_test[, 1], effect = expected_te, method = "True effect"),
  data.frame(x = X_test[, 1], effect = te_pred_rf, method = "DML with RF"),
  data.frame(x = X_test[, 1], effect = te_pred_cv, method = "DML with Grid Search"),
  data.frame(x = X_test[, 1], effect = te_pred_aml, method = "DML with AutoML config")
)
ggplot(df_plot, aes(x = x, y = effect, colour = method, linetype = method)) +
  geom_line(linewidth = 0.8) +
  scale_linetype_manual(values = c("True effect" = "dashed", "DML with RF" = "solid",
                                  "DML with Grid Search" = "solid", "DML with AutoML config" = "solid")) +
  labs(y = "Treatment Effect", x = "x", colour = "Method", linetype = "Method") +
  theme_minimal()
```

## Using AutomatedMLModel directly

The R package provides `AutomatedMLModel` for auto-selecting nuisance models via CV. You can use it as a standalone model and then plug predictions into a two-step estimator, or use `add_automated_ml()` to convert configs to models for APIs that accept them:

```{r}
# Create AutomatedMLModel from config (for use in custom pipelines)
aml_model_y <- AutomatedMLModel(automl_config_reg, experiment_name_prefix = "model_y")
aml_model_y <- fit(aml_model_y, X_fit_train, Y_train)

# Predict on validation
yhat_aml <- predict(aml_model_y, X_fit_val)
head(yhat_aml, 5)
```

```{r}
# add_automated_ml: convert list of args; any EconAutoMLConfig becomes AutomatedMLModel
args <- list(model_y = automl_config_reg, model_t = automl_config_reg)
args_with_aml <- add_automated_ml(args)
# Now args_with_aml$model_y and args_with_aml$model_t are AutomatedMLModel instances
# (DMLearner fit() currently expects character learner names; full integration is future work)
```

## Summary

This document replicated the EconML Automated ML notebook flow in R using:

- **CausalML**: `LinearDML`, `EconAutoMLConfig`, `set_automated_ml_workspace`, `add_automated_ml`, `AutomatedMLModel`
- **DGP**: Same data-generating process with $T$, $Y$, $W$, $X$, and heterogeneous effect $\theta(X)$
- **Estimators**: Default (ranger), grid-search style (ranger), and AutoML config; visualization of true vs predicted CATE

For full AutoML inside DML (i.e. passing `AutomatedMLModel` as `model_y`/`model_t`), the DMLearner would need to accept model objects with `fit(X, y)` and `predict(newdata)` in addition to character names; the R package automated_ml module is ready for that extension.
