# Causal Variational Autoencoder (CausalVAE) 


This tutorial provides a detailed guide to implementing and using Causal Variational Autoencoders (CausalVAE) in R, based on the 2021 CVPR paper "CausalVAE: Disentangled Representation Learning via Neural Structural Causal Models" by Yang et al. We'll cover the concepts, an R/torch implementation, and applications including Average Treatment Effect (ATE) prediction. The code uses the `torch` R package. For demonstration, we'll use the Diabetes dataset loaded from local storage.

## Overview

CausalVAE extends $\beta$-VAEs by incorporating a **Structural Causal Model (SCM)** in the latent space to learn causally disentangled representations. Unlike standard VAEs, which assume independent latents, CausalVAE models dependencies via a **Directed Acyclic Graph (DAG)**, allowing interventions and counterfactual reasoning.

**Key features:** - **Exogenous variables (**$\varepsilon$): Independent noise sources. - **Endogenous variables (**$z$): Causally generated from $\varepsilon$ via a neural SCM (e.g., linear with masks). - **Causal layer**: Enforces DAG structure for causal propagation. - **Benefits**: Robust to interventions, enables ATE estimation in causal inference. - **Applications**: Image generation with causal edits (e.g., CelebA faces), reinforcement learning, biomedical data analysis.

The model learns the causal graph jointly with representations, often requiring weak supervision (e.g., partial graph knowledge) for identifiability.

### How It Works

CausalVAE modifies the VAE architecture:

1.  **Encoder**: $q(\varepsilon \mid x)$ — Maps observations $x$ to independent exogenous $\varepsilon \sim \mathcal{N}(\mu, \sigma)$.

2.  **Causal Layer (SCM)**: Transforms $\varepsilon$ to $z$ via $z = f(z, \varepsilon; \theta)$, where $f$ respects a DAG. In the linear case: $$z = (I - A \odot M)^{-1} \varepsilon$$ but practically implemented as masked autoregressive or additive: $z = \varepsilon + (A \odot M) z$.

    -   $A$: Learnable weights.
    -   $M$: Binary mask for DAG (learned or fixed).

3.  **Decoder**: $p(x \mid z)$ — Reconstructs $x$ from causal $z$.

4.  **Prior**: $p(\varepsilon) = \mathcal{N}(0, I)$.

5.  **Loss**: $\beta$-VAE ELBO plus causal regularizers (DAG penalty, sparsity):

$$\mathcal{L} = \mathbb{E}_{q(\varepsilon|x)} \big[ \log p(x|z) \big] - \beta \, D_{\mathrm{KL}}\big( q(\varepsilon|x) \,\|\, p(\varepsilon) \big) + \gamma \, \big( h(A) + \lambda \|A\|_1 \big)$$

where $h(A)$ is a DAG penalty, e.g. $h(A) = \mathrm{tr}\big(\exp(A \circ A)\big) - d$ (with $d$ the number of latent dimensions).

For **interventions**: set $\mathrm{do}(z_i = \text{value})$ and propagate through the graph.

## Implementation in R

We will use the `RCausalML` package for the implementation CausalVAE in R which defines the model architecture, forward pass, and loss function. We will source this file to access the necessary functions for training and evaluation.


**Standalone functions**

-   `generate_data`-synthetic data generation
-   `estimate_ate_causalvae_ate` - average treatment effect estimation for `CausalVAE_ATE`
-   `loss_function` - VAE + causal loss
-   `train_causalvae` - training loop
-   `tune_hyperparameters`- hyperparameter tuning
-   `print_optimization_summary`-  optimization summary

**`nn_module` classes**

-   `CausalVAE` - main Causal VAE
-   `CausalVAE_ATE` -  CausalVAE with ATE outcome head


### Setup

We will use the `torch` package for the implementation. The code below checks for required packages, loads them, and sets up the device for computation.

```{r}
#| label: setup-packages
#| message: false
#| warning: false

# Check and install required packages (minimal set for speed)
packages <- c("torch", "progress", 
              "ggplot2", "corrplot", 
              "coro", "RCausalML",
              "tidyverse")

# Verify installation
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))

# Load packages with suppressed messages
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```


### Device Setup

```{r}
#| label: device-setup
#| message: false

# Check for GPU availability (torch)
device <- if (torch::cuda_is_available()) "cuda" else "cpu"
cat("Using device:", device, "\n")
set.seed(42)
torch::torch_manual_seed(42L)
```

### Data Generation and Processing

We generate synthetic data with 3D latents $z$ following a causal chain: $\varepsilon_1 \to z_1 = \varepsilon_1$ $z_2 = z_1 + \varepsilon_2$ $z_3 = z_2^2 + \varepsilon_3$

Observations $x$ are a nonlinear mix of $z$. For real data, we'll use the Diabetes dataset.

```{r}
#| label: generate-synthetic-data
#| echo: true

# OPTIMIZED: Vectorized data generation with pre-allocation
generate_data  <- function(n_samples = 5000L, latent_dim = 3L) {
  # Pre-allocate all matrices
  epsilon_mat  <- matrix(rnorm(n_samples * latent_dim), 
                        nrow = n_samples, ncol = latent_dim)
  z_mat  <- matrix(0.0, nrow = n_samples, ncol = latent_dim)
  x_mat  <- matrix(0.0, nrow = n_samples, ncol = latent_dim)
  
  # Vectorized causal generation
  z_mat[, 1]  <- epsilon_mat[, 1]
  z_mat[, 2]  <- z_mat[, 1] + epsilon_mat[, 2]
  z_mat[, 3]  <- z_mat[, 2]^2 + epsilon_mat[, 3]
  
  # Vectorized observation generation
  x_mat[, 1]  <- z_mat[, 1] * z_mat[, 3]
  x_mat[, 2]  <- sin(z_mat[, 2]) + z_mat[, 1]
  x_mat[, 3]  <- z_mat[, 3]^2 + rnorm(n_samples, 0, 0.1)
  
  list(
    x = torch_tensor(x_mat, dtype = torch_float(), device = device),
    z = torch_tensor(z_mat, dtype = torch_float(), device  = device),
    epsilon = torch_tensor(epsilon_mat, dtype = torch_float(), device = device)
  )
}

# Generate synthetic data (reduced from 10000 to 5000 for speed)
synthetic_data  <- generate_data(n_samples = 5000L)
x  <- synthetic_data$x
true_z  <- synthetic_data$z
true_epsilon  <- synthetic_data$epsilon

# Normalize the data
normalize_data  <- function(x) {
  mean_val  <- x$mean(dim = 1L)
  std_val  <- x$std(dim = 1L)
  list(
    x_norm = (x - mean_val) / std_val,
    mean = mean_val,
    std = std_val
  )
}

norm_result  <- normalize_data(x)
x_norm  <- norm_result$x_norm
x_mean  <- norm_result$mean
x_std  <- norm_result$std

# Data Split
n_samples  <- nrow(x_norm)
train_size  <- as.integer(0.8 * n_samples)
val_size  <- as.integer(0.1 * n_samples)

x_train  <- x_norm[1:train_size, ]
x_val  <- x_norm[(train_size + 1):(train_size + val_size), ]
x_test  <- x_norm[(train_size + val_size + 1):n_samples, ]

true_z_train  <- true_z[1:train_size, ]
true_z_val  <- true_z[(train_size + 1):(train_size + val_size), ]
true_z_test  <- true_z[(train_size + val_size + 1):n_samples, ]

cat(sprintf( "Train: %d, Val: %d, Test: %d\n ", 
            nrow(x_train), nrow(x_val), nrow(x_test)))
```

### Model Building

CausalVAE is a variant of Variational Autoencoder (VAE) where latent variables are structured according to a causal Directed Acyclic Graph (DAG) instead of being standard Gaussian.

Main Components: - **Encoder**: A 2-layer MLP that outputs $\mu$, $\log \sigma^2$ for a Gaussian over noise variables $\varepsilon$ - **Reparameterization**: Sampling: $\varepsilon = \mu + \sigma \cdot \mathcal{N}(0,1)$ - **Causal Layer (SCM over latents)**: Learns a soft adjacency and generates latents via causal relationships - **Decoder**: A 2-layer MLP producing mean and log-variance for $p(x | z)$ - **DAG Penalty**: A matrix-exponential constraint ensures that the learned graph is acyclic

Model Flow: 

$$
x \rightarrow \text{Encoder} \rightarrow \varepsilon \rightarrow \text{Causal Layer} \rightarrow z \rightarrow \text{Decoder} \rightarrow \text{Reconstruction}

$$


The model jointly learns both accurate reconstructions and a causal graph over latent variables.

### Model Initialization and Loss Function

Train with the augmented loss: reconstruction, $\beta \cdot D_{\mathrm{KL}}(q(\varepsilon|x) | p(\varepsilon))$, and $\gamma \big( h(A) + \lambda |M|_1 \big)$.


```{r}
#| label: loss-function
#| echo: true
#| warning: false

# This block defines the loss function and sets up the model and optimizer for training the CausalVAE.

# Instantiate the CausalVAE, optimizer, and scheduler
model <- CausalVAE()$to(device = device)
optimizer <- optim_adam(model$parameters, lr = 1e-3, weight_decay = 1e-5)
# Reduce learning rate if validation loss plateaus (for stable convergence)
scheduler <- lr_reduce_on_plateau(optimizer, mode = "min", factor = 0.5, patience = 15)
# Number of training epochs
epochs <- 200L

cat(sprintf("Model: %s\n", class(model)[1]))
cat(sprintf("Device: %s\n", device))
cat(sprintf("Epochs: %d (with early stopping)\n", epochs))
cat(sprintf("LR (init): %.0e\n", 1e-3))
```

### Training

```{r}
#| label: training-loop
#| echo: true
#| warning: false
#| message: false

# Training loop for CausalVAE following best practices from the @10-CausalVAE case study
# See: guidance on debugging, numerical stability, loss tracking, and robust scheduler usage.

train_losses <- numeric(epochs)
val_losses <- numeric(epochs)

# Hyperparameters tuned for stability (reduces post-valley spikes and divergence)
learning_rate <- 2e-4         # Conservative LR to avoid overshooting and loss spikes
min_learning_rate <- 5e-6     # Allow scheduler to anneal very low
weight_decay <- 1e-5          # Slight increase for regularization
batch_size <- 128L            # Appropriate for modern GPUs/CPUs
epochs <- 250L                # Max epochs; early stopping typically stops sooner
warmup <- 50L                 # Gradual structural penalty introduction
lr_warmup_epochs <- 10L        # LR warmup so early updates are gentler
gamma_scale_cap <- 0.35       # Cap structural penalty to avoid post-valley rise
grad_clip_max <- 0.5         # Tighter gradient clipping for numerical stability
loss_skip_threshold <- 50.0   # Skip batch if loss exceeds this (avoids explosion)

optimizer <- optim_adam(model$parameters, lr = learning_rate, weight_decay = weight_decay)
scheduler <- lr_reduce_on_plateau(optimizer, mode = "min", factor = 0.5, patience = 8, min_lr = min_learning_rate)

train_dataset <- torch::tensor_dataset(x_train)
train_loader  <- torch::dataloader(train_dataset, batch_size = batch_size, shuffle = TRUE)

val_dataset   <- torch::tensor_dataset(x_val)
val_loader    <- torch::dataloader(val_dataset, batch_size = batch_size, shuffle = FALSE)

if (!requireNamespace("coro", quietly = TRUE)) {
  stop("package 'coro' is required; install with install.packages('coro')")
}

best_val_loss <- Inf
early_stopping_patience <- 10  # Stop soon after val loss stops improving (best model is often near first valley)
early_stopping_counter <- 0

save_best_model <- function(model, path = ".causalvae_best_model.pt") {
  torch::torch_save(model$state_dict(), path)
}
best_model_path <- tempfile("causalvae_best_", fileext = ".pt")

for (epoch in 1:epochs) {
  # Learning rate warmup: gentler updates in early epochs (only during warmup; after that scheduler controls LR)
  if (epoch <= lr_warmup_epochs) {
    optimizer$param_groups[[1]]$lr <- learning_rate * (epoch / lr_warmup_epochs)
  }

  # Structural penalty (gamma): gradual, capped to avoid post-valley divergence
  gamma_scale <- min(gamma_scale_cap, (epoch + 1) / warmup)

  # --- Training Phase ---
  model$train()
  epoch_train_loss <- 0.0
  batch_count <- 0

  train_batches <- coro::collect(train_loader)
  for (batch in train_batches) {
    x_batch <- if (is.list(batch)) batch[[1]] else batch
    x_batch_on_device <- x_batch$to(device = device)

    optimizer$zero_grad()

    forward_out <- model$forward(x_batch_on_device)
    enc_mu <- forward_out$enc_mu
    enc_logvar <- forward_out$enc_logvar
    dec_mu <- forward_out$dec_mu
    dec_logvar <- forward_out$dec_logvar

    loss <- loss_function(
      x_batch_on_device, dec_mu, dec_logvar, enc_mu, enc_logvar, model, gamma_scale = gamma_scale
    )

    loss_val <- loss$item()
    if (!is.finite(loss_val)) {
      cat(sprintf("Warning: Loss is %s at epoch %d, skipping batch\n",
                  ifelse(is.nan(loss_val), "NaN", "Inf"), epoch))
      next
    }
    if (loss_val > loss_skip_threshold) {
      cat(sprintf("Warning: Loss %.1f > %.1f at epoch %d, skipping batch (numerical stability)\n",
                  loss_val, loss_skip_threshold, epoch))
      next
    }

    loss$backward()
    tryCatch(
      nn_utils_clip_grad_norm_(model$parameters, max_norm = grad_clip_max),
      error = function(e) { optimizer$zero_grad(); cat("Gradient clip error, zeroed grads\n") }
    )
    optimizer$step()
    epoch_train_loss <- epoch_train_loss + loss_val
    batch_count <- batch_count + 1
  }
  
  train_losses[epoch] <- if (batch_count > 0) epoch_train_loss / batch_count else NA
  
  # --- Validation Phase ---
  model$eval()
  epoch_val_loss <- 0.0
  val_batch_count <- 0
  with_no_grad({
    val_batches <- coro::collect(val_loader)
    for (val_batch in val_batches) {
      x_val_batch <- if (is.list(val_batch)) val_batch[[1]] else val_batch
      x_val_batch_on_device <- x_val_batch$to(device = device)
      
      forward_out <- model$forward(x_val_batch_on_device)
      enc_mu <- forward_out$enc_mu
      enc_logvar <- forward_out$enc_logvar
      dec_mu <- forward_out$dec_mu
      dec_logvar <- forward_out$dec_logvar
      
      loss <- loss_function(
        x_val_batch_on_device, dec_mu, dec_logvar, enc_mu, enc_logvar, model, gamma_scale = gamma_scale
      )
      loss_val <- loss$item()
      if (is.finite(loss_val)) {
        epoch_val_loss <- epoch_val_loss + loss_val
        val_batch_count <- val_batch_count + 1
      }
    }
  })
  val_losses[epoch] <- if (val_batch_count > 0) epoch_val_loss / val_batch_count else NA
  
  # Early stopping and robust LR scheduling
  vl <- val_losses[epoch]
  if (!is.finite(vl)) vl <- Inf
  scheduler$step(vl)
  
  if (vl < best_val_loss - 1e-5) {
    best_val_loss <- vl
    early_stopping_counter <- 0
    save_best_model(model, best_model_path)
  } else {
    early_stopping_counter <- early_stopping_counter + 1
  }
  
  if (early_stopping_counter >= early_stopping_patience) {
    cat(sprintf("Early stopping at epoch %d (no improvement for %d epochs)\n", 
                epoch + 1, early_stopping_patience))
    break
  }
  
  # Progress logging every 5 epochs and at epoch 1 (so short runs and early stopping are visible)
  if (epoch == 1 || epoch %% 5 == 0) {
    lr <- optimizer$param_groups[[1]]$lr
    cat(sprintf("Epoch %3d, Train: %.4f, Val: %.4f, lr: %.2e, gamma_scale: %.3f\n",
                epoch, train_losses[epoch], val_losses[epoch], lr, gamma_scale))
  }
}

# Number of epochs actually run, for plotting and reporting
actual_epochs <- epoch

if (file.exists(best_model_path)) {
  model$load_state_dict(torch::torch_load(best_model_path))
}
```

### Visualize Training and Validate Loss

```{r}
#| label: visualize-training
#| echo: true
#| warning: false
#| fig-width: 6
#| fig-height: 5

# Plot training and validation loss (only epochs that were actually run;
# after early stopping, remaining entries are 0 and would show a fake drop to zero)
library(ggplot2)

epoch_df  <- data.frame(
  Epoch = 1:actual_epochs,
  Train_Loss = train_losses[1:actual_epochs],
  Val_Loss = val_losses[1:actual_epochs]
)

# Remove rows with NA (e.g. if all batches were skipped in an epoch)
epoch_df <- epoch_df[!is.na(epoch_df$Train_Loss) & !is.na(epoch_df$Val_Loss), ]

ggplot(epoch_df, aes(x = Epoch)) +
  geom_line(aes(y = Train_Loss, color  =  "Train Loss "), linewidth = 1.5) +
  geom_line(aes(y = Val_Loss, color =  "Val Loss "), linewidth = 1.5) +
  labs(
    title =  "CausalVAE: Training vs Validation Loss ",
    x =  "Epoch ",
    y =  "Loss ",
    color =  " "
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title = element_text(size = 12),
    legend.position =  "best "
  ) 
```

### ATE Prediction Validation

For Average Treatment Effect (ATE): assume treatment $T$ acts on $z_1$ (e.g., shift by 1) and outcome $Y$ depends on $z$. We extend the data generation to include $T$ and $Y$, then add an outcome head $\mathbb{E}[Y \mid z, T]$ and estimate ATE as $\mathbb{E}[Y \mid \mathrm{do}(T=1)] - \mathbb{E}[Y \mid \mathrm{do}(T=0)]$.

```{r}
#| label: ate-data-generation
#| echo: true

generate_data_with_treatment  <- function(n_samples = 5000L, latent_dim = 3L) {
  epsilon_mat  <- matrix(rnorm(n_samples * latent_dim), nrow = n_samples, ncol = latent_dim)
  z_mat  <- matrix(0.0, nrow = n_samples, ncol = latent_dim)
  z_mat[, 1]  <- epsilon_mat[, 1]
  z_mat[, 2]  <- z_mat[, 1] + epsilon_mat[, 2]
  z_mat[, 3]  <- z_mat[, 2]^2 + epsilon_mat[, 3]
  
  T_vec  <- rbinom(n_samples, 1, 0.5)
  Y_vec  <- z_mat[, 3] + T_vec * 0.5 * z_mat[, 2] + rnorm(n_samples, 0, 0.1)
  
  x_mat  <- matrix(0.0, nrow = n_samples, ncol = latent_dim)
  x_mat[, 1]  <- z_mat[, 1] * z_mat[, 3]
  x_mat[, 2]  <- sin(z_mat[, 2]) + z_mat[, 1]
  x_mat[, 3]  <- z_mat[, 3]^2 + rnorm(n_samples, 0, 0.1)
  
  list(
    x = torch_tensor(x_mat, dtype = torch_float(), device = device),
    z = torch_tensor(z_mat, dtype = torch_float(), device = device),
    epsilon = torch_tensor(epsilon_mat, dtype = torch_float(), device = device),
    T = torch_tensor(matrix(T_vec, ncol = 1), dtype = torch_float(), device = device),
    Y = torch_tensor(matrix(Y_vec, ncol = 1), dtype = torch_float(), device = device) 
  )
}

ate_data  <- generate_data_with_treatment(n_samples = 5000L)
x_ate  <- ate_data$x
z_ate  <- ate_data$z
T_ate  <- ate_data$T
Y_ate  <- ate_data$Y

x_ate_norm  <- normalize_data(x_ate)$x_norm

train_size_ate  <- as.integer(0.8 * nrow(x_ate_norm))
val_size_ate  <- as.integer(0.1 * nrow(x_ate_norm))

x_ate_test  <- x_ate_norm[(train_size_ate + val_size_ate + 1):nrow(x_ate_norm), ]
z_ate_test  <- z_ate[(train_size_ate + val_size_ate + 1):nrow(z_ate), ]
T_ate_test  <- T_ate[(train_size_ate + val_size_ate + 1):nrow(T_ate), ]
Y_ate_test  <- Y_ate[(train_size_ate + val_size_ate + 1):nrow(Y_ate), ]

true_ate  <- (0.5 * z_ate_test[, 2])$mean()$item()
```

### ATE Prediction

```{r}
#| label: ate-prediction
#| echo: true
#| warning: false
#| message: false


# The following code block trains a CausalVAE_ATE model to predict outcomes (Y) 
# given features (x) and a treatment indicator (T), as part of an average 
# treatment effect (ATE) estimation pipeline.

# Prepare training data for ATE prediction:
#   - x_ate_train: normalized observed features for training.
#   - T_ate_train: treatment indicator (binary; 0 or 1) for training.
#   - Y_ate_train: observed outcome based on both z and T for training.
x_ate_train  <- x_ate_norm[1:train_size_ate, ]
T_ate_train  <- T_ate[1:train_size_ate, ]
Y_ate_train  <- Y_ate[1:train_size_ate, ]

# Combine (x, T, Y) into a torch Dataset for efficient batching.
ate_dataset  <- torch::tensor_dataset(x_ate_train, T_ate_train, Y_ate_train)
ate_loader  <- torch::dataloader(ate_dataset, batch_size = 256L, shuffle = TRUE)

# Build CausalVAE_ATE for synthetic data (match input/latent_dim as training data).
model_ate  <- CausalVAE_ATE(
  input_dim = dim(x_ate_norm)[2],
  latent_dim = dim(z_ate)[2],
  hidden_dim = 64L,
  beta = 1.0,
  gamma = 1.0,
  lambda_sparsity = 0.1
)$to(device = device)

optimizer_ate  <- optim_adam(model_ate$parameters, lr = 1e-3)

for (epoch in 1:100) {
  model_ate$train()
  train_batches <- coro::collect(ate_loader)

  for (batch in train_batches) {
    x_b  <- batch[[1]]$to(device = device)
    t_b  <- batch[[2]]$to(device = device)
    y_b  <- batch[[3]]$to(device = device)

    optimizer_ate$zero_grad()

    forward_out  <- model_ate$forward(x_b, t_b)
    enc_mu  <- forward_out$enc_mu
    enc_logvar  <- forward_out$enc_logvar
    dec_mu  <- forward_out$dec_mu
    dec_logvar  <- forward_out$dec_logvar
    y_pred  <- forward_out$y_pred

    # Compute the VAE loss (includes reconstruction and KL divergence terms).
    loss_vae  <- loss_function(x_b, dec_mu, dec_logvar, enc_mu, enc_logvar, model_ate)
    loss_mse  <- nnf_mse_loss(y_pred, y_b)

    (loss_vae + loss_mse)$backward()
    tryCatch(nn_utils_clip_grad_norm_(model_ate$parameters, max_norm = 1.0), error = function(e) NULL)
    optimizer_ate$step()
  }

  if ((epoch + 1) %% 50 == 0) {
    cat(sprintf( "Epoch %d, VAE+MSE loss done.\n ", epoch + 1))
  }
}

# Function to estimate ATE for CausalVAE_ATE models
estimate_ate_CausalVAE_ATE <- function(model, n_samples = 500L, treatment_dim = 1L, shift = 1.0, device = "cpu") {
  model$eval()
  z_samples <- torch_randn(n_samples, model$latent_dim, device = device)
  t_ones   <- torch_ones(n_samples, treatment_dim, device = device)
  t_zeros  <- torch_zeros(n_samples, treatment_dim, device = device)

  # outcome_head is a nn_linear module; call directly with concatenated [z, T]
  zt1 <- torch_cat(list(z_samples, t_ones),  dim = 2L)
  zt0 <- torch_cat(list(z_samples, t_zeros), dim = 2L)

  y1 <- as.numeric(model$outcome_head$forward(zt1)$squeeze())
  y0 <- as.numeric(model$outcome_head$forward(zt0)$squeeze())
  mean(y1) - mean(y0)
}

pred_ate  <- estimate_ate_CausalVAE_ATE(model_ate, n_samples = 500L, treatment_dim = 1L, shift = 1.0, device = device)
cat(sprintf( "Predicted ATE: %.4f, True ATE (approx): %.4f\n ", pred_ate, true_ate))

# Optional package helper:
# pred_ate_pkg <- estimate_ate_causalvae_ate(model_ate, n_samples = 500L, treatment_dim = 1L, shift = 1.0, device = device)
```


## Summary and Conclusion

CausalVAE combines variational autoencoders with a structural causal model (SCM) in the latent space to learn intervenable, causally disentangled representations. This tutorial provided a full R/torch implementation and workflow:

**Concepts**: Encoder $q(\varepsilon|x)$, causal layer (DAG), decoder $p(x|z)$, and loss with reconstruction, KL, and DAG/sparsity terms.

**Synthetic data**: Data generated from a known causal chain ($z_1 \to z_2 \to z_3$) and nonlinear observations.

**Model and training**: CausalVAE and CausalVAE_ATE (with outcome head), loss function, training loop with warmup and ReduceLROnPlateau.

**ATE estimation**: $\mathrm{ATE} = \mathbb{E}[Y|\mathrm{do}(T=1)] - \mathbb{E}[Y|\mathrm{do}(T=0)]$ via interventions on the causal layer.

**Takeaways**: For production use, add weak supervision (e.g. partial graph knowledge) for better identifiability and scale to images with convolutional encoder/decoder. Remaining challenges include identifiability of the causal graph and the computational cost of DAG penalties. CausalVAE is a step toward causal AI and more robust, intervenable representations for decision-making.

## Resources

-   Original paper: Yang et al., [CausalVAE: Disentangled Representation Learning via Neural Structural Causal Models](https://arxiv.org/abs/2004.08697).
-   GitHub: [Official implementation](https://github.com/huawei-noah/trustworthyAI/tree/master/research/CausalVAE).
-   Related: Lippe et al. CausalVAE variants; NOTEARS for DAG learning.
-   Tutorials: torch R package documentation; Causal inference with CausalML.
-   Books: "Causal Inference in Statistics" by Pearl for theory.


## Scientific terminology for beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Structural causal model (SCM) | System of equations describing how variables cause each other. | `Z2 = f(Z1, noise)` encodes direct causal influence from `Z1` to `Z2`. |
| DAG constraint | Rule forcing the learned causal graph to have no cycles. | Model penalizes adjacency matrices that imply loops. |
| Soft intervention | Intervention that changes mechanism/parameters without fully replacing equation. | Increasing treatment tendency rather than forcing treatment deterministically. |
| Hard intervention | Setting variable to a fixed value regardless of parents. | `do(T=1)` sets treatment to 1 for everyone. |
| Disentangled latent space | Latent dimensions correspond to separate factors of variation. | One dimension tracks dosage, another tracks baseline severity. |
| Interventional effect | Outcome change caused by manipulating treatment variable. | Compare predicted `E[Y|do(T=1)]` vs `E[Y|do(T=0)]`. |
