# Continuous Optimization Models (NOTEARS)


**NOTEARS** ("DAGs with NO TEARS", Zheng et al., NeurIPS 2018) was the first method to make causal structure learning (discovering a Directed Acyclic Graph / DAG from observational data) a **purely continuous, gradient-based optimization problem**.

Instead of searching the combinatorial space of $2^{O(d^2)}$ possible DAGs (intractable for $d \gtrsim 10$ variables), NOTEARS optimizes a real-valued weighted adjacency matrix $\mathbf{W} \in \mathbb{R}^{d \times d}$ directly. It minimizes a score that measures how well $\mathbf{W}$ explains the data while enforcing **acyclicity** via a smooth differentiable constraint and promoting **sparsity** with an $\ell_1$ penalty.

Under the **linear Structural Equation Model (SEM)** assumption: $$\mathbf{X} = \mathbf{X} \mathbf{W} + \mathbf{Z}$$ (where $\mathbf{Z}$ is independent noise, often Gaussian), the method recovers the causal DAG (up to Markov equivalence in some cases) efficiently.

**Key innovation**: The acyclicity function $$h(\mathbf{W}) = \mathrm{tr}(\exp(\mathbf{W} \odot \mathbf{W})) - d = 0$$ is exactly zero **if and only if** the graph of $\mathbf{W}$ is a DAG. Combined with augmented Lagrangian optimization, this turns the NP-hard problem into a tractable non-convex program solvable by L-BFGS or SGD.

NOTEARS is simple (core algorithm $<$ 60 lines), fast on moderate $d$ ($\leq 100$–200), and remains a foundational baseline for all later continuous causal discovery methods (GOLEM, DAG-GNN, GraN-DAG, etc.).


##  NOTEARS with {RCausalML}

We will implement the linear and nonlinear NOTEARS variants in R using {RCausalML} package. We will apply these to synthetic data (linear and nonlinear SEMs) and the Sachs dataset, evaluating performance against known ground truth DAGs.

The **RCausalML** package implements NOTEARS (NO TEARS) for causal directed acyclic graph (DAG) structure learning from observational data. NOTEARS treats the discrete acyclicity constraint as a continuous constraint $h(W) = 0$ using the trace of the matrix exponential ($h(W) = \mathrm{tr}(e^{W \odot W}) - d$), so the problem is solved with continuous optimization (e.g. augmented Lagrangian with L-BFGS-B for linear, Adam + L-BFGS for nonlinear).

**Linear NOTEARS**: learns a weighted adjacency matrix $W$ by minimizing a loss (e.g. least squares, logistic, Poisson) plus L1 penalty, subject to $h(W) = 0$. Uses `expm` and L-BFGS-B.

**Nonlinear NOTEARS** (R + **torch**): uses MLP or Sobolev basis models to capture nonlinear relationships; acyclicity is enforced via a differentiable $h(W)$. Requires the **torch** package.

**Main functions:**

- `notears_linear` — Linear NOTEARS: L2/logistic/Poisson loss + L1, augmented Lagrangian.
- `notears_nonlinear` — Nonlinear NOTEARS with MLP or Sobolev model (Adam then L-BFGS).
- `NotearsMLP` — Torch module: MLP for nonlinear SEM (when **torch** is installed).
- `NotearsSobolev` — Torch module: Sobolev basis for nonlinear SEM (when **torch** is installed).

**Utilities and simulation:**

- `is_dag` — Check if a weighted adjacency matrix is a DAG.
- `simulate_dag` — Simulate random DAG (Erdős–Rényi or scale-free).
- `simulate_parameter` — Draw edge weights for a DAG.
- `simulate_linear_sem` — Generate data from a linear SEM (Gaussian, exponential, uniform noise).
- `simulate_nonlinear_sem` — Generate data from a nonlinear SEM (MLP or MIM).
- `count_accuracy` — Compare estimated graph to ground truth (FDR, TPR, FPR, SHD, nnz).
- `set_random_seed` — Set R and torch seeds for reproducibility.


**Reference:** Zheng, X., Aragam, B., Ravikumar, P., & Xing, E. P. (2018). DAGs with NO TEARS: Continuous optimization for structure learning. *NeurIPS*. <https://arxiv.org/abs/1803.01422>.

### Setup

```{r}
#| echo: false
#| warning: false
#| message: false

# Install required packages if not already installed
required_packages <- c("expm", "igraph", "torch", "tidyverse")

for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
    library(pkg, character.only = TRUE)
  }
}

# Source the NOTEARS implementation (from package R/)
pkg_root <- if (file.exists("R/notears.R")) "." else if (file.exists("../../../../R/notears.R")) "../../../../" else stop("Run from package root or notebook directory")
source(file.path(pkg_root, "R/notears.R"))
source(file.path(pkg_root, "R/drawing.R"))

# Set random seed for reproducibility
set_random_seed(42)
```

### Laod R Packages

```{r}
#| echo: true
#| warning: false
#| message: false

library(expm)
library(igraph)
library(torch)
library(ggplot2)
library(dplyr)
library(tidyr)

# Source the NOTEARS implementation files
# source("notears.R")
# source("drawing.R")

# Set device: use GPU if available, else fallback to CPU
if (cuda_is_available()) {
  device <- torch_device("cuda")
} else {
  device <- torch_device("cpu")
}
cat("Using device:", as.character(device), "\n")
```

## Part I: Synthetic Data — Linear NOTEARS

### Data and Data Processing

We use synthetic data from the official utilities (ground-truth DAG known for validation).

```{r}
#| echo: true
#| warning: false

#| echo: true
#| warning: false
# ── DATASET 1: Linear Gaussian SEM  (d = 20) ─────────────────────────────────
set_random_seed(42)
n_lin <- 2000; d_lin <- 20; s0_lin <- 20
graph_type <- "ER"; sem_type_lin <- "gauss"
B_true_lin <- simulate_dag(d_lin, s0_lin, graph_type)
W_true_lin <- simulate_parameter(B_true_lin)
X_lin      <- simulate_linear_sem(W_true_lin, n_lin, sem_type_lin)
# Save for reproducibility
write.csv(X_lin,      "X_linear.csv",      row.names = FALSE)
write.csv(W_true_lin, "W_true_linear.csv", row.names = FALSE)
# Center (NOTEARS is scale-sensitive)
X_lin <- scale(X_lin, center = TRUE, scale = FALSE)
# 70 / 15 / 15 split
set.seed(42)
tr1  <- sample(1:n_lin, floor(0.70 * n_lin))
tmp1 <- setdiff(1:n_lin, tr1)
va1  <- sample(tmp1, floor(0.50 * length(tmp1)))
te1  <- setdiff(tmp1, va1)
X_lin_train <- X_lin[tr1, ]; X_lin_val <- X_lin[va1, ]; X_lin_test <- X_lin[te1, ]
cat("Linear SEM    — Train:", dim(X_lin_train),
    "| Val:", dim(X_lin_val), "| Test:", dim(X_lin_test), "\n")
```

### Training NOTEARS linear

```{r}
#| echo: true
#| warning: false
#| message: false
W_est_lin <- notears_linear(
  X           = X_lin_train,
  lambda1     = 0.1,
  loss_type   = "l2",
  max_iter    = 100,
  h_tol       = 1e-8,
  rho_max     = 1e+16,
  w_threshold = 0.3
)
cat("Acyclic (linear SEM):", is_dag(W_est_lin), "\n")
# Validation reconstruction loss: (1/(2n)) ||X_val - X_val W||_F^2
n_val_lin    <- nrow(X_lin_val)
val_loss_lin <- 0.5 / n_val_lin * sum((X_lin_val - X_lin_val %*% W_est_lin)^2)
cat(sprintf("Validation L2 loss (linear model | linear SEM): %.6f\n", val_loss_lin))
B_est_lin <- (abs(W_est_lin) > 0) * 1
```

**Hyperparameter tuning tips**: - $\lambda_1 = 0.01$--$0.5$ (cross-validate on SHD if you have ground truth) - $w_{\mathrm{threshold}} = 0.1$--$0.5$ - For very large $d$, use NOTEARS-MLP variant (in `notears.R`).

### CATE Prediction and Validation

NOTEARS learns **structure**, not effects directly. Once you have the DAG, you can: 1. Identify adjustment sets (backdoor criterion). 2. Estimate **Average Treatment Effect (ATE)** or **Conditional Average Treatment Effect (CATE)** using regression / IPW / DoubleML on the adjustment set.

**Synthetic CATE example** (assume variable 0 = treatment $T$, variable 19 = outcome $Y$):

```{r}
#| echo: true
#| warning: false

# ── True ATE via structural path sum: (I − W)^{-1}[Y, T] ──────────────────
I_lin           <- diag(d_lin)
true_total_lin  <- solve(I_lin - W_true_lin)
true_ATE_lin    <- true_total_lin[20, 1]   # R is 1-indexed
cat(sprintf("True ATE (T → Y, linear SEM):  %.4f\n", true_ATE_lin))

# ── Backdoor-adjusted OLS estimate ────────────────────────────────────────
confounders_idx_lin <- setdiff(1:d_lin, c(1, 20))
T_tr_lin   <- X_lin_train[, 1]
Y_tr_lin   <- X_lin_train[, 20]
C_tr_lin   <- X_lin_train[, confounders_idx_lin]

model_lin  <- lm(Y_tr_lin ~ T_tr_lin + C_tr_lin)
est_ATE_lin <- coef(model_lin)["T_tr_lin"]
cat(sprintf("Estimated ATE:  %.4f  (error: %.4f)\n",
            est_ATE_lin, abs(est_ATE_lin - true_ATE_lin)))

# ── Heterogeneous CATE via interaction model ──────────────────────────────
X_int_lin       <- model.matrix(~ T_tr_lin * C_tr_lin - 1)
model_cate_lin  <- lm(Y_tr_lin ~ X_int_lin - 1)
cat("CATE interaction model fitted. Use predict(model_cate_lin, newdata) on new covariates.\n")

```

**Validation metrics** (with ground truth):

They are standard **evaluation metrics** used when you have a **ground-truth DAG** (B_true: binary adjacency matrix) and a **predicted/estimated DAG** (B_est: binary adjacency matrix learned by the algorithm), and we want to measure how well the predicted graph matches the true causal structure.

-   **fdr** → **False Discovery Rate** Fraction of predicted edges that are **incorrect** (false positives among all positives predicted by the model). Formula: FDR = FP / (TP + FP) → Lower is better (ideally close to 0). High FDR means the model is over-predicting many wrong causal edges (too many spurious links).

-   **tpr** → **True Positive Rate** (also called **Recall** or **Sensitivity** in this setting) Fraction of **true edges** that the model correctly recovered. Formula: TPR = TP / (TP + FN) → Higher is better (ideally close to 1). High TPR means the model finds most of the real causal relationships.

-   **fpr** → **False Positive Rate** Fraction of **non-edges** in the true graph that were incorrectly predicted as edges. Formula: FPR = FP / (FP + TN) → Lower is better (ideally close to 0). High FPR indicates the model is adding many spurious directed edges.

-   **shd** → **Structural Hamming Distance** Total number of edge differences between the predicted graph and the true graph. It counts:

    -   Missing edges (FN)
    -   Extra edges (FP)
    -   Reversed edges (one of the most common errors in equivalence-class methods) → Lower is better (0 = perfect recovery). SHD is one of the most commonly reported overall error metrics in causal discovery papers.

-   **nnz** → **Number of Non-Zero entries** (i.e., number of predicted edges) Simply the count of edges in the estimated graph $B_{\mathrm{est}}$ (sum of the binary matrix). → Used to understand the sparsity of the learned graph. You compare it to the true number of edges in $B_{\mathrm{true}}$. If nnz is much higher than the true edge count → likely high FDR/FPR. If much lower → likely low TPR (missing many true edges).

These metrics treat the problem as **directed edge prediction** (binary classification over all possible directed pairs, excluding self-loops). They are especially useful on synthetic benchmarks where you know the ground truth DAG.

If you're seeing these in your notebook run, they tell you exactly how accurate your learned causal graph is compared to the simulation's true structure. Lower SHD + high TPR + low FDR is the sweet spot most papers aim for.

```{r}
#| echo: true
#| warning: false

acc_lin_result <- count_accuracy(B_true_lin, B_est_lin)
print(acc_lin_result)   # list(fdr = , tpr = , fpr = , shd = , nnz = )

```

SHD (Structural Hamming Distance) $< 5$--$10$ is excellent on this scale.

### Visualization

The **Learned Weighted Adjacency Matrix (NOTEARS)** is the main output of the NOTEARS algorithm: a **$d \times d$ matrix $W$** that NOTEARS fits from data and that encodes the estimated causal graph and edge strengths.

It represents

-   **Rows and columns** both index the same set of **d variables** (e.g. the 11 Sachs proteins).
-   **Entry W\[i, j\]** is the **estimated strength of the directed edge j → i** (from variable j to variable i).
    -   So column j corresponds to the "cause" and row i to the "effect."
-   **Zeros** mean "no direct edge" in the estimated DAG; **non‑zeros** are the learned weights (can be positive or negative).

So the matrix is the **weighted adjacency matrix** of the estimated DAG: which edges exist and how strong they are.

NOTEARS **learns** this matrix by:

1.  **Fitting** it so that the linear SEM $\mathbf{X} = \mathbf{X}\mathbf{W} + \mathbf{Z}$ (or a nonlinear variant) fits the data well (e.g. squared reconstruction error).
2.  **Enforcing** that the graph of $W$ is a **DAG** (no cycles), using the NOTEARS acyclicity constraint.
3.  **Encouraging sparsity** (many zeros) via an L1 penalty so that only a small set of edges is kept.

So "Learned Weighted Adjacency Matrix (NOTEARS)" = the **estimated DAG in matrix form**, with entries giving the estimated direct effect of one variable on another (edge presence and strength).

The **Learned Causal DAG** is the directed acyclic graph (DAG) that NOTEARS infers from the data. It's the same structure as the learned weighted adjacency matrix, but viewed as a graph instead of a matrix.

```{r}
#| echo: true
#| warning: false
#| fig-width: 8
#| fig-height: 6

library(reshape2)

# Weighted adjacency matrix heatmap
W_melt_lin <- melt(W_est_lin)
colnames(W_melt_lin) <- c("Effect", "Cause", "Weight")

ggplot(W_melt_lin, aes(x = Cause, y = Effect, fill = Weight)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Learned Weighted Adjacency Matrix — Linear SEM (NOTEARS Linear)",
       x = "Cause (source)", y = "Effect (destination)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

```{r}
#| echo: true
#| warning: false
#| fig-width: 8
#| fig-height: 6

G_lin <- graph_from_adjacency_matrix(W_est_lin != 0, mode = "directed", weighted = TRUE)
V(G_lin)$name <- 1:d_lin
pos_lin <- spring_layout(G_lin, seed = 42)

plot(G_lin,
     layout             = pos_lin,
     vertex.color       = "lightblue",
     vertex.label.color = "black",
     edge.color         = "gray40",
     edge.arrow.size    = 0.5,
     main               = "Learned Causal DAG — Linear Gaussian SEM (NOTEARS Linear)",
     vertex.size        = 15,
     vertex.label.cex   = 0.8)

```

## Part 2: NOTEARS Nonlinear (MLP and Sobolev) on Sachs

We now fit three variants on the nonlinear SEM and compare against the known ground-truth DAG B_true_nlin

### Data and Data Prpcessing

```{r}

#| echo: true
#| warning: false

# ── Nonlinear SEM simulator (re-synthesized for nonlinear NOTEARS models) ───
# Convention: W[i, j] = weight of directed edge j → i
#             (row = destination / effect, column = source / cause)
#
# sem_type:
#   "mlp"  – tanh(z)      smooth, bounded nonlinearity
#   "mim"  – z * sin(z)   multiplicative interaction (non-monotone)
#   "gp"   – sin(z)       periodic / Gaussian-process-like
#
simulate_nonlinear_sem_custom <- function(W, n, sem_type = "mlp", noise_sd = 0.5) {
  d <- ncol(W)

  # igraph: A[a, b] != 0  =>  edge a -> b
  # Since W[i, j] != 0  <=>  j -> i, we need A = t(W)
  g   <- igraph::graph_from_adjacency_matrix((t(W) != 0) * 1L, mode = "directed")
  ord <- as.integer(igraph::topo_sort(g))   # parents processed before children

  X <- matrix(0.0, nrow = n, ncol = d)

  for (i in ord) {
    pa <- which(W[i, ] != 0)               # parents of node i

    if (length(pa) == 0L) {
      X[, i] <- rnorm(n)                   # root node: Gaussian noise only
    } else {
      z <- as.vector(X[, pa, drop = FALSE] %*% W[i, pa])   # weighted parent sum
      X[, i] <- switch(sem_type,
        mlp = tanh(z)   + rnorm(n, sd = noise_sd),
        mim = z * sin(z)+ rnorm(n, sd = noise_sd),
        gp  = sin(z)    + rnorm(n, sd = noise_sd),
                 z      + rnorm(n)          # fallback: linear
      )
    }
  }
  X
}



# ── DATASET 2: Nonlinear SEM, re-synthesized  (d = 10) ───────────────────────
# d = 10 keeps NotearsMLP / Sobolev runtimes tractable in a notebook
set_random_seed(42)
n_nlin <- 2000; d_nlin <- 10; s0_nlin <- 10
B_true_nlin <- simulate_dag(d_nlin, s0_nlin, graph_type)
W_true_nlin <- simulate_parameter(B_true_nlin)
X_nlin      <- simulate_nonlinear_sem_custom(W_true_nlin, n_nlin, sem_type = "mlp")
write.csv(X_nlin,      "X_nonlinear.csv",      row.names = FALSE)
write.csv(W_true_nlin, "W_true_nonlinear.csv", row.names = FALSE)
X_nlin <- scale(X_nlin, center = TRUE, scale = FALSE)
set.seed(42)
tr2  <- sample(1:n_nlin, floor(0.70 * n_nlin))
tmp2 <- setdiff(1:n_nlin, tr2)
va2  <- sample(tmp2, floor(0.50 * length(tmp2)))
te2  <- setdiff(tmp2, va2)
X_nlin_train <- X_nlin[tr2, ]; X_nlin_val <- X_nlin[va2, ]; X_nlin_test <- X_nlin[te2, ]
cat("Nonlinear SEM — Train:", dim(X_nlin_train),
    "| Val:", dim(X_nlin_val), "| Test:", dim(X_nlin_test), "\n")
```

### Linear NOTEARS — Misspecified Baseline

Fitting a linear model on nonlinear data shows the cost of the linearity assumption.

```{r}
#| echo: true
#| warning: false
#| message: false

W_lin_nlin <- notears_linear(
  X           = X_nlin_train,
  lambda1     = 0.1,
  loss_type   = "l2",
  max_iter    = 100,
  h_tol       = 1e-8,
  rho_max     = 1e+16,
  w_threshold = 0.3
)

cat("Linear NOTEARS on nonlinear SEM — acyclic:", is_dag(W_lin_nlin), "\n")

n_val_nlin      <- nrow(X_nlin_val)
val_loss_lin_nl <- 0.5 / n_val_nlin *
                   sum((X_nlin_val - X_nlin_val %*% W_lin_nlin)^2)
cat(sprintf("Validation L2 loss (linear model | nonlinear SEM): %.6f\n", val_loss_lin_nl))

B_lin_nlin <- (abs(W_lin_nlin) > 0) * 1

```

### NotearsMLP on Nonlinear Structural Equation Model (SEM)

```{r}
#| echo: true
#| warning: false
#| message: false

torch_set_default_dtype(torch_double())
set_random_seed(42)

model_mlp_nlin <- NotearsMLP(d = d_nlin, hidden = 10, bias = TRUE)

W_mlp_nlin <- notears_nonlinear(
  model_mlp_nlin, X_nlin_train,
  lambda1     = 0.01,
  lambda2     = 0.01,
  max_iter    = 150L,
  lbfgs_iter  = 0L,
  w_threshold = 0.3,
  lr          = 1e-2,
  verbose     = FALSE
)

cat("NotearsMLP on nonlinear SEM — acyclic:", is_dag(W_mlp_nlin), "\n")
B_mlp_nlin <- (abs(W_mlp_nlin) > 1e-5) * 1

# Training loss curve
losses_mlp <- attr(W_mlp_nlin, "train_losses")
if (!is.null(losses_mlp) && length(losses_mlp) > 0) {
  ggplot(data.frame(Iteration = seq_along(losses_mlp), Loss = losses_mlp),
         aes(x = Iteration, y = Loss)) +
    geom_line(color = "steelblue", linewidth = 0.8) +
    labs(title = "NotearsMLP — Training Loss (Nonlinear SEM)",
         x = "Iteration", y = "Loss") +
    theme_minimal()
}

```

#### NotearsSobolev on Nonlinear SEM

```{r}
#| echo: true
#| warning: false
#| message: false

set_random_seed(42)

# k = 4 basis functions per variable (sine-based Sobolev expansion)
model_sob_nlin <- NotearsSobolev(d = d_nlin, k = 4L)

W_sob_nlin <- notears_nonlinear(
  model_sob_nlin, X_nlin_train,
  lambda1     = 0.01,
  lambda2     = 0.01,
  max_iter    = 150L,
  lbfgs_iter  = 0L,
  w_threshold = 0.3,
  lr          = 1e-2,
  verbose     = FALSE
)

cat("NotearsSobolev on nonlinear SEM — acyclic:", is_dag(W_sob_nlin), "\n")
B_sob_nlin <- (abs(W_sob_nlin) > 1e-5) * 1

# Training loss curve
losses_sob <- attr(W_sob_nlin, "train_losses")
if (!is.null(losses_sob) && length(losses_sob) > 0) {
  ggplot(data.frame(Iteration = seq_along(losses_sob), Loss = losses_sob),
         aes(x = Iteration, y = Loss)) +
    geom_line(color = "darkorange", linewidth = 0.8) +
    labs(title = "NotearsSobolev — Training Loss (Nonlinear SEM)",
         x = "Iteration", y = "Loss") +
    theme_minimal()
}

```

### Model Comparison

Performance of all models on the nonlinear SEM (d=100, ground truth B_true_nlin known). The linear model on the linear SEM is also shown for reference

```{r}
#| echo: true
#| warning: false

# Accuracy helper — returns NULL with a message if graph is not a DAG
safe_acc <- function(B_true, W_est, B_est, label) {
  if (!is_dag(W_est)) {
    cat(sprintf("  %-30s  NOT a DAG — skipping metrics\n", label))
    return(NULL)
  }
  count_accuracy(B_true, B_est)
}

acc_lin_on_nlin  <- safe_acc(B_true_nlin, W_lin_nlin,  B_lin_nlin,  "Linear NOTEARS (nonlin SEM)")
acc_mlp_on_nlin  <- safe_acc(B_true_nlin, W_mlp_nlin,  B_mlp_nlin,  "NotearsMLP")
acc_sob_on_nlin  <- safe_acc(B_true_nlin, W_sob_nlin,  B_sob_nlin,  "NotearsSobolev")

print_row <- function(label, acc) {
  if (is.null(acc)) {
    cat(sprintf("  %-34s %7s %7s %7s %5s %5s\n",
                label, "--", "--", "--", "--", "--"))
  } else {
    cat(sprintf("  %-34s %7.4f %7.4f %7.4f %5d %5d\n",
                label, acc$fdr, acc$tpr, acc$fpr, acc$shd, acc$nnz))
  }
}

div <- paste(rep("-", 72), collapse = "")
cat("\n", div, "\n")
cat(sprintf("  %-34s %7s %7s %7s %5s %5s\n", "Model", "FDR", "TPR", "FPR", "SHD", "nnz"))
cat(div, "\n")

# Reference: linear model on its native (linear) SEM
cat(sprintf("  %-34s %7.4f %7.4f %7.4f %5d %5d   (d=20, linear SEM)\n",
            "Linear NOTEARS — native SEM *",
            acc_lin_result$fdr, acc_lin_result$tpr, acc_lin_result$fpr,
            acc_lin_result$shd, acc_lin_result$nnz))
cat(div, "\n")

# All three models on nonlinear SEM
print_row("Linear NOTEARS  (misspecified baseline)", acc_lin_on_nlin)
print_row("NotearsMLP      (hidden = 10)",           acc_mlp_on_nlin)
print_row("NotearsSobolev  (k = 4)",                 acc_sob_on_nlin)
cat(div, "\n")
cat(sprintf("  True edges in nonlinear SEM graph: %d  |  d = %d\n",
            sum(B_true_nlin), d_nlin))
cat("  * different d and SEM — shown for qualitative reference only\n")
cat("  Lower FDR / FPR / SHD and higher TPR = better\n\n")

```

### ATE Estimation — Nonlinear SEM

We select the best-performing DAG from the nonlinear comparison (prefer MLP if valid), use it to identify the adjustment set (parents of $Y$, i.e. $\mathrm{PA}(Y) \setminus \{T\}$), then estimate the ATE of variable 1 (treatment $T$) on variable 10 (outcome $Y$). A 500-replicate bootstrap gives the 95% CI.

```{r}
#| echo: true
#| warning: false

T_idx_nlin <- 1L
Y_idx_nlin <- d_nlin   # last variable as outcome

# Choose best available DAG for the adjustment set
if (!is.null(acc_mlp_on_nlin)) {
  W_best_nlin  <- W_mlp_nlin;  label_best <- "NotearsMLP"
} else if (!is.null(acc_sob_on_nlin)) {
  W_best_nlin  <- W_sob_nlin;  label_best <- "NotearsSobolev"
} else {
  W_best_nlin  <- W_lin_nlin;  label_best <- "Linear NOTEARS (fallback)"
}
cat(sprintf("Adjustment set derived from: %s\n", label_best))

# Adjustment set: parents of Y in the learned graph, PA(Y) \ {T}
pa_Y_nlin  <- which(abs(W_best_nlin[Y_idx_nlin, ]) > 1e-5)
conf_nlin  <- setdiff(pa_Y_nlin, T_idx_nlin)
if (length(conf_nlin) == 0L)
  conf_nlin <- setdiff(1:d_nlin, c(T_idx_nlin, Y_idx_nlin))  # fallback: all others

T_n <- X_nlin_train[, T_idx_nlin]
Y_n <- X_nlin_train[, Y_idx_nlin]
C_n <- X_nlin_train[, conf_nlin, drop = FALSE]

train_nlin_df  <- data.frame(Y = Y_n, T = T_n, C_n)
rhs_nlin       <- paste(colnames(C_n), collapse = " + ")
fmla_nlin      <- if (!nzchar(rhs_nlin)) "Y ~ T" else paste("Y ~ T +", rhs_nlin)
model_ate_nlin <- lm(as.formula(fmla_nlin), data = train_nlin_df)
ate_nlin       <- coef(model_ate_nlin)["T"]

# Unadjusted comparison (naive regression, no confounders)
ate_unadj_nlin <- coef(lm(Y_n ~ T_n))["T_n"]

# Bootstrap 95% CI
set.seed(42); n_boot <- 500
ate_boot_nlin <- replicate(n_boot, {
  idx <- sample(nrow(X_nlin_train), replace = TRUE)
  Tb  <- X_nlin_train[idx, T_idx_nlin]
  Yb  <- X_nlin_train[idx, Y_idx_nlin]
  Cb  <- X_nlin_train[idx, conf_nlin, drop = FALSE]
  bdf <- data.frame(Y = Yb, T = Tb, Cb)
  rhs_boot <- paste(colnames(Cb), collapse = " + ")
  fmla_boot <- if (!nzchar(rhs_boot)) "Y ~ T" else paste("Y ~ T +", rhs_boot)
  coef(lm(as.formula(fmla_boot), data = bdf))["T"]
})
ci_lo_nlin <- quantile(ate_boot_nlin, 0.025)
ci_hi_nlin <- quantile(ate_boot_nlin, 0.975)

# Structural ATE from learned W (linear approximation; exact only for linear SEMs)
I_nlin <- diag(d_nlin)
struct_ate_nlin <- tryCatch(
  solve(I_nlin - W_best_nlin)[Y_idx_nlin, T_idx_nlin],
  error = function(e) NA_real_
)

cat(sprintf("\nATE  (T=%d → Y=%d)  on Nonlinear SEM\n", T_idx_nlin, Y_idx_nlin))
cat(sprintf("  Adjusted estimate   : %+.4f\n",            ate_nlin))
cat(sprintf("  Unadjusted estimate : %+.4f\n",            ate_unadj_nlin))
cat(sprintf("  Confounding gap     : %+.4f\n",            ate_nlin - ate_unadj_nlin))
cat(sprintf("  Bootstrap 95%% CI   : [%.4f, %.4f]\n",    ci_lo_nlin, ci_hi_nlin))
cat(sprintf("  Structural ATE (W)  : %+.4f  (linear approx.)\n", struct_ate_nlin))

cat("\n--- Validation ---\n")
if (ci_lo_nlin > 0 || ci_hi_nlin < 0) {
  cat("CI excludes 0  --> significant effect at 95%.\n")
} else {
  cat("CI includes 0  --> no strong evidence of effect.\n")
}
if (abs(ate_nlin - ate_unadj_nlin) > 0.05 * abs(ci_hi_nlin - ci_lo_nlin)) {
  cat("Confounders matter: adjusted and unadjusted estimates differ substantially.\n")
} else {
  cat("Limited confounding: adjusted ≈ unadjusted.\n")
}

```

### Visualization — Nonlinear SEM Models

Side-by-side heatmaps — one panel per model:

```{r}
#| echo: true
#| warning: false
#| fig-width: 14
#| fig-height: 4

library(gridExtra)

make_heatmap <- function(W, title_str, lim = 1) {
  df <- reshape2::melt(W)
  colnames(df) <- c("Effect", "Cause", "Weight")
  ggplot(df, aes(x = Cause, y = Effect, fill = Weight)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0, limits = c(-lim, lim)) +
    labs(title = title_str, x = "Cause", y = "Effect") +
    theme_minimal() +
    theme(axis.text.x  = element_text(angle = 90, hjust = 1, size = 7),
          axis.text.y  = element_text(size = 7),
          plot.title   = element_text(size = 9, face = "bold"))
}

p1 <- make_heatmap(W_lin_nlin,  "Linear NOTEARS\n(nonlinear SEM baseline)")
p2 <- make_heatmap(W_mlp_nlin,  "NotearsMLP\n(nonlinear SEM)")
p3 <- make_heatmap(W_sob_nlin,  "NotearsSobolev  k = 4\n(nonlinear SEM)")

grid.arrange(p1, p2, p3, ncol = 3)

```

### Directed Acyclic Graph (DAG)

```{r}
#| echo: true
#| warning: false
#| fig-width: 16
#| fig-height: 5

par(mfrow = c(1, 4), mar = c(1, 1, 3, 1))

common_args <- list(
  vertex.label.color = "black",
  edge.arrow.size    = 0.5,
  vertex.size        = 22,
  vertex.label.cex   = 0.85
)

# Ground truth
G_gt <- graph_from_adjacency_matrix(B_true_nlin, mode = "directed")
V(G_gt)$name <- 1:d_nlin
do.call(plot, c(list(G_gt, main = "Ground Truth\n(Nonlinear SEM)",
                     vertex.color = "lightgreen", edge.color = "forestgreen"),
               common_args))

# Linear NOTEARS (misspecified)
G_ln <- graph_from_adjacency_matrix(B_lin_nlin, mode = "directed")
V(G_ln)$name <- 1:d_nlin
do.call(plot, c(list(G_ln, main = "Linear NOTEARS\n(misspecified baseline)",
                     vertex.color = "lightblue", edge.color = "gray40"),
               common_args))

# NotearsMLP
B_plot_mlp <- if (!is.null(acc_mlp_on_nlin)) B_mlp_nlin else matrix(0L, d_nlin, d_nlin)
G_mlp <- graph_from_adjacency_matrix(B_plot_mlp, mode = "directed")
V(G_mlp)$name <- 1:d_nlin
do.call(plot, c(list(G_mlp, main = "NotearsMLP",
                     vertex.color = "lightsalmon", edge.color = "tomato3"),
               common_args))

# NotearsSobolev
B_plot_sob <- if (!is.null(acc_sob_on_nlin)) B_sob_nlin else matrix(0L, d_nlin, d_nlin)
G_sob <- graph_from_adjacency_matrix(B_plot_sob, mode = "directed")
V(G_sob)$name <- 1:d_nlin
do.call(plot, c(list(G_sob, main = "NotearsSobolev  (k = 4)",
                     vertex.color = "plum", edge.color = "purple4"),
               common_args))

par(mfrow = c(1, 1))

```

### Interpretation notes for Part I-B:

Linear NOTEARS on nonlinear data is a deliberately misspecified baseline — it typically shows higher SHD and lower TPR than the native-linear setting, revealing how the linearity assumption costs recovery accuracy.

-   NotearsMLP tends to outperform linear NOTEARS on nonlinear data when $n$ is sufficient ($\geq 500$ per variable), at the cost of longer training.

-   NotearsSobolev is competitive with MLP on smooth / periodic-like nonlinearities (the `tanh` and `sin` SEM types here), because its sine-basis expansion naturally approximates such functions.

-   DAG validity (`is_dag`) must be checked before interpreting structural metrics; augmented Lagrangian may occasionally fail to enforce acyclicity on small nonlinear problems — reduce `lr` or increase `max_iter` if this happens.

## Summary and Conclusion

NOTEARS revolutionized causal discovery by turning an intractable combinatorial search into a clean continuous optimization problem. Its architecture (loss + smooth acyclicity + augmented Lagrangian) is elegant and extensible (nonlinear versions, low-rank, etc.).

**Strengths**: - Extremely simple & fast implementation. - State-of-the-art on linear Gaussian data. - Interpretable weighted edges.

**Limitations & best practices**: - Assumes linearity and no hidden confounders. - Sensitive to scaling/units of variables → always center/standardize carefully or use robust variants. - For nonlinear data, use NOTEARS-MLP (in the same repo) or DAG-GNN / GraN-DAG. - Always validate on synthetic data first; combine with constraint-based methods (PC) for robustness.

With this tutorial you can run end-to-end causal structure learning + downstream effect estimation in $<$ 100 lines of code. Start with synthetic data, move to your domain dataset, and iterate on $\lambda_1$ and $w_{\mathrm{threshold}}$.

## Resources

-   **Original Paper**: [DAGs with NO TEARS (Zheng et al., 2018)](https://arxiv.org/abs/1803.01422)
-   **Official Code**: [github.com/xunzheng/notears](https://github.com/xunzheng/notears) (the 60-line implementation)
-   **Extensions**: gCastle (NOTEARS + many variants), CausalNex (NOTEARS + Bayesian networks)
-   **Further Reading**:
    -   "Causal Discovery in Python" Medium series (2023–2024)
    -   GraN-DAG & DAG-GNN papers (for nonlinear)
    -   Benchmark: Sachs dataset, DREAM challenge networks
-   **Advanced Libraries**: `gcastle`, `causal-learn` (for PC/GES), `DoWhy` + `EconML` (for effect estimation on learned graphs)
-   **R Packages**: `expm`, `igraph`, `torch`, `ggplot2`, `causaldata`

## Scientific terminology for beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| NOTEARS | Method turning DAG search into smooth continuous optimization. | Learn weighted adjacency matrix by gradient descent. |
| Adjacency matrix | Matrix encoding graph edges and strengths. | `W[i,j] != 0` means variable `i` is a parent of `j`. |
| Acyclicity function | Differentiable penalty that is zero only for DAGs. | `h(W)=tr(exp(W∘W))-d` in NOTEARS. |
| Sparsity penalty | Encourages fewer edges for interpretable graphs. | L1 regularization sets small edge weights to zero. |
| Thresholding | Converting weighted graph to binary edges using cutoff. | Keep edges with absolute weight above 0.3. |
| SHD (Structural Hamming Distance) | Number of edge additions/deletions/reversals vs true graph. | Lower SHD means learned graph is closer to ground truth. |
