# Graph Neural Network (GNN) Causal Models {.unnumbered}

**Graph Neural Networks (GNNs)** treat multivariate time-series variables as nodes and causal relationships as directed edges in a graph $\mathcal{G} = (\mathcal{V}, \mathcal{E})$. Message passing makes causal structure explicit and inspectable:

$$\mathbf{h}_i^{(\ell+1)} = \mathrm{UPDATE}\!\left(\mathbf{h}_i^{(\ell)},\ \mathrm{AGGREGATE}\left(\{\mathbf{h}_j^{(\ell)} : j \in \mathcal{N}(i)\}\right)\right)$$

Unlike RNNs or Transformers where causal effects hide inside hidden states, GNN causal models make the graph a first-class object that can be regularised, inspected, and interpreted.

## The Three Models

### 1) GVAR — Graph Vector Autoregression

GVAR extends classical VAR with lag-specific learned adjacency matrices and nonlinear GNN transforms:

$$X_i^{(t)} = \sum_{k=1}^{p} \sum_{j=1}^{d} A_{ij}^{(k)} \, \phi\!\left(X_j^{(t-k)}\right) + \epsilon_i^{(t)}$$

- $A^{(k)} \in \mathbb{R}^{d \times d}$: lag-$k$ soft adjacency (sigmoid-parameterised, no self-loops)
- $\phi(\cdot)$: nonlinear per-lag GNN feature map (Linear + ReLU)
- Two stacked message-passing layers aggregate cross-variable information
- Sparsity (L1) + NOTEARS acyclicity penalties encourage parsimonious DAGs:

$$h(A) = \operatorname{tr}(e^{A \odot A}) - d = 0 \iff A \text{ is a DAG}$$

### 2) CausalGNN / CD-GNN

Joint learning: a bilinear graph learner infers a soft DAG from temporal summaries; stacked edge-conditioned message passing (EdgeConv + GRU update) propagates causal information for forecasting:

$$\mathbf{m}_{j \to i}^{(t)} = \mathrm{MSG}\!\left(\mathbf{h}_j^{(t)}, \mathbf{h}_i^{(t)}, e_{ij}\right), \quad \mathbf{h}_i^{(t+1)} = \mathrm{UPDATE}\!\left(\mathbf{h}_i^{(t)}, \sum_{j \in \mathrm{Pa}(i)} \mathbf{m}_{j \to i}^{(t)}\right)$$

### 3) CUTS+ Inspired — Causal Discovery Under Missing Time Series

Designed for datasets with missing values. Learns probabilistic Bernoulli graph edges $q(G) \approx \prod_{ij} \mathrm{Bernoulli}(\pi_{ij})$ and jointly imputes missing values before forecasting. The composite ELBO-style loss is:

$$\mathcal{L} = \underbrace{\mathbb{E}[\log p(X_{\mathrm{obs}} \mid G, X_{\mathrm{miss}})]}_{\text{reconstruction}} + \underbrace{\mathbb{E}[\log p(X_{\mathrm{miss}} \mid G)]}_{\text{imputation}} - \underbrace{\mathrm{KL}[q(G) \| p(G)]}_{\text{graph prior}}$$

## Implementation in R

We use `{RCausalML}`'s `gnn_causal_model()` function (and convenience wrappers `gvar_model()`, `causal_gnn_model()`, `cuts_model()`) to fit all three architectures on S&P 500 sector ETF daily log-return data (2018–2024).

### Load and Check Required Libraries

```{r}
#| label: packages-list
#| warning: false
packages <- c(
  'tidyverse',
  'plyr',
  'RCausalML',
  'torch',
  'tidyquant',
  'reshape2',
  'scales',
  'patchwork',
  'igraph',
  'zoo'
)
```

### Install Missing Packages

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# Install missing packages
#new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
#if(length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: verify-installation
#| warning: false
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: load-required-libraries
#| warning: false
# Find package root (DESCRIPTION file) and load dev version of RCausalML
.pkg_root <- local({
  paths <- c(".", "../..", "../../..", "../../../..")
  found <- Filter(function(p) file.exists(file.path(p, "DESCRIPTION")), paths)
  if (length(found)) normalizePath(found[1]) else NULL
})
if (!is.null(.pkg_root) && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(.pkg_root, quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

```{r}
#| label: setup
#| include: true
set.seed(42)
device_use <- NULL
if (requireNamespace("torch", quietly = TRUE)) {
  device_use <- if (torch::cuda_is_available()) "cuda" else "cpu"
  torch::torch_manual_seed(42)
}
cat(sprintf("Using device: %s\n", device_use))
```

## Data Loading and Preprocessing

This section downloads S&P 500 sector ETF prices, converts them to standardized log-returns, builds lagged input/output arrays, and injects synthetic missingness (15 %) for the CUTS+ demonstration.

> **Note:** If online market data is unavailable, the notebook automatically falls back to synthetic demo data so all downstream cells remain runnable.

```{r}
#| label: data-download
#| warning: false

TICKERS <- c(
  "XLF" = "Financials",
  "XLE" = "Energy",
  "XLK" = "Technology",
  "XLV" = "Healthcare",
  "XLI" = "Industrials",
  "XLY" = "ConsumerDisc",
  "XLP" = "ConsumerStap",
  "XLU" = "Utilities"
)

fetch_close_prices <- function(tickers, start = "2018-01-01", end = "2024-01-01") {
  tryCatch({
    raw <- tidyquant::tq_get(names(tickers), get = "stock.prices",
                              from = start, to = end, complete_cases = TRUE)
    if (is.null(raw) || nrow(raw) == 0) return(NULL)
    raw |>
      dplyr::select(symbol, date, adjusted) |>
      tidyr::pivot_wider(names_from = symbol, values_from = adjusted) |>
      dplyr::arrange(date) |>
      tidyr::drop_na()
  }, error = function(e) NULL)
}

close_wide <- fetch_close_prices(TICKERS)

if (is.null(close_wide) || nrow(close_wide) < 30) {
  message("Warning: market data unavailable; using synthetic demo data.")
  set.seed(42)
  n_steps <- 1600L
  d_syn   <- length(TICKERS)
  A_syn   <- matrix(0, d_syn, d_syn)
  for (i in seq_len(d_syn)) {
    A_syn[i, i] <- 0.45
    if (i > 1)                 A_syn[i, i - 1] <- 0.25
    if (i > 2 && i %% 2 == 1) A_syn[i, i - 2] <- 0.15
  }
  x_syn      <- matrix(0, n_steps, d_syn)
  x_syn[1, ] <- rnorm(d_syn)
  noise_syn  <- 0.15 * matrix(rnorm(n_steps * d_syn), n_steps, d_syn)
  for (t in seq(2L, n_steps)) x_syn[t, ] <- x_syn[t - 1L, ] %*% t(A_syn) + noise_syn[t, ]
  returns_df <- as.data.frame(x_syn)
  colnames(returns_df) <- unname(TICKERS)
  data_np    <- scale(x_syn)
} else {
  price_mat  <- as.matrix(close_wide[, names(TICKERS)])
  log_ret    <- log(price_mat[-1L, ] / price_mat[-nrow(price_mat), ])
  colnames(log_ret) <- unname(TICKERS[colnames(price_mat)])
  returns_df <- as.data.frame(log_ret)
  data_np    <- scale(log_ret)
}

VAR_NAMES <- colnames(data_np)
d         <- ncol(data_np)
Tt        <- nrow(data_np)

cat(sprintf("d=%d variables, T=%d\n", d, Tt))
cat("Variables:", paste(VAR_NAMES, collapse = ", "), "\n")
```

```{r}
#| label: build-lag-dataset
#| warning: false

LAG   <- 10L
AHEAD <- 1L

build_lag_dataset <- function(data, lag, ahead = 1L) {
  N     <- nrow(data) - lag - ahead + 1L
  x_seq <- array(NA_real_, dim = c(N, lag, ncol(data)))
  y_mat <- matrix(NA_real_, nrow = N, ncol = ncol(data))
  for (t in seq_len(N)) {
    x_seq[t, , ] <- data[t:(t + lag - 1L), , drop = FALSE]
    y_mat[t,  ]  <- data[t + lag + ahead - 1L, ]
  }
  list(x_seq = x_seq, y_mat = y_mat)
}

lag_data <- build_lag_dataset(data_np, LAG, AHEAD)
X_seq    <- lag_data$x_seq
Y_mat    <- lag_data$y_mat

split  <- floor(0.8 * nrow(X_seq))
X_tr   <- X_seq[seq_len(split), , ]
X_val  <- X_seq[(split + 1L):nrow(X_seq), , ]
Y_tr   <- Y_mat[seq_len(split), ]
Y_val  <- Y_mat[(split + 1L):nrow(Y_mat), ]

cat(sprintf("Train: (%d, %d, %d)  Val: (%d, %d, %d)\n",
            dim(X_tr)[1], dim(X_tr)[2], dim(X_tr)[3],
            dim(X_val)[1], dim(X_val)[2], dim(X_val)[3]))
```

```{r}
#| label: synthetic-missingness
#| warning: false

# Inject 15 % synthetic missingness for CUTS+ demo (same as Python notebook)
set.seed(0L)
MISS_RATE <- 0.15
miss_mask <- array(runif(prod(dim(X_seq))) < MISS_RATE, dim = dim(X_seq))
X_miss    <- X_seq
X_miss[miss_mask] <- 0.0
X_miss_tr  <- X_miss[seq_len(split), , ]
X_miss_val <- X_miss[(split + 1L):nrow(X_seq), , ]

cat(sprintf("Missing rate: %.1f%%\n", 100 * mean(miss_mask)))
cat(sprintf("Missing tensor shape: (%d, %d, %d)\n", dim(X_miss)[1], dim(X_miss)[2], dim(X_miss)[3]))
```

## Model Architectures Overview

All three architectures are implemented as R torch `nn_module` objects inside `{RCausalML}` and are accessible through the single entry-point `gnn_causal_model()`.

### GVAR Architecture

1. **GVARGraphLearner**: $p$ lag-specific $(d \times d)$ sigmoid-parameterised adjacency matrices; diagonal forced to zero (no self-loops); L1 sparsity penalty.
2. **GVARMessagePass** (×2): per-lag linear transform + ReLU → batch-matmul with $A^{(k)}$ → concatenate → aggregate linear layer.
3. **Prediction head**: input-projected features (1 → hidden) pass through two stacked GNN layers then a two-layer GELU MLP producing $d$-vector forecasts.
4. **Losses**: MSE + sparsity (L1) + DAG (NOTEARS acyclicity).

### CausalGNN Architecture

1. **Per-variable GRU encoder**: each variable's $T$-step sequence is encoded to a `hidden`-dim state.
2. **CausalGraphLearner**: bilinear edge scoring on temporal summaries → soft DAG $\hat{A} \in [0,1]^{d \times d}$.
3. **EdgeConvLayer** (×`n_gnn_layers`): concatenates source/target features + edge weight → two-layer MLP → GRU update → layer norm.
4. **Losses**: MSE + DAG acyclicity + L1 sparsity.

### CUTS+ Architecture

```
Input x_t (d vars, possibly missing)
      │
      ▼
Imputation Network (values + mask → lag*d)
      │
      ▼
Per-variable GRU Encoder
      │
      ▼
Variational Graph  q(G) = Bernoulli(π_ij)
      │
      ▼
EdgeConv Message Passing
      │
      ▼
Forecast ŷ (d variables)
```

Losses: MSE + DAG acyclicity + KL($q(G) \| \text{Bernoulli}(0.1)$)

## Fit All Three Models

`gnn_causal_model()` is the single entry point for all three architectures. Internally it:

- Builds windowed `(lag, d)` input/output tensors (with optional synthetic missingness).
- Trains each model with AdamW + cosine LR schedule + early stopping.
- Extracts the $d \times d$ causal matrix from each trained model.

```{r}
#| label: fit-all-models
#| warning: false

cat("Fitting GVAR, CausalGNN, and CUTS+-like models ...\n")

gnn_fit <- gnn_causal_model(
  data         = data_np,
  lag          = LAG,
  models       = c("gvar", "causalgnn", "cuts"),
  hidden       = 48L,
  n_gnn_layers = 3L,
  lam_sparse   = 0.003,
  lam_dag      = 0.08,
  lam_kl       = 1e-3,
  miss_rate    = MISS_RATE,
  epochs       = 60L,
  lr           = 4e-4,
  patience     = 12L,
  batch_size   = 64L,
  val_split    = 0.2,
  device       = device_use,
  verbose      = TRUE
)

cat("\nTraining complete.\n")
cat("Validation MSE:\n")
print(round(gnn_fit$val_mse, 6))
```

### Validation MSE Summary Table

```{r}
#| label: val-mse-table
#| warning: false

val_mse_df <- data.frame(
  Model   = c("GVAR", "CausalGNN", "CUTS+like"),
  Val_MSE = unname(gnn_fit$val_mse)
) |> dplyr::arrange(Val_MSE)

print(val_mse_df)
```

## Training Loss Curves

```{r}
#| label: plot-training-loss
#| warning: false

hist_list <- gnn_fit$histories

make_loss_df <- function(nm) {
  h <- hist_list[[nm]]
  if (is.null(h)) return(NULL)
  n_ep <- sum(h$val != 0)
  data.frame(
    epoch = seq_len(n_ep),
    val   = h$val[seq_len(n_ep)],
    train = h$train[seq_len(n_ep)],
    model = nm
  )
}

loss_df <- dplyr::bind_rows(lapply(names(hist_list), make_loss_df))
loss_df$model <- factor(loss_df$model,
                         levels = c("gvar", "causalgnn", "cuts"),
                         labels = c("GVAR", "CausalGNN", "CUTS+like"))

ggplot(loss_df, aes(x = epoch, y = val, colour = model)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(
    "GVAR"      = "#1b9e77",
    "CausalGNN" = "#d95f02",
    "CUTS+like" = "#7570b3"
  )) +
  labs(
    title  = "Validation Loss Curves — GNN Causal Models",
    x      = "Epoch",
    y      = "MSE",
    colour = NULL
  ) +
  theme_minimal()
```

### Training vs. Validation Loss per Model

```{r}
#| label: plot-train-val-loss
#| warning: false
#| fig-width: 12
#| fig-height: 4

plot_model_loss <- function(nm, label, colour) {
  h <- hist_list[[nm]]
  if (is.null(h)) return(NULL)
  n_ep <- sum(h$val != 0)
  df <- data.frame(
    epoch = seq_len(n_ep),
    Train = h$train[seq_len(n_ep)],
    Val   = h$val[seq_len(n_ep)]
  ) |> tidyr::pivot_longer(-epoch, names_to = "split", values_to = "loss")
  ggplot(df, aes(x = epoch, y = loss, colour = split)) +
    geom_line(linewidth = 0.85) +
    scale_colour_manual(values = c("Train" = colour, "Val" = "grey40")) +
    labs(title = label, x = "Epoch", y = "MSE", colour = NULL) +
    theme_minimal(base_size = 9)
}

p1 <- plot_model_loss("gvar",      "GVAR",      "#1b9e77")
p2 <- plot_model_loss("causalgnn", "CausalGNN", "#d95f02")
p3 <- plot_model_loss("cuts",      "CUTS+like", "#7570b3")

patchwork::wrap_plots(p1, p2, p3, ncol = 3)
```

## Causal Matrix Visualization

For each model, the learned causal matrix $C \in \mathbb{R}^{d \times d}$ encodes variable-to-variable influence:

- **GVAR**: lag-averaged soft adjacency $\sum_k A^{(k)}$, where $A^{(k)}_{ij}$ encodes the causal influence of $j$ on $i$ at lag $k$.
- **CausalGNN**: bilinear soft edge scores $\hat{A}_{ij}$ reflecting the inferred directed dependency strength.
- **CUTS+like**: Bernoulli variational parameters $\pi_{ij} = \sigma(\text{logit}_{ij})$, the posterior edge probabilities.

```{r}
#| label: extract-causal-matrices
#| warning: false

C_gvar   <- causal_matrix_gnn(gnn_fit, model = "gvar")
C_cgnn   <- causal_matrix_gnn(gnn_fit, model = "causalgnn")
C_cuts   <- causal_matrix_gnn(gnn_fit, model = "cuts")

cat("GVAR causal matrix (first 3 rows):\n")
print(round(C_gvar[1:3, ], 4))
cat("\nDiagonal (should be ~0 — no self-loops):", round(diag(C_gvar), 3), "\n")
```

### Heatmap Comparison

```{r}
#| label: heatmap-comparison
#| warning: false
#| fig-width: 15
#| fig-height: 5

plot_causal_heatmap <- function(C, title, var_names) {
  A <- as.matrix(C)
  diag(A) <- NA_real_
  rownames(A) <- var_names
  colnames(A) <- var_names
  df <- reshape2::melt(A, na.rm = TRUE)
  colnames(df) <- c("Target", "Source", "Weight")
  df$Target <- factor(df$Target, levels = rev(var_names))
  df$Source <- factor(df$Source, levels = var_names)
  ggplot(df, aes(x = Source, y = Target, fill = Weight)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = round(Weight, 2)), size = 2.5, colour = "black") +
    scale_fill_gradient(low = "white", high = "#D7191C",
                        na.value = "grey90", name = "Weight") +
    labs(title = title, x = "Source (cause)", y = "Target (effect)") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

p_gvar <- plot_causal_heatmap(C_gvar, "GVAR — Lag-Averaged Adjacency",        VAR_NAMES)
p_cgnn <- plot_causal_heatmap(C_cgnn, "CausalGNN — Bilinear Edge Scores",      VAR_NAMES)
p_cuts <- plot_causal_heatmap(C_cuts, "CUTS+like — Bernoulli Edge Posteriors",  VAR_NAMES)

patchwork::wrap_plots(p_gvar, p_cgnn, p_cuts, ncol = 3)
```

### Top Causal Edges

```{r}
#| label: top-causal-edges
#| warning: false

top_edges <- function(C, names, k = 10L) {
  n   <- length(names)
  idx <- which(row(C) != col(C), arr.ind = FALSE)
  edges <- data.frame(
    source = names[col(C)[idx]],
    target = names[row(C)[idx]],
    weight = C[idx]
  )
  edges[order(edges$weight, decreasing = TRUE), ][seq_len(min(k, nrow(edges))), ]
}

for (label_nm in list(list("GVAR", C_gvar), list("CausalGNN", C_cgnn), list("CUTS+like", C_cuts))) {
  cat(sprintf("\nTop 10 edges for %s (source -> target):\n", label_nm[[1]]))
  te <- top_edges(label_nm[[2]], VAR_NAMES, 10L)
  for (i in seq_len(nrow(te))) {
    cat(sprintf("  %-14s -> %-14s | %.3f\n", te$source[i], te$target[i], te$weight[i]))
  }
}
```

### Causal Graph Network Diagrams

```{r}
#| label: network-diagram-comparison
#| warning: false

draw_gnn_igraph <- function(C, title, var_names, threshold = 0.12,
                             node_colour = "steelblue") {
  M     <- as.matrix(C)
  M[is.na(M)] <- 0.0
  A_bin <- (M > threshold) * 1L
  diag(A_bin) <- 0L
  g <- igraph::graph_from_adjacency_matrix(
    A_bin, mode = "directed", weighted = NULL, diag = FALSE
  )
  igraph::V(g)$name        <- var_names
  igraph::V(g)$color       <- node_colour
  igraph::V(g)$label       <- var_names
  igraph::V(g)$size        <- 28
  igraph::V(g)$label.color <- "white"
  igraph::V(g)$label.cex   <- 0.8
  igraph::E(g)$color       <- "grey50"
  igraph::E(g)$arrow.size  <- 0.6

  old_par <- par(mar = c(0, 0, 2, 0))
  plot(g,
       layout = igraph::layout_in_circle(g),
       main   = title,
       vertex.frame.color = "white")
  par(old_par)
}

par(mfrow = c(1, 3))
draw_gnn_igraph(C_gvar, "GVAR (threshold > 0.12)",      VAR_NAMES, 0.12, "steelblue")
draw_gnn_igraph(C_cgnn, "CausalGNN (threshold > 0.12)", VAR_NAMES, 0.12, "darkorange")
draw_gnn_igraph(C_cuts, "CUTS+like (threshold > 0.12)", VAR_NAMES, 0.12, "mediumpurple")
par(mfrow = c(1, 1))
```

## Graph Statistics

```{r}
#| label: graph-statistics
#| warning: false

graph_stats_gnn <- function(C, threshold = 0.12, name = "") {
  M     <- as.matrix(C)
  M[is.na(M)] <- 0.0
  A_bin <- (M > threshold) * 1L
  diag(A_bin) <- 0L
  g    <- igraph::graph_from_adjacency_matrix(
    A_bin, mode = "directed", weighted = NULL, diag = FALSE
  )
  d_n  <- igraph::gorder(g)
  e_n  <- igraph::gsize(g)
  dens <- if (d_n > 1) e_n / (d_n * (d_n - 1)) else 0.0
  data.frame(
    model          = name,
    nodes          = d_n,
    edges          = e_n,
    density        = round(dens, 4),
    is_dag         = igraph::is_dag(g),
    max_in_degree  = if (d_n > 0) max(igraph::degree(g, mode = "in"))  else 0L,
    max_out_degree = if (d_n > 0) max(igraph::degree(g, mode = "out")) else 0L
  )
}

stats_all <- do.call(rbind, list(
  graph_stats_gnn(C_gvar, 0.12, "GVAR"),
  graph_stats_gnn(C_cgnn, 0.12, "CausalGNN"),
  graph_stats_gnn(C_cuts, 0.12, "CUTS+like")
))
rownames(stats_all) <- NULL

cat("Graph statistics (threshold = 0.12):\n")
print(stats_all)
```

## Model Validation: Prediction Error on Held-Out Data

```{r}
#| label: validation-mse
#| warning: false

pred_gvar <- predict(gnn_fit, model = "gvar",      newdata = X_val)
pred_cgnn <- predict(gnn_fit, model = "causalgnn", newdata = X_val)
pred_cuts <- predict(gnn_fit, model = "cuts",      newdata = X_val)

mse_gvar <- mean((pred_gvar - Y_val)^2)
mse_cgnn <- mean((pred_cgnn - Y_val)^2)
mse_cuts <- mean((pred_cuts - Y_val)^2)

val_df <- data.frame(
  model   = c("GVAR", "CausalGNN", "CUTS+like"),
  val_mse = c(mse_gvar, mse_cgnn, mse_cuts)
)

cat("Validation MSE (next-step reconstruction):\n")
print(val_df)

ggplot(val_df, aes(x = reorder(model, val_mse), y = val_mse, fill = model)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  scale_fill_manual(values = c(
    "GVAR"      = "#1b9e77",
    "CausalGNN" = "#d95f02",
    "CUTS+like" = "#7570b3"
  )) +
  coord_flip() +
  labs(
    title = "Validation MSE by Model",
    x     = NULL,
    y     = "Mean Squared Error"
  ) +
  theme_minimal()
```

### Per-Variable Prediction Error

```{r}
#| label: per-variable-mse
#| warning: false

per_var_mse <- data.frame(
  variable  = VAR_NAMES,
  GVAR      = colMeans((pred_gvar - Y_val)^2),
  CausalGNN = colMeans((pred_cgnn - Y_val)^2),
  CUTSlike  = colMeans((pred_cuts - Y_val)^2)
)

cat("Per-variable MSE:\n")
print(data.frame(
  variable = per_var_mse$variable,
  round(per_var_mse[, -1], 5)
))

pv_long <- tidyr::pivot_longer(per_var_mse, -variable,
                                names_to = "model", values_to = "mse")

ggplot(pv_long, aes(x = variable, y = mse, fill = model)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(
    "GVAR"      = "#1b9e77",
    "CausalGNN" = "#d95f02",
    "CUTSlike"  = "#7570b3"
  )) +
  labs(
    title = "Per-Variable Validation MSE",
    x     = "Sector",
    y     = "MSE",
    fill  = "Model"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## CUTS+ Missing Data Analysis

CUTS+-like's imputation network fills in zero-coded missing values before encoding. We analyse prediction quality under missingness by comparing CUTS+ (which sees a corrupted input) against GVAR (which sees the same zero-filled input without explicit imputation).

```{r}
#| label: cuts-missing-data-analysis
#| warning: false
#| fig-width: 10
#| fig-height: 4

# Predict on the missing-data validation set
pred_cuts_miss <- predict(gnn_fit, model = "cuts", newdata = X_miss_val)
pred_gvar_miss <- predict(gnn_fit, model = "gvar", newdata = X_miss_val)

mse_cuts_miss <- colMeans((pred_cuts_miss - Y_val)^2)
mse_gvar_miss <- colMeans((pred_gvar_miss - Y_val)^2)

miss_cmp <- data.frame(
  variable = rep(VAR_NAMES, 2),
  mse      = c(mse_cuts_miss, mse_gvar_miss),
  method   = rep(c("CUTS+like (imputation)", "GVAR (no imputation)"), each = d)
)

ggplot(miss_cmp, aes(x = variable, y = mse, fill = method)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(
    "CUTS+like (imputation)" = "#7570b3",
    "GVAR (no imputation)"   = "#1b9e77"
  )) +
  labs(
    title = sprintf("Per-Variable MSE on %.0f%% Missing Input: CUTS+like vs. GVAR", 100 * MISS_RATE),
    x     = "Sector",
    y     = "MSE",
    fill  = NULL
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: cuts-imputation-gain
#| warning: false

imputation_gain <- data.frame(
  variable        = VAR_NAMES,
  MSE_GVAR_miss   = round(mse_gvar_miss, 5),
  MSE_CUTS_miss   = round(mse_cuts_miss, 5),
  imputation_gain = round(mse_gvar_miss - mse_cuts_miss, 5)
)
cat("Imputation gain (positive = CUTS+ helps):\n")
print(imputation_gain)
```

## Causal Matrix Comparison Across Models

```{r}
#| label: causal-matrix-correlation
#| warning: false

# Compare off-diagonal entries across model pairs
off_diag <- function(C) {
  M <- as.matrix(C); diag(M) <- NA
  M[!is.na(M)]
}

gvar_v  <- off_diag(C_gvar)
cgnn_v  <- off_diag(C_cgnn)
cuts_v  <- off_diag(C_cuts)

corr_mat <- matrix(c(
  1,                      cor(gvar_v, cgnn_v),   cor(gvar_v, cuts_v),
  cor(cgnn_v, gvar_v),   1,                      cor(cgnn_v, cuts_v),
  cor(cuts_v, gvar_v),   cor(cuts_v, cgnn_v),   1
), nrow = 3, dimnames = list(c("GVAR","CausalGNN","CUTS+like"),
                              c("GVAR","CausalGNN","CUTS+like")))
cat("Pairwise Pearson correlation of causal matrices (off-diagonal entries):\n")
print(round(corr_mat, 4))
```

## Interpretation Notes

- **GVAR** yields explicit lag-specific adjacency matrices, making it the closest to classical Granger causality in a GNN framework. Its learned mask is directly interpretable as a sparse directed influence graph.
- **CausalGNN** jointly learns the graph topology and nonlinear dynamics using bilinear edge scoring, which can capture asymmetric dependencies that linear methods miss.
- **CUTS+like** is designed for missing-data scenarios. Its variational graph posterior produces probabilistic edge weights, and its imputation network recovers missing observations using graph-aware context.

In practice, combining these models is often powerful:

1. Use **GVAR** for a sparse, lag-structured Granger-style causal graph.
2. Use **CausalGNN** for nonlinear, jointly discovered structure.
3. Use **CUTS+like** when input data have substantial missingness or uneven sampling.

## Summary and Conclusions

In this notebook, we explored three GNN-based causal discovery architectures — **GVAR**, **CausalGNN**, and **CUTS+-like** — on S&P 500 sector ETF data and compared their learned causal graphs.

**Key Findings:**

- All three models achieve comparable validation MSE, demonstrating that explicit graph structure constraints need not hurt predictive performance.
- **GVAR** produces sparse, lag-structured adjacency matrices that resemble a multi-lag Granger-causal graph; its L1 + NOTEARS penalties promote DAG-consistent structure.
- **CausalGNN** recovers both symmetric and asymmetric dependencies via nonlinear bilinear edge scoring; the learned DAG often differs meaningfully from linear GVAR, highlighting nonlinear relationships.
- The **CUTS+-like** model's imputation network provides measurable benefit on missing-data inputs, and its Bernoulli edge posteriors offer uncertainty-aware causal graphs.

**Conclusions and Recommendations:**

- GNN causal models unify temporal representation learning and graph structure discovery, offering richer modeling power than purely linear baselines.
- Visualization of learned adjacency matrices and ranking of discovered edges help practitioners interpret results and guide domain validation.
- For robust application: apply thresholding, stability selection, and bootstrap confidence intervals before treating discovered edges as causal hypotheses.
- The NOTEARS acyclicity penalty provides a differentiable mechanism to encourage DAG structure, but should be combined with domain knowledge for final graph selection.
- Further work could include larger financial universes, formal bootstrap uncertainty on causal matrices, and integration with intervention-based validation.

## Resources

### Primary References

- **GVAR (2022):** [Learning Granger Causal Feature Representations](https://arxiv.org/abs/2206.10311) — lag-specific graph autoregression for Granger-causal discovery.
- **CausalGNN / CD-GNN (2022):** [Causal Discovery with Graph Neural Networks](https://arxiv.org/abs/2003.12769) — joint graph learning and dynamics for multivariate time series.
- **CUTS+ (2023):** [CUTS+: High-dimensional Causal Discovery from Irregular Time-Series](https://arxiv.org/abs/2305.05890) — variational causal discovery under missingness and irregular sampling.
- **NOTEARS (2018):** [DAGs with NO TEARS: Continuous Optimization for Structure Learning](https://arxiv.org/abs/1803.01422) — Zheng et al.; the acyclicity penalty used in all three models.

### Background Texts

- **Causality (Judea Pearl):** [Causality: Models, Reasoning and Inference](https://bayes.cs.ucla.edu/BOOK-2K/) — definitive reference for structural causal models.
- **Elements of Causal Inference (Peters, Janzing & Schölkopf):** [MIT Press](https://mitpress.mit.edu/9780262037310/elements-of-causal-inference/) — rigorous treatment of SCMs and identifiability.
- **The Book of Why (J. Pearl & D. Mackenzie)** — accessible introduction to causal reasoning.
- **Graph Neural Networks Review:** [Exploring Causal Learning through GNNs](https://arxiv.org/html/2311.14994) — comprehensive review of GNN approaches to causal discovery.

### Software and Datasets

- **R torch:** <https://torch.mlverse.org/> — PyTorch bindings for R used by `{RCausalML}`.
- **tidyquant:** <https://business-science.github.io/tidyquant/> — tidy finance data retrieval.
- **{RCausalML}:** `gnn_causal_model()`, `causal_matrix_gnn()`, `predict.gnn_causal_model()` — main API for this notebook.

### Companion Notebooks in This Series

- `05-04-01-DeepCausalML-timeseries-neural-granger-cusuality-r.qmd` — Neural Granger Causality (cMLP, cLSTM, SRU, NRI) on the same sector ETF universe.
- `05-04-02-DeepCausalML-timeseries-structural-causal-model-SMC-r.qmd` — Deep Structural Causal Models: DeepSCM, DECI, DynoTEARS on sector ETF data.
- `05-04-03-DeepCausalML-attension-transformer-r.qmd` — Attention-Based/Transformer Causal Models: TCDF, CausalTransformer, TFT on sector ETF data.
- `05-04-04-DeepCausalML-timeseries-RNN-LSTM-causaML-r.qmd` — RNN/LSTM Causal Models: CausalLSTM, RETAIN, Intervention-Aware RNN on sector ETF data.

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Graph $\mathcal{G}=(\mathcal{V},\mathcal{E})$ | A set of nodes $\mathcal{V}$ (variables) connected by directed edges $\mathcal{E}$ (causal links). | A company org-chart: people are nodes, reporting lines are edges. |
| Message passing | Each node aggregates information from its neighbours across edges, then updates its own state. | In a phone tree: each person relays a message from their callers before summarising to the boss. |
| Soft adjacency $A_{ij} \in [0,1]$ | A continuous weight encoding how strongly node $j$ causally influences node $i$. | A water pipe diameter: larger = more flow; zero = pipe is closed. |
| NOTEARS penalty $h(A)$ | A differentiable function that equals zero if and only if $A$ encodes a DAG; penalises cycles. | A traffic rule that charges a fine proportional to how many circular road loops exist. |
| DAG (Directed Acyclic Graph) | A graph with directed edges and no cycles — required for causal identification. | A family tree: parents come before children; no person can be their own ancestor. |
| Sparsity penalty (L1) | A regularisation term that shrinks small edge weights to zero, selecting only important edges. | Like decluttering: keeping only the items used daily and discarding the rest. |
| EdgeConv layer | Computes messages by combining source, target, and edge-weight features, then aggregates into node updates. | Each pair of friends shares a combined message; each person tallies all messages they receive. |
| Variational graph $q(G)$ | A probability distribution over graphs, learned to match the observed data. | Instead of drawing one map, you draw a distribution over possible maps with edge confidence scores. |
| KL divergence to prior | A penalty that keeps the inferred edge probabilities close to a sparse baseline ($p=0.1$). | Penalising a guess that deviates too far from the prior belief that most pairs are unrelated. |
| Imputation network | A neural network that predicts missing values from observed values and a mask before encoding. | An autocomplete system that fills in blanks in a time series based on surrounding context. |
| Lag $k$ | How many time steps in the past an input is taken from. | Predicting today's returns using data from $k$ days ago. |
| AdamW | Adaptive gradient-descent optimizer with decoupled weight decay regularisation. | A GPS navigator that adapts turn-by-turn suggestions while penalising large detours. |
| Cosine annealing | A learning-rate schedule that decays the LR following a cosine curve, slowing training as convergence nears. | Gradually applying the brakes as you approach a stop sign. |
