# CausalDiscrepancy VAE: Discrepancy-Regularized Causal Representation Learning {.unnumbered}

`CausalDiscrepancyVAE` extends a standard VAE objective with an explicit distribution-matching penalty between treatment groups in latent space. The practical goal is to reduce treatment-group imbalance in representation space while preserving predictive structure for outcomes. In this notebook, we convert the complete `causalDiscrepanct_VAE.ipynb` workflow into R and run it with `{RCausalML}`'s torch implementation: `causal_discrepancy_vae()`.

## Overview

A Variational Autoencoder learns to encode data $x$ into a latent space $z$ and reconstruct it, optimizing the Evidence Lower Bound (ELBO):

$$
\mathcal{L}_{\text{ELBO}} = \mathbb{E}_{q_\phi(z|x)}\left[\log p_\theta(x|z)\right] - D_{\text{KL}}\left(q_\phi(z|x) \,\|\, p(z)\right)
$$

The first term is the reconstruction loss - how faithfully the decoder recovers $x$. The second is a regularization term that keeps the approximate posterior $q_\phi(z|x)$ close to a prior $p(z)$ (typically a standard Gaussian).

This works well for generative modeling, but it has no notion of causality - it cannot distinguish correlation from intervention.

## The Causal Extension: Adding MMD

The CausalDiscrepancy VAE adds a third term to the objective:

$$
\mathcal{L}_{\text{CD-VAE}} = \mathcal{L}_{\text{ELBO}} - \lambda \cdot \text{MMD}\!\left(\hat{p}_{\text{int}},\, p_{\text{int}}\right)
$$

where:

- $\hat{p}_{\text{int}}$ is the model's predicted distribution under a virtual (simulated) intervention
- $p_{\text{int}}$ is the real distribution observed under an actual intervention (e.g., from a Perturb-Seq experiment)
- $\lambda$ controls how strongly this causal alignment is enforced

The model is thus penalized whenever its simulated interventions produce outputs that look different from real experimental interventions.

## What is Maximum Mean Discrepancy (MMD)?

MMD is a kernel-based distance between two probability distributions. Given samples from two distributions $P$ and $Q$, it measures how different they are by comparing their mean embeddings in a Reproducing Kernel Hilbert Space (RKHS):

$$
\text{MMD}^2(P, Q) = \left\|\, \mathbb{E}_{x \sim P}[\phi(x)] - \mathbb{E}_{y \sim Q}[\phi(y)] \,\right\|^2_{\mathcal{H}}
$$

In practice, using the kernel trick with a kernel $k$ (e.g., RBF / Gaussian), this expands to:

$$
\text{MMD}^2(P, Q) = \mathbb{E}_{x,x' \sim P}[k(x,x')] - 2\,\mathbb{E}_{x \sim P,\, y \sim Q}[k(x,y)] + \mathbb{E}_{y,y' \sim Q}[k(y,y')]
$$

Key properties that make it useful here:

| Property | Why it matters |
|---|---|
| Non-parametric | No assumption about the shape of either distribution |
| Differentiable | Can be backpropagated through during training |
| Operates on samples | Works even with finite batches of cells/observations |
| MMD = 0 iff $P = Q$ | Provides a proper distributional test |

## Virtual Interventions in the Latent Space

The "causal" part of the model relies on do-calculus intuition. A virtual intervention $\text{do}(X_i = v)$ is simulated by:

1. Encoding observed (unperturbed) cells into the latent space $z \sim q_\phi(z|x)$
2. Modifying a specific latent dimension (or a learned intervention subspace) corresponding to gene $i$ - this is the virtual perturbation
3. Decoding the modified $\tilde{z}$ back into gene expression space to get $\hat{x}_{\text{int}}$

This produces a predicted post-intervention distribution $\hat{p}_{\text{int}}$. The MMD term then checks: does this simulated perturbation actually look like real CRISPR knockout data?

```text
Unperturbed cell x
        |
        v
   Encoder q_phi(z|x)
        |
        v
   Latent code z
        |
   [modify z_i]  <- virtual intervention do(gene_i = v)
        |
        v
   Decoder p_theta(x|z_tilde)
        |
        v
   x_hat_int  -->  MMD(x_hat_int, x_int_real)  <- real CRISPR data
```

## Application to Perturb-Seq / CRISPR Data

Perturb-Seq combines CRISPR-based genetic perturbations with single-cell RNA sequencing. Each cell has:

- A known gene knockout (the intervention)
- A full transcriptome readout (the outcome)

This creates a dataset with both observational and (partial) interventional structure - exactly what CausalDiscrepancy VAE exploits.

### Why standard VAEs fail here

A regular VAE trained on Perturb-Seq data would learn to reconstruct gene expression profiles, but it conflates the statistical pattern of knockouts with genuine causal effects. It cannot answer: "If I knock out gene X in a cell where it was never knocked out, what happens?"

### What CD-VAE does differently

- The latent space is structured so that specific dimensions encode causal factors - ideally individual genes or gene programs
- The MMD penalty forces the model's predicted response to gene $X$ knockout to match the observed response, for every knockout in the training set
- Generalization then allows predicting the effect of unseen or combinatorial perturbations

## Full Training Objective (Expanded)

$$
\mathcal{L} = \underbrace{\mathbb{E}_{q_\phi(z|x)}\left[\log p_\theta(x|z)\right]}_{\text{reconstruction}} - \underbrace{D_{\text{KL}}\left(q_\phi(z|x) \,\|\, p(z)\right)}_{\text{latent regularization}} - \underbrace{\lambda \sum_k \text{MMD}\!\left(\hat{p}_{\text{int}}^{(k)},\, p_{\text{int}}^{(k)}\right)}_{\text{causal discrepancy penalty}}
$$

The sum over $k$ runs over all perturbation conditions present in the training data (e.g., each distinct gene knockout). Each term asks: for perturbation $k$, does the model's simulation match reality?

## Strengths and Limitations

### Strengths

- Leverages real interventional data as a supervisory signal, which most causal VAEs lack
- MMD is computationally tractable even in high dimensions (e.g., ~20,000 genes)
- Naturally handles distributional shift between observational and interventional regimes
- Can generalize to out-of-distribution perturbations if the latent causal structure is well-learned

### Limitations

- Requires some real interventional data - it's not purely observational causal discovery
- Identifying which latent dimensions correspond to which genes is non-trivial; often requires auxiliary supervision or disentanglement constraints
- MMD is sensitive to kernel choice; a poorly chosen kernel may fail to capture biologically relevant distributional differences
- Combinatorial perturbations grow exponentially, so full coverage of the training signal is rarely possible

The core insight is elegant: rather than inferring causal structure purely from observational data (which is notoriously hard), the model uses the discrepancy between its causal predictions and real experimental outcomes as a direct training signal - anchoring the latent space to physical interventions rather than statistical associations alone.

## Implementation in R

We use the installed `{RCausalML}` package implementations of `causal_discrepancy_vae()` and `predict.causal_discrepancy_vae()`.

### Setup

### Load and Check Required Libraries

```{r}
#| label: packages-list
#| warning: false
packages <- c(
  "tidyverse",
  "RCausalML",
  "torch",
  "causaldata",
  "ForCausality",
  "mlbench"
)
```

### Install Missing Packages

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
# if (length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: verify-installation
#| warning: false
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: load-required-libraries
#| warning: false
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

if (!exists("causal_discrepancy_vae", where = asNamespace("RCausalML"), mode = "function")) {
  stop("RCausalML::causal_discrepancy_vae() is not available. Reinstall/update RCausalML.")
}

```

```{r}
#| label: setup
#| include: true
run_fast <- TRUE
device_use <- NULL
if (requireNamespace("torch", quietly = TRUE)) {
  device_use <- if (torch::cuda_is_available()) "cuda" else "cpu"
}

set.seed(42)
if (requireNamespace("torch", quietly = TRUE)) torch::torch_manual_seed(42)

# Hyperparameters (mapped from Python tutorial)
LATENT_DIM <- 16L
HIDDEN_DIM <- 128L
BATCH_SIZE <- if (run_fast) 256L else 128L
EPOCHS <- if (run_fast) 20L else 80L
LR <- 1e-3
BETA <- 1.0
LAMBDA_MMD <- 5.0
LAMBDA_Y <- 10.0
N_KERNEL <- 5L
```

## Fit CausalDiscrepancyVAE on IHDP Dataset

The data loading section follows the same IHDP pipeline style used in the CEVAE notebook, including fallback to synthetic data if remote files are unavailable.

### Dataset

```{r load-data}
base_url <- "https://raw.githubusercontent.com/uber/causalml/master/docs/examples/data"
cols <- c("treatment", "y_factual", "y_cfactual", "mu0", "mu1", paste0("X", 0:24))
df <- NULL
for (i in 1:9) {
  url <- sprintf("%s/ihdp_npci_%d.csv", base_url, i)
  tmp <- tryCatch({
    read.csv(url, header = FALSE)
  }, error = function(e) NULL)
  if (!is.null(tmp) && nrow(tmp) > 0) {
    colnames(tmp) <- cols[seq_len(ncol(tmp))]
    if (is.null(df)) df <- tmp else df <- rbind(df, tmp)
  }
}
if (!is.null(df) && nrow(df) > 0) {
  replications <- if (run_fast) 2L else 10L
  df <- do.call(rbind, replicate(replications, df, simplify = FALSE))
  message("Loaded IHDP (all 9 files × ", replications, " replications): ", nrow(df), " × ", ncol(df))
} else {
  df <- NULL
}

if (is.null(df) || nrow(df) == 0) {
  d <- tryCatch(
    synthetic_data(mode = 1, n = 5000, p = 25, sigma = 1.0, adj = 0),
    error = function(e) synthetic_data(mode = 1, n = 5000, p = 25, sigma = 1.0)
  )
  df <- as.data.frame(d$X)
  colnames(df) <- paste0("X", 0:(ncol(df) - 1))
  df$treatment <- d$w
  df$y_factual <- d$y
  df$y_cfactual <- ifelse(d$w == 1, d$b - 0.5 * d$tau, d$b + 0.5 * d$tau)
  df$mu0 <- d$b - 0.5 * d$tau
  df$mu1 <- d$b + 0.5 * d$tau
  message("Using synthetic fallback: ", nrow(df), " × ", ncol(df))
}

# Feature order: binary first (Python indices 6-24 -> R columns 7-25), then continuous (0-5 -> 1-6)
binfeats <- 7:25
contfeats <- 1:6
perm <- c(binfeats, contfeats)
xcols <- paste0("X", 0:24)
X <- as.matrix(df[, xcols][, perm])
treatment <- as.integer(df$treatment)
y <- as.numeric(df$y_factual)
y_cf <- as.numeric(df$y_cfactual)
mu0 <- as.numeric(df$mu0)
mu1 <- as.numeric(df$mu1)
tau <- ifelse(df$treatment == 1, df$y_factual - df$y_cfactual, df$y_cfactual - df$y_factual)

# Train/validation split (80/20, random_state=1 like Python)
n <- nrow(X)
set.seed(1)
ite <- sample(n, size = round(0.2 * n))
itr <- setdiff(seq_len(n), ite)
X_train <- X[itr, ]
treatment_train <- treatment[itr]
y_train <- y[itr]
tau_train <- tau[itr]
mu_0_train <- mu0[itr]
mu_1_train <- mu1[itr]
X_val <- X[ite, ]
treatment_val <- treatment[ite]
y_val <- y[ite]
tau_val <- tau[ite]
mu_0_val <- mu0[ite]
mu_1_val <- mu1[ite]
cat("Train size:", length(itr), "| Val size:", length(ite), "\n")
```

###  Helper Function Conversions


```{r helper-functions}
preprocess_features <- function(train_x, test_x) {
  # Standardize continuous x1-x6; keep binary x7-x25 unchanged
  scaled_train <- train_x
  scaled_test <- test_x
  means <- colMeans(train_x[, 20:25, drop = FALSE])
  sds <- apply(train_x[, 20:25, drop = FALSE], 2, sd)
  sds[sds == 0] <- 1
  scaled_train[, 20:25] <- scale(train_x[, 20:25, drop = FALSE], center = means, scale = sds)
  scaled_test[, 20:25] <- scale(test_x[, 20:25, drop = FALSE], center = means, scale = sds)
  list(train_x = scaled_train, test_x = scaled_test, means = means, sds = sds)
}

build_training_arrays <- function(x, t, y, subsample = 5000L, seed = 42L) {
  set.seed(seed)
  idx <- sample.int(nrow(x), size = min(subsample, nrow(x)), replace = FALSE)
  list(
    x = x[idx, , drop = FALSE],
    t = as.numeric(t[idx]),
    y = as.numeric(y[idx]),
    idx = idx
  )
}

estimate_ate_cdvae <- function(model, x_eval, mu0_true, mu1_true) {
  y0_hat <- predict(model, x_eval, type = "mu0")
  y1_hat <- predict(model, x_eval, type = "mu1")
  ite_hat <- y1_hat - y0_hat
  ite_true <- mu1_true - mu0_true
  ate_pred <- mean(ite_hat)
  ate_true <- mean(ite_true)
  pehe <- sqrt(mean((ite_hat - ite_true)^2))
  list(
    ite_hat = ite_hat,
    ite_true = ite_true,
    ate_pred = ate_pred,
    ate_true = ate_true,
    ate_error = abs(ate_true - ate_pred),
    pehe = pehe
  )
}

plot_training_history_cdvae <- function(history) {
  h <- dplyr::bind_rows(lapply(seq_along(history$train), function(i) {
    as.data.frame(history$train[[i]]) |>
      dplyr::mutate(epoch = i)
  }))
  h_long <- tidyr::pivot_longer(h, cols = c(total, kl, mmd), names_to = "component", values_to = "value")
  ggplot(h_long, aes(epoch, value, color = component)) +
    geom_line(linewidth = 0.8) +
    scale_y_log10() +
    facet_wrap(~component, scales = "free_y", nrow = 1) +
    labs(
      title = "CD-VAE Training History",
      x = "Epoch",
      y = "Loss (log scale)"
    ) +
    theme_minimal()
}

plot_latent_space_cdvae <- function(model, x_data, t_data) {
  z_mu <- predict(model, x_data, type = "latent")
  pc <- prcomp(z_mu, center = TRUE, scale. = TRUE)
  z2 <- pc$x[, 1:2, drop = FALSE]
  dfz <- data.frame(
    PC1 = z2[, 1],
    PC2 = z2[, 2],
    treatment = factor(t_data, levels = c(0, 1), labels = c("Control", "Treated"))
  )
  ggplot(dfz, aes(PC1, PC2, color = treatment)) +
    geom_point(alpha = 0.35, size = 1.2) +
    labs(
      title = "Latent Space (PCA of encoder mean)",
      x = "PC1",
      y = "PC2"
    ) +
    theme_minimal()
}

plot_ite_analysis_cdvae <- function(ite_hat, ite_true, ate_pred, ate_true) {
  df <- data.frame(ite_true = ite_true, ite_hat = ite_hat)
  p1 <- ggplot(df, aes(ite_true, ite_hat)) +
    geom_point(alpha = 0.35, size = 1.2, color = "#5C6BC0") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    geom_hline(yintercept = ate_pred, linetype = "dotted", color = "#FF9800") +
    geom_vline(xintercept = ate_true, linetype = "dotted", color = "#4CAF50") +
    labs(title = "Predicted vs True ITE", x = "True ITE", y = "Predicted ITE") +
    theme_minimal()

  dfh <- dplyr::bind_rows(
    data.frame(value = ite_true, type = "True ITE"),
    data.frame(value = ite_hat, type = "Predicted ITE")
  )
  p2 <- ggplot(dfh, aes(value, fill = type)) +
    geom_histogram(position = "identity", alpha = 0.5, bins = 40) +
    geom_vline(xintercept = ate_true, linetype = "dashed", color = "#4CAF50") +
    geom_vline(xintercept = ate_pred, linetype = "dashed", color = "#FF5722") +
    labs(title = "ITE Distribution", x = "ITE", y = "Count") +
    theme_minimal()

  list(scatter = p1, hist = p2)
}

plot_mmd_convergence_cdvae <- function(history) {
  h <- dplyr::bind_rows(lapply(seq_along(history$train), function(i) {
    as.data.frame(history$train[[i]]) |>
      dplyr::mutate(epoch = i)
  }))
  ggplot(h, aes(epoch, mmd)) +
    geom_line(color = "#E91E63", linewidth = 0.9) +
    labs(title = "MMD Causal Alignment Convergence", x = "Epoch", y = "MMD") +
    theme_minimal()
}
```

### Feature Scaling and Subsampling 

```{r feature-scaling}
proc <- preprocess_features(X_train, X_val)
X_train_scaled <- proc$train_x
X_val_scaled <- proc$test_x

sub_train <- build_training_arrays(
  x = X_train_scaled,
  t = treatment_train,
  y = y_train,
  subsample = if (run_fast) 5000L else nrow(X_train_scaled),
  seed = 42L
)
sub_val <- build_training_arrays(
  x = X_val_scaled,
  t = treatment_val,
  y = y_val,
  subsample = if (run_fast) 2000L else nrow(X_val_scaled),
  seed = 42L
)

X_train_cd <- sub_train$x
treatment_train_cd <- sub_train$t
y_train_cd <- sub_train$y
X_val_cd <- sub_val$x

cat("Train covariate matrix shape:", paste(dim(X_train_cd), collapse = " × "), "\n")
cat("Validation covariate matrix shape:", paste(dim(X_val_cd), collapse = " × "), "\n")
```

### Fit CausalDiscrepancyVAE Model

```{r cdvae-fit}
cdvae <- RCausalML::causal_discrepancy_vae(
  X = X_train_cd,
  treatment = treatment_train_cd,
  y = y_train_cd,
  latent_dim = LATENT_DIM,
  hidden_dim = HIDDEN_DIM,
  beta = BETA,
  lambda_mmd = LAMBDA_MMD,
  lambda_y = LAMBDA_Y,
  n_kernels = N_KERNEL,
  num_epochs = as.integer(EPOCHS),
  batch_size = as.integer(BATCH_SIZE),
  learning_rate = LR,
  weight_decay = 1e-4,
  verbose = !run_fast,
  device = device_use
)

history <- cdvae$history
cat("CD-VAE model fitted.\n")
cat("Latent dimension:", cdvae$latent_dim, "\n")
```

### Training Diagnostics

```{r training-history-plot, fig.height=4, fig.width=12}
plot_training_history_cdvae(history)
```

```{r mmd-convergence-plot, fig.height=4, fig.width=8}
plot_mmd_convergence_cdvae(history)
```

### ATE and ITE Evaluation

```{r ate-ite-eval}
eval_idx <- sub_val$idx
ate_eval <- estimate_ate_cdvae(
  model = cdvae,
  x_eval = X_val_scaled[eval_idx, , drop = FALSE],
  mu0_true = mu_0_val[eval_idx],
  mu1_true = mu_1_val[eval_idx]
)

cat("ATE Estimation\n")
cat("  True ATE :", round(ate_eval$ate_true, 4), "\n")
cat("  Pred ATE :", round(ate_eval$ate_pred, 4), "\n")
cat("  ATE Error:", round(ate_eval$ate_error, 4), "\n")
cat("  sqrt(PEHE):", round(ate_eval$pehe, 4), "\n")
```

```{r ite-analysis-plots, fig.height=5, fig.width=12}
plots_ite <- plot_ite_analysis_cdvae(
  ite_hat = ate_eval$ite_hat,
  ite_true = ate_eval$ite_true,
  ate_pred = ate_eval$ate_pred,
  ate_true = ate_eval$ate_true
)
plots_ite$scatter
plots_ite$hist
```

### Latent Space Visualization

```{r latent-plot, fig.height=5, fig.width=7}
plot_latent_space_cdvae(
  model = cdvae,
  x_data = X_train_cd,
  t_data = treatment_train_cd
)
```

### Final Training Summary

```{r training-summary}
last_train <- history$train[[length(history$train)]]
cat("\n", strrep("=", 60), "\n", sep = "")
cat("CD-VAE Training Complete\n")
cat(strrep("=", 60), "\n", sep = "")
cat("Final train loss:", round(last_train$total, 4), "\n")
cat("Final recon     :", round(last_train$recon, 4), "\n")
cat("Final KL        :", round(last_train$kl, 4), "\n")
cat("Final MMD       :", round(last_train$mmd, 6), "\n")
cat("Final y loss    :", round(last_train$y, 4), "\n")
cat("True ATE        :", round(ate_eval$ate_true, 4), "\n")
cat("Pred ATE        :", round(ate_eval$ate_pred, 4), "\n")
cat("ATE Error       :", round(ate_eval$ate_error, 4), "\n")
cat(strrep("=", 60), "\n", sep = "")
```

## Summary and Conclusions

`CausalDiscrepancyVAE` is a practical bridge between deep latent-variable modeling and causal effect estimation under covariate imbalance. In this R conversion, we preserved the full Python workflow design while delegating model internals to `{RCausalML}`'s torch implementation for package-safe training and prediction.

- The notebook follows the same end-to-end stages as the original Python tutorial: IHDP loading, preprocessing, training, evaluation, and visualization.
- All key Python helper functions were converted into R utilities (`preprocess_features`, `build_training_arrays`, `estimate_ate_cdvae`, and diagnostics plotting functions).
- Model fitting uses `causal_discrepancy_vae()` with explicit control over `beta`, `lambda_mmd`, `lambda_y`, and latent architecture.
- Evaluation reports ATE error and PEHE from known potential outcomes and adds latent-space and MMD convergence diagnostics.

## Resources

- [CausalML IHDP/CEVAE example data source](https://raw.githubusercontent.com/uber/causalml/master/docs/examples/data/)
- [CEVAE paper (Louizos et al., 2017)](https://papers.nips.cc/paper_files/paper/2017/file/94b5bde6de888ddf9cde6748ad2523d1-Paper.pdf)
- [RCausalML package repository](https://github.com/zia207/RCausalML)

## Scientific terminology for beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Discrepancy loss | Penalty encouraging two distributions to be similar. | MMD term pushes treated and control latent distributions closer. |
| Domain alignment | Making representations from different groups comparable. | Aligning treated/control latent points before outcome modeling. |
| Latent regularization | Constraints on hidden variables to avoid overfitting or collapse. | KL and discrepancy terms stabilize latent space learning. |
| Potential outcome head | Output branch for one treatment potential outcome. | One head predicts `Y(0)` and another predicts `Y(1)`. |
| Covariate shift | Input distribution changes between groups or splits. | Treated group has older population than control group. |
| Model diagnostics | Checks to assess if training and causal predictions are reliable. | Inspect PEHE, ATE error, latent plots, and convergence curves. |
