# Uplift Trees — Regression

In this notebook we use **synthetic data** with a **continuous outcome** to demonstrate uplift modeling with tree-based methods in R.  The workflow mirrors the spirit of the [Uplift Trees Example with Synthetic Data](https://causalml.readthedocs.io/en/latest/examples/uplift_trees_with_synthetic_data.html) from the Python CausalML docs, adapted to RCausalML’s regression behavior.


## Scientific Terminology for Beginners

In regression uplift, outcomes are continuous rather than binary. These terms are used throughout the notebook.

- **Continuous outcome**: A numeric target such as revenue, time, cost, or test score.
- **Treatment effect**: Difference in expected outcome between treated and control groups.
- **CATE**: $\tau(x)=\mathbb{E}[Y\mid T=1,X=x]-\mathbb{E}[Y\mid T=0,X=x]$, the expected effect for a profile $x$.
- **T-learner**: A method that fits separate outcome models for treatment and control, then subtracts predictions.
- **Interaction tree**: A tree that finds subgroups where treatment effects differ significantly.
- **Gain curve**: A ranking diagnostic that shows cumulative benefit when targeting high-predicted-uplift units first.

### Simple example

Suppose a tutoring program (**treatment**) aims to improve exam scores (**continuous outcome**).

- Predicted score without tutoring for a student profile: 72.
- Predicted score with tutoring for the same profile: 79.
- Estimated CATE is **+7 points**.

A positive CATE suggests that profile is likely to benefit from treatment.

## Overview

Uplift modeling with **continuous outcomes** is often called **uplift regression** or **heterogeneous treatment effect regression**. The goal is to estimate the **conditional average treatment effect (CATE)** $\tau(x) = \mathbb{E}[Y \mid T=1, X=x] - \mathbb{E}[Y \mid T=0, X=x]$ as a function of covariates $X$. This is a natural extension of uplift modeling from binary outcomes (conversion/no conversion) to settings where the outcome is a real-valued metric (e.g. revenue, time spent, etc.).


## Implemention in R

For **binary** outcomes, `uplift_rf_kl()` and `uplift_rf_multi()` use **divergence-based uplift trees** (KL and related criteria). For **continuous** outcomes, the same entry points fit a **T-learner random forest** (`ranger`): separate outcome models under treatment and control, with predicted CATE $\hat{\tau}(x) = \hat{\mu}_1(x) - \hat{\mu}_0(x)$. That is still a standard uplift / heterogeneous treatment effect estimator, but you no longer get a single KL-split tree per arm from `uplift_rf_kl()`.

To **visualize an explicit regression tree** on treatment–control contrasts, we use **interaction trees** (`interaction_tree()`) and optionally **causal inference trees** (`causal_inference_tree()`), which partition on covariates using test statistics for mean outcome differences between arms (see package help: `?interaction_tree`, `?causal_inference_tree`). We evaluate ranking with **cumulative gain** (`plot_gain()`), and use **kernel SHAP** (`kernelshap`, `shapviz`) to explain heterogeneity in predicted uplift, following the same pattern as the classification uplift notebook.

### Setup

### Load and check required libraries

```{r}
#| label: setup-packages
packages <- c(
  "tidyverse",
  "RCausalML",
  "rpart.plot",
  "shapviz",
  "kernelshap"
)
```

### Install missing packages (if needed)

```{r}
#| label: install-missing-packages
#| warning: false
#| error: false
# new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
# if (length(new_packages)) install.packages(new_packages)
```

### Load required libraries

```{r}
#| label: load-required-libraries
if (file.exists("DESCRIPTION") && requireNamespace("devtools", quietly = TRUE)) {
  try(devtools::load_all(".", quiet = TRUE), silent = TRUE)
}
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Simulated regression uplift data

We generate data with `make_uplift_regression()` in `R/dataset.R`. The construction matches `make_uplift_classification()` (same feature blocks and treatment-specific uplift structure), but the outcome is **continuous**: baseline mean plus treatment uplift plus Gaussian noise.

```{r}
#| label: make-data

out <- make_uplift_regression(
  treatment_name = c("control", "treatment1", "treatment2", "treatment3"),
  y_name = "outcome",
  n_samples = 4000,
  n_regression_features = 10,
  n_regression_informative = 5,
  n_uplift_increase_dict = list(treatment1 = 5, treatment2 = 5, treatment3 = 5),
  n_uplift_decrease_dict = list(treatment1 = 3, treatment2 = 3, treatment3 = 3),
  delta_uplift_increase_dict = list(treatment1 = 0.4, treatment2 = 0.6, treatment3 = 0.8),
  delta_uplift_decrease_dict = list(treatment1 = -0.35, treatment2 = -0.35, treatment3 = -0.35),
  sigma = 1,
  random_seed = 111
)
df <- out$data
x_names <- out$X_names
```

```{r}
#| label: df-head

head(df, 5)
```

Mean outcome and sample size by arm:

```{r}
#| label: pivot

agg <- aggregate(outcome ~ treatment_group_key, data = df,
  FUN = function(z) c(mean = mean(z), sd = stats::sd(z), size = length(z))
)
agg <- do.call(data.frame, agg)
names(agg) <- c("treatment_group_key", "mean", "sd", "size")
agg <- rbind(
  agg,
  data.frame(
    treatment_group_key = "All",
    mean = mean(df$outcome),
    sd = stats::sd(df$outcome),
    size = nrow(df)
  )
)
print(agg)
```

### Multi-arm uplift: T-learner forests per treatment

`uplift_rf_multi()` fits one `uplift_rf_kl()` object per non-control arm (control vs `treatment_k`). With a continuous outcome, each of those is a **T-learner `ranger` ensemble**, not a KL uplift tree forest.

```{r}
#| label: split

set.seed(111)
idx_test <- sample(nrow(df), size = round(0.2 * nrow(df)))
df_train <- df[-idx_test, ]
df_test <- df[idx_test, ]
```

### Shallow T-learner per arm (analogous to “one tree” depth cap)

For regression, “`n_trees = 1`” means **one tree in each `ranger` model** (treated and control) for that arm—useful for a fast sanity check, not for production accuracy.

```{r}
#| label: multi-shallow

clf_shallow <- uplift_rf_multi(
  as.matrix(df_train[, x_names]),
  df_train$treatment_group_key,
  df_train$outcome,
  control_name = "control",
  n_trees = 1,
  min_node_size = 100,
  max_depth = 4
)
X_test <- as.matrix(df_test[, x_names])
p_shallow <- predict(clf_shallow, X_test, full_output = TRUE)
df_res_shallow <- p_shallow[, c("control", clf_shallow$treatment_names), drop = FALSE]
head(df_res_shallow, 5)
```

Here `control` and `treatment*` columns are **predicted mean outcomes** $\hat{\mu}_0(x)$ and $\hat{\mu}_k(x)$ under that arm’s training subsample (stacked control-vs-$k$), not probabilities.

### Full multi-arm model (more trees per T-learner)

```{r}
#| label: uplift-rf-multi

uplift_model <- uplift_rf_multi(
  as.matrix(df_train[, x_names]),
  df_train$treatment_group_key,
  df_train$outcome,
  control_name = "control",
  n_trees = 50,
  min_node_size = 50,
  max_depth = 8
)
df_res <- predict(uplift_model, X_test, full_output = TRUE)
cat("Shape:", nrow(df_res), "x", ncol(df_res), "\n")
head(df_res)
```

Predicted **uplift matrix** (columns = treatment arms, entries = $\hat{\mu}_k(x) - \hat{\mu}_0(x)$ for that arm’s model):

```{r}
#| label: y-pred

y_pred <- predict(uplift_model, X_test)
result <- as.data.frame(y_pred)
names(result) <- uplift_model$treatment_names
head(result)
```

### Uplift (gain) curve — continuous outcome

As in the classification notebook, we build a **synthetic subpopulation** (units assigned to the model’s best arm or to control), then plot cumulative gain using **observed** mean outcome contrasts sorted by predicted uplift.

```{r}
#| label: best-treatment

uplift_cols <- uplift_model$treatment_names
best_treatment <- ifelse(
  apply(result[uplift_cols], 1, function(r) all(r < 0)),
  "control",
  uplift_cols[apply(result[uplift_cols], 1, which.max)]
)
actual_is_best <- as.integer(df_test$treatment_group_key == best_treatment)
actual_is_control <- as.integer(df_test$treatment_group_key == "control")
```

```{r}
#| label: synthetic

synthetic <- (actual_is_best == 1) | (actual_is_control == 1)
synth <- result[synthetic, , drop = FALSE]
```

```{r}
#| label: auuc-metrics

auuc_metrics <- synth
auuc_metrics$is_treated <- 1 - actual_is_control[synthetic]
auuc_metrics$outcome <- df_test$outcome[synthetic]
auuc_metrics$uplift_model <- apply(synth, 1, max)
auuc_metrics <- auuc_metrics[, c("is_treated", "outcome", "uplift_model")]
head(auuc_metrics)
```

```{r}
#| label: plot-gain
#| figure.width: 6
#| figure.height: 5
plot_gain(auuc_metrics, outcome_col = "outcome", treatment_col = "is_treated")
```

### Binary subset (control vs treatment1): interpretable regression trees

For a **single transparent tree** with a continuous outcome, fit an **interaction tree** (or CIT) on the subset with only control and one treatment. These methods split to maximize evidence of heterogeneous treatment effects on the **mean** outcome (not KL divergence on probabilities).

```{r}
#| label: binary-uplift-data

set.seed(111)
df_bin <- df[df$treatment_group_key %in% c("control", "treatment1"), ]
df_bin$w_bin <- as.integer(df_bin$treatment_group_key != "control")
X_all <- as.matrix(df_bin[, x_names])
colnames(X_all) <- x_names
y_all <- df_bin$outcome
w_all <- df_bin$w_bin
idx_bin <- sample(nrow(X_all), size = floor(0.8 * nrow(X_all)))
X_train <- X_all[idx_bin, , drop = FALSE]
y_train <- y_all[idx_bin]
w_train <- w_all[idx_bin]
X_test <- X_all[-idx_bin, , drop = FALSE]
y_test <- y_all[-idx_bin]
w_test <- w_all[-idx_bin]
```

### Interaction tree: string and plot

```{r}
#| label: it-tree-string

it_fit <- interaction_tree(
  X_train, w_train, y_train,
  min_node_size = 200,
  max_depth = 4
)
cat(uplift_tree_string(it_fit$tree, it_fit$X_names))
```

Leaf labels show **$\tau \approx \mathbb{E}[Y \mid T{=}1] - \mathbb{E}[Y \mid T{=}0]$** in that region.

```{r}
#| label: it-tree-plot
#| fig.width: 8
#| fig.height: 8
uplift_tree_plot(it_fit, main = "Interaction tree — continuous outcome (control vs treatment1)")
```

### Optional: causal inference tree (CIT)

```{r}
#| label: cit-tree-plot
#| fig.width: 8
#| fig.height: 8
cit_fit <- causal_inference_tree(
  X_train, w_train, y_train,
  min_node_size = 200,
  max_depth = 4
)
uplift_tree_plot(cit_fit, main = "Causal inference tree (CIT) — continuous outcome")
```

### Train vs test mean CATE (T-learner on the same binary sample)

`uplift_rf_kl()` on continuous data fits a **T-learner ranger**; use it for accurate CATE on the binary sample and compare train/test means.

```{r}
#| label: binary-tlearner

uplift_bin_rf <- uplift_rf_kl(
  X_train, w_train, y_train,
  n_trees = 100,
  min_node_size = 80,
  max_depth = 8
)
pred_tr <- predict(uplift_bin_rf, X_train)
pred_te <- predict(uplift_bin_rf, X_test)
cat(
  "Mean predicted CATE — train:", round(mean(pred_tr), 4),
  " test:", round(mean(pred_te), 4), "\n"
)
```

**Note:** There is **no** `forest$trees` list for this object, so `uplift_forest_plot()` does not apply. Use **variable importance** from `ranger` or **SHAP** on the predicted CATE instead.

```{r}
#| label: ranger-importance

imp0 <- uplift_bin_rf$fit_0$variable.importance
imp1 <- uplift_bin_rf$fit_1$variable.importance
imp <- sort((imp0 + imp1) / 2, decreasing = TRUE)
print(round(head(imp, 8), 4))
```

### One control + multiple treatments: interaction tree on “any treatment”

As in the classification notebook, we can code **any non-control** as treated and fit one interaction tree to summarize heterogeneity in the **average** lift of treatment packages vs control on the continuous outcome.

```{r}
#| label: multi-data

out_multi <- make_uplift_regression(
  treatment_name = c("control", "treatment1", "treatment2", "treatment3"),
  y_name = "outcome",
  n_samples = 4000,
  n_regression_features = 10,
  n_regression_informative = 5,
  n_uplift_increase_dict = list(treatment1 = 5, treatment2 = 5, treatment3 = 5),
  n_uplift_decrease_dict = list(treatment1 = 3, treatment2 = 3, treatment3 = 3),
  delta_uplift_increase_dict = list(treatment1 = 0.4, treatment2 = 0.6, treatment3 = 0.8),
  delta_uplift_decrease_dict = list(treatment1 = -0.35, treatment2 = -0.35, treatment3 = -0.35),
  sigma = 1,
  random_seed = 111
)
df_multi <- out_multi$data
X_names_multi <- out_multi$X_names
```

```{r}
#| label: multi-split

set.seed(111)
idx_m <- sample(nrow(df_multi), size = floor(0.8 * nrow(df_multi)))
df_multi$w_multi <- as.integer(df_multi$treatment_group_key != "control")
X_multi <- as.matrix(df_multi[, X_names_multi])
y_multi <- df_multi$outcome
w_multi <- df_multi$w_multi
```

```{r}
#| label: multi-it

it_multi <- interaction_tree(
  X_multi[idx_m, ], w_multi[idx_m], y_multi[idx_m],
  min_node_size = 200,
  max_depth = 3
)
```

```{r}
#| label: multi-tree-plot
#| fig.width: 8
#| fig.height: 8
uplift_tree_plot(
  it_multi,
  main = "Interaction tree: any treatment vs control (continuous outcome)"
)
```

### Distribution of predicted CATE (binary T-learner)

```{r}
#| label: hist-cate

uf <- uplift_rf_kl(
  X_all, w_all, y_all,
  n_trees = 80,
  min_node_size = 40,
  max_depth = 8
)
pred_cate <- predict(uf, X_all)
```

```{r}
#| label: plot-cate
#| fig.width: 8
#| fig.height: 6
hist(
  pred_cate,
  breaks = 30,
  main = "T-learner forest: predicted CATE (continuous outcome)",
  xlab = "Predicted treatment effect",
  col = "steelblue",
  border = NA
)
abline(v = mean(pred_cate), lty = 2, col = "red")
legend("topright", legend = sprintf("Mean = %.3f", mean(pred_cate)), lty = 2, col = "red", bty = "n")
```

### Kernel SHAP for predicted uplift (CATE)

We explain $\hat{\tau}(X)$ from the **binary T-learner** `uplift_bin_rf` on a subset of the training data (full training rows can be slow).

```{r}
#| label: compute-shap-uplift

set.seed(2026)

n_explain <- min(100L, nrow(X_train))
bg_size <- min(100L, nrow(X_train))
X_explain <- X_train[seq_len(n_explain), , drop = FALSE]
bg_idx <- sample.int(nrow(X_train), bg_size)
bg_X <- X_train[bg_idx, , drop = FALSE]

pred_fun <- function(object, newdata) {
  as.vector(predict(object, newdata))
}

library(future)
plan(multisession, workers = parallel::detectCores() - 1)

shap_ksh <- kernelshap(
  object = uplift_bin_rf,
  X = X_explain,
  bg_X = bg_X,
  pred_fun = pred_fun,
  bg_w = NULL,
  verbose = FALSE
)
plan(sequential)  # reset after

shp_uplift <- shapviz(shap_ksh, X = X_explain)
```

```{r}
#| label: shap-importance-beeswarm-uplift
#| fig.width: 7
#| fig.height: 5.5

sv_importance(shp_uplift, kind = "beeswarm")
sv_importance(shp_uplift, show_numbers = TRUE)
```

```{r}
#| label: shap-dependence-uplift
#| fig.width: 8
#| fig.height: 6

important_vars <- x_names[seq_len(min(4L, length(x_names)))]
sv_dependence(shp_uplift, v = important_vars, share_y = TRUE)
```

```{r}
#| label: shap-dependence-single-uplift
if (length(x_names) >= 2) {
  sv_dependence(shp_uplift, v = x_names[[2]], color_var = x_names[[1]])
}
```

```{r}
#| label: shap-waterfall-single-uplift
#| fig.width: 7
#| fig.height: 4.5

preds_explain <- pred_fun(uplift_bin_rf, X_explain)
row_id_max <- which.max(preds_explain)

sv_waterfall(shp_uplift, row_id = row_id_max) +
  ggplot2::ggtitle("SHAP waterfall — highest estimated CATE in explain set")

sv_force(shp_uplift, row_id = row_id_max)
```

```{r}
#| label: shap-test-set-uplift
n_te <- min(10L, nrow(X_test))
shap_test <- kernelshap(
  uplift_bin_rf,
  X = X_test[seq_len(n_te), , drop = FALSE],
  bg_X = bg_X,
  pred_fun = pred_fun,
  verbose = FALSE
)

shp_test <- shapviz(shap_test, X = X_test[seq_len(n_te), , drop = FALSE])
sv_importance(shp_test, kind = "beeswarm")
```

### Interpretation notes

- **Positive SHAP** pushes the predicted **CATE** up; **negative SHAP** pushes it down.
- Explanations target **heterogeneity in** $\hat{\tau}(X)$, not the raw outcome margin $\mathbb{E}[Y \mid X]$.
- **Kernel SHAP** is model-agnostic and works with the T-learner predictions; tree-specific explainers target other implementations (e.g. **ranger**-focused tools).

## Summary

With **continuous** outcomes, `uplift_rf_multi()` and `uplift_rf_kl()` in RCausalML implement **T-learner random forests** (per arm or binary), while **interaction trees** and **CIT** give **explicit partition rules** for treatment effect heterogeneity on the mean. Cumulative **gain curves** and **SHAP** extend directly from the classification uplift tutorial, with outcome column `outcome` instead of `conversion`.

## References

- [Uplift Trees Example with Synthetic Data (Python CausalML)](https://causalml.readthedocs.io/en/latest/examples/uplift_trees_with_synthetic_data.html)
- Gutierrez and Gérardy (2017), [Causal inference and uplift modelling: A review of the literature](http://proceedings.mlr.press/v67/gutierrez17a/gutierrez17a.pdf)
