# 5. Instrumental Forest {.unnumbered}

An **Instrumental Forest** is a machine learning method, specifically an extension of the Generalized Random Forests (GRF) framework, used to estimate **conditional local average treatment effects (LATE)** in observational data studies or experiments with imperfect compliance. It combines the principles of instrumental variable (IV) analysis with random forest techniques to address issues of endogeneity and heterogeneity in treatment effects. It is particularly useful in causal inference for estimating how a treatment impacts an outcome when influenced by an instrument, accounting for confounding variables.

## Overview

An Instrumental Forest is designed to estimate treatment effects using an **instrumental variable (Z)**, which is a variable that affects the treatment assignment (W) but does not directly affect the outcome (Y) except through the treatment. The forest estimates the conditional local average treatment effect, denoted as **τ(X)**, which represents the treatment effect for specific covariate values (X). Formally, it estimates:

$$ τ(X) = \frac{Cov[Y, Z | X = x]}{ Cov[W, Z | X = x]} $$

Where:

-   $Y$: Outcome variable (e.g., health outcome, sales, etc.).
-   $W$: Treatment assignment (e.g., whether a patient received a drug).
-   $Z$ Instrument (e.g., a randomized encouragement to take the drug).
-   $X$: Covariates (e.g., patient characteristics like age, gender).
-   $Cov[Y, Z | X = x]$: Covariance between the outcome and instrument given covariates.
-   $Cov[W, Z | X = x]$: Covariance between the treatment and instrument given covariates.

When the instrument $Z$ and treatment $W$ are the same, the Instrumental Forest reduces to a Causal Forest, a related method for estimating causal effects without instruments.

::: callout-note

### Instrumental variable (IV)

An **instrumental variable (IV)** is a tool used in statistics and econometrics to estimate causal relationships when a randomized controlled trial is not feasible, or when there is concern about endogeneity—i.e., when the treatment or exposure variable is correlated with unobserved factors that also affect the outcome. IV methods help isolate the causal effect of a treatment (or independent variable) on an outcome by leveraging an external variable, called the instrument, that influences the treatment but does not directly affect the outcome except through the treatment.

The **Conditional Local Average Treatment Effect (LATE)** is a causal effect measure used in instrumental variable (IV) analysis to estimate the effect of a treatment (or exposure) on an outcome for a specific subpopulation, known as the "compliers," conditional on a set of covariates $X$. It extends the Local Average Treatment Effect (LATE) by allowing the treatment effect to vary across different values of covariates, making it particularly useful in heterogeneous treatment effect estimation.
:::

## Scientific Terminology for Beginners

This glossary introduces key scientific terms used in this notebook with simple examples.

### 1) Instrument (IV)

- **Explanation**: A variable that changes treatment assignment but affects the outcome only through treatment.
- **Example**: Distance to a training center influences participation but not wages directly.

### 2) Endogeneity

- **Explanation**: Treatment is correlated with unobserved factors that also affect the outcome.
- **Example**: More motivated people may both choose treatment and earn higher income.

### 3) LATE

- **Explanation**: Treatment effect for compliers affected by the instrument.
- **Example**: Effect of training among people who attend only when encouraged.

### 4) Compliers

- **Explanation**: Individuals whose treatment behavior changes because of the instrument.
- **Example**: People who take a course only if they receive a voucher.

### 5) Exogeneity

- **Explanation**: The instrument is independent of unmeasured outcome determinants.
- **Example**: Random encouragement assignment is exogenous.

### 6) Exclusion Restriction

- **Explanation**: The instrument has no direct effect on the outcome.
- **Example**: Voucher affects income only through training participation.

### 7) Heterogeneous Effects

- **Explanation**: Effects differ across covariate-defined subgroups.
- **Example**: Program gains are larger for younger participants.

### 8) Honest Splitting

- **Explanation**: Use separate data for tree splitting and treatment-effect estimation.
- **Example**: Half-sample defines leaves, remaining sample estimates IV effects.

### How Does an Instrumental Forest Work?

Instrumental Forests build on decision trees and random forests to estimate heterogeneous treatment effects while addressing endogeneity (i.e., when the treatment assignment is correlated with unobserved factors affecting the outcome). Here’s a step-by-step explanation of how it works:

1.  Input Data

-   `Covariates (X)`: Features or characteristics of the units (e.g., patients, customers).
-   `Outcome (Y)`: The variable of interest (e.g., recovery rate, revenue).
-   `Treatment (W)`: The intervention or exposure (binary or continuous).
-   `Instrument (Z)`: A variable that influences W but is uncorrelated with unmeasured confounders affecting Y.
-   Optional: Pre-estimated values like $\hat{Y}$ (expected outcome), $\hat{W}$ (treatment propensity), and $\hat{Z}$ (instrument propensity) can be provided; otherwise, they are estimated using regression forests.

2.  Tree Construction

-   The algorithm constructs multiple decision trees, each splitting the data based on covariates (X) to maximize differences in the estimated treatment effect $τ(X)$.

-   Unlike standard random forests, Instrumental Forests use an `instrumental variable splitting criterion`. Splits are chosen to maximize the heterogeneity in the treatment effect while accounting for the instrument’s role in isolating exogenous variation in $W$.

-   Parameters like `min.node.size` (minimum observations per tree leaf), `sample.fraction` (fraction of data used per tree), and `mtry` (number of features considered for splitting) control the tree structure.

-   `Honesty`: The algorithm can use “honest splitting” (separating data for split selection and effect estimation) to reduce bias. The `honesty.fraction` parameter determines the proportion of data used for splitting.

3.  Instrumental Variable Estimation

-   For each leaf in a tree, the algorithm estimates $τ(X)$ using the instrumental variable formula:

$$ \tau(X) = \frac{\text{Cov}[Y - \hat{Y}, Z - \hat{Z} | X = x]}{\text{Cov}[W - \hat{W}, Z - \hat{Z} | X = x]} $$

-   This isolates the causal effect of $W$ on $Y$ by leveraging the exogenous variation induced by Z, mitigating bias from unmeasured confounders.

4.  Forest Aggregation

-   Multiple trees are grown (a “forest”), each using a random subset of the data and features to ensure robustness and reduce overfitting.

-   The final estimate for $τ(X)$ at a given covariate value X = x is the average of the treatment effect estimates across all trees where x falls into a leaf.

5.  Prediction

-   For new data points (or out-of-bag training samples), the forest predicts $τ(X)$ by averaging the estimates from all trees.

-   Optionally, confidence intervals can be computed if the $ci.group.size$ parameter is set to at least 2, allowing for uncertainty quantification.

6.  Tuning

-   Instrumental Forests are sensitive to hyperparameters (e.g., `min.node.size`, `alpha`, `imbalance.penalty`). Tuning these parameters is critical for accurate estimation, as suboptimal settings can lead to poor performance, especially in IV settings where effects are often weak.

-   Tuning can be automated using methods like cross-validation to minimize the “debiased error” (a measure of estimation accuracy).[ ](https://github.com/grf-labs/grf/issues/501)

7.  Output

-   A trained Instrumental Forest object containing the estimated treatment effects **τ(X)** for each data point, along with metadata like the seed, number of threads, and tuning parameters.
-   Predictions can be made for new data or out-of-bag (OOB) samples, providing estimates of the treatment effect conditional on covariates.

### Key Features and Considerations

- `Endogeneity Correction`: [By using an instrument, the forest corrects for biases due to unmeasured confounders, unlike standard regression or causal forests](https://pubsonline.informs.org/doi/abs/10.1287/mnsc.2021.4084)

- `Heterogeneous Effects`:[It captures how treatment effects vary across subgroups defined by covariates (X), making it suitable for personalized medicine, targeted marketing, etc](https://pubsonline.informs.org/doi/abs/10.1287/mnsc.2021.4084)

-   `Challenges`:

-   Requires a valid instrument ($Z$ must affect $W$ but not $Y$ directly, except through $W$).

-   Sensitive to tuning parameters, which can significantly impact performance.[ ](https://github.com/grf-labs/grf/issues/501)

-   Computationally intensive, especially for large datasets or when tuning is involved.

-   `Applications`: [Used in fields like economics, medicine, and policy evaluation to estimate causal effects in observational data studies, such as the effect of early surgery on patient outcomes or laparoscopic colectomy in hospitals](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-022-01663-0)[ ](https://pubsonline.informs.org/doi/abs/10.1287/mnsc.2021.4084)

### Between Instrumental Forest and Causal Forests

Both Instrumental Forests and Causal Forests are machine learning methods within the `Generalized Random Forests (GRF)` framework, implemented in the `{grf}` package in R, designed for estimating causal effects\* in observational data or experimental settings. They aim to estimate `heterogeneous treatment effects` (i.e., how the effect of a treatment varies across subgroups defined by covariates). However, they differ in their assumptions, use cases, and how they handle endogeneity. Below is a detailed comparison of Instrumental Forests`and Causal Forests`, including their differences, similarities, and when to use each.

| **Aspect** | **Causal Forest** | **Instrumental Forest** |
|------------------|---------------------------|---------------------------|
| `Purpose` | Estimates CATE for a treatment assuming unconfoundedness. | Estimates LATE using an instrumental variable to address endogeneity. |
| `Key Assumption` | Unconfoundedness: Treatment assignment $W$ is independent of potential outcomes $Y(0), Y(1)$ given covariates $X$. | Valid instrument: $Z$ affects $W$, not $Y$ directly (exclusion restriction), and is uncorrelated with confounders (independence). |
| `Treatment (W)` | Binary or continuous, assumed exogenous (random or conditionally random given $X$). | Binary or continuous, potentially endogenous (affected by unobserved confounders). |
| `Instrument (Z)` | Not required (assumes $W = Z$). | Required: A variable that influences $W$ but not $Y$ except through $W$. |
| `Endogeneity` | Cannot handle unobserved confounding. | Explicitly addresses unobserved confounding via the instrument. |
| `Estimation Formula` | $\tau(X) = E[Y(1) - Y(0) | X = x]$ | $\tau(X) = \frac{\text{Cov}[Y - \hat{Y}, Z - \hat{Z} | X = x]}{\text{Cov}[W - \hat{W}, Z - \hat{Z} | X = x]}$ |
| `Use Case` | Randomized experiments or observational studies with strong unconfoundedness. | Observational studies with endogeneity or imperfect compliance in experiments. |
| `Output` | CATE for the entire treated population. | LATE for the subpopulation of compliers (those whose ( W ) is influenced by ( Z )). |


## Instrumental Forest with R

This tutorial demonstrates how to implement an Instrumental Forest using the `{grf}`  package wrapped with {RCausalML} package in R, focusing on estimating treatment effects in a hypothetical scenario using the lung dataset. We will use `card` dataset for focusing on estimating the **causal effect of education on wages** using proximity to college as an instrument—a classic IV application (Card, 1995).

### Load and Check Required Libraries

```{r}
#| label: ivf-packages
packages <- c('tidyverse',
              'plyr',
              'RCausalML',
              'haven',           # for read_dta()
              'car',             # for linearHypothesis()
              'kernelshap',
              'shapviz',
              'Metrics'
         )
```

### Install Missing Packages

```{r}
#| label: ivf-install-missing
#| warning: false
#| error: false
# Install missing packages
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if (length(new_packages)) install.packages(new_packages)
```

### Verify Installation

```{r}
#| label: ivf-verify-installation
# Verify installation
cat("Installed packages:\n")
print(sapply(packages, requireNamespace, quietly = TRUE))
```

### Load Required Libraries

```{r}
#| label: ivf-load-libraries
# Load packages with suppressed messages
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
```

### Check Loaded Packages

```{r}
#| label: ivf-check-loaded-packages
# Check loaded packages
cat("Successfully loaded packages:\n")
print(search()[grepl("package:", search())])
```

### Data: Returns to Education (Card Dataset)

We'll estimate the causal effect of **education on wages** using proximity to college as an instrument—a classic IV application (Card, 1995).

Key variables:

- `lwage`   = log(wage) — outcome (Y)
- `educ`    = years of education — endogenous treatment (W)
- `nearc4`  = binary: grew up near 4-year college (instrument Z)
- `exper`   = potential work experience
- `expersq` = experience squared
- `black`   = binary: Black
- `smsa`    = binary: lives in SMSA (standard metropolitan statistical area)
- `south`   = binary: lives in South

### Load and Prepare Data

```{r}
#| label: ivf-load-card-data
# Direct download from known reliable source (Stata .dta format)
card_url <- "https://storage.googleapis.com/causal-inference-mixtape.appspot.com/card.dta"
card <- read_dta(card_url)

# Create experience squared (often used in Mincerian wage equations)

card <- card %>%
  mutate(expersq = exper^2) %>%
  select(lwage, educ, nearc4, exper, expersq, black, smsa, south) %>%
  na.omit()

# Convert to matrix for grf (X must be numeric matrix)

X <- card %>%
  select(exper, expersq, black, smsa, south) %>%
  as.matrix()

Y <- as.numeric(card$lwage)   # Outcome: log wage
W <- as.numeric(card$educ)    # Treatment: years of education
Z <- as.numeric(card$nearc4)  # Instrument: near 4-year college
```

### Descriptive Statistics & First-Stage Check

```{r}
#| label: ivf-first-stage-regression
#| fig.width: 7
#| fig.height: 5
# First-stage regression: Does instrument affect education?
first_stage <- lm(educ ~ nearc4 + exper + expersq + black + smsa + south,
                  data = card)
summary(first_stage)$coefficients["nearc4", 1:4]

# F-statistic for instrument strength (should be >10 for strong IV)
linearHypothesis(first_stage, "nearc4 = 0")
```

```{r}
#| label: ivf-first-stage-boxplot
#| fig.width: 7
#| fig.height: 5
# Visualize first stage
ggplot(card, aes(x = factor(nearc4), y = educ)) +
  geom_boxplot(fill = c("#E69F00", "#56B4E9")) +
  labs(title = "First Stage: College Proximity Increases Years of Education",
       x = "Grew up near 4-year college (nearc4)", y = "Years of Education (educ)") +
  theme_minimal()
```

### Training an Instrumental Forest

We use `instrumental_forest()` from the `{grf}` package to estimate heterogeneous returns to education (conditional LATE τ(X)).

```{r}
#| label: ivf-train-instrumental-forest
# Train the instrumental forest
iv_forest <- instrumental_forest(
  X = X,
  Y = Y,
  W = W,
  Z = Z,
  num.trees = 1000,          # Reduced for speed; use 2000+ in production
  honesty = TRUE,
  tune.parameters = "all"    # Auto-tune important hyperparameters
)

iv_forest
```

### Predicting Treatment Effects

```{r}
#| label: ivf-predict-oob
# Out-of-bag predictions (for training data)
iv_pred <- predict(iv_forest)
tau_hat <- iv_pred$predictions  # Estimated conditional LATE τ(X)

# Average treatment effect (doubly robust / average over forest)
ate <- average_treatment_effect(iv_forest)
cat("Average Treatment Effect (LATE):", round(ate$estimate, 3),
    "+/-", round(1.96 * ate$std.err, 3), "\n")
```

### Visualizing Results

```{r}
#| label: ivf-plot-tau-vs-exper
#| fig.width: 6
#| fig.height: 5
# Plot heterogeneous effects vs experience (example)
plot(card$exper, tau_hat, pch = 16, col = rgb(0,0,0,0.3),
     xlab = "Potential Experience (exper)", ylab = "Estimated Return to Education (τ(X))",
     main = "Heterogeneous Returns to Schooling by Experience")
abline(h = ate$estimate, col = "red", lwd = 2, lty = 2)
legend("topright", legend = "Average LATE", col = "red", lwd = 2, lty = 2)
```

### Predicting Treatment Effects for New Data

```{r}
#| label: ivf-predict-newdata
# Example: Predict for a hypothetical individual
new_data <- data.frame(
  exper = 10,
  expersq = 10^2,
  black = 0,
  smsa = 1,
  south = 0
)

new_pred <- predict(iv_forest, newdata = as.matrix(new_data))
cat("Predicted return to education (LATE) for this profile:", round(new_pred$predictions, 3), "\n")
```

### Evaluate Model Performance

```{r}
#| label: ivf-debiased-error
# Debiased error from tuning (lower = better)
tuning_info <- iv_forest$debiased.error
cat("Average debiased error (tuning):", round(mean(tuning_info, na.rm = TRUE), 4), "\n")
```

::: callout-note
- In practice, verify instrument validity (relevance + exclusion) carefully.
- Education is treated as continuous here (common in Card-style applications).
- Increase `num.trees` and tune more aggressively for production use.
- Check first-stage F-stat >10 (rule of thumb for strong instrument).
:::

### Compute Kernel SHAP values for CATE using shapviz

```{r}
#| label: compute-kernel-shap-iv
#| fig.width: 7
#| fig.height: 5
#| eval: true

library(kernelshap)
library(shapviz)
library(future)

#  Parallel backend 
n_cores <- max(1L, parallel::detectCores() - 1L)
plan(multisession, workers = n_cores)

#  Scale sizes to data
set.seed(42)
n_total <- nrow(X)

if (n_total > 5000L) {
  n_explain <- 400L
  n_bg      <- 80L

} else if (n_total > 1000L) {
  n_explain <- 250L
  n_bg      <- 60L

} else {
  n_explain <- min(150L, n_total)
  n_bg      <- min(50L,  n_total)
}

# Non-overlapping explain / background sets 
if (n_total >= n_explain + n_bg) {
  all_idx     <- sample(n_total)
  explain_idx <- all_idx[seq_len(n_explain)]
  bg_idx      <- all_idx[seq(n_explain + 1L, n_explain + n_bg)]

} else {
  # Small dataset — overlap allowed
  explain_idx <- sample(n_total, n_explain, replace = FALSE)
  bg_idx      <- sample(n_total, n_bg,      replace = FALSE)
}

X_explain <- X[explain_idx, , drop = FALSE]
bg_small  <- X[bg_idx,      , drop = FALSE]

# Prediction wrapper 
pred_tau <- function(object, newdata) {
  predict(object, newdata = newdata)$predictions
}

#  permshap instead of kernelshap (5–20x faster, same output) 
shap_iv <- permshap(
  object   = iv_forest,
  X        = X_explain,
  bg_X     = bg_small,
  pred_fun = pred_tau,
  verbose  = FALSE     # flip to TRUE to monitor progress
)

# Reset workers 
plan(sequential)
```

### Visualise SHAP values for the Instrumental Forest

```{r}
#| label: plot-shap-iv
# Wrap for plotting
shp_iv <- shapviz(shap_iv)

# Visualise 
sv_importance(shp_iv, kind = "both") +
  ggtitle("SHAP Importance: Drivers of Heterogeneous Returns to Education")

sv_dependence(shp_iv, v = "exper") +
  ggtitle("SHAP Dependence: Experience")

sv_dependence(shp_iv, v = "smsa") +
  ggtitle("SHAP Dependence: Metropolitan Area")

```


## Summary and Conclusion

Instrumental Forest is a powerful tool for estimating treatment effects in the presence of endogeneity, leveraging the strengths of random forests and instrumental variable analysis. It allows researchers to estimate heterogeneous treatment effects while addressing confounding issues, making it suitable for various applications in causal inference. This tutorial provided a step-by-step guide to implementing an Instrumental Forest using the `{grf}` package in R, demonstrating how to estimate heterogeneous returns to schooling using the classic Card (1995) dataset and college proximity as an instrument. At the end of the tutorial, we summarized the results, including the average treatment effect and SHAP explanations of heterogeneity. The code can be adapted to other real-world datasets with appropriate instruments and covariates.

## Resources

1. [Athey, Susan, Julie Tibshirani, and Stefan Wager. "Generalized Random Forests." Annals of Statistics, 47(2), 2019](https://grf-labs.github.io/grf/reference/instrumental_forest.html)
2. [Chernozhukov, Victor, and Christian Hansen. "The reduced form: A simple approach to inference with weak instruments." Economics Letters, 100(1), 2008.](https://github.com/grf-labs/grf/issues/501)
3. [Wang, Guihua, et al. "An Instrumental Variable Forest Approach for Detecting Heterogeneous Treatment Effects in Observational Studies." Management Science, 2021](https://pubsonline.informs.org/doi/abs/10.1287/mnsc.2021.4084)
4. Card, David (1995). "Using Geographic Variation in College Proximity to Estimate the Return to Schooling." In Aspects of Labour Market Behaviour: Essays in Honour of John Vanderkamp.

This version keeps the original format, structure, markdown/callouts, and theoretical content intact while fully transitioning the empirical example to the **Card dataset** for a realistic, interpretable IV application.
