# 5. Double Machine Learning with Instrumental Variables {.unnumbered}

In this tutorial we implement **DML-IV** in R using **RCausalML**’s native `DMLIVLearner()` (defined in the package source file `R/DMLearner.R`). We use **synthetic data** to show why a naive regression of the outcome on the treatment is biased, then fit **DML IV** and **standard DML (without the instrument)** with several first-stage learners—**ordinary least squares (`"lm"`)**, **random forest (`"ranger"`)**, and **gradient boosting (`"xgb"`)**—and compare **bias relative to the true effect**, **uncertainty**, and **runtime**. **Part B** repeats the same model comparison on the **401(k) IV** sample (`fetch_401k(..., instrument = TRUE)` from **DoubleML**), with **eligibility** (`e401`) instrumenting **participation** (`p401`). The statistical ideas align with the [DoubleML R: Basic Instrumental Variables Calculation](https://docs.doubleml.org/stable/examples/R_double_ml_basic_iv.html) and with **two-stage least squares (2SLS)** / sieve IV perspectives (e.g., [EconML Sieve 2SLS](https://www.pywhy.org/EconML/spec/estimation/two_sls.html)); here the implementation is **RCausalML only**—no `DoubleML` package required.


## Overview

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Instrumental Variable (IV) | A variable that affects treatment but does not directly affect the outcome. | "Lottery assignment changes program attendance but not earnings directly." |
| Endogeneity | When treatment is correlated with unobserved factors affecting outcome. | "More motivated people choose treatment and also perform better." |
| PLIV (Partially Linear IV model) | DML model for causal effects with an instrument and covariates. | "Estimate effect of education on wages using distance-to-college as IV." |
| Relevance (IV condition) | Instrument must strongly predict treatment. | "If IV barely changes treatment, estimates become unstable." |
| Exclusion restriction | Instrument affects outcome only through treatment. | "The instrument should not have its own direct path to the outcome." |
| Local Average Treatment Effect (LATE) | Effect for compliers, i.e., people whose treatment changes because of IV. | "Effect among students who attend college only because of distance changes." |

**Double machine learning with instrumental variables (DML-IV)** combines IV identification with double/debiased machine learning: an instrument $Z$ addresses endogeneity of the treatment $D$, while ML (or linear models) and cross-fitting estimate nuisance functions without overfitting bias. RCausalML’s `DMLIVLearner()` builds cross-fitted $\hat{\mathbb{E}}[Y \mid X,W]$, $\hat{\mathbb{E}}[T \mid X,W]$, and $\hat{\mathbb{E}}[T \mid X,W,Z]$, forms the DMLIV pseudo-outcome, and fits a **linear** CATE in $X$ (see `?DMLIVLearner`).

**Key idea**: DML-IV uses cross-fitting and flexible first stages to estimate nuisances, then combines them in a pseudo-outcome regression on $X$ for the effect of $D$ on $Y$ (e.g., a **local average treatment effect**, LATE, under standard IV assumptions).

In the basic IV DAG:

- **$Z$** is the instrument (assumed to affect $D$, and to affect $Y$ only through $D$).
- **$C$** are unobserved confounders (e.g., motivation, ability).
- **$D$** is the treatment or decision variable.
- **$Y$** is the outcome.

A **naive** regression of $Y$ on $D$ is biased because $D$ is correlated with $C$. **2SLS** and **sieve 2SLS** project $D$ onto $Z$ (and covariates) in the first stage; **DML IV** in RCausalML generalizes the nuisance steps with optional ML (`"ranger"`, `"glmnet"`, `"xgb"`) and cross-fitting (`cv`).

**Standard DML** (`LinearDML`) uses the same cross-fitted nuisance idea **without** an instrument. It does **not** correct for bias from **unobserved** confounders that affect both $D$ and $Y$; we include it here so you can see **IV vs non-IV** side by side for each learner.

## Implementation in R

The workflow is:

1. **Simulate IV data** with a binary instrument, binary treatment, continuous outcome, and unobserved confounders.
2. **Optional nuisance covariates** `W`: extra columns for the first stages so tree methods (`"ranger"`, `"xgb"`) have features to learn (with only an intercept, random forests are not meaningful).
3. **Naive estimation**: Show that a simple mean comparison is biased.
4. **Fit `DMLIVLearner()`** (with $Z$) and **`LinearDML()`** (without $Z$) for each first-stage string (`"lm"`, `"ranger"`, `"xgb"`).
5. **Compare** estimates to the true effect, standard errors, and elapsed time.

### Setup and Libraries

```{r}
#| label: setup-packages
#| code-summary: "Packages and optional dependency messages"
packages <- c("RCausalML",
              "tidyverse")
if (!requireNamespace("DoubleML", quietly = TRUE)) {
  message("Install suggested package 'DoubleML' for Part B (`fetch_401k`): install.packages(\"DoubleML\")")
}
if (!requireNamespace("xgboost", quietly = TRUE)) {
  message("Install optional package 'xgboost' for the `\"xgb\"` nuisance rows: install.packages(\"xgboost\")")
}
if (!requireNamespace("kernelshap", quietly = TRUE) || !requireNamespace("shapviz", quietly = TRUE)) {
  message("Install optional packages 'kernelshap' and 'shapviz' for Part B SHAP: install.packages(c(\"kernelshap\", \"shapviz\"))")
}
```

### Install Missing Packages

```{r}
#| label: install-packages
#| code-summary: "Install missing packages (chunk eval disabled by default)"
#| eval: false
#| warning: false
#| error: false
new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)
```

### Load Libraries and Options

```{r}
#| label: load-libraries
#| code-summary: "Load libraries (RCausalML, tidyverse, optional DoubleML / SHAP)"
invisible(lapply(packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))
if (requireNamespace("DoubleML", quietly = TRUE)) {
  suppressPackageStartupMessages(library(DoubleML))
}
if (requireNamespace("kernelshap", quietly = TRUE)) {
  suppressPackageStartupMessages(library(kernelshap))
}
if (requireNamespace("shapviz", quietly = TRUE)) {
  suppressPackageStartupMessages(library(shapviz))
}
```

## Part A: DML-IV with Synthetic Data

### IV Data-Generation

We generate synthetic data compatible with the IV DAG: a binary unobserved confounder $C$, a binary instrument $Z$, a binary treatment $D$ (influenced by $Z$ and $C$), and a continuous outcome $Y$. The target is the **decision effect** (impact of $D$ on $Y$).

We also draw **noise covariates** `W` (not in the structural outcome equation) so that **random forest** and **XGBoost** nuisances in `DMLIVLearner(..., W = W)` and `LinearDML(..., X = W)` see a nontrivial design matrix. The second stage for IV remains a **constant** CATE (`X = NULL`); for non-IV DML we summarize performance by the **sample mean** of the fitted CATE $\hat\theta(W_i)$ (heterogeneity driven only by `W`).


```{r}
#| label: iv-data-generate
#| code-summary: "Simulate synthetic IV data (instrument, treatment, outcome, W)"
n <- 10000
decision_effect <- -2
instrument_effect <- 0.7

set.seed(1234)

confounder <- rbinom(n, 1, 0.3)
instrument <- rbinom(n, 1, 0.5)
decision <- as.numeric(runif(n) <= instrument_effect * instrument + 0.4 * confounder)
outcome <- 30 + decision_effect * decision + 10 * confounder + rnorm(n, sd = 2)

W_ctrl <- matrix(stats::rnorm(n * 3L), nrow = n, ncol = 3L)
colnames(W_ctrl) <- paste0("W", 1:3)

df <- data.frame(instrument, decision, outcome)
```

### Naive Estimation (Biased)

Estimating the effect of `decision` on `outcome` by the difference in mean outcomes between decision groups ignores endogeneity. Because the confounder $C$ affects both $D$ and $Y$, this difference is **biased** (we expect a value near 1 here, not the true effect $-2$).

```{r}
#| label: naive-ate
#| code-summary: "Naive ATE (difference in mean outcome by treatment)"
naive_ate <- mean(df[df$decision == 1, "outcome"]) - mean(df[df$decision == 0, "outcome"])
naive_ate
```

The naive estimate is far from the true `decision_effect = -2`. We need IV (or DML-IV) to recover the causal effect.

### Comparison: IV (`DMLIVLearner`) vs non-IV (`LinearDML`) for `lm`, `ranger`, and `xgb`

- **With IV**: `DMLIVLearner` uses `Z = instrument`, first stages on `(W)` (and implicit intercept via `dml_iv_XW`), and a **linear** second stage on `X = NULL` (constant effect). We read $\hat\theta$ and its SE from `summary(fitted$model_cate$model)`.
- **Without IV**: `LinearDML` fits the usual **partially linear DML** with the same `Y`, `T`, and nuisance features `X = W_ctrl` (CATE may vary with `W`; we report the **mean** $\frac{1}{n}\sum_i \hat\theta(W_i)$ as a scalar summary and an SE $\sqrt{\widehat{\mathrm{Var}}(\hat\theta(W_i))/n}$).

Both use `discrete_treatment = TRUE` for IV (binary $T$ nuisances); `LinearDML` treats binary treatment via its default binary path. Cross-fitting uses `cv = 3` and `num.trees = 200` for ranger to keep the vignette build fast.

```{r}
#| label: dml-iv-vs-linear-dml
#| code-summary: "Compare DML-IV vs LinearDML (lm, ranger, xgb nuisances)"
#| message: false
#| warning: false
truth <- decision_effect
cv_fold <- 3L
rf_trees <- 200L

learners <- c("lm", "ranger", "xgb")
if (!requireNamespace("xgboost", quietly = TRUE)) {
  learners <- setdiff(learners, "xgb")
}

se_mean_cate <- function(fit, X) {
  te <- predict(fit, X)
  sqrt(stats::var(as.vector(te), na.rm = TRUE) / length(as.vector(te)))
}

run_one <- function(learner, use_iv) {
  tm <- system.time({
    if (use_iv) {
      mod <- DMLIVLearner(
        model_y_xw = learner,
        model_t_xw = learner,
        model_t_xwz = learner,
        fit_cate_intercept = FALSE,
        discrete_treatment = TRUE,
        cv = cv_fold,
        random_state = 1234L
      )
      fit_obj <- fit(
        mod,
        Y = df$outcome,
        T = df$decision,
        Z = df$instrument,
        X = NULL,
        W = W_ctrl,
        num.trees = rf_trees
      )
      sm <- summary(fit_obj$model_cate$model)$coefficients
      est <- as.numeric(sm[1L, "Estimate"])
      se <- as.numeric(sm[1L, "Std. Error"])
    } else {
      mod <- LinearDML(
        model_y = learner,
        model_t = learner,
        fit_cate_intercept = TRUE,
        n_fold = cv_fold,
        seed = 1234L
      )
      fit_obj <- fit(
        mod,
        X = W_ctrl,
        treatment = df$decision,
        y = df$outcome,
        num.trees = rf_trees
      )
      est <- mean(as.vector(predict(fit_obj, W_ctrl)))
      se <- se_mean_cate(fit_obj, W_ctrl)
    }
  })
  tibble::tibble(
    learner = learner,
    instrumental_variable = use_iv,
    estimate = est,
    std_error = se,
    abs_bias = abs(est - truth),
    time_sec = as.numeric(tm["elapsed"])
  )
}

cmp <- purrr::map_dfr(learners, function(lrn) {
  dplyr::bind_rows(
    run_one(lrn, use_iv = TRUE),
    run_one(lrn, use_iv = FALSE)
  )
})

cmp <- cmp |>
  dplyr::mutate(
    learner = dplyr::case_when(
      learner == "lm" ~ "lm",
      learner == "ranger" ~ "ranger (random forest)",
      learner == "xgb" ~ "xgboost",
      TRUE ~ as.character(learner)
    ),
    instrumental_variable = ifelse(instrumental_variable, "yes (DML-IV)", "no (Linear DML)")
  ) |>
  dplyr::arrange(learner, dplyr::desc(instrumental_variable))

cmp
```

```{r}
#| label: tbl-comparison-results
#| code-summary: "Table: synthetic-data comparison (hidden code)"
#| echo: false
#| tbl-cap: "Point estimates, standard errors, absolute bias vs true effect, and elapsed time. IV-based rows should be near the true effect; non-IV rows stay biased by unobserved confounding."
knitr::kable(
  cmp,
  digits = 3,
  col.names = c(
    "First-stage learner",
    "Uses instrument Z",
    "Estimate",
    "Std. error",
    "|Bias|",
    "Time (s)"
  )
)
```

```{r}
#| label: fig-abs-bias
#| code-summary: "Figure: absolute bias by learner (IV vs non-IV)"
#| fig-cap: "Absolute bias |estimate - true effect| for each learner, split by whether the instrument is used."
ggplot2::ggplot(cmp, ggplot2::aes(x = learner, y = abs_bias, fill = instrumental_variable)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(width = 0.8), width = 0.75) +
  ggplot2::labs(
    x = NULL,
    y = "Absolute bias",
    fill = NULL
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "top")
```

**Takeaway**: For every first-stage learner, **DML-IV** (instrument used) tracks the true `decision_effect` of **-2**, while **Linear DML without the instrument** remains near the **naive** confounded comparison (around **1**). Random forest and XGBoost nuisances behave sensibly once `W` is supplied; **absolute bias** and **runtime** let you compare methods on this simulation.

### Single IV fit (reference)

The following reproduces the minimal **linear** nuisance specification from the DoubleML-style workflow (`lm` only, no `W`).

```{r}
#| label: iv-fit-lm-reference
#| code-summary: "Reference DML-IV fit (linear nuisances, no W)"
iv_mod_lm <- DMLIVLearner(
  model_y_xw = "lm",
  model_t_xw = "lm",
  model_t_xwz = "lm",
  fit_cate_intercept = FALSE,
  discrete_treatment = TRUE,
  cv = 5L,
  random_state = 1234L
)

iv_fit_lm <- fit(
  iv_mod_lm,
  Y = df$outcome,
  T = df$decision,
  Z = df$instrument,
  X = NULL
)

summary(iv_fit_lm$model_cate$model)
```

For richer DGPs, set `model_*` to `"ranger"`, `"glmnet"`, or `"xgb"` (defaults map `"auto"` to `"ranger"`). For a **nonlinear** CATE in $X$, see `NonParamDMLIVLearner()` in the same source file.

### DML-IV Results and Summary

The **linear second stage** is stored as `iv_fit$model_cate$model` (an `lm` object). The coefficient named `X1` is the estimated constant effect when `X` is omitted (internal column of ones).

```{r}
#| label: iv-summary-coefficients
#| code-summary: "Second-stage coefficients (DML-IV lm reference)"
sm <- summary(iv_fit_lm$model_cate$model)$coefficients
data.frame(
  estimate = sm[, "Estimate"],
  std_error = sm[, "Std. Error"],
  t_value = sm[, "t value"],
  p_value = sm[, "Pr(>|t|)"]
)
```

The **DML-IV estimate** for the effect of `decision` on `outcome` is close to the true value $-2$, with a standard error from the final `lm`. The naive estimate was biased; the IV-based DML estimate recovers the estimand under the usual IV assumptions.

### Comparison and Interpretation

| Estimator   | Estimate (approx.) | Interpretation |
|------------|--------------------|----------------|
| Naive      | $\approx 1$        | Biased (confounding) |
| DMLIV (`DMLIVLearner`)   | $\approx -1.9$     | Constant CATE / LATE-style estimand |

The DMLIV construction targets an IV estimand comparable in spirit to the **LATE** when $Z$ and $D$ are binary; with only a constant effect in $X$, the second-stage coefficient summarizes the average causal effect recovered by the instrument.

## Part B: DML-IV with the 401(k) data

We use the **401(k) dataset** bundled with **DoubleML** (`fetch_401k()`; Chernozhukov et al., 2018, 2020) in the **instrumental-variables** form used in the replication code: **outcome** $Y$ is `net_tfa` (net total financial assets), **endogenous treatment** $D$ is `p401` (participation in a 401(k)), and **instrument** $Z$ is `e401` (employer offers a plan). Covariates $X$ are age, income, family size, education, pension and household indicators (`age`, `inc`, `fsize`, `educ`, `db`, `marr`, `twoearn`, `pira`, `hown`). Under standard IV assumptions, DML-IV targets a **LATE**-style estimand for the effect of participation on wealth, using eligibility as excluded variation.

We fit on an **80% / 20%** train split **stratified on `p401`** so both folds keep similar participation rates, and **compare all models** from Part A on the training rows: **naive** difference in mean `net_tfa` by `p401`, **`DMLIVLearner()`** with $Z =$ `e401` (nuisances on covariates `W`; constant second stage, `X = NULL`), and **`LinearDML()`** without the instrument (partially linear DML for the effect of `p401` on `net_tfa` given $X$ only—**not** the same causal estimand as IV if participation is endogenous). First-stage learners are again **`"lm"`**, **`"ranger"`**, and **`"xgb"`** where available.

### Data

```{r}
#| label: real-load-data
#| code-summary: "Load 401(k) data (instrument = TRUE: p401, e401)"
#| message: false
#| warning: false
if (!requireNamespace("DoubleML", quietly = TRUE)) {
  stop("Install DoubleML for fetch_401k: install.packages(\"DoubleML\")")
}

df_401k <- fetch_401k(return_type = "data.frame", instrument = TRUE)

# Y = net_tfa; D = p401 (participation); Z = e401 (eligibility); X = covariates
x_cols_401k <- c("age", "inc", "fsize", "educ", "db", "marr", "twoearn", "pira", "hown")

cat("Dimensions:", nrow(df_401k), "x", ncol(df_401k), "\n")
head(df_401k)
```

### Train–test split and model comparison (`lm`, `ranger`, `xgb`)

```{r}
#| label: real-dml-401k-compare
#| code-summary: "401(k): train split, naive ATE, DML-IV vs LinearDML loop"
#| message: false
#| warning: false
X_401k <- as.matrix(df_401k[, x_cols_401k, drop = FALSE])
y_401k <- df_401k$net_tfa
t_401k <- df_401k$p401
z_401k <- df_401k$e401

set.seed(3141)
train_frac <- 0.8
i1 <- which(t_401k == 1L)
i0 <- which(t_401k == 0L)
n1_tr <- max(1L, floor(train_frac * length(i1)))
n0_tr <- max(1L, floor(train_frac * length(i0)))
train_idx <- c(sample(i1, n1_tr), sample(i0, n0_tr))
test_idx <- setdiff(seq_along(t_401k), train_idx)

W_tr <- X_401k[train_idx, , drop = FALSE]
y_tr <- y_401k[train_idx]
t_tr <- t_401k[train_idx]
z_tr <- z_401k[train_idx]

cat(
  "Train n =", length(train_idx), "| Test n =", length(test_idx),
  "| mean(p401) train =", round(mean(t_tr), 3),
  "| mean(e401) train =", round(mean(z_tr), 3), "\n"
)

cv_fold <- 3L
rf_trees <- 200L

learners_401k <- c("lm", "ranger", "xgb")
if (!requireNamespace("xgboost", quietly = TRUE)) {
  learners_401k <- setdiff(learners_401k, "xgb")
}

naive_401k <- mean(y_tr[t_tr == 1L]) - mean(y_tr[t_tr == 0L])

se_mean_cate <- function(fit, X) {
  te <- predict(fit, X)
  sqrt(stats::var(as.vector(te), na.rm = TRUE) / length(as.vector(te)))
}

run_one_401k <- function(learner, use_iv) {
  tm <- system.time({
    if (use_iv) {
      mod <- DMLIVLearner(
        model_y_xw = learner,
        model_t_xw = learner,
        model_t_xwz = learner,
        fit_cate_intercept = FALSE,
        discrete_treatment = TRUE,
        discrete_instrument = TRUE,
        cv = cv_fold,
        random_state = 3141L
      )
      fit_obj <- fit(
        mod,
        Y = y_tr,
        T = t_tr,
        Z = z_tr,
        X = NULL,
        W = W_tr,
        num.trees = rf_trees
      )
      sm <- summary(fit_obj$model_cate$model)$coefficients
      est <- as.numeric(sm[1L, "Estimate"])
      se <- as.numeric(sm[1L, "Std. Error"])
    } else {
      mod <- LinearDML(
        model_y = learner,
        model_t = learner,
        fit_cate_intercept = TRUE,
        n_fold = cv_fold,
        seed = 3141L
      )
      fit_obj <- fit(
        mod,
        X = W_tr,
        treatment = t_tr,
        y = y_tr,
        num.trees = rf_trees
      )
      est <- mean(as.vector(predict(fit_obj, W_tr)))
      se <- se_mean_cate(fit_obj, W_tr)
    }
  })
  tibble::tibble(
    learner = learner,
    instrumental_variable = use_iv,
    estimate = est,
    std_error = se,
    time_sec = as.numeric(tm["elapsed"])
  )
}

cmp_401k <- purrr::map_dfr(learners_401k, function(lrn) {
  dplyr::bind_rows(
    run_one_401k(lrn, use_iv = TRUE),
    run_one_401k(lrn, use_iv = FALSE)
  )
})

cmp_401k <- cmp_401k |>
  dplyr::mutate(
    learner = dplyr::case_when(
      learner == "lm" ~ "lm",
      learner == "ranger" ~ "ranger (random forest)",
      learner == "xgb" ~ "xgboost",
      TRUE ~ as.character(learner)
    ),
    instrumental_variable = ifelse(
      instrumental_variable,
      "yes (DML-IV)",
      "no (Linear DML)"
    )
  ) |>
  dplyr::arrange(learner, dplyr::desc(instrumental_variable))

cmp_naive <- tibble::tibble(
  learner = "naive (mean diff.)",
  instrumental_variable = "—",
  estimate = naive_401k,
  std_error = NA_real_,
  time_sec = NA_real_
)

cmp_401k_all <- dplyr::bind_rows(cmp_naive, cmp_401k)

cmp_401k_all
```

### Model comparison

```{r}
#| label: tbl-401k-comparison
#| code-summary: "Table: 401(k) comparison (hidden code)"
#| echo: false
#| tbl-cap: "401(k) training sample: naive mean difference, DML-IV (e401 instruments p401), and Linear DML without Z for each nuisance learner. Estimates are in the same units as net_tfa; standard errors are omitted for the naive row."
knitr::kable(
  cmp_401k_all,
  digits = 3,
  col.names = c(
    "Model / first-stage learner",
    "Uses instrument Z (e401)",
    "Estimate",
    "Std. error",
    "Time (s)"
  )
)
```

### Visualize the Results

```{r}
#| label: fig-401k-estimates
#| code-summary: "Figure: 401(k) point estimates with error bars"
#| fig-cap: "Point estimates on the training subsample: naive mean difference and, for each first-stage learner, DML-IV vs Linear DML without the instrument."
est_plot <- cmp_401k_all |>
  dplyr::filter(learner != "naive (mean diff.)") |>
  dplyr::mutate(
    series = paste(learner, instrumental_variable, sep = " — ")
  )
est_plot <- dplyr::bind_rows(
  cmp_401k_all |>
    dplyr::filter(learner == "naive (mean diff.)") |>
    dplyr::mutate(series = learner),
  est_plot
)

ggplot2::ggplot(est_plot, ggplot2::aes(x = series, y = estimate, fill = series)) +
  ggplot2::geom_col(width = 0.75) +
  ggplot2::geom_errorbar(
    ggplot2::aes(ymin = estimate - std_error, ymax = estimate + std_error),
    width = 0.2,
    na.rm = TRUE
  ) +
  ggplot2::coord_flip() +
  ggplot2::labs(x = NULL, y = "Estimate (net_tfa units)", fill = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

**Reading the table**: **DML-IV** rows use eligibility `e401` as an instrument for participation `p401`, with nuisances on the covariate matrix `W`. **Linear DML** rows ignore `e401` and estimate a partially linear effect of `p401` on `net_tfa` given $X$ (summary is the mean predicted CATE on the training fold). The **naive** row is the raw mean outcome difference by `p401` on the same training rows. Magnitudes and ordering across learners illustrate stability of the IV pipeline vs the non-IV and naive comparisons on real data (there is no simulation truth here).

### SHAP Analysis

This section shows how to compute and interpret SHAP values for **predicted** conditional treatment effects $\hat{\tau}(X)$ from the IV and partial-linear DML fits. We use:

- **`explain_cate()`** — wrapper integrating **{RCausalML}** with **{kernelshap}** / permutation SHAP  
- **{shapviz}** — importance, dependence, waterfall, and force plots  

These summaries are **descriptive** (how the fitted surface attributes $\hat{\tau}(X)$ to covariates), not additional causal claims. [SHAP](https://github.com/shap/shap) is defined for **model predictions**; see `R/shapviz_integration.R` for the implementation.

The comparison table above uses a **scalar** DML-IV fit (`X = NULL`). For SHAP we need $\hat{\tau}(X)$ that **varies with covariates**, so we refit **`DMLIVLearner()`** with a **linear heterogeneous** second stage (`X =` covariates, `W = NULL`) while keeping $Z =$ `e401` and $D =$ `p401`, and we pair that with **`LinearDML()`** (heterogeneous CATE in $X$). The **naive** mean difference is not a covariate-based predictor, so it is **omitted**.

Because SHAP is expensive, we use **permutation SHAP** (`use_permshap = TRUE`) on a **small explanation subsample** and a **background** sample from the training fold. **Use non-overlapping rows** for explanation vs background when the pool is large enough, to avoid leakage between the points being explained and the reference distribution. When **{future}** is available, **`permshap(..., parallel = TRUE)`** can use multiple workers (see `?kernelshap::permshap`).

#### 1. SHAP estimation

SHAP values describe how each covariate contributes to the **individual** predicted CATE on the explanation rows — i.e. which features push $\hat{\tau}(X)$ up or down for that observation.

```{r}
#| label: shap-401k-has
#| code-summary: "Detect kernelshap and shapviz"
#| message: false
#| warning: false
#| dependson: real-dml-401k-compare

has_shap <- requireNamespace("kernelshap", quietly = TRUE) &&
  requireNamespace("shapviz", quietly = TRUE)
```

```{r}
#| label: shap-401k-fit
#| code-summary: "SHAP: refit DML-IV (heterogeneous X) and LinearDML on training data"
#| message: false
#| warning: false
#| dependson: shap-401k-has

if (has_shap) {
  W_tr_n <- W_tr
  colnames(W_tr_n) <- x_cols_401k

  fit_shap_dmliv <- function(learner) {
    mod <- DMLIVLearner(
      model_y_xw = learner,
      model_t_xw = learner,
      model_t_xwz = learner,
      fit_cate_intercept = TRUE,
      discrete_treatment = TRUE,
      discrete_instrument = TRUE,
      cv = cv_fold,
      random_state = 3141L
    )
    fit(
      mod,
      Y = y_tr,
      T = t_tr,
      Z = z_tr,
      X = W_tr_n,
      W = NULL,
      num.trees = rf_trees
    )
  }

  fit_shap_ldml <- function(learner) {
    mod <- LinearDML(
      model_y = learner,
      model_t = learner,
      fit_cate_intercept = TRUE,
      n_fold = cv_fold,
      seed = 3141L
    )
    fit(mod, X = W_tr_n, treatment = t_tr, y = y_tr, num.trees = rf_trees)
  }

  fits_iv <- stats::setNames(lapply(learners_401k, fit_shap_dmliv), learners_401k)
  fits_pl <- stats::setNames(lapply(learners_401k, fit_shap_ldml), learners_401k)
}
```

```{r}
#| label: shap-401k-pool
#| code-summary: "SHAP: distinct moderator rows + non-overlapping explain vs background"
#| message: false
#| warning: false
#| dependson: shap-401k-fit

if (!has_shap) {
  message("Skipping SHAP: install.packages(c(\"kernelshap\", \"shapviz\"))")
} else {
  W_pool <- as.data.frame(W_tr)
  colnames(W_pool) <- x_cols_401k
  X_shap_pool <- W_pool[!duplicated(W_pool), , drop = FALSE]

  set.seed(3150L)
  nexplain <- min(80L, nrow(X_shap_pool))
  nbg <- min(50L, nrow(X_shap_pool))

  if (nrow(X_shap_pool) > nexplain + nbg) {
    idxexplain <- sample.int(nrow(X_shap_pool), nexplain)
    idxbg <- sample(setdiff(seq_len(nrow(X_shap_pool)), idxexplain), nbg)
  } else {
    idxexplain <- sample.int(nrow(X_shap_pool), nexplain)
    idxbg <- sample.int(nrow(X_shap_pool), nbg)
  }

  X_explain <- X_shap_pool[idxexplain, , drop = FALSE]
  bg_X_df <- X_shap_pool[idxbg, , drop = FALSE]
}
```

```{r}
#| label: shap-401k-explain
#| code-summary: "SHAP: permshap via explain_cate (tryCatch + reset future)"
#| message: false
#| warning: false
#| dependson: shap-401k-pool

if (has_shap && exists("X_explain")) {
  n_workers <- if (requireNamespace("future", quietly = TRUE)) {
    min(4L, max(1L, future::availableCores() - 1L))
  } else {
    1L
  }
  use_parallel <- n_workers > 1L && requireNamespace("future", quietly = TRUE)

  fg_prev <- getOption("future.globals.maxSize", default = 500 * 1024^2)
  if (use_parallel) {
    options(future.globals.maxSize = max(2 * 1024^3, fg_prev))
    future::plan(future::multisession, workers = n_workers)
  }

  explain_one <- function(fit_obj, label) {
    tryCatch(
      {
        args <- list(
          fit_obj,
          X = X_explain,
          bg_X = bg_X_df,
          use_permshap = TRUE,
          verbose = FALSE,
          max_iter = 60L,
          tol = 0.02,
          parallel = use_parallel,
          parallel_args = list(packages = c("RCausalML", "ranger", "xgboost"))
        )
        if (!requireNamespace("xgboost", quietly = TRUE)) {
          args$parallel_args$packages <- setdiff(args$parallel_args$packages, "xgboost")
        }
        do.call(explain_cate, args)
      },
      error = function(e) {
        warning("SHAP failed for ", label, ": ", conditionMessage(e))
        NULL
      }
    )
  }

  tryCatch(
    {
      ks_iv <- mapply(
        explain_one,
        fits_iv,
        paste0("DML-IV (", names(fits_iv), ")"),
        SIMPLIFY = FALSE
      )
      ks_pl <- mapply(
        explain_one,
        fits_pl,
        paste0("Linear DML (", names(fits_pl), ")"),
        SIMPLIFY = FALSE
      )
      shp_iv <- lapply(ks_iv, function(k) {
        if (!is.null(k)) shapviz::shapviz(k) else NULL
      })
      shp_pl <- lapply(ks_pl, function(k) {
        if (!is.null(k)) shapviz::shapviz(k) else NULL
      })
    },
    finally = {
      if (use_parallel) {
        future::plan(future::sequential)
        options(future.globals.maxSize = fg_prev)
      }
    }
  )
}
```

#### 2. Visualize SHAP values

**2.1 Beeswarm importance** — distribution of SHAP values per feature; large |SHAP| means a strong influence on $\hat{\tau}(X)$ for the explanation sample. Below we show **ranger** nuisances side by side (DML-IV vs Linear DML).

```{r}
#| label: fig-shap-401k-beeswarm-ranger
#| code-summary: "Figure: SHAP beeswarm — ranger DML-IV vs Linear DML"
#| fig-cap: "Beeswarm SHAP summary for predicted $\\hat{\\tau}(X)$ (ranger nuisances): DML-IV vs Linear DML."
#| fig-width: 8
#| fig-height: 5
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_iv") && exists("shp_pl")) {
  if (!is.null(shp_iv[["ranger"]]) && !is.null(shp_pl[["ranger"]])) {
    b1 <- shapviz::sv_importance(shp_iv[["ranger"]], kind = "beeswarm") +
      ggplot2::labs(title = "DML-IV (ranger)")
    b2 <- shapviz::sv_importance(shp_pl[["ranger"]], kind = "beeswarm") +
      ggplot2::labs(title = "Linear DML (ranger)")
    if (requireNamespace("patchwork", quietly = TRUE)) {
      patchwork::wrap_plots(b1, b2, ncol = 2L)
    } else {
      print(b1)
      print(b2)
    }
  }
}
```

**2.2 Bar importance (mean |SHAP|)** — good for ranking features. **All** nuisance learners: DML-IV and Linear DML panels. Values are contributions to predicted $\hat{\tau}(X)$ on the explanation subsample, not to the scalar ATE in the table above.

```{r}
#| label: fig-shap-401k-dmliv
#| code-summary: "Figure: SHAP bar importance — DML-IV (hidden code)"
#| fig-cap: "SHAP variable importance for predicted heterogeneous CATE from DML-IV (e401 instruments p401), linear second stage in covariates."
#| fig-width: 10
#| fig-height: 3.5
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_iv") && length(shp_iv)) {
  plots_iv <- lapply(seq_along(shp_iv), function(i) {
    if (is.null(shp_iv[[i]])) {
      return(NULL)
    }
    nm <- names(shp_iv)[i]
    shapviz::sv_importance(shp_iv[[i]], max_display = 9L, show_numbers = TRUE) +
      ggplot2::labs(title = paste0("DML-IV (", nm, ")")) +
      ggplot2::theme(plot.title = ggplot2::element_text(size = 10))
  })
  plots_iv <- Filter(Negate(is.null), plots_iv)
  if (length(plots_iv)) {
    if (requireNamespace("patchwork", quietly = TRUE)) {
      patchwork::wrap_plots(plots_iv, ncol = 3L)
    } else {
      for (g in plots_iv) print(g)
    }
  }
}
```

```{r}
#| label: fig-shap-401k-lineardml
#| code-summary: "Figure: SHAP bar importance — Linear DML (hidden code)"
#| fig-cap: "SHAP variable importance for predicted CATE from Linear DML (no instrument), same nuisance learners."
#| fig-width: 10
#| fig-height: 3.5
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_pl") && length(shp_pl)) {
  plots_pl <- lapply(seq_along(shp_pl), function(i) {
    if (is.null(shp_pl[[i]])) {
      return(NULL)
    }
    nm <- names(shp_pl)[i]
    shapviz::sv_importance(shp_pl[[i]], max_display = 9L, show_numbers = TRUE) +
      ggplot2::labs(title = paste0("Linear DML (", nm, ")")) +
      ggplot2::theme(plot.title = ggplot2::element_text(size = 10))
  })
  plots_pl <- Filter(Negate(is.null), plots_pl)
  if (length(plots_pl)) {
    if (requireNamespace("patchwork", quietly = TRUE)) {
      patchwork::wrap_plots(plots_pl, ncol = 3L)
    } else {
      for (g in plots_pl) print(g)
    }
  }
}
```

#### 3. SHAP dependence plots

How each feature drives attributed CATE on the explanation sample (colour may encode another variable). **Ranger** nuisances: full covariate set, shared $y$-axis for comparability.

```{r}
#| label: fig-shap-401k-dependence-ranger
#| code-summary: "Figure: SHAP dependence — all covariates, ranger (hidden code)"
#| fig-cap: "SHAP dependence for ranger-based DML-IV and Linear DML (predicted $\\hat{\\tau}(X)$)."
#| fig-width: 8
#| fig-height: 8
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_iv") && exists("shp_pl")) {
  if (!is.null(shp_iv[["ranger"]]) && !is.null(shp_pl[["ranger"]])) {
    xv <- colnames(shp_iv[["ranger"]]$X)
    d1 <- shapviz::sv_dependence(shp_iv[["ranger"]], v = xv, share_y = TRUE) +
      ggplot2::labs(title = "DML-IV (ranger)")
    d2 <- shapviz::sv_dependence(shp_pl[["ranger"]], v = xv, share_y = TRUE) +
      ggplot2::labs(title = "Linear DML (ranger)")
    if (requireNamespace("patchwork", quietly = TRUE)) {
      patchwork::wrap_plots(d1, d2, ncol = 1L)
    } else {
      print(d1)
      print(d2)
    }
  }
}
```

*Example read*: if SHAP for `age` increases with `age`, the fitted surface attributes higher $\hat{\tau}(X)$ to older units on this subsample (exploratory; not a standalone causal statement).

#### 4. SHAP for a single prediction

Waterfall and force plots show how the baseline and each feature contribution build up to one row’s predicted CATE. We use the **ranger** DML-IV object when available.

```{r}
#| label: fig-shap-401k-single
#| code-summary: "Figure: waterfall and force for one row (ranger DML-IV)"
#| fig-cap: "Single-row SHAP decomposition for predicted $\\hat{\\tau}(X)$ (ranger DML-IV)."
#| fig-width: 6
#| fig-height: 5
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_iv") && !is.null(shp_iv[["ranger"]])) {
  shapviz::sv_waterfall(shp_iv[["ranger"]], row_id = 1L)
  shapviz::sv_force(shp_iv[["ranger"]], row_id = 1L)
}
```

#### 5. Subgroup-specific SHAP (optional)

Example: waterfall for the first training row with **`marr == 1`** (married), if that column exists in the explanation matrix.

```{r}
#| label: fig-shap-401k-subgroup
#| code-summary: "Figure: waterfall for first married row (ranger DML-IV)"
#| fig-width: 6
#| fig-height: 5
#| echo: false
#| dependson: shap-401k-explain

if (has_shap && exists("shp_iv") && !is.null(shp_iv[["ranger"]])) {
  sx <- shp_iv[["ranger"]]$X
  if ("marr" %in% names(sx)) {
    idx_m <- which(sx$marr == 1L)
    row_show <- if (length(idx_m) > 0L) idx_m[1L] else 1L
  } else {
    row_show <- 1L
  }
  shapviz::sv_waterfall(shp_iv[["ranger"]], row_id = row_show)
}
```

**Reading the plots**: **DML-IV** and **Linear DML** surfaces can differ in level and shape; treat comparisons as **exploratory**, not as a test of IV validity. If **patchwork** is not installed, panels print sequentially.


## Summary and Conclusion

This tutorial used **synthetic data** (Part A) and the **401(k) IV sample** from `fetch_401k(..., instrument = TRUE)` (Part B) to illustrate **double machine learning with instrumental variables** using **RCausalML**’s `DMLIVLearner()` and `fit()` from `R/DMLearner.R`, and contrasted it with **`LinearDML()`** **without** an instrument under the same nuisance learners. The logic parallels **ML-based 2SLS** and the DoubleML **IIVM** story: flexible nuisances, cross-fitting, and a final stage for the structural parameter.

| Step | Description |
|------|-------------|
| **IV structure** | Instrument $Z$, treatment $D$, outcome $Y$; unobserved confounders $C$ make naive comparisons biased |
| **`DMLIVLearner()`** | Choose `model_y_xw`, `model_t_xw`, `model_t_xwz`, `discrete_treatment`, `cv`, `random_state` |
| **`fit(Y, T, Z, X, W)`** | Cross-fitted nuisances and linear CATE in `X` (use `X = NULL` for constant effect; set `fit_cate_intercept` consistently) |
| **`LinearDML()` (no IV)** | Same `Y`, `T`, and covariates `X` / `W`; **does not** use $Z$—biased when unobserved confounding drives $D$ and $Y$ |
| **Inference** | `summary(fitted$model_cate$model)` for OLS standard errors on the IV second stage |
| **Nonlinear CATE** | `NonParamDMLIVLearner()` for ranger/`lm` final stage on pseudo-outcomes |
| **SHAP (401(k))** | `explain_cate()` + **shapviz** on predicted $\hat{\tau}(X)$ (refit DML-IV with heterogeneous $X$); non-overlapping explain vs background when possible; beeswarm/bar importance, dependence, waterfall/force, optional subgroup row |


## References

- Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W., & Robins, J. (2018). Double/debiased machine learning for treatment and structural parameters. *The Econometrics Journal*, 21(1), C1–C68. <https://doi.org/10.1111/ectj.12097>

- Ruiz de Villa, A. (2024). *Causal Inference for Data Science*. Manning Publications.

- [DoubleML R: Basic Instrumental Variables Calculation](https://docs.doubleml.org/stable/examples/R_double_ml_basic_iv.html) (conceptual reference; this notebook uses RCausalML)

- [DoubleML User Guide](https://docs.doubleml.org/stable/guide/index.html)

- [EconML: Sieve 2SLS Instrumental Variable Estimation](https://www.pywhy.org/EconML/spec/estimation/two_sls.html)
