# 7. Weighted Double Machine Learning

Double Machine Learning (DML) fits outcome and treatment with flexible ML, then uses a linear model to predict outcome residuals from treatment residuals. The [EconML Weighted Double ML notebook](https://github.com/py-why/EconML/blob/main/notebooks/Weighted%20Double%20Machine%20Learning%20Examples.ipynb) demonstrates **summarized data**: for each segment (unique covariate profile), the data are summarized into one or two rows with **frequency weights** (`freq_weight`) and **sample variance** (`sample_var`). Estimators then use weighted regression for correctness.

The **RCausalML** R package does not currently support `freq_weight` or `sample_var` in `fit()`. This notebook therefore uses the **raw (unsummarized) data** with the same DGP, so that the workflow (LinearDML, effect by segment, coefficients, polynomial CATE, confidence intervals, CausalForestDML) is directly comparable. When using summarized data in practice, you can either expand the data by repeating rows according to segment counts (approximate weighting) or use EconML in Python with native weighted DML.


## Overview

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Sample weight | Importance assigned to each observation in estimation. | "Underrepresented groups get larger weights in the analysis." |
| Inverse Probability Weighting (IPW) | Reweights data by inverse treatment probability to balance groups. | "If treatment probability is 0.2, that treated unit gets weight 1/0.2=5." |
| Weighted ATE | Average treatment effect computed with observation weights. | "Population-representative effect using survey weights." |
| Covariate balance | Similar distribution of covariates across treatment groups after weighting. | "Age and income look comparable after reweighting." |
| Stabilized weights | Modified IPW weights to reduce variance from extreme values. | "Use bounded weights to avoid one observation dominating results." |
| Positivity | Every covariate profile has a non-zero chance of each treatment level. | "For all X values, both treated and control cases exist." |

**Weighted Double Machine Learning (weighted DML)** is an extension of the standard Double/Debiased Machine Learning (DML) framework for causal inference. It enables estimation of *weighted* average treatment effects (or effects in specific subgroups/populations) while preserving DML’s key advantages: robustness to high-dimensional confounders, flexibility with modern machine learning models for nuisance parameters, and √N-consistency via Neyman orthogonality and cross-fitting.

Standard DML (Chernozhukov et al., 2018) targets parameters like the average treatment effect (ATE) in partially linear or interactive regression models by “partialling out” confounding via machine learning. Weighted DML simply generalizes this by incorporating a user-specified weight function ω (which can depend on Y, D, X or be data-driven) into the score/estimating equation. This lets you target:
- Average Treatment Effect on the Treated (ATTE)
- Group Average Treatment Effects (GATEs) or subgroup effects
- Effects in specific subpopulations (e.g., compliers in an IV setting, or treated in period 1 for dynamic treatments)
- Dynamic/sequential treatment effects with subgroup weighting

It is implemented in libraries like DoubleML (Python/R) and appears in applied papers (e.g., for dynamic treatments or reweighting to compliers).

### Quick Recap: How Standard DML Works

DML solves a moment condition (score equation) that is *orthogonal* (insensitive) to estimation errors in “nuisance” functions. This allows ML models (random forests, neural nets, boosted trees, etc.) to estimate nuisances at slower rates (n⁻¹/⁴) while still getting √N-consistent estimates of the causal parameter θ.

For the **interactive regression model (IRM)** with binary treatment D and outcome Y (controlling for high-dimensional X):
- Nuisance functions (estimated via ML with cross-fitting):
  - m₀(X) = E[D | X] (propensity score)
  - g₀(d, X) = E[Y | D = d, X] (conditional outcome regressions for d = 0,1)
- Orthogonal score for ATE (θ₀ = E[g₀(1,X) − g₀(0,X)]):

$$
  \psi(W; \theta, \eta) = \left( g_0(1,X) - g_0(0,X) \right) + \frac{D(Y - g_0(1,X))}{m_0(X)} - \frac{(1-D)(Y - g_0(0,X))}{1-m_0(X)} - \theta
$$

  (or equivalent forms; the exact expression varies slightly by model).
  
- **Cross-fitting algorithm** (K-fold):

  1. Split sample into K folds.
  2. For each fold k: train nuisances on the other K−1 folds, predict on fold k.
  3. Compute the score on the held-out fold.
  4. Average scores across all observations → solve for θ̂ (usually via simple averaging or weighted least squares on residuals).
This “debiases” the estimator and prevents overfitting bias.

### How Weighted DML Works 

Weighted DML replaces the target with a *weighted* version:

$$
\theta_0 = \mathbb{E}\left[ \left( g_0(1,X) - g_0(0,X) \right) \omega(Y,D,X) \right]
$$
(or normalized by E[ω] if you want a weighted average). The weight function ω determines the estimand:

- ω = 1 → standard ATE
- ω = D / p(D=1|X) → ATTE (weights treated units by inverse propensity)
- ω = 1{G = k} → GATE for group k
- ω = predicted Pr(complier | X) (from a first-stage ML model) → approximates Local Average Treatment Effect (LATE) for compliers
- For dynamic/sequential treatments: ω incorporates subgroup indicators S (e.g., treated in period 1) and nested propensity/outcome models

The **score function is multiplied/adjusted by ω** (or its conditional expectation ω̄(X) = E[ω | X] when needed). In DoubleMLIRM, you pass `weights` (a vector or dictionary) at initialization; the package automatically builds the weighted score and handles cross-fitting.

**Step-by-step procedure** (same cross-fitting spirit as standard DML):
1. **Define the target weights** ω (fixed vector, function of observed data, or predicted via a separate ML step, e.g., random forest for complier probability).
2. **Estimate nuisances** (m₀(X), g₀(d,X)) with ML + cross-fitting (exactly as in standard DML).
3. **Form the weighted orthogonal score** ψ_weighted = ω × (standard score terms) − θ (or the precise formulation for the chosen model).
4. **Solve the empirical moment condition**: average the weighted scores over the sample (or solve a weighted regression on residualized outcomes/treatments).
5. **Inference**: Use the influence function / asymptotic variance from the scores (or bootstrap) to get standard errors and CIs. The estimator remains doubly robust and Neyman-orthogonal.

For **dynamic/sequential treatments** (multiple periods), the score becomes more nested: it uses sequential propensity scores p_{d_t}(history, X) and nested conditional means μ and ν to avoid modeling the full distribution of time-varying covariates. Weighting by g(X₀) = Pr(S=1 | X₀) then targets subgroup-specific dynamic effects.

**Why it works**:

- **Neyman orthogonality** still holds → errors in estimating nuisances or the weights do not affect first-order bias in θ̂.
- **Cross-fitting** prevents overfitting in both the main model and any ML step used to construct weights.
- **Double robustness** properties are preserved (or enhanced when weights are inverse-propensity-type).

### Practical Advantages and Use Cases
- **Target policy-relevant subpopulations** (e.g., reweight controls to look like “compliers” in an IV experiment so observational DML approximates the experimental LATE).
- **Heterogeneous effects** (GATEs by subgroups).
- **Dynamic causal effects** with subgroup weighting.
- **Robustness** (e.g., combining DML with IPW-style weights brings it closer to a fully doubly-robust estimator).


In short, weighted DML is “DML with a weighting function plugged into the orthogonal score.” It keeps all the power of standard DML but lets you flexibly reweight to answer more targeted causal questions—exactly what you need when the average effect across the whole population isn’t the one you care about.



**Contents**

1. [Generating raw data](#1-generating-raw-data)
2. [Applying LinearDML](#2-applying-lineardml)
3. [Treatment effect by segment and confidence interval](#3-treatment-effect-by-segment-and-confidence-interval)
4. [Coefficients of the linear CATE model](#4-coefficients-of-the-linear-cate-model)
5. [Non-linear CATE with polynomial features](#5-non-linear-cate-with-polynomial-features)
6. [Confidence intervals of coefficients](#6-confidence-intervals-of-coefficients)
7. [CATE and intervals for all segment combinations](#7-cate-and-intervals-for-all-segment-combinations)
8. [Non-linear CATE with CausalForestDML](#8-non-linear-cate-with-causalforestdml)
9. [Summary](#9-summary)

### Setup

```{r setup}
#| label: setup
library(RCausalML)
library(ggplot2)
set.seed(123)
```

### Generating Raw Data

We use the same DGP as the EconML notebook: binary features (columns 1--4 for heterogeneity, rest as controls), an imbalanced binary treatment depending on the first feature, and an outcome with treatment-effect heterogeneity and heteroskedastic errors.

```{r dgp}
#| label: dgp
n <- 10000L   # number of raw samples
d <- 10L      # number of binary features + 1

# Binary features; first 4 used for heterogeneity, rest as controls
X_raw <- matrix(rbinom(n * d, 1, 0.5), n, d)
# First column is treatment: imbalanced A/B test
log_odds <- qlogis(pmin(pmax(X_raw[, 2], 0.001), 0.999))  # expit(X[,2])-like
X_raw[, 1] <- rbinom(n, 1, plogis(log_odds))
# Outcome: heterogeneity from first binary feature, confounding, heteroskedastic errors
y_raw <- (-1 + 2 * X_raw[, 2]) * X_raw[, 1] + X_raw[, 2] +
  (1 * X_raw[, 2] + 1) * rnorm(n, 0, 1)
```

**Note:** In the Python notebook, the data are then summarized per segment (unique `X` row) with two copies per segment, yielding `X_sum`, `y_sum`, `n_sum`, and `var_sum`. Estimation uses `freq_weight=n_sum` and `sample_var=var_sum`. Here we keep the raw data and fit on it.

## Applying LinearDML

We fit **LinearDML** with binary treatment. Covariates: first column is treatment, columns 2--5 are effect modifiers (A, B, C, D), columns 6--10 are controls. The R API uses a single design matrix: we pass effect modifiers and controls together as `X`; the CATE is linear in the effect modifiers. For a closer match to the Python setup (CATE in A--D only), we pass only the four effect modifiers as `X`; the first stage then uses only those four (no separate W in the current R API).

```{r linear-dml}
#| label: linear-dml
# Treatment = first column; X = effect modifiers (cols 2--5); W = controls (cols 6--10)
treatment <- X_raw[, 1]
X_effect <- X_raw[, 2:5, drop = FALSE]
W_control <- X_raw[, 6:d, drop = FALSE]
colnames(X_effect) <- c("A", "B", "C", "D")
# Full X for first stage (effect modifiers + controls). CATE will be linear in X_effect.
X_full <- cbind(X_effect, W_control)
colnames(X_full) <- c(colnames(X_effect), paste0("W", seq_len(ncol(W_control))))

est <- LinearDML(
  model_y = "ranger",
  model_t = "ranger",
  fit_cate_intercept = TRUE,
  n_fold = 3L,
  seed = 123L
)
# Fit on full X so first stage uses all covariates; CATE is linear in all columns.
# For CATE linear only in A--D, we would fit on X_effect only (weaker first stage).
est <- fit(est, X_full, treatment, y_raw)
```

### Treatment Effect by Segment and Confidence Interval

Treatment effect for a specific segment (e.g. features A=1, B=0, C=0, D=0). We pass one row of covariates for that segment; `predict()` returns the CATE at that point. For a confidence interval we use `estimate_ate()` over the whole sample or bootstrap; segment-level intervals are not built-in in the R DMLearner.

```{r effect-segment}
#| label: effect-segment
# Segment with features [1, 0, 0, 0] (A=1, B=0, C=0, D=0); use mean controls for W
x_segment <- matrix(c(1, 0, 0, 0), nrow = 1)
colnames(x_segment) <- c("A", "B", "C", "D")
# For prediction we need full X (effect modifiers + controls). Use sample means for W.
W_mean <- matrix(colMeans(W_control), nrow = 1, ncol = ncol(W_control), byrow = TRUE)
newdata_segment <- cbind(x_segment, W_mean)
colnames(newdata_segment) <- colnames(X_full)

effect_segment <- predict(est, newdata_segment)
print(paste0("Treatment effect for segment [A=1, B=0, C=0, D=0]: ", round(effect_segment, 4)))
```

```{r ate-ci}
#| label: ate-ci
ate_ci <- estimate_ate(est, X_full, treatment, y_raw, return_ci = TRUE)
print("ATE with 95% confidence interval:")
print(ate_ci)
```

### Coefficients of the Linear CATE Model

The linear CATE model is $\theta(X) = \text{intercept} + A\,\beta_A + B\,\beta_B + C\,\beta_C + D\,\beta_D + \ldots$. We report the estimated coefficients and intercept (and feature names for the effect-modifier part).

```{r coef-names}
#| label: coef-names
coef_est <- coef(est)
int_est <- intercept(est)
cnames <- if (!is.null(names(coef_est))) names(coef_est) else paste0("V", seq_along(coef_est))
print("Coefficients of the linear CATE model (effect modifiers + controls):")
print(data.frame(feature = cnames, coef = round(as.numeric(coef_est), 4)))
print(paste0("Intercept: ", round(int_est, 4)))
```

## Non-linear CATE with Polynomial Features

We can use polynomial (e.g. interaction-only) features in the final stage by building a design matrix with interactions and passing it as `X`. Here we use **LinearDML** with interaction-only polynomial features of degree 2 (all main effects and pairwise products of A--D) plus the control columns, so the CATE is non-linear in the original effect modifiers. For a much larger polynomial basis, **SparseLinearDML** (Lasso final stage) is the usual choice in **RCausalML**.

```{r polynomial-cate}
#| label: polynomial-cate
# Interaction-only polynomial features for effect modifiers (A--D)
poly_X <- stats::model.matrix(~ .^2 - 1, data = as.data.frame(X_effect))
# Drop intercept-like column if present; keep interactions and main effects
X_poly_full <- cbind(poly_X, W_control)
colnames(X_poly_full) <- paste0("V", seq_len(ncol(X_poly_full)))

est_poly <- LinearDML(
  model_y = "ranger",
  model_t = "ranger",
  fit_cate_intercept = TRUE,
  n_fold = 3L,
  seed = 123L
)
est_poly <- fit(est_poly, X_poly_full, treatment, y_raw)
```

### Confidence Intervals of Coefficients

We extract the underlying `lm` object from the **polynomial** `LinearDML` final stage (`est_poly`) and report confidence intervals for the intercept and coefficients (e.g. at 99% level).

```{r coef-interval}
#| label: coef-interval
m <- est_poly$model_cate
if (!is.null(m$model) && inherits(m$model, "lm")) {
  ci_int <- stats::confint(m$model, level = 0.99)
  all_coef <- stats::coef(m$model)
  point_int <- as.numeric(all_coef[1])
  point_coef <- as.numeric(all_coef[-1])
  lower_int <- ci_int[1, 1]
  upper_int <- ci_int[1, 2]
  lower_coef <- ci_int[-1, 1]
  upper_coef <- ci_int[-1, 2]
  n_coef <- length(point_coef)
} else {
  point_int <- intercept(est_poly)
  point_coef <- coef(est_poly)
  n_coef <- length(point_coef)
  lower_int <- upper_int <- lower_coef <- upper_coef <- NA_real_
}
```

```{r fig-coef-ci}
#| label: fig-coef-ci
#| fig-cap: "Point estimates and 99% confidence intervals for intercept and polynomial CATE coefficients (est_poly)"
idx <- seq_len(n_coef + 1)
pts <- c(point_int, point_coef)
lo <- c(lower_int, lower_coef)
hi <- c(upper_int, upper_coef)
labels <- c("intercept", if (!is.null(est_poly$X_names) && length(est_poly$X_names) == n_coef) est_poly$X_names else paste0("V", seq_len(n_coef)))
df_coef <- data.frame(
  param = factor(idx),
  point = pts,
  lower = lo,
  upper = hi,
  label = labels
)
ggplot(df_coef, aes(x = param, y = point)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, colour = "steelblue", linewidth = 0.8) +
  geom_point(size = 3) +
  scale_x_discrete(labels = df_coef$label) +
  labs(y = "Coefficient", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### CATE and Intervals for All Segment Combinations

We compute the CATE for all 16 segments (all combinations of A, B, C, D in $\{0,1\}$), with point estimates and (where available) confidence intervals. Segment-level intervals are approximated using the asymptotic variance of the linear CATE model (predict at each segment and use the linear combination of coefficient variances).

```{r cate-all-segments}
#| label: cate-all-segments
lst <- expand.grid(A = 0:1, B = 0:1, C = 0:1, D = 0:1)
feat_names_abbrev <- c("A", "B", "C", "D")
segment_labels <- apply(lst, 1, function(r) paste(feat_names_abbrev[r > 0], collapse = " "))
segment_labels[segment_labels == ""] <- "(none)"

W_mean_all <- matrix(colMeans(W_control), nrow = nrow(lst), ncol = ncol(W_control), byrow = TRUE)
newdata_all <- cbind(as.matrix(lst), W_mean_all)
colnames(newdata_all) <- colnames(X_full)

point_cate <- predict(est, newdata_all)
# Approximate CI using the lm variance: for linear CATE, Var(theta(x)) = x' V x
if (!is.null(est$model_cate$model) && inherits(est$model_cate$model, "lm")) {
  V <- stats::vcov(est$model_cate$model)
  # Design for each segment: intercept + x (effect modifiers + W means)
  D_seg <- cbind(1, newdata_all)
  se_cate <- sqrt(diag(D_seg %*% V %*% t(D_seg)))
  z <- qnorm(0.995)
  lower_cate <- point_cate - z * se_cate
  upper_cate <- point_cate + z * se_cate
} else {
  lower_cate <- upper_cate <- rep(NA_real_, length(point_cate))
}
```

```{r fig-cate-segments}
#| label: fig-cate-segments
#| fig-cap: "CATE and 99% intervals for all segment combinations"
stat_sig <- (lower_cate > 0) | (upper_cate < 0)
df_seg <- data.frame(
  segment = factor(segment_labels, levels = segment_labels),
  point = point_cate,
  lower = lower_cate,
  upper = upper_cate,
  significant = stat_sig
)
ggplot(df_seg, aes(x = segment, y = point, colour = significant)) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.25, linewidth = 0.7) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("FALSE" = "red", "TRUE" = "black"), labels = c("insig", "stat_sig")) +
  labs(y = "CATE", x = "Segment", colour = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position = "bottom")
```

## Non-linear CATE with CausalForestDML

The Python notebook uses **CausalForestDML** with `sample_weight=n_sum` for summarized data. The R **CausalForestDML** does not support sample weights; we fit on the raw data for comparison.

```{r forest-dml}
#| label: forest-dml
if (requireNamespace("grf", quietly = TRUE)) {
  est_forest <- CausalForestDML(
    model_y = "ranger",
    model_t = "ranger",
    n_fold = 3L,
    seed = 123L,
    num_trees = 1000L,
    min_node_size = 2L
  )
  est_forest <- fit(est_forest, X_full, treatment, y_raw)
  cate_forest <- predict(est_forest, newdata_all)
  print("CausalForestDML fitted (no sample weights in R). CATE for first 5 segments:")
  print(round(cate_forest[1:5], 4))
} else {
  message("Package 'grf' not installed; CausalForestDML skipped.")
  cate_forest <- rep(NA_real_, nrow(newdata_all))
}
```

## Summary

- **Weighted DML in EconML** uses summarized data with `freq_weight` (segment counts) and `sample_var` (within-segment variance) for correct inference. **RCausalML** currently fits on **raw data** only; for summarized data you can expand rows by count or use EconML in Python.
- We replicated the **DGP** (binary features, imbalanced treatment, heterogeneous and heteroskedastic outcome) and fitted **LinearDML** for segment-specific effects, **coefficients**, **polynomial CATE** (`est_poly`), **coefficient and CATE intervals**, and **CausalForestDML**.
- **Tree interpretation** of the CATE (e.g. EconML’s `SingleTreeCateInterpreter`) is not implemented in R; see the [EconML weighted DML notebook](https://github.com/py-why/EconML/blob/main/notebooks/Weighted%20Double%20Machine%20Learning%20Examples.ipynb) for that feature.

## Resources

- [EconML — Weighted Double Machine Learning Examples](https://github.com/py-why/EconML/blob/main/notebooks/Weighted%20Double%20Machine%20Learning%20Examples.ipynb) (Python; `freq_weight`, `sample_var`)
- Chernozhukov, Chetverikov, Demirer, Duflo, Hansen, Newey, Robins (2018). *Double/debiased machine learning for treatment and structural parameters*. [Econometrics Journal](https://doi.org/10.1111/ectj.12097)
- **RCausalML** DML implementations: `R/DMLearner.R` (`LinearDML`, `SparseLinearDML`, `CausalForestDML`)
