# 6. Dynamic Double Machine Learning

**Dynamic Double Machine Learning (Dynamic DML)** is a causal inference method designed to estimate **heterogeneous treatment effects** (also called conditional average treatment effects, or CATE) in **dynamic, sequential settings** where treatments are assigned over multiple time periods according to an adaptive policy.

It is implemented in the EconML library (in `econml.panel.dml.DynamicDML`) and extends the classic **Double Machine Learning (DML)** framework to handle **panel data** or time-series data with repeated interventions (e.g., A/B tests, policy rollouts, or observational studies with time-varying treatments).

## Overview

## Scientific Terminology for Beginners

| Term | Simple explanation | Beginner example |
|---|---|---|
| Dynamic treatment effect | Treatment effect that changes over time. | "A job program has small effects at 3 months, larger at 12 months." |
| Lag | A past value used to explain current outcomes. | "Use last month's sales to predict this month's sales." |
| Panel data | Repeated observations of the same units over time. | "Monthly income records for the same households." |
| State dependence | Current outcome depends on previous outcome states. | "Unemployment today depends on unemployment last quarter." |
| Time-varying confounder | A confounder that changes over time and affects treatment/outcome. | "Health status changes yearly and influences both treatment and costs." |
| Sequential ignorability | At each time point, treatment is as-if random after conditioning on history. | "Given past covariates and outcomes, treatment timing is unconfounded." |

Standard DML works great for **static** (one-time) treatments with high-dimensional or complex confounders. However, many real-world scenarios involve **dynamic treatments**:
- Treatments in period $t$ depend on past treatments and evolving states.

- Outcomes accumulate over time.

- Confounders (variables that affect both treatment decisions and outcomes) are high-dimensional or have unknown functional forms.

Dynamic DML solves this by using machine learning to flexibly model the “nuisance” relationships (confounding and dynamics) while delivering **robust, semi-parametric estimates** of how treatments in different periods affect the final outcome — with heterogeneity driven by initial conditions.

### How Dynamic DML Works

It assumes the data follow a **Markov decision process**-like structure with observed dynamic confounders. Here’s the core logic:

1. **Data Structure**  
   You provide **panel data** indexed by unit (e.g., individual, firm) and time $t = 1, \ldots, m$:

   $$
   \{X_t, W_t, T_t, Y_t\}_{t=1}^m
   $$

   - $X_t$: State variables at time $t$ (can be high-dimensional).
   - $W_t$: Additional confounders/controls at time $t$.
   - $T_t$: Treatment(s) assigned at time $t$ (can be continuous, binary, or multi-valued).
   - $Y_t$: Outcome observed at time $t$.  
   A `groups` array tells the model which rows belong to the same unit across time.

2. **Underlying Structural Model**  
   The method is built on these equations (the data-generating process):

   $$
   XW_t = A \cdot T_{t-1} + B \cdot XW_{t-1} + \eta_t
   $$
   (state/confounder evolution)

   $$
   T_t = p(T_{t-1}, XW_t, \zeta_t)
   $$
   (treatment policy — how treatments are chosen adaptively)

   $$
   Y_t = \theta_0(X_0)' T_t + \mu' XW_t + \epsilon_t
   $$
   (outcome equation)

   - $XW_t$ is the concatenation of states and confounders.
   - $\theta_0(X_0)$ is the **heterogeneous treatment effect** you want to estimate — it depends only on the **initial state** $X_0$.
   - The effect of treatments from all periods is ultimately measured on the **final-period outcome**.
   - Errors $\eta_t, \zeta_t, \epsilon_t$ are assumed mean-independent of the covariates (standard DML exogeneity).

3. **Estimation via Double Machine Learning (the “Double” part)**  

   Dynamic DML uses the classic DML trick to get **unbiased** estimates even when nuisance functions are estimated with ML (which can overfit):

   - **Step A: Nuisance estimation**  
     Use flexible ML models to estimate the two “nuisance” functions:
     - Treatment model: $\hat{p}(T_t \mid T_{t-1}, XW_t)$ (or conditional expectation if continuous).
     - Outcome model: $\hat{E}[Y_t \mid T_t, XW_t, \text{past}]$.

   - **Step B: Residualization & orthogonalization**  
     Create residuals that “partial out” the confounding. Because of the dynamic structure, this is done recursively across time periods while respecting the panel structure.

   - **Step C: Cross-fitting**  
     Data are split into folds. Nuisance models are trained on one fold and used to predict residuals on the held-out fold (prevents overfitting and ensures valid inference).

   - **Step D: Solve the moment condition**  
     The final estimator solves a score equation that isolates $\theta_0(X_0)$, yielding consistent and asymptotically normal estimates even if the ML nuisance models are slightly misspecified (as long as they converge fast enough).

The result is an estimate of $\theta_0(X_0)$ for any new $X_0$, telling you how much the sequence of treatments would change the final outcome for units with different initial characteristics.

### Key Features & Parameters in EconML’s DynamicDML

- `model_t`: ML model for the treatment (nuisance).
- `model_y`: ML model for the outcome (nuisance).
- `X`: Initial or time-varying state features.
- `W`: Additional high-dimensional controls.
- `groups`: Panel identifier.
- Supports continuous, binary, or multi-valued treatments.

### Comparison to Standard DML

| Aspect                  | Standard DML                  | Dynamic DML                          |
|-------------------------|-------------------------------|--------------------------------------|
| Treatment               | One-time (static)             | Sequential & adaptive over time      |
| Data                    | Cross-sectional or simple panel | Full panel with time dynamics        |
| Confounding             | Static controls               | Dynamic (time-varying) confounders   |
| Effect estimated        | CATE at a point in time       | Cumulative effect on final outcome   |
| Heterogeneity           | Function of any covariates    | Driven by **initial state** $X_0$ |

### Assumptions & Limitations

- **All confounders observed**: No unobserved time-varying confounding.
- Markovian dynamics: Future depends on the past only through the current state.
- Overlap/positivity: Treatment probability is bounded away from 0 and 1.
- ML models must converge at reasonable rates (standard DML condition).

It is computationally heavier than static DML because it processes the full panel structure and dynamic nuisance models.

In short, Dynamic DML lets you answer questions like:  
“If two units start with different initial characteristics, how much better (or worse) will the final outcome be if we follow this dynamic treatment policy versus that one — after properly controlling for all the evolving confounders using machine learning?”  

### Example applications of Dynamic Double ML include:

- The effect of **investments on revenue** at companies that receive investments at regular intervals ([see more](https://arxiv.org/abs/2103.08390))
- The effect of **prices on demand** in stores where prices change over time
- The effect of **income on health outcomes** in people who receive yearly income

The preferred data format is **balanced panel data**: each panel is one entity (e.g. company, store, or person) and rows are time points. Example:

| Company | Year | Features | Investment | Revenue |
|---------|------|----------|------------|---------|
| A       | 2018 | ...      | \$1,000     | \$10,000 |
| A       | 2019 | ...      | \$2,000     | \$12,000 |
| A       | 2020 | ...      | \$3,000     | \$15,000 |
| B       | 2018 | ...      | \$0         | \$5,000  |
| ...     | ...  | ...      | ...        | ...     |

When passing data to the estimator, the entity column corresponds to the `groups` argument at fit time. The time index can be inferred from the order of rows within each group.

## Implimantion in R

This notebook demonstrates **DynamicDMLearner** (native R, no Python dependency) on synthetic panel data, mirroring the [EconML Dynamic Double ML Examples](https://github.com/microsoft/EconML/blob/main/notebooks/Dynamic%20Double%20Machine%20Learning%20Examples.ipynb).

**Contents**

* `Part A`: Dynamic DML with Synthetic Data

1. [Example usage with average treatment effects](#1-example-usage-with-average-treatment-effects)
2. [Example usage with heterogeneous treatment effects](#2-example-usage-with-heterogeneous-treatment-effects)

* `Part B`: Hospital clinical deterioration (data pipeline, risk model, Dynamic DML, policy curves, SHAP)

### Setup

```{r setup}
#| label: setup
library(RCausalML)
library(ggplot2)
set.seed(123)
```

## Part A: Dynamic DML with Synthetic Data

### Usage with Average Treatment Effects

#### Data Generation Process (DGP)

We use a simple **panel data generating process**: each unit has one row per period; treatments in each period affect the **final-period outcome** $Y_T$. With $T_t$ = treatment(s) at time $t$, $X_t$ = features/controls, and a linear CATE:

$$
Y_T = \sum_{t=1}^{T} \theta_t' T_t + \gamma' W + \eta
$$

We generate balanced panels with `n_panels` units and `n_periods` time periods; each period has `n_treatments` treatment variables. The outcome is observed at the last period (and repeated per row for compatibility with the fit API).

```{r dgp-params}
#| label: dgp-params
n_panels   <- 800L   # number of panels (units)
n_periods  <- 3L     # time periods per panel
n_treatments <- 2L   # treatments per period
n_w        <- 20L    # control variables
n_x        <- 100L   # total features (for Section 2)
s_x        <- 10L    # support size for controls
```

 Generate panel data with dynamic treatments and a final-period outcome that depends on all treatments and controls. The true effect vector has one coefficient per treatment dimension (period-specific effects).

```{r dgp-gen}
#| label: dgp-gen
# One row per (unit, period). Outcome Y is the final-period outcome for that unit.
n <- n_panels * n_periods
groups <- rep(seq_len(n_panels), each = n_periods)

# True effects: (n_periods * n_treatments) coefficients
true_effect <- c(0.04, 0.07, 0.32, 0.24, 0.13, 0.61)

# Controls W (time-varying): same dimension for all rows
W <- matrix(rnorm(n * n_w), n, n_w)
colnames(W) <- paste0("W", seq_len(n_w))

# Treatments T: (n_periods * n_treatments) columns; for row in period p, fill period 1..p, rest 0
T_mat <- matrix(0, n, n_periods * n_treatments)
for (i in seq_len(n_panels)) {
  idx <- ((i - 1) * n_periods + 1):(i * n_periods)
  for (p in seq_len(n_periods)) {
    row <- (i - 1) * n_periods + p
    T_mat[row, ((p - 1) * n_treatments + 1):(p * n_treatments)] <- rnorm(n_treatments, 0.5, 0.5)
  }
}
colnames(T_mat) <- paste0("T", seq_len(n_periods * n_treatments))

# Outcome at last period: Y = T %*% theta + W %*% gamma + noise (per unit)
gamma_w <- runif(n_w, 0, 0.3)
Y_last <- numeric(n_panels)
for (i in seq_len(n_panels)) {
  idx <- ((i - 1) * n_periods + 1):(i * n_periods)
  T_i <- T_mat[idx, , drop = FALSE]
  # Use period-specific T for this unit (row n_periods has all treatments)
  Ti_full <- T_i[n_periods, ]
  W_i <- W[idx[n_periods], , drop = FALSE]
  Y_last[i] <- sum(true_effect * Ti_full) + as.numeric(W_i %*% gamma_w) + rnorm(1, 0, 0.5)
}
# Repeat last-period outcome for each row of the panel (for fit API)
Y <- rep(Y_last, each = n_periods)
```

#### Train Estimator

We fit **DynamicDMLearner** with cross-fitting. With no $X$ we use an intercept-only CATE (one effect per treatment dimension). We set `fit_cate_intercept = FALSE` when $X$ is omitted so the design matrix has a single column and the estimated coefficient for each treatment is the period effect.

```{r est-create-fit}
#| label: est-create-fit
est <- DynamicDMLearner(
  model_y = "ranger", model_t = "ranger", cv = 3L, random_state = 123L,
  fit_cate_intercept = FALSE  # intercept-only when X = NULL (avoids collinearity)
)
est <- fit(est, Y = Y, T = T_mat, X = NULL, W = W, groups = groups)
```

Average treatment effect: we approximate the "default policy" effect by averaging predicted CATE over the sample (or using a baseline vs target policy).

```{r ate-default}
#| label: ate-default
# Predicted CATE at each row (intercept-only model: same for all)
cate_pred <- predict(est, newdata = NULL)
ate_approx <- mean(cate_pred, na.rm = TRUE)
print(paste0("Average effect (mean predicted CATE): ", round(ate_approx, 2)))
```

Effect of moving from baseline policy (all zeros) to target policy (all ones). With 6 treatment dimensions, the effect is the sum of the estimated period effects (since T1 - T0 = 1 for each).

```{r effect-policy}
#| label: effect-policy
baseline_policy <- matrix(0, 1, n_periods * n_treatments)
target_policy   <- matrix(1, 1, n_periods * n_treatments)
# Effect = sum(theta * (T1 - T0)); with T1 - T0 = 1 for all, effect = sum(period effects)
intercept_est_full <- vapply(est$model_cate$models, function(fit) as.numeric(stats::coef(fit)[1]), numeric(1))
eff_policy <- sum(intercept_est_full * as.numeric(target_policy - baseline_policy))
print(paste0("Effect of target vs baseline policy (sum of period effects): ", round(eff_policy, 2)))
```

Period treatment effects: the R **DynamicDMLearner** stores one linear model per treatment column. With `fit_cate_intercept = FALSE` and no $X$, the single coefficient per model is the marginal effect for that period/treatment.

```{r period-effects}
#| label: period-effects
m <- est$model_cate
if (m$type == "lm_multi") {
  intercepts <- vapply(m$models, function(fit) as.numeric(stats::coef(fit)[1]), numeric(1))
} else {
  intercepts <- as.numeric(stats::coef(m$model)[1])
}
dim(intercepts) <- c(n_treatments, n_periods)
for (p in seq_len(n_periods)) {
  print(paste0("Marginal effect of treatments in period ", p, " on final outcome: ",
               paste(round(intercepts[, p], 4), collapse = ", ")))
}
```

#### Performance visualization

We compare estimated intercepts (period effects) to the true effect vector.

```{r conf-int}
#| label: conf-int
if (est$model_cate$type == "lm_multi") {
  conf_ints <- t(vapply(est$model_cate$models, function(fit) {
    ci <- stats::confint(fit, parm = 1, level = 0.95)
    c(ci[1], ci[2])
  }, numeric(2)))
  conf_lower <- conf_ints[, 1]
  conf_upper <- conf_ints[, 2]
  intercept_est <- vapply(est$model_cate$models, function(fit) as.numeric(stats::coef(fit)[1]), numeric(1))
} else {
  intercept_est <- as.numeric(stats::coef(est$model_cate$model)[1])
  ci <- stats::confint(est$model_cate$model, parm = 1, level = 0.95)
  conf_lower <- ci[1]; conf_upper <- ci[2]
}
```

```{r fig-viz-ate}
#| label: fig-viz-ate
#| fig-cap: "Estimated vs true period treatment effects (DynamicDMLearner)"
idx_est <- seq_len(length(true_effect))
df_plot <- data.frame(
  param = factor(idx_est),
  estimate = if (length(intercept_est) == length(true_effect)) intercept_est else rep(intercept_est[1], length(true_effect)),
  true = true_effect,
  lower = if (length(conf_lower) == length(true_effect)) conf_lower else NA_real_,
  upper = if (length(conf_upper) == length(true_effect)) conf_upper else NA_real_
)
# If we have one intercept only, plot it once
if (length(intercept_est) != length(true_effect)) {
  df_plot$estimate <- intercept_est[1]
  df_plot$lower <- NA
  df_plot$upper <- NA
}
ggplot(df_plot, aes(x = param)) +
  geom_errorbar(aes(ymin = lower, ymax = upper, y = estimate), width = 0.2, colour = "steelblue", linewidth = 0.8) +
  geom_point(aes(y = estimate, colour = "DynamicDMLearner"), size = 3) +
  geom_point(aes(y = true, colour = "Ground truth"), size = 3, shape = 18) +
  geom_vline(xintercept = seq(2.5, 6, by = 2), linetype = "dashed", alpha = 0.4) +
  scale_x_discrete(labels = paste0("theta_", rep(0:(n_periods - 1), each = n_treatments))) +
  labs(y = "Effect", x = "Parameter", colour = "") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

###  Usage with Heterogeneous Treatment Effects

We can estimate **treatment effect heterogeneity** with respect to a subset of features $X$ (e.g. time-invariant unit characteristics). Pass $X$ at fit time; the CATE is then $\theta(X) = \phi(X)' \beta + \text{intercept}$.

#### DGP with heterogeneity

We add features $X$ and let the effect depend on them: $\theta(X) = \theta_0 + \theta_1' X$.

```{r dgp-het-params}
#| label: dgp-het-params
het_strength <- 0.5
n_hetero    <- 2L   # number of effect modifiers
n_x_use     <- 5L   # use first 5 X columns for heterogeneity
```

```{r dgp-het-gen}
#| label: dgp-het-gen
set.seed(12)
X_full <- matrix(rnorm(n * n_x_use), n, n_x_use)
colnames(X_full) <- paste0("X", seq_len(n_x_use))

# Heterogeneous effect: true_effect + X[, 1:n_hetero] %*% het_coef

het_coef <- matrix(het_strength * runif(n_hetero * (n_periods * n_treatments)), n_hetero, n_periods * n_treatments)
true_effect_het <- true_effect  # baseline
# For each unit, Y_last = T %*% (true_effect + het_coef' %*% x) + ...
Y_last_het <- numeric(n_panels)
for (i in seq_len(n_panels)) {
  idx <- ((i - 1) * n_periods + 1):(i * n_periods)
  x_i <- X_full[idx[1], 1:n_hetero, drop = FALSE]
  theta_i <- true_effect + as.numeric(x_i %*% het_coef)
  Ti_full <- T_mat[idx[n_periods], ]
  W_i <- W[idx[n_periods], , drop = FALSE]
  Y_last_het[i] <- sum(theta_i * Ti_full) + as.numeric(W_i %*% gamma_w) + rnorm(1, 0, 0.5)
}
Y_het <- rep(Y_last_het, each = n_periods)
```

#### Train estimator with X

```{r est-het-fit}
#| label: est-het-fit
est_het <- DynamicDMLearner(model_y = "ranger", model_t = "ranger", cv = 3L, random_state = 1L)
est_het <- fit(est_het, Y = Y_het, T = T_mat, X = X_full, W = W, groups = groups)
```

Average treatment effect for a test set of units:

```{r ate-het-test}
#| label: ate-het-test
X_test_idx <- seq(1, min(25, n_panels), by = 3)
X_test <- X_full[(X_test_idx[1] - 1) * n_periods + 1, , drop = FALSE]
for (j in seq_along(X_test_idx)[-1]) {
  row_j <- (X_test_idx[j] - 1) * n_periods + 1
  X_test <- rbind(X_test, X_full[row_j, , drop = FALSE])
}
cate_het <- predict(est_het, newdata = X_test)
print(paste0("Average effect (mean CATE on test set): ", round(mean(cate_het), 2)))
```

Effect of target vs baseline policy for each test unit:

```{r effect-het-test}
#| label: effect-het-test
eff_het <- vapply(seq_len(nrow(X_test)), function(i) {
  effect(est_het, newdata = X_test[i, , drop = FALSE],
         T0 = as.numeric(baseline_policy), T1 = as.numeric(target_policy))
}, numeric(1))
print("Effect of target vs baseline for test units:")
print(round(eff_het, 4))
```

Coefficients (intercept and slopes on $X$): the CATE model is linear in $\phi(X)$; we have one such model per treatment column.

```{r coef-intercept-het}
#| label: coef-intercept-het
m_het <- est_het$model_cate
if (m_het$type == "lm_multi") {
  intercept_het <- vapply(m_het$models, function(fit) {
    cf <- stats::coef(fit)
    if ("intercept" %in% names(cf)) cf["intercept"] else cf[1]
  }, numeric(1))
  coef_het <- vapply(m_het$models, function(fit) {
    cf <- stats::coef(fit)
    unname(cf[names(cf) != "intercept"])
  }, FUN.VALUE = numeric(n_x_use))
} else {
  intercept_het <- stats::coef(m_het$model)["intercept"]
  cf <- stats::coef(m_het$model)
  coef_het <- unname(cf[names(cf) != "intercept"])
}
print("Intercepts (period/treatment effects):")
print(round(intercept_het, 4))
```

#### Performance visualization

Plot estimated vs true parameters (intercept + first heterogeneity coefficients) when identifiable from the DGP.

```{r fig-viz-het}
#| label: fig-viz-het
#| fig-cap: "Heterogeneous Dynamic DML: estimated vs true parameters"
param_hat <- c(intercept_het, as.numeric(coef_het))
n_params <- length(param_hat)
df_het <- data.frame(
  param = factor(seq_len(n_params)),
  estimate = param_hat
)
ggplot(df_het, aes(x = param, y = estimate)) +
  geom_point(colour = "steelblue", size = 3) +
  labs(y = "Coefficient", x = "Parameter") +
  theme_minimal()
```

## Part B: Use case — hospital clinical deterioration

This [dataset](https://www.kaggle.com/datasets/tarekmasryo/hospital-deterioration-dataset) is a simulated hospital cohort for early-warning modelling. Each admission has up to 72 hourly rows with vitals, labs, static patient attributes, and labels such as **deterioration in the next 12 hours** (`deterioration_next_12h`).

**Workflow (below).** (1) Rebuild `hospital_deterioration_ml_ready.csv` by merging `vitals_timeseries.csv`, `labs_timeseries.csv`, and `patients.csv`, and attach the hourly outcome from `hospital_deterioration_hourly_panel.csv` (same row keys as the published panel). (2) Add **3-hour rolling means** and **hour-to-hour deltas** for selected vitals and labs. (3) Define a modelling feature set, **split by `patient_id`**, and run **EDA**. (4) Fit a **supervised risk model** (random forest classifier) for **ROC** and **precision–recall** curves on `deterioration_next_12h`. (5) Fit **DynamicDMLearner** on a **three-hour panel** with time-varying **nurse alerts** as the treatment and the hour-2 deterioration label as the **final-period outcome** (repeated on each period row, as in Part A). (6) Summarize **heterogeneous effects**, **baseline vs target alert policies**, and **threshold vs recall / precision / alert rate** using the risk model scores (a simple “balanced” policy picks a threshold on that curve). (7) **SHAP** on predicted CATE for interpretability.

**Caution.** Nurse alerts are not randomized; Dynamic DML estimates are only meaningful under **sequential ignorability** and correct specification of controls. Treat this block as a **methods template**, not a causal audit of alert efficacy.

```{r hospital-data-dir}
#| label: hospital-data-dir
# Package root: prefer project render; fall back to relative path from this notebook
candidates <- c(
  file.path(getwd(), "data", "hospital-deterioration-dataset"),
  normalizePath(file.path("..", "..", "..", "..", "data", "hospital-deterioration-dataset"), mustWork = FALSE),
  "/home/zia207/Websites/R_Website/Causal_Inference_R/Cursor_R/RCausalML/data/hospital-deterioration-dataset"
)
ok <- file.exists(file.path(candidates, "vitals_timeseries.csv"))
data_dir <- candidates[which(ok)[1L]]
stopifnot(length(data_dir) == 1L, dir.exists(data_dir))
```

### Data processing

1. **Merge** vitals and labs on `patient_id` and `hour_from_admission`; join **patient-level** columns from `patients.csv`.
2. **Outcome** `deterioration_next_12h` is taken from `hospital_deterioration_hourly_panel.csv` so labels stay aligned with the bundled panel definition.
3. **Rolling means** (including the current hour, up to the last three observations per admission) for vitals and labs listed in the code chunk.
4. **Deltas** = current hour minus previous hour, within patient (`NA` for the first hour).
5. **Write** `hospital_deterioration_ml_ready.csv` (now includes `patient_id`, static outcomes such as `baseline_risk_score` / `los_hours` from `patients.csv`, engineered `_roll3` and `_dh` columns). *Rendering this chunk recomputes the full file and can take about one minute.*

```{r hospital-build-ml-ready}
#| label: hospital-build-ml-ready
#| cache: true
vitals <- read.csv(file.path(data_dir, "vitals_timeseries.csv"), check.names = FALSE)
labs <- read.csv(file.path(data_dir, "labs_timeseries.csv"), check.names = FALSE)
patients <- read.csv(file.path(data_dir, "patients.csv"), check.names = FALSE)

d_hosp <- merge(vitals, labs, by = c("patient_id", "hour_from_admission"), sort = FALSE)
d_hosp <- merge(d_hosp, patients, by = "patient_id", sort = FALSE)
yref <- read.csv(file.path(data_dir, "hospital_deterioration_hourly_panel.csv"), check.names = FALSE)
yref <- yref[, c("patient_id", "hour_from_admission", "deterioration_next_12h")]
d_hosp <- merge(d_hosp, yref, by = c("patient_id", "hour_from_admission"), sort = FALSE)

roll3_mean <- function(x) {
  vapply(seq_along(x), function(i) mean(x[max(1L, i - 2L):i]), numeric(1))
}

d_hosp <- d_hosp[order(d_hosp$patient_id, d_hosp$hour_from_admission), , drop = FALSE]

num_engineer <- c(
  "heart_rate", "respiratory_rate", "spo2_pct", "temperature_c",
  "systolic_bp", "diastolic_bp", "oxygen_flow", "mobility_score",
  "wbc_count", "lactate", "creatinine", "crp_level", "hemoglobin", "sepsis_risk_score"
)
for (nm in num_engineer) {
  d_hosp[[paste0(nm, "_roll3")]] <- ave(d_hosp[[nm]], d_hosp$patient_id, FUN = roll3_mean)
  d_hosp[[paste0(nm, "_dh")]] <- ave(d_hosp[[nm]], d_hosp$patient_id, FUN = function(z) c(NA_real_, diff(z)))
}

out_csv <- file.path(data_dir, "hospital_deterioration_ml_ready.csv")
write.csv(d_hosp, out_csv, row.names = FALSE)
message("Wrote ", nrow(d_hosp), " rows, ", ncol(d_hosp), " columns to ", normalizePath(out_csv))
```

**Feature schema (modelling).** Core predictors match the clinical feature set you listed: `hour_from_admission`, vitals (`heart_rate`, `respiratory_rate`, `spo2_pct`, `temperature_c`, `systolic_bp`, `diastolic_bp`, `oxygen_device`, `oxygen_flow`, `mobility_score`, `nurse_alert`), labs (`wbc_count`, `lactate`, `creatinine`, `crp_level`, `hemoglobin`, `sepsis_risk_score`), and static covariates (`age`, `gender`, `comorbidity_index`, `admission_type`), plus the engineered `_roll3` and `_dh` fields. Categorical columns are expanded to dummies below. Rows with missing deltas are filled with 0 before matrix construction.

```{r hospital-feature-matrices}
#| label: hospital-feature-matrices
d_risk <- d_hosp
d_risk[is.na(d_risk)] <- 0

base_risk <- c(
  "hour_from_admission", "heart_rate", "respiratory_rate", "spo2_pct", "temperature_c",
  "systolic_bp", "diastolic_bp", "oxygen_flow", "mobility_score", "nurse_alert",
  "wbc_count", "lactate", "creatinine", "crp_level", "hemoglobin", "sepsis_risk_score",
  "age", "comorbidity_index"
)
roll_dh <- c(paste0(num_engineer, "_roll3"), paste0(num_engineer, "_dh"))
rhs <- paste(c(base_risk, roll_dh, "gender", "admission_type", "oxygen_device"), collapse = " + ")
X_risk <- model.matrix(as.formula(paste("~", rhs)), data = d_risk)
X_risk <- X_risk[, -1L, drop = FALSE]
colnames(X_risk) <- make.names(colnames(X_risk))
y_risk <- d_risk$deterioration_next_12h
patient_id_risk <- d_risk$patient_id
```

### Train / test split and exploratory analysis

```{r hospital-train-test}
#| label: hospital-train-test
uid <- unique(patient_id_risk)
set.seed(123)
train_uid <- sample(uid, size = floor(0.8 * length(uid)), replace = FALSE)
test_uid <- setdiff(uid, train_uid)
tr <- patient_id_risk %in% train_uid
te <- patient_id_risk %in% test_uid
X_tr <- X_risk[tr, , drop = FALSE]
y_tr <- y_risk[tr]
X_te <- X_risk[te, , drop = FALSE]
y_te <- y_risk[te]
# Cap training rows so the notebook renders in reasonable time (remove to use all training rows).
max_rf_train_rows <- 60000L
if (nrow(X_tr) > max_rf_train_rows) {
  set.seed(124)
  sub_tr <- sample.int(nrow(X_tr), max_rf_train_rows)
  X_tr <- X_tr[sub_tr, , drop = FALSE]
  y_tr <- y_tr[sub_tr]
}
```

```{r hospital-eda}
#| label: hospital-eda
#| fig-cap: "Hourly prevalence of deterioration in the next 12 hours (full cohort)"
eda_df <- aggregate(deterioration_next_12h ~ hour_from_admission, data = d_hosp, mean)
ggplot(eda_df, aes(x = hour_from_admission, y = deterioration_next_12h)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(size = 1.5, colour = "steelblue") +
  labs(x = "Hour from admission", y = "P(deterioration in next 12h)") +
  theme_minimal()
```

### Fit and validate model

**Supervised risk model (ROC and PR).** Random forest **classification** with `probability = TRUE` gives a risk score for `deterioration_next_12h` using the engineered feature matrix. ROC and precision–recall curves are computed on the **test** split (patient-level holdout).

```{r hospital-rf-roc-pr}
#| label: hospital-rf-roc-pr
#| fig-cap: "Test-set ROC and precision–recall curves for deterioration_next_12h"
library(ranger)
rf_risk <- ranger(
  dependent.variable.name = "y",
  data = data.frame(X_tr, y = factor(y_tr, levels = c(0, 1))),
  probability = TRUE,
  classification = TRUE,
  num.trees = 300L,
  min.node.size = 10L,
  seed = 123L,
  num.threads = max(1L, parallel::detectCores() - 1L)
)
pred_te <- predict(rf_risk, data.frame(X_te))$predictions[, 2]
threshold_grid <- seq(0.001, 0.999, length.out = 200)
pr_at <- function(scores, labels, tau) {
  pred <- as.integer(scores >= tau)
  tp <- sum(pred == 1 & labels == 1)
  fp <- sum(pred == 1 & labels == 0)
  fn <- sum(pred == 0 & labels == 1)
  prec <- if (tp + fp > 0) tp / (tp + fp) else NA_real_
  rec <- if (tp + fn > 0) tp / (tp + fn) else NA_real_
  alert_rate <- mean(pred == 1)
  c(precision = prec, recall = rec, alert_rate = alert_rate)
}
curve_df <- as.data.frame(t(vapply(threshold_grid, function(tau) pr_at(pred_te, y_te, tau), numeric(3))))
curve_df$tau <- threshold_grid
roc_df <- data.frame(
  FPR = sapply(threshold_grid, function(tau) sum(pred_te >= tau & y_te == 0) / max(1, sum(y_te == 0))),
  TPR = sapply(threshold_grid, function(tau) sum(pred_te >= tau & y_te == 1) / max(1, sum(y_te == 1))),
  tau = threshold_grid
)
p1 <- ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_path(linewidth = 0.8, colour = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, alpha = 0.4) +
  coord_equal() +
  labs(title = "ROC (test)", x = "False positive rate", y = "True positive rate") +
  theme_minimal()
p2 <- ggplot(curve_df, aes(x = recall, y = precision)) +
  geom_path(linewidth = 0.8, colour = "darkred") +
  labs(title = "Precision–recall (test)", x = "Recall", y = "Precision") +
  theme_minimal()
print(p1)
print(p2)
```

**Dynamic Double ML.** We keep patients with hours 0, 1, and 2; each patient contributes three rows (panel structure via `groups`). **Treatment** `T` is the hourly binary `nurse_alert` (one column, time-varying). **Outcome** `Y` is `deterioration_next_12h` at **hour 2**, repeated on all three rows, as in Part A. **$W$** holds rich hourly vitals and labs (rolling means and deltas) plus `oxygen_device` dummies — `nurse_alert` is **not** included in $W$ because it is $T$. **$X$** is `age`, `comorbidity_index`, `gender`, and `admission_type` (reference-level contrasts; intercept handled by `fit_cate_intercept = TRUE`). For speed, we subsample patients (`max_patients_dml`).

```{r hospital-dml-panel}
#| label: hospital-dml-panel
max_patients_dml <- 4000L
h3 <- d_hosp[d_hosp$hour_from_admission %in% 0:2, ]
ok <- tapply(h3$hour_from_admission, h3$patient_id, function(h) all(0:2 %in% h))
pid_ok <- names(ok)[as.logical(ok)]
set.seed(7)
pid_sub <- sample(pid_ok, size = min(max_patients_dml, length(pid_ok)), replace = FALSE)
sub <- h3[h3$patient_id %in% pid_sub, ]
sub <- sub[order(sub$patient_id, sub$hour_from_admission), ]
stopifnot(identical(as.integer(table(sub$patient_id)), rep(3L, length(pid_sub))))

T_dyn <- matrix(sub$nurse_alert, ncol = 1L)
colnames(T_dyn) <- "nurse_alert"
y2 <- sub[sub$hour_from_admission == 2, c("patient_id", "deterioration_next_12h")]
Y_dyn <- y2$deterioration_next_12h[match(sub$patient_id, y2$patient_id)]
groups_dyn <- sub$patient_id

sub[is.na(sub)] <- 0
W_base <- c(
  "hour_from_admission", "heart_rate", "respiratory_rate", "spo2_pct", "temperature_c",
  "systolic_bp", "diastolic_bp", "oxygen_flow", "mobility_score",
  "wbc_count", "lactate", "creatinine", "crp_level", "hemoglobin", "sepsis_risk_score",
  paste0(num_engineer, "_roll3"), paste0(num_engineer, "_dh")
)
W_dyn <- model.matrix(~ 0 + ., data = sub[, c(W_base, "oxygen_device"), drop = FALSE])
colnames(W_dyn) <- make.names(colnames(W_dyn))

Xh <- model.matrix(~ age + comorbidity_index + gender + admission_type, data = sub)
X_het <- Xh[, -1L, drop = FALSE]
colnames(X_het) <- make.names(colnames(X_het))
```

```{r hospital-dml-fit}
#| label: hospital-dml-fit
#| cache: true
est_hosp <- DynamicDMLearner(
  model_y = "ranger", model_t = "ranger", cv = 3L, random_state = 42L,
  fit_cate_intercept = TRUE
)
est_hosp <- fit(est_hosp, Y = Y_dyn, T = T_dyn, X = X_het, W = W_dyn, groups = groups_dyn)
```

### Heterogeneous effects, baseline policy, and threshold performance

```{r hospital-heterogeneity-policy}
#| label: hospital-heterogeneity-policy
X_u <- X_het[!duplicated(as.data.frame(X_het)), , drop = FALSE]
if (nrow(X_u) > 200L) {
  set.seed(3)
  X_u <- X_u[sample.int(nrow(X_u), 200L), , drop = FALSE]
}
cate_u <- predict(est_hosp, newdata = as.matrix(X_u))
summary(cate_u)

eff_demo <- vapply(seq_len(min(20L, nrow(X_u))), function(i) {
  effect(est_hosp, newdata = as.matrix(X_u[i, , drop = FALSE]), T0 = 0, T1 = 1)
}, numeric(1))
print(round(eff_demo, 4))
```

**Threshold vs recall, precision, and alert rate** uses the **supervised** risk score `pred_te` (clinical alerting is naturally tied to a risk threshold). A **balanced** operating point is illustrated by maximizing **F1** on the test grid (you can replace this with cost-weighted or constraint-based rules).

```{r hospital-threshold-curve}
#| label: hospital-threshold-curve
#| fig-cap: "Test-set threshold trade-offs for the risk model (balanced point = max F1 on this grid)"
f1 <- with(curve_df, 2 * precision * recall / pmax(precision + recall, 1e-9))
curve_df$f1 <- f1
balanced_idx <- which.max(replace(f1, is.na(f1), -Inf))
tau_star <- curve_df$tau[balanced_idx]

tdf <- rbind(
  transform(curve_df[, c("tau", "precision")], metric = "precision", value = precision)[, c("tau", "metric", "value")],
  transform(curve_df[, c("tau", "recall")], metric = "recall", value = recall)[, c("tau", "metric", "value")],
  transform(curve_df[, c("tau", "alert_rate")], metric = "alert_rate", value = alert_rate)[, c("tau", "metric", "value")]
)
ggplot(tdf, aes(x = tau, y = value, colour = metric)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = tau_star, linetype = 2, alpha = 0.6) +
  labs(
    title = "Threshold vs precision, recall, and alert rate (test)",
    x = "Threshold on predicted P(deterioration)",
    y = "Metric value",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Here is a clean, tutorial‑ready, debugged, and more intuitive rewrite of your SHAP section.  
No content is removed — just reorganized, clarified, and made easier to understand.

### SHAP Analysis

This section shows how to compute and interpret SHAP values for CATE predictions from a causal model (e.g., a causal forest or DynamicDMLearner). We use:

• `explain_cate()` — wrapper integrating {RCausalML} with {kernelshap}/{permshap}  
• {shapviz} — visualization toolkit for importance, dependence, waterfall, and force plots  

Because SHAP is computationally expensive, we use subsamples of the training data for explanation and for background integration.

#### 1. SHAP estimation

SHAP values explain how each covariate contributes to the individual CATE, i.e.,  
“Which features make the model believe treatment is more beneficial (or more harmful) for this person?”

We compute SHAP values using `explain_cate()`:

 X = rows you want to explain  
 bgX = background (integration) distribution  
 Use non-overlapping sets when possible to avoid leakage

```{r}
#| label: shap-estimation-setup
library(future)

# Distinct moderator rows from the hospital Dynamic DML fit (same columns as X in fit)
X_shap_pool <- as.data.frame(X_het[!duplicated(as.data.frame(X_het)), , drop = FALSE])

plan(multisession, workers = max(1L, parallel::detectCores() - 1L))

set.seed(42)

nexplain <- min(80L, nrow(X_shap_pool))
nbg <- min(50L, nrow(X_shap_pool))

if (nrow(X_shap_pool) > nexplain + nbg) {
  idxexplain <- sample(nrow(X_shap_pool), nexplain)
  idxbg <- sample(setdiff(seq_len(nrow(X_shap_pool)), idxexplain), nbg)
} else {
  idxexplain <- sample(nrow(X_shap_pool), nexplain)
  idxbg <- sample(nrow(X_shap_pool), nbg)
}

Xexplain <- X_shap_pool[idxexplain, , drop = FALSE]
bg_X_df <- X_shap_pool[idxbg, , drop = FALSE]
```

```{r}
#| label: shap-estimation

fg_prev <- getOption("future.globals.maxSize", default = 500 * 1024^2)
options(future.globals.maxSize = max(2 * 1024^3, fg_prev))
ks <- tryCatch(
  explain_cate(
    est_hosp,
    X = Xexplain,
    bg_X = bg_X_df,
    use_permshap = TRUE,
    parallel = TRUE,
    verbose = FALSE,
    parallel_args = list(packages = c("ranger", "RCausalML"))
  ),
  finally = {
    future::plan(future::sequential)
    options(future.globals.maxSize = fg_prev)
  }
)
```

#### 2. Visualize SHAP values
2.1 Beeswarm importance plot

Shows the distribution of SHAP values for each feature:  
large absolute values = feature strongly influences the CATE.

```{r}
#| label: shp-importance-beeswarm
#| fig-width: 6
#| fig-height: 5

shp <- shapviz::shapviz(ks)
shapviz::sv_importance(shp, kind = "beeswarm")
```

2.2 Bar importance plot

Shows average |SHAP| per feature — good for ranking.

```{r}
#| label: shap-importance-plot
#| fig-width: 6
#| fig-height: 5

shapviz::sv_importance(shp, show_numbers = TRUE)
```

#### 3. SHAP dependence plots

Shows how each feature drives the treatment effect (colour often encodes another variable automatically).

```{r}
#| label: shap-dependence-plots
#| fig-width: 8
#| fig-height: 8

xvars <- colnames(shp$X)
shapviz::sv_dependence(shp, v = xvars, share_y = TRUE)
```

Interpretation example:  
• If SHAP increases with age, the treatment effect is stronger for older patients.

#### 4. SHAP for a single prediction

Waterfall/force plots illustrate why the model assigns a high or low CATE for one individual.

```{r}
#| label: shap-single-prediction
#| fig-width: 6
#| fig-height: 5

shapviz::sv_waterfall(shp, row_id = 1)
shapviz::sv_force(shp, row_id = 1)
```

This explains:  
• baseline CATE  
• feature-by-feature contribution  
• final individualized CATE

#### 5. Subgroup-specific SHAP (optional)

Example: show waterfall plot for the first female (sex == 2) observation.

```{r}
#| label: shap-female-waterfall
#| fig-width: 6
#| fig-height: 5

gcols <- grep("^gender", names(shp$X), value = TRUE)
if (length(gcols) > 0) {
  idx_g <- which(rowSums(as.matrix(shp$X[, gcols, drop = FALSE])) > 0)
  row_show <- if (length(idx_g) > 0) idx_g[1] else 1L
} else if ("sex" %in% names(shp$X)) {
  idx_female <- which(shp$X$sex == 2)
  row_show <- if (length(idx_female) > 0) idx_female[1] else 1L
} else {
  row_show <- 1L
}

shapviz::sv_waterfall(shp, row_id = row_show)
```



## Summary and conclusion

This notebook demonstrates **dynamic double machine learning** with **DynamicDMLearner** from the **RCausalML** package on panel-style rows (one row per unit and time).

- **Part A — synthetic data.** First, a spec **without covariates** $X$ recovers **average period-wise** treatment effects and compares them to a known DGP. Then covariates $X$ are added for **heterogeneous** effects, with CATE-style predictions and policy-style contrasts on a test split.
- **Part B — hospital deterioration.** The workflow builds `hospital_deterioration_ml_ready.csv` from vitals, labs, and patients (rolling and delta features), uses a **patient-level** train/test split, and visualizes deterioration prevalence over time. A **random forest** scores short-term risk (ROC and precision–recall on the test set). **DynamicDMLearner** is fit on an hourly panel with **nurse alerts** as treatment $T$; `effect()` summarizes **heterogeneous CATE** and **alert on vs off** contrasts, and a final plot shows **threshold–performance** trade-offs for the risk model.
- **SHAP (optional).** `explain_cate()` with **shapviz** interprets drivers of individualized CATEs (beeswarm/bar importance, dependence, waterfall, and force plots).


## Resources

**RCausalML (`DynamicDMLearner`).** After `library(RCausalML)`, see `?DynamicDMLearner` for arguments, the panel layout (outcome repeated on each row within a unit, `groups` for unit IDs), and methods such as `fit`, `predict`, and `effect`. The implementation is **native R**: flexible nuisance learners plus a **linear final stage** for CATE / treatment contrasts. It is meant to mirror the **ideas** of dynamic DML on panels; details and defaults need not match EconML line-for-line.

**EconML (Python reference).** Microsoft’s [`DynamicDML`](https://econml.azurewebsites.net/_autosummary/econml.panel.dml.DynamicDML.html) class documents the full sequential nuisance structure, options, and inference-oriented API. The [Dynamic Double Machine Learning Examples](https://github.com/microsoft/EconML/blob/main/notebooks/Dynamic%20Double%20Machine%20Learning%20Examples.ipynb) notebook parallels many of the workflows in this chapter.

**Theory.** Dynamic double / debiased ML for sequential treatments is laid out in [Lewis & Syrgkanis (2021), *Double/Debiased Machine Learning for Dynamic Treatment Effects via g-Estimation*](https://arxiv.org/abs/2002.07285) (arXiv:2002.07285). For the original double/debiased ML framework in a static setting, see [Chernozhukov et al. (2018)](https://arxiv.org/abs/1608.00060).

**SHAP for CATE (optional sections).** This notebook uses **RCausalML**’s `explain_cate()` together with the [**kernelshap**](https://CRAN.R-project.org/package=kernelshap) and [**shapviz**](https://CRAN.R-project.org/package=shapviz) packages for computation and plots.
