---
title: "Getting Started with RCausalML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with RCausalML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

**RCausalML** is an R port of the Python
[CausalML](https://causalml.readthedocs.io/en/latest/about.html) and
[EconML](https://www.microsoft.com/en-us/research/project/econml/) packages.
It provides a unified interface for estimating the **Conditional Average
Treatment Effect (CATE)** from experimental or observational data, covering:

| Module | Algorithms |
|---|---|
| **Meta-Learners** | S, T, X, R, DR, TMLE, Domain Adaptation |
| **Double ML** | LinearDML, SparseLinearDML, KernelDML, NonParamDML, CausalForestDML |
| **IV Learners** | OrthoIV, DMLIV, NonParamDMLIV, DRIV, 2SLS |
| **DoubleML Bridges** | PLR, PLIV, DiD-DML, 401k/Bonus datasets |
| **Uplift Trees/RF** | KL, ED, Chi, CTS splits; multi-arm |
| **GRF wrappers** | Causal, Survival, Multi-arm, Instrumental |
| **Deep Causal Nets** | DragonNet, TARNet, CFRNet, CEVAE, GANITE, CausalGAN, … |
| **Causal Structure** | NOTEARS, DAGMA, CASTLE, DAG-GNN, GraN-DAG |
| **Policy Learning** | DRPolicyTree, DRPolicyForest, LRSRegressor |

---

## Installation

```r
# From GitHub (recommended)
# install.packages("devtools")
devtools::install_github("zia207/RCausalML")
```

Core R dependencies are installed automatically. Neural models additionally
require the `torch` package:

```r
install.packages("torch")
torch::install_torch()
```

---

## Quick Example: T-Learner

The T-Learner fits separate outcome models for treated and control units and
estimates CATE as the difference.

```{r tlearner, eval = FALSE}
library(RCausalML)

# Simulate a simple randomised trial
set.seed(42)
n   <- 500
X   <- matrix(rnorm(n * 5), n, 5)
W   <- rbinom(n, 1, 0.5)
tau <- 2 + X[, 1]            # heterogeneous treatment effect
Y   <- tau * W + rnorm(n)

# Fit the T-Learner
tl  <- TLearner$new(learner = "ranger")
tl$fit(X, W, Y)

# Predict CATE on new data
X_test <- matrix(rnorm(100 * 5), 100, 5)
cate   <- tl$predict(X_test)
head(cate)
```

---

## Quick Example: Linear DML

Double Machine Learning (DML) removes the confounding bias of nuisance models
via cross-fitting before estimating the treatment effect.

```{r dml, eval = FALSE}
library(RCausalML)

set.seed(42)
n    <- 1000
X    <- matrix(rnorm(n * 10), n, 10)
W    <- as.numeric(X[, 1] > 0) + rnorm(n, sd = 0.1)
Y    <- 2 * W + X[, 1] + rnorm(n)

dml  <- LinearDML$new(
  model_y      = "ranger",
  model_t      = "ranger",
  n_folds      = 5
)
dml$fit(X, W, Y)

ate  <- dml$effect()
cat("Estimated ATE:", round(ate, 3), "\n")
```

---

## Quick Example: Uplift Random Forest

Uplift forests directly optimise the difference in outcomes between treated and
control units at each split.

```{r uplift, eval = FALSE}
library(RCausalML)

# Simulate uplift dataset
df <- make_uplift_classification(
  n_samples  = 2000,
  treatment_name = "treatment"
)

# KL-based uplift random forest
uf <- uplift_rf_kl(
  formula   = ~ . - treatment - outcome,
  data      = df,
  treatment = df$treatment,
  outcome   = df$outcome
)

# Gain and Qini curves
gain_df <- gain_curve(uf, df$treatment, df$outcome)
plot_gain(gain_df)
```

---

## Quick Example: Causal Forest (GRF)

```{r grf, eval = FALSE}
library(RCausalML)

set.seed(42)
n  <- 800
X  <- matrix(rnorm(n * 10), n, 10)
W  <- rbinom(n, 1, 0.4)
Y  <- (1 + X[, 1]) * W + rnorm(n)

cf  <- grf_causal_forest(X, Y, W)

# Average treatment effect
grf_average_treatment_effect(cf)

# Variable importance
grf_variable_importance(cf)
```

---

## Quick Example: DragonNet (Deep)

DragonNet jointly estimates the propensity score and the potential outcomes
using a shared representation and a targeted regularisation term.

```{r dragonnet, eval = FALSE}
library(RCausalML)

set.seed(42)
n   <- 600
X   <- matrix(rnorm(n * 8), n, 8)
W   <- rbinom(n, 1, plogis(X[, 1]))
Y   <- (0.5 + X[, 2]) * W + rnorm(n)

net <- dragonnet(X, W, Y,
  hidden_size = 64,
  epochs      = 100,
  batch_size  = 64
)

preds <- predict(net, X)
head(preds)
```

---

## Package Architecture

```
RCausalML/
├── R/
│   ├── meta_learners.R          # S/T/X/R/DR/TMLE learners
│   ├── DMLearner.R              # DML + DoubleML bridges
│   ├── iv.R                     # IV learners
│   ├── causalForest.R           # GRF wrappers + GNN causal models
│   ├── causalTree.R             # Tree-based causal methods
│   ├── causalDeepNet.R          # Neural causal models (DragonNet, …)
│   ├── causalXGBoost.R          # CXGBoost R6 class
│   ├── uplift_trees.R           # Uplift tree utilities
│   ├── uplift_tree_core.R       # Core uplift splitting logic
│   ├── policy_learner.R         # Policy learning
│   ├── cate_interpreter.R       # CATE / policy interpretation
│   ├── propensity.R             # Propensity score estimation
│   ├── match.R                  # Matching methods
│   ├── notears.R                # NOTEARS DAG learning
│   ├── automated_ml.R           # AutoML nuisance models
│   ├── synthetic_data.R         # Data generators
│   └── utils.R                  # Shared utilities
└── inst/notebooks/              # 59 Quarto tutorial notebooks
```

---

## Further Reading

- [Reference documentation](https://zia207.github.io/RCausalML/reference/index.html)
- [Tutorial notebooks](https://github.com/zia207/RCausalML/tree/main/inst/notebooks)
- [Python CausalML](https://causalml.readthedocs.io/en/latest/)
- [Microsoft EconML](https://econml.azurewebsites.net/)
- [DoubleML R package](https://docs.doubleml.org/stable/guide/r.html)
